### 2020-05-18 paper.tex
Adding paper

%% 
%% Copyright 2007-2019 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

%%\documentclass[preprint,12pt,authoryear]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
\documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

\journal{Astronomy and Computing}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{jax-cosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

\author{jax-cosmo contributors}

\address{}

\begin{abstract}
With jax-cosmo, Fisher forecasts are accurate and instantaneous, inference is lightning-fast, survey optimization is trivial.
\end{abstract}

%%Graphical abstract
%\begin{graphicalabstract}
%\includegraphics{grabs}
%\end{graphicalabstract}

%%Research highlights
%\begin{highlights}
%\item Research highlight 1
%\item Research highlight 2
%\end{highlights}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

%% \linenumbers

%% main text

\section{Introduction}

\section{JAX: Autograd meets XLA}

\section{jax-cosmo architecture}

\section{Demo}

\subsection{Instantaneous Fisher Forecasts}

\subsection{Survey Optimization by FoM Maximization}

\subsection{Massive Optimal Compression in 3 Lines}

\subsection{Lightning fast inference with BayesFast}

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
%%  \bibliographystyle{elsarticle-harv} 
%%  \bibliography{<your bibdatabase>}

%% else use the following coding to input the bibitems directly in the
%% TeX file.

\begin{thebibliography}{00}

%% \bibitem[Author(year)]{label}
%% Text of bibliographic item

\bibitem[ ()]{}

\end{thebibliography}
\end{document}

\endinput
%%
%% End of file `elsarticle-template-harv.tex'.

### 2020-07-27 paper.tex
update paper and gitignore

@@
 \usepackage{amssymb}
+\usepackage{amsmath}
 %% The amsthm package provides extended theorem environments
 %% \usepackage{amsthm}
-
+\usepackage{aas_macros}
 %% The lineno packages adds line numbers. Start line numbering with
@@
 
-\section{jax-cosmo architecture}
+\section{Architecture}
 
-\section{Demo}
+\section{Validation against the Core Cosmology Library (CCL)}
+
+\section{Example of applications}
 
@@
 \subsection{Massive Optimal Compression in 3 Lines}
+
+\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
+
+Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
+samples from high dimensional parameter spaces.  It was introduced in \citet{duane}
+and developed extensively since.  See \citep{betancourt} for a full review; we describe
+very basic features here.
+
+HMC samples spaces by generating particle tracjectories through the space, using the log-posterior
+as the negative potential energy of a particle at each point $p$ in the space.  At each sample a trajectory are initialized
+with a random velocity $q$, and then integrated using Hamilton's equations with a numerical integration:
+\begin{align}
+\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial {\cal P}}{\mathrm{d} q}\\
+\frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
+\end{align}
+where $M$ is a mass matrix which should be set to approximate the covariance of the posterior. This is also
+used to set the scale of the random initial velocities.
+
+Formally, the set of $n_\mathrm{dim}$ moment are treated as new parameters, and after 
+some number of integration steps a final point in the trajectory is compared to the initial one,
+and a Metropolis-Hastings acceptance criterion on the total energy $p^T M^{-1} p + {\cal}P(q)$ is applied.
+If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
+a relaxation to the integration accuracy.
+
+That gradients $\frac{\partial {\cal P}}{\mathrm{d} q}$ can be estimated using finite differencing,
+but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
+in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
+calculating a Hessian, as in JAX-Cosmo, makes it feasible and efficient.
+
+Below we show an example set of HMC constraints using JAX-Cosmo to constrain a simulated
+Dark Energy Survey Year 1 (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
+five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
+decomposition,
+of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
+edge of the parameter space, as recommended in \citet{afshar}.  We use a fixed number (25) of integration steps and 
+a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
+far from optimal.  We compare to the highly-optimized \emph{Cobaya} 
+Metropolis-Hastings implementation \citep{cobaya1,cobaya2}, initializing it with the Fisher matrix as a proposal and
+tuning thereafter.
+
+The Hamiltonian sampler required about half the number of samples as the Metropolis to reach approximately the same
+convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
+on a CPU.
+This proof-of-principle example can certainly be hugely improved using various extensions to HMC such as No U-Turn Sampling \citep{nuts} or sampling from the entire trajectories \cite{betancourt}.
+On next generation GPU high-performance systems, the advantage will be large.
+
+\begin{figure}
+\includegraphics[width=0.95\columnwidth]{figures/hmc.pdf}
+\caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
+Monte Carlo sampler with JAX-Cosmo (blue), compared to the Cobaya Metropolis-Hastings sampler (orange).}
+\end{figure}
+
+We note that a full Boltzmann code accurate enough to calculate the CMB spectra to Planck \citep{planck18}
+or future \citep{s4,simons} accuracy is currently beyond the scope of this project.  
+If this is needed then HMC could use a fast approximation to a CMB theory spectrum, and importance sampling applied to the complete chain.
 
@@
 %%
-%%  \bibliographystyle{elsarticle-harv} 
-%%  \bibliography{<your bibdatabase>}
+\section*{Bibliography}
+\bibliographystyle{elsarticle-harv} 
+\bibliography{refs}
 
@@
 
-\begin{thebibliography}{00}
-
-%% \bibitem[Author(year)]{label}
-%% Text of bibliographic item
-
-\bibitem[ ()]{}
-
-\end{thebibliography}
 \end{document}

### 2020-08-21 paper.tex

@@
 samples from high dimensional parameter spaces.  It was introduced in \citet{duane}
-and developed extensively since.  See \citep{betancourt} for a full review; we describe
+and developed extensively since.  See \citet{betancourt} for a full review; we describe
 very basic features here.
@@
 HMC samples spaces by generating particle tracjectories through the space, using the log-posterior
-as the negative potential energy of a particle at each point $p$ in the space.  At each sample a trajectory are initialized
-with a random velocity $q$, and then integrated using Hamilton's equations with a numerical integration:
+as the negative potential energy of a particle at each point $q$ in the space.  At each sample a trajectory are initialized
+with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration:
 \begin{align}
-\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial {\cal P}}{\mathrm{d} q}\\
+\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial \log{\cal P}}{\mathrm{d} q}\\
 \frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
@@
 some number of integration steps a final point in the trajectory is compared to the initial one,
-and a Metropolis-Hastings acceptance criterion on the total energy $p^T M^{-1} p + {\cal}P(q)$ is applied.
+and a Metropolis-Hastings acceptance criterion on the total energy $H = p^T M^{-1} p + \log{\cal P}(q)$ is applied.
 If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
@@
 
-That gradients $\frac{\partial {\cal P}}{\mathrm{d} q}$ can be estimated using finite differencing,
+That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
 but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
@@
 Below we show an example set of HMC constraints using JAX-Cosmo to constrain a simulated
-Dark Energy Survey Year 1 (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
+Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
 five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
-decomposition,
+decomposition
 of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
@@
 a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
-far from optimal.  We compare to the highly-optimized \emph{Cobaya} 
-Metropolis-Hastings implementation \citep{cobaya1,cobaya2}, initializing it with the Fisher matrix as a proposal and
-tuning thereafter.
+far from optimal.  We compare to the highly-optimized {\sc Cobaya} 
+Metropolis-Hastings implementation \citep{cobaya2,cobaya1}, initializing it with the Fisher matrix as a proposal and
+tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.
 
-The Hamiltonian sampler required about half the number of samples as the Metropolis to reach approximately the same
+The Hamiltonian sampler required around half the number of samples as the Metropolis (70,000 vs 140,000) to reach approximately the same
 convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower

### 2022-01-22 paper.tex
Updates from Overleaf
SVI+NeutraReparam start

@@
 \journal{Astronomy and Computing}
+
+%%%% new command (JEC)
+\newcommand{\nn}{\nonumber}
+\newcommand{\numpyro}{\texttt{NumPyro}}
+\newcommand{\bydef}{\overset{def.}{\equiv}}
 
@@
 
+%
+\subsection{Use of SVI \& Neutral Reparametrisation for NUTS}
+%
+Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps.
+If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
+The SVI aims to get an approximation of this distribution, ie.
+$p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
+To do so, one uses the Kullback-Leibler divergence of the two distributions $KL(q(z;\lambda)||p(z|\mathcal{D}))$ leading to the following relation
+\begin{align}
+\log p(\mathcal{D}) &= \mathtt{ELBO} +  KL(q(z;\lambda)||p(z|\mathcal{D})) \label{eq-ELBO} \\
+\mathrm{with} \ \mathtt{ELBO} &\equiv -\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right] + \mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right] 
+\end{align}
+which defines the evidence lower bound (ELBO) that one tends to maximize to get the $\lambda$ values and the  highest lower bound on the logarithm of the marginal probability of the observations $\log p(\mathcal{D})$.
+So, the optimal variational distribution satisfies
+\begin{equation}
+q(z;\lambda^\ast) = \underset{q(z;\lambda)}{\mathrm{argmax}}\  \mathtt{ELBO} = 
+\underset{\lambda}{\mathrm{argmin}}\ \mathcal{L}(\lambda)
+\end{equation}
+The function $\mathcal{L}(\lambda)$ is the cost function used in practice. It is composed of two parts:
+\begin{equation}
+\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right]}_{model}
+\label{eq-loss-svi-1}
+\end{equation}
+where the \textit{guide}, in the \numpyro\ word, may be a multi-variate gaussian distribution (MVN) for instance.
+
+Using the auto-differentation tool, one can use "black-box" guides (aka \textit{automatic differentiation variational inference}).
+As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
+function, taking derivatives, and running a gradient-based or coordinate-ascent optimization.
+First is defined a bijective differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
+One can develop "AutoGuides" (\numpyro\ terminology) that can be adapted to the user models.
+Then, the cost function reads
+\begin{equation}
+\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log p(\xi,\mathcal{D}) \right]}_{model}
+\label{eq-loss-svi-2}
+\end{equation}
+with
+\begin{equation}
+p(\xi,\mathcal{D}) \bydef p(T^{-1}(\xi),\mathcal{D}) |J_{T^{-1}}(\xi)|
+\end{equation}
+which exhibit the Jacobian of the $T^{-1}$ transformation. The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
+
+Now, one is faced with the evaluation of the expectations during gradient descent.
+The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}).
+Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ ($\lambda=(\mu, \Sigma)$).
+One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg.
+$\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$.
+The Jacobien of this distribution is 1, so the cost function reads
+\begin{multline}
+-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
+\label{eq-loss-svi-3}
+\end{multline}
+where $\mathbb{H}(q)$ is the Shannon entropy of the gaussian distribution which is independent of the model, so its gradient can be computed once for all and reused in user model. Then,
+to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
+\begin{multline}
+-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
++ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
+\right\}
++ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
+\label{eq-loss-svi-4}
+\end{multline}
+For the use case of the MVN (auto)guide, one gets
+\begin{align}
+-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
++ \nabla_\xi \log|J_{T^{-1}}(\xi)|
+\right\} \nn \\
+-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
+&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
+\right\}+ (L^{-1})^T
+\end{align}
+Reminding that $z=T^{-1}(\xi)$ and $\xi=S^{-1}(\zeta)$, the expectation used in the above expression is computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$.
 %% The Appendices part is started with the command \appendix;

### 2022-01-23 paper.tex
improve SVI+NUTS Neural Param with text and figure

@@
 \subsection{Use of SVI \& Neutral Reparametrisation for NUTS}
+\subsubsection{Brief introduction on the methods}
 %
@@
 If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
-The SVI aims to get an approximation of this distribution, ie.
+The VI aims to get an approximation of this distribution, ie.
 $p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
@@
 \end{align}
-Reminding that $z=T^{-1}(\xi)$ and $\xi=S^{-1}(\zeta)$, the expectation used in the above expression is computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$.
+The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI.
+
+If the SVI method can be used as is to get $z$ i.i.d.
+samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
+From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
+\begin{equation}
+q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
+\end{equation}
+which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg.
+$F_\lambda=T^{-1}\circ S^{-1}_\lambda$).
+So, one can use a HMC sampler (eg.
+NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$.
+Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples.
+The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.
+%
+\subsubsection{Example}
+%
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200.pdf}
+\caption{Constraints on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in..., and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
+\label{fig_svi_nuts_1}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x1000.pdf}
+\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 1 chain of 200 warm followed by 1000 samples for the NUTS sampler after SVI optimization.}
+\label{fig_svi_nuts_2}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/SVI_NUTS_10x200x1000.pdf}
+\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 10 chains of 200 warm followed by 1000 samples for the NUTS sampler after SVI optimization.}
+\label{fig_svi_nuts_3}
+\end{figure}
+
 %% The Appendices part is started with the command \appendix;

### 2022-01-23 paper.tex
improve SVI+NUTS Neural Param with text and a table on ESS

@@
 Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
-samples from high dimensional parameter spaces.  It was introduced in \citet{duane}
+samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
 and developed extensively since.  See \citet{betancourt} for a full review; we describe
@@
 of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
-edge of the parameter space, as recommended in \citet{afshar}.  We use a fixed number (25) of integration steps and 
+edge of the parameter space, as recommended in \citep{NIPS2015_8303a79b}.  We use a fixed number (25) of integration steps and 
 a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
@@
 %
-\subsubsection{Example}
+\subsubsection{DES-Y1 Example}
 %
+Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI.
+The Adam optimizer has been used with a learning rate set to $10^{-3}$.
+Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler.
+The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a V100 GPU.
+Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (vectorized) with 1000 samples each.
+The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
+Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running.
+The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
+The remarkable outcome is that within few hours on a single GPU, one can figure out progressively if the contours have a meaningful appearance before launching multi-GPU multi-chains NUTS sampling to get high statistics Markov Chain with high effective sample size (ESS).
+Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}....
 \begin{figure}
@@
 \end{figure}
+
+\begin{table}[htb]
+\caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
+(1) are computed from MCMC obtained by the the NUTS sampler described..., while (2) are the results of SVI followed by NUTS sampler after a Neural Transform  with 10 chains of 1000 sampling each after 200 warmup samples.}
+\label{tab-ESS-NUTS_SVI-1}
+ \centering
+\begin{tabular}{ccccccccccc}
+\hline
+         & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$     & $n_s$ & $A$ \\
+\hline
+(1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ \\
+(2) & $38.7$ &  $24.8$     & $16.0$     & $18.6$ & $43.3$ & $39.5$ & $21.7$ \\
+\end{tabular}
+\end{table}
 

### 2022-01-24 paper.tex
add a figure to show bias correlation. Start Appendix for DES-Y1 paameters...

@@
 The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
+The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.
+
 The remarkable outcome is that within few hours on a single GPU, one can figure out progressively if the contours have a meaningful appearance before launching multi-GPU multi-chains NUTS sampling to get high statistics Markov Chain with high effective sample size (ESS).
@@
 \includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200.pdf}
-\caption{Constraints on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in..., and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
+\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 22 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in..., and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
 \label{fig_svi_nuts_1}
@@
 
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200_bis.pdf}
+\caption{In the same conditions of Fig.~\ref{fig_svi_nuts_1} but for the five lens galaxy bias which are highly correlated.}
+\label{fig_svi_nuts_4}
+\end{figure}
+
 \begin{table}[htb]
@@
 \hline
-         & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$     & $n_s$ & $A$ \\
+         & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$ & $n_s$ & $A$ & $\eta$\\
 \hline
-(1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ \\
-(2) & $38.7$ &  $24.8$     & $16.0$     & $18.6$ & $43.3$ & $39.5$ & $21.7$ \\
+(1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\
+(2) & $38.7$ &  $24.8$     & $16.0$     & $18.6$ & $43.3$ & $39.5$ & $21.7$ & $16.6$\\
+\hline
 \end{tabular}
@@
 %% appendix sections are then done as normal sections
-%% \appendix
+\appendix
 
-%% \section{}
-%% \label{}
+\section{Description of the DES-Y1}
+\label{app-DESY1}
+\begin{table}[htb]
+\caption{Priors of the 22 variables of the DES-Y1 of the 3x2pt likelihood (number counts and shear).}
+\label{tab-DESY1}
+ \centering
+\begin{tabular}{ccccccccccc}
+\hline
+ parameter &  prior \\
+ \hline
+  \multicolumn{2}{c}{\textbf{Cosmology}} \\
+$\Omega_b$ & $\mathcal{U}[0.10, 0.9]$ \\
+$\Omega_c$ & $\mathcal{U}[0.40, 1.0]$ \\
+$\sigma_8$ & $\mathcal{U}[0.03, 0.07]$ \\
+$w_0$ & $\mathcal{U}[-2.00, -0.33]$ \\
+$h$ & $\mathcal{U}[0.55, 0.91]$ \\
+$n_s$ & $\mathcal{U}[0.87, 1.07]$ \\
+\multicolumn{2}{c}{\textbf{Intrinsic Alignment}} \\
+$A$ & $\mathcal{U}[-5,5]$ \\
+$\eta$ &$\mathcal{U}[-5,5]$ \\
+\multicolumn{2}{c}{\textbf{Lens Galaxy Bias}} \\
+$(b_i)_{i:1,\dots,5}$ & $\mathcal{U}[0.8,3.0]$ \\
+\multicolumn{2}{c}{\textbf{Shear Calibration Systematics}} \\
+$(m_i)_{i:1,\dots,4}$ & $\mathcal{N}[0.012,0.023]$ \\
+\multicolumn{2}{c}{\textbf{Source photo-$z$ shift}} \\
+$dz_1$ & $\mathcal{N}[0.001,0.016]$ \\
+$dz_2$ & $\mathcal{N}[-0.019,0.013]$ \\
+$dz_3$ & $\mathcal{N}[0.009,0.011]$ \\
+$dz_4$ & $\mathcal{N}[-0.018,0.022]$ \\
+\hline
+\end{tabular}
+\end{table}
 

### 2022-01-24 paper.tex
SVI DES-Y1 example text update.

@@
 The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.
+Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}.
+On can see that the NUTS sampling with a \texttt{max\_tree\_depth} parameter of 5 after the SVI and Neutral Parametrisation gives same order of efficiencies (lower by a factor 2 for some parameters) compared to the NUTS sampling with block mass matrix and \texttt{max\_tree\_depth} parameter of 7.
 
 The remarkable outcome is that within few hours on a single GPU, one can figure out progressively if the contours have a meaningful appearance before launching multi-GPU multi-chains NUTS sampling to get high statistics Markov Chain with high effective sample size (ESS).
-Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}....
+
 \begin{figure}
@@
 \caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
-(1) are computed from MCMC obtained by the the NUTS sampler described..., while (2) are the results of SVI followed by NUTS sampler after a Neural Transform  with 10 chains of 1000 sampling each after 200 warmup samples.}
+(1) are computed from 10 parallel chains of 200 warm up samples followed by 1000 samples each obtained by NUTS sampler after the several uniform distributions (Tab.~\ref{tab-DESY1}) transformed in an homogeneous uniform distribution in the $[-5,5]$ interval, while (2) are the results of SVI followed by NUTS sampler after a Neural Transform  with 10 chains of 200 warm up samples followed by 1000 samples each.
+In method (1) the \texttt{max\_tree\_depth} parameter is set to 7 while it is set to 5 in the second method.}
 \label{tab-ESS-NUTS_SVI-1}

### 2022-02-11 paper.tex
Updates from Overleaf

@@
 %% The amssymb package provides various useful mathematical symbols
+\usepackage{minted}
 \usepackage{amssymb}
@@
 
-\section{JAX: Autograd meets XLA}
+% \section{JAX: Autograd meets XLA}
 
 \section{Architecture}
+% Explain how the code is structured, mention some tricks of autodiff if we have some
+% Follows a bit the structure of the CCL library
 
-\section{Validation against the Core Cosmology Library (CCL)}
+\subsection{Validation against the Core Cosmology Library (CCL)}
 
 \section{Example of applications}
+% illustrates why autodiff is interesting from several points of view
+
+\subsection{DES Y1 setting}
+% Here we describe the data and 3x2pt model for DES Y1
 
@@
 
+\begin{minted}{python}
+def likelihood(cosmo):
+  # Compute mean and covariance of angular Cls, 
+  # for specific probes
+  mu, cov = gaussian_cl_covariance_and_mean(cosmo, 
+                    ell, probes)
+  # Return likelihood value
+  return gaussian_log_likelihood(data, mu, cov)
+
+# Compute derivatives of the likelihood with 
+# respect to cosmological parameters
+g = jax.grad(likelihood)(cosmo)
+
+# Compute Fisher matrix of cosmological 
+# parameters
+F = - jax.hessian(likelihood)(cosmo)
+\end{minted}
+
+% Plot showing error in Jacobian computation by finite diff
+
+% Plot showing error on some parameters
+
 \subsection{Survey Optimization by FoM Maximization}
+% Needs to mention the tomo challenge, mention that we can backpropagate through a NN and the cosmology model
+% Maybe this can be only mentioned in the discussion.
 
 \subsection{Massive Optimal Compression in 3 Lines}
+
+% Here we can implement MOPED for the compression of
+
+\section{Gradient-based fast cosmological inference}
+
+\subsection{Variationnal Inference}
+
+\subsection{Hamiltonian Monte Carlo}
+
+\subsubsection{Introduction}
+
+% Show Joe's vanilla HMC results against Cobaya
+
+\subsubsection{NUTS}
+
+% Show the advantage of using NUTS 
+% look at difference in efficiency in terms of how many times we need to call the model.
+
+\subsubsection{Neural Reparametrisation}
+
+\subsection{Discussion}
+
+\section{Conclusion}
 

### 2022-02-11 paper.tex
Made a few modification

@@
 \usepackage{aas_macros}
+\usepackage{fontawesome}
 %% The lineno packages adds line numbers. Start line numbering with
@@
 %% \usepackage{lineno}
+
+\usepackage{xcolor}
+\definecolor{xlinkcolor}{cmyk}{1,1,0,0}
+\usepackage[colorlinks = true,
+            linkcolor = xlinkcolor,
+            urlcolor  = xlinkcolor,
+            citecolor = xlinkcolor,
+            anchorcolor = xlinkcolor]{hyperref}
+
+\newcommand{\nblink}[1]{\href{https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/blob/master/notebooks/#1.ipynb}{\faFileCodeO}}
+\newcommand{\github}{\href{https://github.com/DifferentiableUniverseInitiative/jax\_cosmo}{\faGithub}}
 
@@
 \begin{abstract}
-With jax-cosmo, Fisher forecasts are accurate and instantaneous, inference is lightning-fast, survey optimization is trivial.
+With jax-cosmo, Fisher forecasts are accurate and instantaneous, inference is lightning-fast, survey optimization is trivial. \github
 \end{abstract}
@@
 
-\section{Architecture}
+\section{Modeling and implementation}
+
+In this section, we describe the cosmological modeling provided by jax-cosmo, and its implementation in JAX.
+
 % Explain how the code is structured, mention some tricks of autodiff if we have some
@@
 
+\subsection{Background cosmology}
+% Usual stuff
+
+\subsection{Growth of perturbations}
+% The interesting point to mention here is the differentiable growth ODE
+
+\subsection{Matter power spectrum}
+% Here we describe the EH power spectrum, mention possible extensions with emulators
+
+\subsection{Tracers}
+% Here we describe the kernels, like in CCL
+
+\subsection{Angular power-spectra}
+% Limber integration, this can be an opportunity to talk about differentiation through integrals as well.
+
 \subsection{Validation against the Core Cosmology Library (CCL)}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/halofit_pk.png}
+    \caption{Comparison of matter power spectrum between CCL and jax-cosmo.
+\textbf{Example of validation plot} \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo/blob/master/docs/notebooks/CCL_comparison.ipynb}{\faFileCodeO}}
+    \label{fig:halofit_comparison}
+\end{figure}
 
@@
 % Here we describe the data and 3x2pt model for DES Y1
+
+In this section, we adopt as a setting a DES Y1 3x2pt-like analysis.
 
@@
 \section{Gradient-based fast cosmological inference}
-
-\subsection{Variationnal Inference}
-
-\subsection{Hamiltonian Monte Carlo}
-
-\subsubsection{Introduction}
-
-% Show Joe's vanilla HMC results against Cobaya
-
-\subsubsection{NUTS}
-
-% Show the advantage of using NUTS 
-% look at difference in efficiency in terms of how many times we need to call the model.
-
-\subsubsection{Neural Reparametrisation}
-
-\subsection{Discussion}
-
-\section{Conclusion}
-
-\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
-
-Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
-samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
-and developed extensively since.  See \citet{betancourt} for a full review; we describe
-very basic features here.
-
-HMC samples spaces by generating particle tracjectories through the space, using the log-posterior
-as the negative potential energy of a particle at each point $q$ in the space.  At each sample a trajectory are initialized
-with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration:
-\begin{align}
-\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial \log{\cal P}}{\mathrm{d} q}\\
-\frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
-\end{align}
-where $M$ is a mass matrix which should be set to approximate the covariance of the posterior. This is also
-used to set the scale of the random initial velocities.
-
-Formally, the set of $n_\mathrm{dim}$ moment are treated as new parameters, and after 
-some number of integration steps a final point in the trajectory is compared to the initial one,
-and a Metropolis-Hastings acceptance criterion on the total energy $H = p^T M^{-1} p + \log{\cal P}(q)$ is applied.
-If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
-a relaxation to the integration accuracy.
-
-That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
-but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
-in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
-calculating a Hessian, as in JAX-Cosmo, makes it feasible and efficient.
-
-Below we show an example set of HMC constraints using JAX-Cosmo to constrain a simulated
-Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
-five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
-decomposition
-of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
-edge of the parameter space, as recommended in \citep{NIPS2015_8303a79b}.  We use a fixed number (25) of integration steps and 
-a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
-far from optimal.  We compare to the highly-optimized {\sc Cobaya} 
-Metropolis-Hastings implementation \citep{cobaya2,cobaya1}, initializing it with the Fisher matrix as a proposal and
-tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.
-
-The Hamiltonian sampler required around half the number of samples as the Metropolis (70,000 vs 140,000) to reach approximately the same
-convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
-on a CPU.
-This proof-of-principle example can certainly be hugely improved using various extensions to HMC such as No U-Turn Sampling \citep{nuts} or sampling from the entire trajectories \cite{betancourt}.
-On next generation GPU high-performance systems, the advantage will be large.
-
-\begin{figure}
-\includegraphics[width=0.95\columnwidth]{figures/hmc.pdf}
-\caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
-Monte Carlo sampler with JAX-Cosmo (blue), compared to the Cobaya Metropolis-Hastings sampler (orange).}
-\end{figure}
-
-We note that a full Boltzmann code accurate enough to calculate the CMB spectra to Planck \citep{planck18}
-or future \citep{s4,simons} accuracy is currently beyond the scope of this project.  
-If this is needed then HMC could use a fast approximation to a CMB theory spectrum, and importance sampling applied to the complete chain.
-
-\subsection{Lightning fast inference with BayesFast}
-
 %
-\subsection{Use of SVI \& Neutral Reparametrisation for NUTS}
-\subsubsection{Brief introduction on the methods}
+In the following sections we briefly review the statistical methods used to make some cosmological parameter inference.
+Namely we have used \textit{Stochastic Variational Inference} (aka SVI), \textit{Hybrid Hamiltonian Monte Carlo} (aka HMC), a variant of HMC knows as \textit{No-U-Turn} (aka NUTS), and finally the use of  \textit{Neutral Reparametrization}.
+%
+\subsection{Stochastic Variational Inference}
 %
@@
 
+\subsection{Hamiltonian Monte Carlo}
+
+\subsubsection{Introduction}
+
+% Show Joe's vanilla HMC results against Cobaya
+Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
+samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
+and developed extensively since.  See \citet{betancourt} for a full review; we describe
+very basic features here.
+
+HMC samples spaces by generating particle tracjectories through the space, using the log-posterior
+as the negative potential energy of a particle at each point $q$ in the space.  At each sample a trajectory are initialized
+with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration:
+\begin{align}
+\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial \log{\cal P}}{\mathrm{d} q}\\
+\frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
+\end{align}
+where $M$ is a mass matrix which should be set to approximate the covariance of the posterior. This is also
+used to set the scale of the random initial velocities.
+
+Formally, the set of $n_\mathrm{dim}$ moment are treated as new parameters, and after 
+some number of integration steps a final point in the trajectory is compared to the initial one,
+and a Metropolis-Hastings acceptance criterion on the total energy $H = p^T M^{-1} p + \log{\cal P}(q)$ is applied.
+If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
+a relaxation to the integration accuracy.
+
+That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
+but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
+in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
+calculating a Hessian, as in JAX-Cosmo, makes it feasible and efficient.
+
+Below we show an example set of HMC constraints using JAX-Cosmo to constrain a simulated
+Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
+five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
+decomposition
+of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
+edge of the parameter space, as recommended in \citep{NIPS2015_8303a79b}.  We use a fixed number (25) of integration steps and 
+a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
+far from optimal.  We compare to the highly-optimized {\sc Cobaya} 
+Metropolis-Hastings implementation \citep{cobaya2,cobaya1}, initializing it with the Fisher matrix as a proposal and
+tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.
+
+The Hamiltonian sampler required around half the number of samples as the Metropolis (70,000 vs 140,000) to reach approximately the same
+convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
+on a CPU.
+This proof-of-principle example can certainly be hugely improved using various extensions to HMC such as No U-Turn Sampling \citep{nuts} or sampling from the entire trajectories \cite{betancourt}.
+On next generation GPU high-performance systems, the advantage will be large.
+
+\begin{figure}
+\includegraphics[width=0.95\columnwidth]{figures/hmc.pdf}
+\caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
+Monte Carlo sampler with JAX-Cosmo (blue), compared to the Cobaya Metropolis-Hastings sampler (orange).}
+\end{figure}
+
+We note that a full Boltzmann code accurate enough to calculate the CMB spectra to Planck \citep{planck18}
+or future \citep{s4,simons} accuracy is currently beyond the scope of this project.  
+If this is needed then HMC could use a fast approximation to a CMB theory spectrum, and importance sampling applied to the complete chain.
+
+\subsubsection{NUTS}
+%
+% Show the advantage of using NUTS 
+% look at difference in efficiency in terms of how many times we need to call the model.
+
+A variant of the traditional HMC sampler has been introduced in reference \citep{nuts}.
+
+\subsubsection{Neural Reparametrisation}
 If the SVI method can be used as is to get $z$ i.i.d.
@@
 The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.
+
+\subsection{Discussion}
+
+\section{Conclusion}
+
+\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
+
+\subsection{Lightning fast inference with BayesFast}
+
+%
 %

### 2022-02-11 paper.tex
test github de jec
More on NUTS

@@
 \end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the gaussian distribution which is independent of the model, so its gradient can be computed once for all and reused in user model. Then,
+where $\mathbb{H}(q)$ is the Shannon entropy of the Gaussian distribution which is independent of the model, so its gradient can be computed once for all and reused in user model. Then,
 to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
@@
 The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI.
-
+Even, if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to  a NUTS sampler which is described in section \ref{sec-Neural-Reparametrisation} as \textit{Neural Reparametrisation}.
+%
 \subsection{Hamiltonian Monte Carlo}
@@
 A variant of the traditional HMC sampler has been introduced in reference \citep{nuts}.
-
+It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$.
+Namely, the root algorithm of HMC deals with the \textit{leapfrog} integration which requires to set two main parameters: $\varepsilon$ the step size and $L$ the number of steps.
+As stated by the authors, NUTS  uses a recursive algorithm to build a set of likely candidate samples that spans a wide swath of the target distribution, stopping automatically when it starts to double back and retrace its steps.
+So, in a automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
+The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls a each generation.
+Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
+%
 \subsubsection{Neural Reparametrisation}
+\label{sec-Neural-Reparametrisation}
+%
 If the SVI method can be used as is to get $z$ i.i.d.
@@
 The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.
-
-\subsection{Discussion}
-
-\section{Conclusion}
-
-\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
-
-\subsection{Lightning fast inference with BayesFast}
-
-%
-%
-\subsubsection{DES-Y1 Example}
-%
-Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI.
-The Adam optimizer has been used with a learning rate set to $10^{-3}$.
-Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler.
-The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a V100 GPU.
-Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (vectorized) with 1000 samples each.
-The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
-Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running.
-The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
-The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.
-Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}.
-On can see that the NUTS sampling with a \texttt{max\_tree\_depth} parameter of 5 after the SVI and Neutral Parametrisation gives same order of efficiencies (lower by a factor 2 for some parameters) compared to the NUTS sampling with block mass matrix and \texttt{max\_tree\_depth} parameter of 7.
-
-The remarkable outcome is that within few hours on a single GPU, one can figure out progressively if the contours have a meaningful appearance before launching multi-GPU multi-chains NUTS sampling to get high statistics Markov Chain with high effective sample size (ESS).
 
@@
 
+\subsection{Discussion}
+Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI.
+The Adam optimizer has been used with a learning rate set to $10^{-3}$.
+Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler.
+The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a V100 GPU.
+Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (vectorized) with 1000 samples each.
+The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
+Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running.
+The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
+The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.
+Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}.
+On can see that the NUTS sampling with a \texttt{max\_tree\_depth} parameter of 5 after the SVI and Neutral Parametrisation gives same order of efficiencies (lower by a factor 2 for some parameters) compared to the NUTS sampling with block mass matrix and \texttt{max\_tree\_depth} parameter of 7.
+
+The remarkable outcome is that within few hours on a single GPU, one can figure out progressively if the contours have a meaningful appearance before launching multi-GPU multi-chains NUTS sampling to get high statistics Markov Chain with high effective sample size (ESS).
+
 \begin{table}[htb]
@@
 \end{table}
+
+\section{Conclusion}
+
+\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
+
+\subsection{Lightning fast inference with BayesFast}
+
+%
+%
+\subsubsection{DES-Y1 Example}
+%
 

### 2022-02-15 paper.tex
NUTS/SVI + Discussion starts. Table 1 on ESS/n_eff revisted with run (0) added (aka Francois run)

@@
 \subsubsection{NUTS}
+\label{sec-NUTS}
 %
@@
 The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls a each generation.
+
 Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
+So, We have setup \texttt{NUTS} sampler Using the same DES-Y1 modeling as in previous section, with
+\begin{itemize}
+    \item first a reparametrisation of the latent variable uniform priors (Tab.~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$;
+    \item second for the warm-up phase we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
+This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in figure \ref{fig_svi_nuts_4}.
+\end{itemize}
+We have run two sampling runs with different settings although similar.
+The first run labelled "(0)" has been undertaken on a single  NVidia Titan Xp (12GB RAM)  with 16 chains of 1,000 samples each (\texttt{vectorized} option) after a warm-up phase consisting of 200 samples and the \texttt{max\_tree\_depth} has been set to  7 (ie.
+maximum of 128 steps for each iteration).
+The second run labelled "(1)" was done with a NVidia V100 (32GB RAM) with 10 chains of 1,000 samples each after a warm-up phase consisting of 200 samples and the \texttt{max\_tree\_depth} has been set to  5 (ie.
+maximum of 32 steps for each iteration).
+
+The results in terms of sampling efficiency are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1}.
+The confidential level iso-contours are all in agreement with the \texttt{cobaya} sampler with  significantly larger efficiency.
+{\color{red} FIGURE?}  It is worth to point out that the mass matrix structuring is a very effective tuning to increase the sampling efficiency (gain of about a factor 2).
+The elapse time was 20 hours (5 hours) for run "0" ("1") showing a linear scaling with respect to the number of steps per iteration, while the sampling efficiencies of both runs are very similar, so the user is highly encouraged to decrease the \texttt{max\_tree\_depth} parameter.
+
+%  JEC: the 16 chains X 1000 (max_tree_depth=7) is the run done by Francois. 20 heures sur un GPU TitanXp avec 12 GB de ram 
+% 5h sur V100 10 chains x 1000 (max_tree_depth=5)
+
 %
@@
 
-\subsection{Discussion}
 Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI.
@@
 Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler.
-The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a V100 GPU.
-Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (vectorized) with 1000 samples each.
+The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a Nvidia V100 GPU (32 GB RAM).
+Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (\texttt{vectorized} option) with 1000 samples each.
 The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
 Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running.
-The results of the evolution of the contours compared to the direct NUTS sampling lasting 20 hours on the same GPU are shown on the figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
+The results of the evolution of the contours compared to the NUTS sampling  run "(1)" (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
 The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.
-Results upon ESS are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}.
-On can see that the NUTS sampling with a \texttt{max\_tree\_depth} parameter of 5 after the SVI and Neutral Parametrisation gives same order of efficiencies (lower by a factor 2 for some parameters) compared to the NUTS sampling with block mass matrix and \texttt{max\_tree\_depth} parameter of 7.
+Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}.
+On can see SVI followed by Neutral Parametrisation gives also high efficiency without the need by the user to setup a particular a priori mass matrix structure (use \texttt{dense} matrix option).
+%
+\subsection{Discussion}
+%
+The question that one may naturally arises are what is the HMC/NUTS gain compared to the well known standard \texttt{cobaya} sampler? 
+One may argue that number of efficient sample per  model call is a right metric to consider, ie. if one demands $N_s$ samples then  
+\begin{equation}
+    \eta = \frac{n_{eff}}{#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
+\end{equation}
+with $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation.
+Then, for \texttt{cobaya} one roughly gets $\varepsilon\approx 1\%$ with $n_{steps}=1$ while with the NUTS sampler $\varepsilon\approx 50\%$ but with $n_{steps}=2^5$ or more.
+So, the $\eta$ values for both methods are at the end of the day comparable (ie.
+about $1\%$).
+So, at first sight it seems that there might be no gain to use HMC/NUTS sampler compared to standard MCMC sampler.
 
-The remarkable outcome is that within few hours on a single GPU, one can figure out progressively if the contours have a meaningful appearance before launching multi-GPU multi-chains NUTS sampling to get high statistics Markov Chain with high effective sample size (ESS).
-
+But the above metric is one aspect of sample generation.
+One should have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
+several days on dedicated infrastructure.
+Moreover, the user is faced to a dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, 2015arXiv151007727O}.
+So, the authors think that it is worth considering methods that by construction are more efficient.
+Moreover, using SVI with reparametrisation offers a way to undertake a progressive validation of the model with rather modest sample size (eg.
+200 samples) by producing good quality marginal contours in few hours.
+This validation phase can save time before producing sizeable batch for final analysis.
+%
 \begin{table}[htb]
 \caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
-(1) are computed from 10 parallel chains of 200 warm up samples followed by 1000 samples each obtained by NUTS sampler after the several uniform distributions (Tab.~\ref{tab-DESY1}) transformed in an homogeneous uniform distribution in the $[-5,5]$ interval, while (2) are the results of SVI followed by NUTS sampler after a Neural Transform  with 10 chains of 200 warm up samples followed by 1000 samples each.
-In method (1) the \texttt{max\_tree\_depth} parameter is set to 7 while it is set to 5 in the second method.}
+{\color{red} 
+(0) "F.L run" are computed from NUTS sampler with 16 parallel chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (see text) The numbers are from \texttt{n\_eff/16,000};
+}
+(1) "J.E run" are computed from NUTS sampler 10 with parallel chains of 1,000 samples each and 200 warm-up nd \texttt{max\_tree\_depth=5} (see text); (2) are the results of SVI followed by NUTS sampler of setup (1) after a Neural Transform (see text).}
 \label{tab-ESS-NUTS_SVI-1}
@@
 \hline
+% (0)  JEC extrait de https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/blob/master/notebooks/VectorizedNumPyro_block.ipynb
+\color{red}(0) & $51.5$ &  $49.5$     & $39.3$     & $39.3$ & $52.0$ &  $50.2$ & $69.8$ & $46.2$\\
+% (1) & (2) from JEC independant runs @ CCIN2P3
 (1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\

### 2022-02-16 paper.tex
Clarify numbers are JEC has run a 16 chain x 1,000 samples + 200, max-tree-depth=7 and block mass matrix, identical run as Francois did. So SVI, NUTS and  Neural parametrisation sections are coherent.

@@
 \end{itemize}
-We have run two sampling runs with different settings although similar.
-The first run labelled "(0)" has been undertaken on a single  NVidia Titan Xp (12GB RAM)  with 16 chains of 1,000 samples each (\texttt{vectorized} option) after a warm-up phase consisting of 200 samples and the \texttt{max\_tree\_depth} has been set to  7 (ie.
-maximum of 128 steps for each iteration).
-The second run labelled "(1)" was done with a NVidia V100 (32GB RAM) with 10 chains of 1,000 samples each after a warm-up phase consisting of 200 samples and the \texttt{max\_tree\_depth} has been set to  5 (ie.
-maximum of 32 steps for each iteration).
+We have setup a NUTS sampling using 16 chains of 1,000 samples each (\texttt{vectorized} option) after a warm-up phase consisting of 200 samples and the \texttt{max\_tree\_depth} has been set to 7 by default (ie.
+128 steps for each iteration).
+The job has been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or NVidia V100 (32GB RAM) at the CCIN2P3 Center.
 
 The results in terms of sampling efficiency are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1}.
-The confidential level iso-contours are all in agreement with the \texttt{cobaya} sampler with  significantly larger efficiency.
-{\color{red} FIGURE?}  It is worth to point out that the mass matrix structuring is a very effective tuning to increase the sampling efficiency (gain of about a factor 2).
-The elapse time was 20 hours (5 hours) for run "0" ("1") showing a linear scaling with respect to the number of steps per iteration, while the sampling efficiencies of both runs are very similar, so the user is highly encouraged to decrease the \texttt{max\_tree\_depth} parameter.
+The confidential level iso-contours are all in agreement with the \texttt{cobaya} sampler and are reported as the blue curves in figures presented in section \ref{sec-Neural-Reparametrisation}.
+It is worth to point out that the mass matrix structuring is a very effective when NUTS is run stand-alone as it increases the sampling efficiency by about a factor 2.
+The elapse time was 20 hours for the default setting (\texttt{max\_tree\_depth=7}).
+Anticipating the section \ref{sec-Neural-Reparametrisation} devoted to the Neural Reparametrisation which is an effective alternative, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
+This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Tab.~\ref{tab-ESS-NUTS_SVI-1}), so the user is highly encouraged to tune this critical parameter.
 
@@
 \includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200.pdf}
-\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 22 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in..., and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
+\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 22 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in section \ref{sec-NUTS}, and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
 \label{fig_svi_nuts_1}
@@
 \includegraphics[width=\columnwidth]{figures/SVI_NUTS_10x200x1000.pdf}
-\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 10 chains of 200 warm followed by 1000 samples for the NUTS sampler after SVI optimization.}
+\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 10 chains of 200 warm followed by 1000 samples for the NUTS sampler (\texttt{max\_tree\_depth=5}) after SVI optimization.}
 \label{fig_svi_nuts_3}
@@
 Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (\texttt{vectorized} option) with 1000 samples each.
+All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
+
 The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
-Notice that the \texttt{max\_tree\_depth} parameter has been set to 5 instead of the default value of 10 to speed up running.
-The results of the evolution of the contours compared to the NUTS sampling  run "(1)" (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
-The  agreement is also rather good for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.
-Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1}.
-On can see SVI followed by Neutral Parametrisation gives also high efficiency without the need by the user to setup a particular a priori mass matrix structure (use \texttt{dense} matrix option).
+The results of the evolution of the 2D parameters contour levels compared to the default NUTS sampling (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
+The agreement is rather good even for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.
+Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Parametrisation gives high efficiency as well without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option).
 %
@@
 The question that one may naturally arises are what is the HMC/NUTS gain compared to the well known standard \texttt{cobaya} sampler? 
-One may argue that number of efficient sample per  model call is a right metric to consider, ie. if one demands $N_s$ samples then  
+In one hand, one may argue considering the elapse time of execution that the number of efficient sample per model call is a right metric to consider: 
 \begin{equation}
-    \eta = \frac{n_{eff}}{#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
+    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
 \end{equation}
-with $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation.
-Then, for \texttt{cobaya} one roughly gets $\varepsilon\approx 1\%$ with $n_{steps}=1$ while with the NUTS sampler $\varepsilon\approx 50\%$ but with $n_{steps}=2^5$ or more.
-So, the $\eta$ values for both methods are at the end of the day comparable (ie.
+with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation.
+Then, for \texttt{cobaya} one roughly gets $\varepsilon\approx 1\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more.
+At the end of the day, the $\eta$ values for both methods are comparable (ie.
 about $1\%$).
-So, at first sight it seems that there might be no gain to use HMC/NUTS sampler compared to standard MCMC sampler.
+So, at first sight considering the elapse time that there might be no gain to use HMC/NUTS sampler compared to standard MCMC sampler.
 
-But the above metric is one aspect of sample generation.
+On the other hand, the above metric is one aspect of the sample generation.
 One should have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
@@
 So, the authors think that it is worth considering methods that by construction are more efficient.
-Moreover, using SVI with reparametrisation offers a way to undertake a progressive validation of the model with rather modest sample size (eg.
+Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of the model with rather modest sample size (eg.
 200 samples) by producing good quality marginal contours in few hours.
-This validation phase can save time before producing sizeable batch for final analysis.
+In practice, this validation phase can save time before producing sizeable batch for final analysis.
+The authors have not investigated very high dimensional ($O(10^2)$ parameters) as well as multi-modal analysis but first HMC/NUTS sampling exploit the geometry of the typical set of the posterior distribution contrary to the standard random walk of Metropolis-Hasting sampling and further more using reparametrisation (eg.
+\cite{2019arXiv190303704H}) one can adapt to poor geometry cases.
 %
@@
 \caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
-{\color{red} 
-(0) "F.L run" are computed from NUTS sampler with 16 parallel chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (see text) The numbers are from \texttt{n\_eff/16,000};
-}
-(1) "J.E run" are computed from NUTS sampler 10 with parallel chains of 1,000 samples each and 200 warm-up nd \texttt{max\_tree\_depth=5} (see text); (2) are the results of SVI followed by NUTS sampler of setup (1) after a Neural Transform (see text).}
+%{\color{red} 
+%(0) "F.L run" are computed from NUTS sampler with 16 %parallel chains of 1,000 samples each and 200 warm-up %and \texttt{max\_tree\_depth=7} (see text) The numbers %are from \texttt{n\_eff/16,000};
+%}
+(1): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Sec.~\ref{sec-NUTS}); (2) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 10 chains of 1,000 samples each and 200 warm-up (Sec.~\ref{sec-Neural-Reparametrisation}).}
 \label{tab-ESS-NUTS_SVI-1}
@@
 % (0)  JEC extrait de https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/blob/master/notebooks/VectorizedNumPyro_block.ipynb
-\color{red}(0) & $51.5$ &  $49.5$     & $39.3$     & $39.3$ & $52.0$ &  $50.2$ & $69.8$ & $46.2$\\
+%\color{red}(0) & $51.5$ &  $49.5$     & $39.3$     & $39.3$ & $52.0$ &  $50.2$ & $69.8$ & $46.2$\\
 % (1) & (2) from JEC independant runs @ CCIN2P3

### 2022-02-17 paper.tex
Rename Neural Reparametrisation into Neural Transport. Add in the introduction an list of some JAX based libraries.

@@
 \newcommand{\numpyro}{\texttt{NumPyro}}
-\newcommand{\bydef}{\overset{def.}{\equiv}}
+\newcommand{\bydef}{:=}
 
@@
 
+Concerning JAX ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing, we can list some libraries that the authors have used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
+
 \subsection{Background cosmology}
@@
 In the following sections we briefly review the statistical methods used to make some cosmological parameter inference.
-Namely we have used \textit{Stochastic Variational Inference} (aka SVI), \textit{Hybrid Hamiltonian Monte Carlo} (aka HMC), a variant of HMC knows as \textit{No-U-Turn} (aka NUTS), and finally the use of  \textit{Neutral Reparametrization}.
+Namely we have used \textit{Stochastic Variational Inference} (aka SVI), \textit{Hybrid Hamiltonian Monte Carlo} (aka HMC), a variant of HMC knows as \textit{No-U-Turn} (aka NUTS), and finally the use of  \textit{Neural Transport}.
 %
@@
 \end{equation}
-where the \textit{guide}, in the \numpyro\ word, may be a multi-variate gaussian distribution (MVN) for instance.
+where the \textit{guide} in the \numpyro\ library may be a multi-variate Gaussian distribution (MVN) for instance.
 
@@
 The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI.
-Even, if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to  a NUTS sampler which is described in section \ref{sec-Neural-Reparametrisation} as \textit{Neural Reparametrisation}.
+Even, if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to  a NUTS sampler which is described in section \ref{sec-Neural-Reparametrisation} as \textit{Neural Transport}.
 %
@@
 The elapse time was 20 hours for the default setting (\texttt{max\_tree\_depth=7}).
-Anticipating the section \ref{sec-Neural-Reparametrisation} devoted to the Neural Reparametrisation which is an effective alternative, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
+Anticipating the section \ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective alternative, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
 This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Tab.~\ref{tab-ESS-NUTS_SVI-1}), so the user is highly encouraged to tune this critical parameter.
@@
 %
-\subsubsection{Neural Reparametrisation}
+\subsubsection{Neural Transport}
 \label{sec-Neural-Reparametrisation}

### 2022-02-17 paper.tex
move JAX ecosystem paragraph

@@
 \section{Introduction}
+Concerning JAX ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing, we can list some libraries that the authors have used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
 
@@
 % Follows a bit the structure of the CCL library
-
-Concerning JAX ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing, we can list some libraries that the authors have used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
 

### 2022-02-17 paper.tex
Start Introducton: MCMC usage & tools in Cosmological Inference,  Autodiff in ML & Statistics, Implementation as Stan/Tensorflow and recent JAX. Why JAX. and JAX ecosystem.

--- 2022-02-17 paper.tex
@@
 \section{Introduction}
+Monte Carlo Markov Chain Bayesian inference is a widely used technique in Cosmology since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for CMB analysis thanks to software as  \texttt{CosmoMC}\footnote{\url{https://cosmologist.info/cosmomc/readme.html}} \citep{2002PhRvD..66j3511L} as well as   \texttt{Cobaya}\footnote{\url{https://cobaya.readthedocs.io}} \citep{2019ascl.soft10019T,2021JCAP...05..057T} which is the baseline option for the current on going cosmological experiments.
+
+Although these package does not take benefit of recent advance in automatic differentiation (aka \textit{autodiff}) \citep{2015arXiv150205767G, 2018arXiv181105031M} which is been widely used in neural network optimization and more generally for machine learning and computational statistics as for instance Hamiltonian Monte Carlo sampling and Variational Inference, two subjects covered in this paper.
+The \textit{autodiff} has been implemented in widely used libraries as for instance \texttt{Stan}\footnote{\url{https://mc-stan.org/users/documentation}} \citep{JSSv076i01} and \texttt{TensorFlow}\footnote{\url{https://www.tensorflow.org/}}  \citep{tensorflow2015-whitepaper}.
+
+Recently, the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} brings a rapid rise as it can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions which is quite attractive to speed up the learning process.
+\texttt{JAX} offers an easy palatalization mechanism, the Just-In-Time compilation and optimization code for CPU, GPU and TPU hardware which make it very attractive.
+Concerning \texttt{JAX} ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing, we can list some libraries that the authors have used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
 

### 2022-02-17 paper.tex
suppress compilation warnings

@@
 \section{Introduction}
-Monte Carlo Markov Chain Bayesian inference is a widely used technique in Cosmology since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for CMB analysis thanks to software as  \texttt{CosmoMC}\footnote{\url{https://cosmologist.info/cosmomc/readme.html}} \citep{2002PhRvD..66j3511L} as well as   \texttt{Cobaya}\footnote{\url{https://cobaya.readthedocs.io}} \citep{2019ascl.soft10019T,2021JCAP...05..057T} which is the baseline option for the current on going cosmological experiments.
+Monte Carlo Markov Chain Bayesian inference is a widely used technique in Cosmology since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for past and current analysis\footnote{see for instance \url{https://ui.adsabs.harvard.edu/abs/2021JCAP...05..057T/citations}.} thanks to software as  \texttt{CosmoMC}\footnote{\url{https://cosmologist.info/cosmomc/readme.html}} \citep{2002PhRvD..66j3511L} as well as   \texttt{Cobaya}\footnote{\url{https://cobaya.readthedocs.io}} \citep{2019ascl.soft10019T,2021JCAP...05..057T}.
 
@@
 \texttt{JAX} offers an easy palatalization mechanism, the Just-In-Time compilation and optimization code for CPU, GPU and TPU hardware which make it very attractive.
-Concerning \texttt{JAX} ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing, we can list some libraries that the authors have used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
+Although \texttt{JAX} library contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as exemple libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing.
+We can list some libraries that the authors have used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
+
+In this context we have developed the \texttt{jax-cosmo} library \url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
     \centering
@@
 a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
-far from optimal.  We compare to the highly-optimized {\sc Cobaya} 
-Metropolis-Hastings implementation \citep{cobaya2,cobaya1}, initializing it with the Fisher matrix as a proposal and
+far from optimal.  We compare to the highly-optimized \texttt{Cobaya} 
+Metropolis-Hastings implementation  %\citep{cobaya2,cobaya1} JEC it may preferable to have a single up to date reference in JCAP
+\citep{2019ascl.soft10019T,2021JCAP...05..057T}, initializing it with the Fisher matrix as a proposal and
 tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.
@@
 \caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
-Monte Carlo sampler with JAX-Cosmo (blue), compared to the Cobaya Metropolis-Hastings sampler (orange).}
+Monte Carlo sampler with JAX-Cosmo (blue), compared to the \texttt{Cobaya} Metropolis-Hastings sampler (orange).}
 \end{figure}
@@
 The results in terms of sampling efficiency are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1}.
-The confidential level iso-contours are all in agreement with the \texttt{cobaya} sampler and are reported as the blue curves in figures presented in section \ref{sec-Neural-Reparametrisation}.
+The confidential level iso-contours are all in agreement with the \texttt{Cobaya} sampler and are reported as the blue curves in figures presented in section \ref{sec-Neural-Reparametrisation}.
 It is worth to point out that the mass matrix structuring is a very effective when NUTS is run stand-alone as it increases the sampling efficiency by about a factor 2.
@@
 %
-The question that one may naturally arises are what is the HMC/NUTS gain compared to the well known standard \texttt{cobaya} sampler? 
+The question that one may naturally arises are what is the HMC/NUTS gain compared to the well known standard \texttt{Cobaya} sampler? 
 In one hand, one may argue considering the elapse time of execution that the number of efficient sample per model call is a right metric to consider: 

### 2022-02-17 paper.tex
INtro:  begin a  description of the paper

@@
 \newcommand{\bydef}{:=}
+\newcommand{\jaxcosmo}{\texttt{jax-cosmo}}
 
@@
 
-\title{jax-cosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library}
+\title{\jaxcosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library}
 
@@
 
-\author{jax-cosmo contributors}
+\author{\jaxcosmo\ contributors}
 
@@
 \begin{abstract}
-With jax-cosmo, Fisher forecasts are accurate and instantaneous, inference is lightning-fast, survey optimization is trivial. \github
+With \jaxcosmo, Fisher forecasts are accurate and instantaneous, inference is lightning-fast, survey optimization is trivial. \github
 \end{abstract}
@@
 
-In this context we have developed the \texttt{jax-cosmo} library \url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}
+In this context we have developed the \texttt{jax-cosmo} library available on GitHub repository \github\footnote{ \url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}} described in section \ref{sec-jaxcosmo-design}.
+We show how to use this library in different use-cases as rapid and numerically stable Fisher forecasts (Sec.~\ref{sec-fisher-forecast}), optimisation strategy as in the LSST-DESC 3x2pt tomography optimization challenge \citep{2021OJAp....4E..13Z} (Sec.~\ref{sec-FOM-optimisation}), and more advanced gradient-based cosmological inference with Stochastic Variational Inference (Sec.~\ref{sec-SVI}) as well as Hamiltonian Monte carlo and its variant No-U-Turn (Sec.\ref{sec-NUTS}) and a mixture of these methods detailed in section \ref{sec-Neural-Reparametrisation}.
 
@@
 \section{Modeling and implementation}
-
+\label{sec-jaxcosmo-design}
 In this section, we describe the cosmological modeling provided by jax-cosmo, and its implementation in JAX.
@@
 \subsection{Instantaneous Fisher Forecasts}
-
+\label{sec-fisher-forecast}
 \begin{minted}{python}
@@
 \subsection{Survey Optimization by FoM Maximization}
+\label{sec-FOM-optimisation}
 % Needs to mention the tomo challenge, mention that we can backpropagate through a NN and the cosmology model
@@
 \subsection{Stochastic Variational Inference}
+\label{sec-SVI}
 %
@@
 in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
-calculating a Hessian, as in JAX-Cosmo, makes it feasible and efficient.
+calculating a Hessian, as in \jaxcosmo, makes it feasible and efficient.
 
-Below we show an example set of HMC constraints using JAX-Cosmo to constrain a simulated
+Below we show an example set of HMC constraints using \jaxcosmo\ to constrain a simulated
 Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
@@
 \caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
-Monte Carlo sampler with JAX-Cosmo (blue), compared to the \texttt{Cobaya} Metropolis-Hastings sampler (orange).}
+Monte Carlo sampler with \jaxcosmo\ (blue), compared to the \texttt{Cobaya} Metropolis-Hastings sampler (orange).}
 \end{figure}
@@
 
-A variant of the traditional HMC sampler has been introduced in reference \citep{nuts}.
+The No-U-Turn (aka NUTS))  variant of the traditional HMC sampler has been introduced in reference \citep{nuts}.
 It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$.

### 2022-02-18 paper.tex
Add more MCMC packages CosmoSIS, MontePython...

@@
 \section{Introduction}
-Monte Carlo Markov Chain Bayesian inference is a widely used technique in Cosmology since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for past and current analysis\footnote{see for instance \url{https://ui.adsabs.harvard.edu/abs/2021JCAP...05..057T/citations}.} thanks to software as  \texttt{CosmoMC}\footnote{\url{https://cosmologist.info/cosmomc/readme.html}} \citep{2002PhRvD..66j3511L} as well as   \texttt{Cobaya}\footnote{\url{https://cobaya.readthedocs.io}} \citep{2019ascl.soft10019T,2021JCAP...05..057T}.
+Monte Carlo Markov Chain Bayesian inference has been widely used in Cosmology since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for past and current analysis thanks to packages as  \texttt{CosmoMC}\footnote{\url{https://cosmologist.info/cosmomc/readme.html}} \citep{2002PhRvD..66j3511L}, \texttt{CosmoSIS}\footnote{\url{http://bitbucket.org/joezuntz/cosmosis}} \citep{2015A&C....12...45Z}, \texttt{MontePython}\footnote{\url{https://github.com/brinckmann/montepython_public}} \citep{2019PDU....24..260B} as well as   \texttt{Cobaya}\footnote{\url{https://cobaya.readthedocs.io}} \citep{2019ascl.soft10019T,2021JCAP...05..057T}.
+See for instance the list of citations of these popular packages.
+Note that some direct optimization of likelihood function has also been successfully undertaken for instance for some Planck analysis \citep{2014A&A...566A..54P},
 
-Although these package does not take benefit of recent advance in automatic differentiation (aka \textit{autodiff}) \citep{2015arXiv150205767G, 2018arXiv181105031M} which is been widely used in neural network optimization and more generally for machine learning and computational statistics as for instance Hamiltonian Monte Carlo sampling and Variational Inference, two subjects covered in this paper.
-The \textit{autodiff} has been implemented in widely used libraries as for instance \texttt{Stan}\footnote{\url{https://mc-stan.org/users/documentation}} \citep{JSSv076i01} and \texttt{TensorFlow}\footnote{\url{https://www.tensorflow.org/}}  \citep{tensorflow2015-whitepaper}.
+Although the above mentioned MCMC packages are widely used and are very valuable, they do not take benefit of recent advance in automatic differentiation (aka \textit{autodiff}) \citep{2015arXiv150205767G, 2018arXiv181105031M} which is widely used in neural network optimization and more generally for machine learning and computational statistics as for instance Hamiltonian Monte Carlo sampling and Variational Inference, two subjects covered in this paper.
+The \textit{autodiff} has been implemented in widely used libraries as for instance \texttt{Stan}\footnote{\url{https://mc-stan.org/users/documentation}} \citep{JSSv076i01} and \texttt{TensorFlow}\footnote{\url{https://www.tensorflow.org/}}  \citep{tensorflow2015-whitepaper} or \texttt{PyTorch}\footnote{\url{https://pytorch.org/docs/}} \citep{NEURIPS2019_9015}.
 
 Recently, the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} brings a rapid rise as it can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions which is quite attractive to speed up the learning process.
-\texttt{JAX} offers an easy palatalization mechanism, the Just-In-Time compilation and optimization code for CPU, GPU and TPU hardware which make it very attractive.
-Although \texttt{JAX} library contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as exemple libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing.
-We can list some libraries that the authors have used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
+\texttt{JAX} offers an easy palatalization mechanism (\texttt{vmap}), the Just-In-Time compilation (\texttt{jit}) and optimization code for CPU, GPU and TPU hardware thanks to \texttt{XLA}\footnote{\url{https://www.tensorflow.org/xla}} which overall make this new library very attractive.
+Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing.
+We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
 
@@
 \bibliographystyle{elsarticle-harv} 
+\typeout{}
 \bibliography{refs}

### 2022-02-18 paper.tex
Updates from Overleaf

@@
 Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler.
-The SVI optimization requires 20,000 steps and has been performed in about 2 hours on a Nvidia V100 GPU (32 GB RAM).
+
+The SVI optimization has been performed with 20,000 steps to ensure a good convergence of the ELBO loss but no particular tuning of the optimizer has been undertaken to decrease the number of steps.
+It takes about 2 hours on a Nvidia V100 GPU (32 GB RAM).
 Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (\texttt{vectorized} option) with 1000 samples each.
 All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
+The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
 
-The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
-The results of the evolution of the 2D parameters contour levels compared to the default NUTS sampling (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
-The agreement is rather good even for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters as show on figure \ref{fig_svi_nuts_4}.
-Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Parametrisation gives high efficiency as well without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option).
+The results of the evolution of the 2D parameter contours for some cosmological parameters compared to the default NUTS sampling (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
+The agreement is quite good even with a rather low number of samples (ie.
+200) as shown on figure \ref{fig_svi_nuts_1} and even for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters (Fig.~\ref{fig_svi_nuts_4}).
+It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a worse behaviour.
+Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transprt gives high efficiency as well without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option).
 %

### 2022-03-04 paper.tex
relecture de la discussion

@@
 It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a worse behaviour.
-Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transprt gives high efficiency as well without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option).
+Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency like the default NUTS settings, without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option).
 %
@@
 %
-The question that one may naturally arises are what is the HMC/NUTS gain compared to the well known standard \texttt{Cobaya} sampler? 
-In one hand, one may argue considering the elapse time of execution that the number of efficient sample per model call is a right metric to consider: 
+The key question is what is the HMC/NUTS gain compared to the well known standard \texttt{Cobaya} sampler? 
+In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
 \begin{equation}
@@
 Then, for \texttt{cobaya} one roughly gets $\varepsilon\approx 1\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more.
-At the end of the day, the $\eta$ values for both methods are comparable (ie.
+At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie.
 about $1\%$).
-So, at first sight considering the elapse time that there might be no gain to use HMC/NUTS sampler compared to standard MCMC sampler.
+So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.
 
-On the other hand, the above metric is one aspect of the sample generation.
-One should have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
+On the other hand, the above metric is may be too crude to cover all aspects of the sample generation.
+One should probably have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
 several days on dedicated infrastructure.
-Moreover, the user is faced to a dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, 2015arXiv151007727O}.
-So, the authors think that it is worth considering methods that by construction are more efficient.
-Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of the model with rather modest sample size (eg.
-200 samples) by producing good quality marginal contours in few hours.
+Moreover, the user is faced to the dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, 2015arXiv151007727O}.
+
+So, the authors think that it is worth considering methods like HMC/NUTS that by construction are more efficient.
+Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of the model with rather modest sample set (eg.
+starting with 200 samples) producing good enough marginal contours in few hours.
 In practice, this validation phase can save time before producing sizeable batch for final analysis.
-The authors have not investigated very high dimensional ($O(10^2)$ parameters) as well as multi-modal analysis but first HMC/NUTS sampling exploit the geometry of the typical set of the posterior distribution contrary to the standard random walk of Metropolis-Hasting sampling and further more using reparametrisation (eg.
-\cite{2019arXiv190303704H}) one can adapt to poor geometry cases.
+The authors have not investigated very high dimensional ($O(10^2)$ parameters) as well as multi-modal analysis, but the key argument in favour of HMC/NUTS sampling is that it exploit the geometry of the typical set of the posterior distribution contrary to the standard random walk of Metropolis-Hasting sampling.
+Furthermore, using reparametrisation  one can adapt to poor geometry cases (eg.
+\cite{2019arXiv190303704H}).
 %

### 2022-03-04 paper.tex
polish

@@
 
-Although the above mentioned MCMC packages are widely used and are very valuable, they do not take benefit of recent advance in automatic differentiation (aka \textit{autodiff}) \citep{2015arXiv150205767G, 2018arXiv181105031M} which is widely used in neural network optimization and more generally for machine learning and computational statistics as for instance Hamiltonian Monte Carlo sampling and Variational Inference, two subjects covered in this paper.
+Even if the above mentioned MCMC packages are widely used and are very valuable, they do not take benefit of recent advance in automatic differentiation (aka \textit{autodiff}) \citep{2015arXiv150205767G, 2018arXiv181105031M} which is widely used in neural network optimization and more generally for machine learning and computational statistics as for instance Hamiltonian Monte Carlo sampling and Variational Inference, two subjects covered in this paper.
 The \textit{autodiff} has been implemented in widely used libraries as for instance \texttt{Stan}\footnote{\url{https://mc-stan.org/users/documentation}} \citep{JSSv076i01} and \texttt{TensorFlow}\footnote{\url{https://www.tensorflow.org/}}  \citep{tensorflow2015-whitepaper} or \texttt{PyTorch}\footnote{\url{https://pytorch.org/docs/}} \citep{NEURIPS2019_9015}.

### 2022-05-25 paper.tex
(JEC) merge the Cobaya/NUTS/SVI in one figure. Reduce HMC paragraphe. Starts to rephrase the comparison in the SVI paragraph.

@@
 %
-In the following sections we briefly review the statistical methods used to make some cosmological parameter inference.
-Namely we have used \textit{Stochastic Variational Inference} (aka SVI), \textit{Hybrid Hamiltonian Monte Carlo} (aka HMC), a variant of HMC knows as \textit{No-U-Turn} (aka NUTS), and finally the use of  \textit{Neural Transport}.
+In the following sections we review some statistical methods used to make the cosmological parameter inference with \jaxcosmo.
+We have used the \textit{Stochastic Variational Inference} (aka SVI), a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (aka NUTS), and finally NUTS after a \textit{Neural Transport} SVI.
+All  the methods have been implemented using the \numpyro Probabilistic Programming Language (PPL).
 %
@@
 \end{align}
-which defines the evidence lower bound (ELBO) that one tends to maximize to get the $\lambda$ values and the  highest lower bound on the logarithm of the marginal probability of the observations $\log p(\mathcal{D})$.
-So, the optimal variational distribution satisfies
+which defines the \textit{evidence lower bound} (aka ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
 \begin{equation}
@@
 The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}).
-Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ ($\lambda=(\mu, \Sigma)$).
+Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
+a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
 One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg.
 $\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$.
-The Jacobien of this distribution is 1, so the cost function reads
+The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
 \begin{multline}
@@
 \end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the Gaussian distribution which is independent of the model, so its gradient can be computed once for all and reused in user model. Then,
+where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model. Then,
 to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
@@
 
-\subsubsection{Introduction}
+\subsubsection{Vanilla HMC}
 
@@
 
-HMC samples spaces by generating particle tracjectories through the space, using the log-posterior
-as the negative potential energy of a particle at each point $q$ in the space.  At each sample a trajectory are initialized
-with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration:
+HMC samples spaces by generating particle trajectories through the space, using the log-posterior as the negative potential energy of a particle at each point $q$ in the space.
+Associated to $q$, we introduce an auxiliary $p$ variable in the spirit of Hamiltonian momentum such that
+\begin{equation}
+- \log{\cal P}(q) = V(q) + cte; \quad H(q,p) = V(q) + U(p)
+\end{equation}
+where $U(p)$ is a sort of \textit{kinetic} term defined by 
+\begin{equation}
+U(p) = p^T M^{-1} p
+\end{equation}
+At each sample, a trajectory is initialized with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration (so called \textit{leapfrog} algorithm):
 \begin{align}
@@
 \end{align}
-where $M$ is a mass matrix which should be set to approximate the covariance of the posterior. This is also
-used to set the scale of the random initial velocities.
+where $M$ is a mass matrix which should be set to approximate the covariance of the posterior. This is also used to set the scale of the random initial velocities.
 
@@
 some number of integration steps a final point in the trajectory is compared to the initial one,
-and a Metropolis-Hastings acceptance criterion on the total energy $H = p^T M^{-1} p + \log{\cal P}(q)$ is applied.
+and a Metropolis-Hastings acceptance criterion on the total energy $H(q,p)$ is applied.
 If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
@@
 
-Below we show an example set of HMC constraints using \jaxcosmo\ to constrain a simulated
-Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
-five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
-decomposition
-of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
-edge of the parameter space, as recommended in \citep{NIPS2015_8303a79b}.  We use a fixed number (25) of integration steps and 
-a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
-far from optimal.  We compare to the highly-optimized \texttt{Cobaya} 
-Metropolis-Hastings implementation  %\citep{cobaya2,cobaya1} JEC it may preferable to have a single up to date reference in JCAP
-\citep{2019ascl.soft10019T,2021JCAP...05..057T}, initializing it with the Fisher matrix as a proposal and
-tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.
+The tricky point is that the \textit{leapfrog} algorithm needs tuning to set the number of steps as well as the step size of integration.
+The next section addresses this problem thanks to the No-U-Turn HMC version.
 
-The Hamiltonian sampler required around half the number of samples as the Metropolis (70,000 vs 140,000) to reach approximately the same
-convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
-on a CPU.
+%Below we show an example set of HMC constraints using \jaxcosmo\ to constrain a simulated
+%Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
+%five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
+%decomposition
+%of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
+%edge of the parameter space, as recommended in \citep{NIPS2015_8303a79b}.  We use a fixed number (25) of integration steps and 
+%a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
+%far from optimal.  We compare to the highly-optimized \texttt{Cobaya} 
+%Metropolis-Hastings implementation  %\citep{cobaya2,cobaya1} JEC it may preferable to have a single up to date reference in JCAP
+%\citep{2019ascl.soft10019T,2021JCAP...05..057T}, initializing it with the Fisher matrix as a proposal and
+%tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.
+%
+%The Hamiltonian sampler required around half the number of samples as the Metropolis (70,000 vs 140,000) to reach approximately the same
+%convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
+%on a CPU.
 This proof-of-principle example can certainly be hugely improved using various extensions to HMC such as No U-Turn Sampling \citep{nuts} or sampling from the entire trajectories \cite{betancourt}.
 On next generation GPU high-performance systems, the advantage will be large.
-
-\begin{figure}
-\includegraphics[width=0.95\columnwidth]{figures/hmc.pdf}
-\caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
-Monte Carlo sampler with \jaxcosmo\ (blue), compared to the \texttt{Cobaya} Metropolis-Hastings sampler (orange).}
-\end{figure}
-
-We note that a full Boltzmann code accurate enough to calculate the CMB spectra to Planck \citep{planck18}
-or future \citep{s4,simons} accuracy is currently beyond the scope of this project.  
-If this is needed then HMC could use a fast approximation to a CMB theory spectrum, and importance sampling applied to the complete chain.
+%
+%\begin{figure}
+%\includegraphics[width=0.95\columnwidth]{figures/hmc.pdf}
+%\caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
+%Monte Carlo sampler with \jaxcosmo\ (blue), compared to the \texttt{Cobaya} Metropolis-Hastings sampler (orange).}
+%\end{figure}
+%
+%
+%We note that a full Boltzmann code accurate enough to calculate the CMB spectra to Planck \citep{planck18}
+%or future \citep{s4,simons} accuracy is currently beyond the scope of this project.  
+%If this is needed then HMC could use a fast approximation to a CMB theory spectrum, and importance sampling applied to the complete chain.
 
@@
 Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
-So, We have setup \texttt{NUTS} sampler Using the same DES-Y1 modeling as in previous section, with
+After the forward model creation for the DES-Y1 3x2pts exercise described in \ref{app-DESY1}, we have performed
 \begin{itemize}
@@
 
-The results in terms of sampling efficiency are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1}.
-The confidential level iso-contours are all in agreement with the \texttt{Cobaya} sampler and are reported as the blue curves in figures presented in section \ref{sec-Neural-Reparametrisation}.
+The results in terms of sampling efficiency are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1}. The confidential level iso-contours are all in agreement with the highly-optimized \texttt{Cobaya} 
+Metropolis-Hastings implementation  
+%\citep{cobaya2,cobaya1} JEC it may preferable to have a single up to date reference in JCAP
+\citep{2019ascl.soft10019T,2021JCAP...05..057T} (see figure \ref{fig_cobaya_NUTS_SVI}).
 It is worth to point out that the mass matrix structuring is a very effective when NUTS is run stand-alone as it increases the sampling efficiency by about a factor 2.
@@
 
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/Cobaya-NUTS-SVI.pdf}
+\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red), and a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green).}
+\label{fig_cobaya_NUTS_SVI}
+\end{figure}
 %
@@
 
-\begin{figure}
-\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200.pdf}
-\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 22 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in section \ref{sec-NUTS}, and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
-\label{fig_svi_nuts_1}
-\end{figure}
+%\begin{figure}
+%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200.pdf}
+%\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 22 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in section \ref{sec-NUTS}, and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
+%\label{fig_svi_nuts_1}
+%\end{figure}
 
-\begin{figure}
-\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x1000.pdf}
-\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 1 chain of 200 warm followed by 1000 samples for the NUTS sampler after SVI optimization.}
-\label{fig_svi_nuts_2}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\columnwidth]{figures/SVI_NUTS_10x200x1000.pdf}
-\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 10 chains of 200 warm followed by 1000 samples for the NUTS sampler (\texttt{max\_tree\_depth=5}) after SVI optimization.}
-\label{fig_svi_nuts_3}
-\end{figure}
+%\begin{figure}
+%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x1000.pdf}
+%\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 1 chain of 200 warm followed by 1000 samples for the NUTS sampler after SVI optimization.}
+%\label{fig_svi_nuts_2}
+%\end{figure}
+%
+%\begin{figure}
+%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_10x200x1000.pdf}
+%\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 10 chains of 200 warm followed by 1000 samples for the NUTS sampler (\texttt{max\_tree\_depth=5}) after SVI optimization.}
+%\label{fig_svi_nuts_3}
+%\end{figure}
 
@@
 %
-\subsubsection{DES-Y1 Example}
+\subsection{DES-Y1 Example}
 %
@@
 \begin{table}[htb]
-\caption{Priors of the 22 variables of the DES-Y1 of the 3x2pt likelihood (number counts and shear).}
+\caption{Priors of the 21 variables of the DES-Y1 of the 3x2pt likelihood (number counts and shear).}
 \label{tab-DESY1}

### 2022-05-25 paper.tex
Updates from Overleaf

--- 2022-03-04 paper.tex
@@
 
-HMC samples spaces by generating particle tracjectories through the space, using the log-posterior
+HMC samples spaces by generating particle trajectories through the space, using the log-posterior
 as the negative potential energy of a particle at each point $q$ in the space.  At each sample a trajectory are initialized
@@
 
-That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
-but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
+That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing, but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
 in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically

### 2022-05-26 paper.tex
update NUTS/SVI/Neutra paragraphs, Table of ESS. Description of DES-Y1 code exemple of forward/inference modelling using numpyro. Update bib

--- 2022-05-25 paper.tex
@@
 Recently, the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} brings a rapid rise as it can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions which is quite attractive to speed up the learning process.
-\texttt{JAX} offers an easy palatalization mechanism (\texttt{vmap}), the Just-In-Time compilation (\texttt{jit}) and optimization code for CPU, GPU and TPU hardware thanks to \texttt{XLA}\footnote{\url{https://www.tensorflow.org/xla}} which overall make this new library very attractive.
+\texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), the Just-In-Time compilation (\texttt{jit}) and optimization code for CPU, GPU and TPU hardware thanks to \texttt{XLA}\footnote{\url{https://www.tensorflow.org/xla}} which overall make this new library very attractive.
 Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing.
@@
 
+%
 \subsection{Validation against the Core Cosmology Library (CCL)}
-
+To illustrate the different features available with the present version of the library, we have written a  notebook to make  comparison with the well established  Core Cosmology Library \citep{2019ApJS..242....2C}.
+As an example, the Figure \ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function \citep{2012ApJ...761..152T, 2003MNRAS.341.1311S}.
+%
 \begin{figure}[ht]
@@
     \includegraphics[width=\columnwidth]{figures/halofit_pk.png}
-    \caption{Comparison of matter power spectrum between CCL and jax-cosmo.
-\textbf{Example of validation plot} \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo/blob/master/docs/notebooks/CCL_comparison.ipynb}{\faFileCodeO}}
+    \caption{Comparison of matter power spectrum between CCL and jax-cosmo. \textbf{Example of validation plot} \nblink{CCL_comparison}}
     \label{fig:halofit_comparison}
@@
 \label{sec-fisher-forecast}
-\begin{minted}{python}
+\begin{minted}[fontsize=\footnotesize]{python}
 def likelihood(cosmo):
@@
 In the following sections we review some statistical methods used to make the cosmological parameter inference with \jaxcosmo.
-We have used the \textit{Stochastic Variational Inference} (aka SVI), a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (aka NUTS), and finally NUTS after a \textit{Neural Transport} SVI.
+We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (aka NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI).
 All  the methods have been implemented using the \numpyro Probabilistic Programming Language (PPL).
 %
-\subsection{Stochastic Variational Inference}
-\label{sec-SVI}
-%
-Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps.
-If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
-The VI aims to get an approximation of this distribution, ie.
-$p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
-To do so, one uses the Kullback-Leibler divergence of the two distributions $KL(q(z;\lambda)||p(z|\mathcal{D}))$ leading to the following relation
-\begin{align}
-\log p(\mathcal{D}) &= \mathtt{ELBO} +  KL(q(z;\lambda)||p(z|\mathcal{D})) \label{eq-ELBO} \\
-\mathrm{with} \ \mathtt{ELBO} &\equiv -\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right] + \mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right] 
-\end{align}
-which defines the \textit{evidence lower bound} (aka ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
-\begin{equation}
-q(z;\lambda^\ast) = \underset{q(z;\lambda)}{\mathrm{argmax}}\  \mathtt{ELBO} = 
-\underset{\lambda}{\mathrm{argmin}}\ \mathcal{L}(\lambda)
-\end{equation}
-The function $\mathcal{L}(\lambda)$ is the cost function used in practice. It is composed of two parts:
-\begin{equation}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right]}_{model}
-\label{eq-loss-svi-1}
-\end{equation}
-where the \textit{guide} in the \numpyro\ library may be a multi-variate Gaussian distribution (MVN) for instance.
-
-Using the auto-differentation tool, one can use "black-box" guides (aka \textit{automatic differentiation variational inference}).
-As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
-function, taking derivatives, and running a gradient-based or coordinate-ascent optimization.
-First is defined a bijective differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
-One can develop "AutoGuides" (\numpyro\ terminology) that can be adapted to the user models.
-Then, the cost function reads
-\begin{equation}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log p(\xi,\mathcal{D}) \right]}_{model}
-\label{eq-loss-svi-2}
-\end{equation}
-with
-\begin{equation}
-p(\xi,\mathcal{D}) \bydef p(T^{-1}(\xi),\mathcal{D}) |J_{T^{-1}}(\xi)|
-\end{equation}
-which exhibit the Jacobian of the $T^{-1}$ transformation. The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
-
-Now, one is faced with the evaluation of the expectations during gradient descent.
-The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}).
-Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
-a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
-One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg.
-$\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$.
-The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
-\begin{multline}
--\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
-\label{eq-loss-svi-3}
-\end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model. Then,
-to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
-\begin{multline}
--\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
-+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
-\right\}
-+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
-\label{eq-loss-svi-4}
-\end{multline}
-For the use case of the MVN (auto)guide, one gets
-\begin{align}
--\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
-+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
-\right\} \nn \\
--\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
-&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
-\right\}+ (L^{-1})^T
-\end{align}
-The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI.
-Even, if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to  a NUTS sampler which is described in section \ref{sec-Neural-Reparametrisation} as \textit{Neural Transport}.
 %
 \subsection{Hamiltonian Monte Carlo}
-
+%
 \subsubsection{Vanilla HMC}
-
+%
 % Show Joe's vanilla HMC results against Cobaya
@@
     \item second for the warm-up phase we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
-This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in figure \ref{fig_svi_nuts_4}.
+This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_svi_nuts_4}.
 \end{itemize}
-We have setup a NUTS sampling using 16 chains of 1,000 samples each (\texttt{vectorized} option) after a warm-up phase consisting of 200 samples and the \texttt{max\_tree\_depth} has been set to 7 by default (ie.
+The NUTS sampler have been running with 16 chains of 1,000 samples each (\texttt{vectorized} option) after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie.
 128 steps for each iteration).
-The job has been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or NVidia V100 (32GB RAM) at the CCIN2P3 Center.
+The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or a NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
+The elapse time for these experiments was 20 hours.
 
-The results in terms of sampling efficiency are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1}. The confidential level iso-contours are all in agreement with the highly-optimized \texttt{Cobaya} 
-Metropolis-Hastings implementation  
-%\citep{cobaya2,cobaya1} JEC it may preferable to have a single up to date reference in JCAP
-\citep{2019ascl.soft10019T,2021JCAP...05..057T} (see figure \ref{fig_cobaya_NUTS_SVI}).
-It is worth to point out that the mass matrix structuring is a very effective when NUTS is run stand-alone as it increases the sampling efficiency by about a factor 2.
-The elapse time was 20 hours for the default setting (\texttt{max\_tree\_depth=7}).
-Anticipating the section \ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective alternative, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
-This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Tab.~\ref{tab-ESS-NUTS_SVI-1}), so the user is highly encouraged to tune this critical parameter.
+The results in terms of relative effective sample sizes (aka "ESS") are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
+One notices the dramatic improvement of the ESS by about a factor 10 using the NUTS sampler compared to the highly-optimized \texttt{Cobaya}  Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T} which has last about 40 hours on CPU to generate 70,000 samples.
+The CL contours are also in very good agreement.
+It is worth to point out that the mass matrix structuring operation is a very effective as it increases the sampling efficiency by about a factor 2.
+Anticipating the section \ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
+This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Tab.~\ref{tab-ESS-NUTS_SVI-1}), then the user is highly encouraged to tune this critical parameter.
+%
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/Cobaya-NUTS-SVI200.png}
+\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue, 70,000 samples), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red; 16,000 samples), and a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green, 200 samples).}
+\label{fig_cobaya_NUTS_SVI}
+\end{figure*}
+\subsection{Stochastic Variational Inference}
+\label{sec-SVI}
+%
+Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps.
+If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
+The VI aims to get an approximation of this distribution, ie.
+$p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
+To do so, one uses the Kullback-Leibler divergence of the two distributions $KL(q(z;\lambda)||p(z|\mathcal{D}))$ leading to the following relation
+\begin{align}
+\log p(\mathcal{D}) &= \mathtt{ELBO} +  KL(q(z;\lambda)||p(z|\mathcal{D})) \label{eq-ELBO} \\
+\mathrm{with} \ \mathtt{ELBO} &\equiv -\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right] + \mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right] 
+\end{align}
+which defines the \textit{evidence lower bound} (aka ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
+\begin{equation}
+q(z;\lambda^\ast) = \underset{q(z;\lambda)}{\mathrm{argmax}}\  \mathtt{ELBO} = 
+\underset{\lambda}{\mathrm{argmin}}\ \mathcal{L}(\lambda)
+\end{equation}
+The function $\mathcal{L}(\lambda)$ is the cost function used in practice. It is composed of two parts:
+\begin{equation}
+\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right]}_{model}
+\label{eq-loss-svi-1}
+\end{equation}
+where the \textit{guide} in the \numpyro\ library may be a multi-variate Gaussian distribution (MVN) for instance.
 
-%  JEC: the 16 chains X 1000 (max_tree_depth=7) is the run done by Francois. 20 heures sur un GPU TitanXp avec 12 GB de ram 
-% 5h sur V100 10 chains x 1000 (max_tree_depth=5)
+Using the auto-differentation tool, one can use "black-box" guides (aka \textit{automatic differentiation variational inference}).
+As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
+function, taking derivatives, and running a gradient-based or coordinate-ascent optimization.
+First is defined a bijective differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
+One can develop "AutoGuides" (\numpyro\ terminology) that can be adapted to the user models.
+Then, the cost function reads
+\begin{equation}
+\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log p(\xi,\mathcal{D}) \right]}_{model}
+\label{eq-loss-svi-2}
+\end{equation}
+with
+\begin{equation}
+p(\xi,\mathcal{D}) \bydef p(T^{-1}(\xi),\mathcal{D}) |J_{T^{-1}}(\xi)|
+\end{equation}
+which exhibit the Jacobian of the $T^{-1}$ transformation. The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
 
-\begin{figure}
-\includegraphics[width=\columnwidth]{figures/Cobaya-NUTS-SVI.pdf}
-\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red), and a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green).}
-\label{fig_cobaya_NUTS_SVI}
-\end{figure}
+Now, one is faced with the evaluation of the expectations during gradient descent.
+The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}).
+Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
+a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
+One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg.
+$\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$.
+The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
+\begin{multline}
+-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
+\label{eq-loss-svi-3}
+\end{multline}
+where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model. Then,
+to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
+\begin{multline}
+-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
++ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
+\right\}
++ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
+\label{eq-loss-svi-4}
+\end{multline}
+For the use case of the MVN (auto)guide, one gets
+\begin{align}
+-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
++ \nabla_\xi \log|J_{T^{-1}}(\xi)|
+\right\} \nn \\
+-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
+&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
+\right\}+ (L^{-1})^T
+\end{align}
+The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI.
+As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX.
+Now even if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to a NUTS sampler which is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}.
 %
@@
 
+%\begin{figure}
+%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200_bis.pdf}
+%\caption{
+%Constraints (30\%, 68\%, 90\%) on the five lens galaxy bias.
+In blue the NUTS sampling described in section \ref{sec-NUTS}, and in red a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
+%\label{fig_svi_nuts_4}
+%\end{figure}
+
 \begin{figure}
-\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200_bis.pdf}
-\caption{In the same conditions of Fig.~\ref{fig_svi_nuts_1} but for the five lens galaxy bias which are highly correlated.}
-\label{fig_svi_nuts_4}
+\centering
+\includegraphics[width=\columnwidth]{figures/Cobaya-NUTS-SVI200_bis.png}
+\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but for the highly correlated five lens galaxy bias.}
+\label{fig_cobaya_NUTS_SVI_bis}
 \end{figure}
 
-Using the same DES-Y1 simulation as in previous section, we have used the \numpyro\ library to first use a simple Multivariate Normal distribution as \textit{guide} to approximate the true posterior (\textit{model}) to feed a SVI.
-The Adam optimizer has been used with a learning rate set to $10^{-3}$.
-Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler.
+Using the same DES-Y1 simulation as in previous section, we use a MVN distribution as \textit{guide} to approximate the true posterior (\textit{model}) and feed a SVI.
+The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$.
+We have stopped the optimization after 20,000 steps to ensure a good convergence of the ELBO loss, but no particular tuning of the optimizer has been undertaken to decrease the number of steps.
+It takes 2 hours on the NVidia V100 GPU.
 
-The SVI optimization has been performed with 20,000 steps to ensure a good convergence of the ELBO loss but no particular tuning of the optimizer has been undertaken to decrease the number of steps.
-It takes about 2 hours on a Nvidia V100 GPU (32 GB RAM).
-Concerning the NUTS running on the same GPU, the chains have run 200 warm samples, and then 200 samples, 1000 samples on a single chain, and finally we have run 10 chains (\texttt{vectorized} option) with 1000 samples each.
+Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler, the chains have run 200 warm samples, and then 200 samples ("low" schema), 1000 samples on a single chain, and finally we have run 10 chains  with 1000 samples each.
 All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
 The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
+Naturally, more samples lead to better contour precision.
+But what is illustrative is the "low" schema results as shown in the figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
+Notice the good results even for the highly correlated lens bias parameters.
+It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a rather poor behaviour with only 200 samples.
 
-The results of the evolution of the 2D parameter contours for some cosmological parameters compared to the default NUTS sampling (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}.
-The agreement is quite good even with a rather low number of samples (ie.
-200) as shown on figure \ref{fig_svi_nuts_1} and even for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters (Fig.~\ref{fig_svi_nuts_4}).
-It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a worse behaviour.
-Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency like the default NUTS settings, without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option).
+Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to use these schema to ease the model development in its early phase.
 %
@@
 %
-The key question is what is the HMC/NUTS gain compared to the well known standard \texttt{Cobaya} sampler? 
+The key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
 In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
@@
 with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation.
-Then, for \texttt{cobaya} one roughly gets $\varepsilon\approx 1\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more.
+Then, for \texttt{Cobaya} one roughly gets $\varepsilon\approx 3\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more.
 At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie.
@@
 \begin{table}[htb]
-\caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
+\caption{Results of the \textit{relative effective sample size} (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}. The ESS can be larger than 100\% in some cases. 
 %{\color{red} 
@@
 %}
-(1): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Sec.~\ref{sec-NUTS}); (2) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 10 chains of 1,000 samples each and 200 warm-up (Sec.~\ref{sec-Neural-Reparametrisation}).}
+(a): results from Cobaya 70,000 samples;
+(b): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Sec.~\ref{sec-NUTS}); 
+(c) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 1 chain of 200 samples and 200 warm-up (Sec.~\ref{sec-Neural-Reparametrisation}).}
 \label{tab-ESS-NUTS_SVI-1}
@@
 \hline
-% (0)  JEC extrait de https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/blob/master/notebooks/VectorizedNumPyro_block.ipynb
-%\color{red}(0) & $51.5$ &  $49.5$     & $39.3$     & $39.3$ & $52.0$ &  $50.2$ & $69.8$ & $46.2$\\
-% (1) & (2) from JEC independant runs @ CCIN2P3
-(1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\
-(2) & $38.7$ &  $24.8$     & $16.0$     & $18.6$ & $43.3$ & $39.5$ & $21.7$ & $16.6$\\
+(a) &  $3.1$ & $2.5$       & $2.9$      & $2.9$  & $2.6$  & $3.1$  & $3.1$ & $2.8$ \\  
+(b) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\
+(c) & $84.0$ &  $28.0$     & $26.0$     & $20.5$ & $80.0$ & $110.5$ & $58.5$ & $29.5$\\
 \hline
@@
 
-\section{Description of the DES-Y1}
+\section{Description of the DES-Y1 exercise}
 \label{app-DESY1}
+From the DES server\footnote{\url{http://desdr-server.ncsa.illinois.edu/despublic/y1a1_files/chains/2pt_NG_mcal_1110.fit}} we have extracted the $N(z)$ distributions of the sources and normalize them to $[1.47, 1.46, 1.50, 0.73]$ effective number of sources per $\mathrm{arcmin}^2$ for each of the four redshift bins, respectively.
+The lens distributions $N(z)$ for five redshift bins are also extracted.
+The corresponding shapes of the different distributions are performed thanks to \texttt{jc.redshift.kde\_nz} function ("\texttt{jc}" hereafter stands for \texttt{jax\_cosmo}) and are presented in Figure \ref{fig-DESY1-src-lens-redshift}.
+\begin{figure}
+\centering
+\includegraphics[height=3cm]{figures/DESY1-source-redshift.png}
+\includegraphics[height=3cm]{figures/DESY1-lens-redshift.png}
+\caption{Distributions of the sources and lenses for the different redshift bins considered.}
+\label{fig-DESY1-src-lens-redshift}
+\end{figure}
 \begin{table}[htb]
@@
 
+Then, using the \numpyro\ PPL, we have set up a \textit{forward model} using the prior description  Table \ref{tab-DESY1} as for instance the $\Omega_c$ parameter
+\begin{minted}[fontsize=\footnotesize]{python}
+Omega_c = numpyro.sample('Omega_c', Uniform(0.1, 0.9))
+\end{minted}
+Then, we generate mock angular power spectra $C_\ell$ from the auto \& cross correlations of the Number Counts and Weak Lensing probes computed with the \texttt{gaussian\_cl\_covariance\_and\_mean} function in the \texttt{jc.angular\_cl} module.
+One reads
+\begin{minted}[fontsize=\footnotesize]{python}
+cl, C = gaussian_cl_covariance_and_mean(cosmo, ell, probes, 
+                               f_sky=0.25, sparse=True)
+\end{minted} 
+with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} a series of 50 angular modes.
+The generation of the mock data yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
+\begin{minted}[fontsize=\footnotesize]{python}
+fiducial_model = condition(model,
+    {'Omega_c':0.2545, 'sigma8':0.801, 
+    'h':0.682, 'Omega_b':0.0485, 'w0':-1.,'n_s':0.971,
+     'A':0.5,'eta':0.,
+     'm1':0.0,'m2':0.0,'m3':0.0,'m4':0.0,
+     'dz1':0.0,'dz2':0.0,'dz3':0.0,'dz4':0.0,
+     'b1':1.2,'b2':1.4,'b3':1.6,'b4':1.8,'b5':2.0
+      })
+
+with seed(rng_seed=42):
+    data, P, C = fiducial_model()
+\end{minted}
+The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generates the model $C_\ell$ from the priors
+\begin{minted}[fontsize=\footnotesize]{python}
+cl = jc.angular_cl.angular_cl(cosmo, ell, probes).flatten()
+return numpyro.sample('cl', MultivariateNormal(cl, 
+                            precision_matrix=P,
+                            covariance_matrix=C))
+\end{minted}
+These model $C_\ell$ are then conditioned by the mock $C_ell$ as followed
+\begin{minted}[fontsize=\footnotesize]{python}
+observed_model = condition(model, {'cl': data})
+\end{minted}
+Then, as an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed
+\begin{minted}[fontsize=\footnotesize]{python}
+nuts_kernel = numpyro.infer.NUTS(observed_model,
+                     step_size=1e-1, 
+                     init_strategy=numpyro.infer.init_to_median,
+                     dense_mass=True,
+                     max_tree_depth=7)
+
+mcmc = numpyro.infer.MCMC(nuts_kernel, 
+                          num_warmup=200, 
+                          num_samples=1000,
+                          num_chains=16,
+                          chain_method='vectorized',
+                          progress_bar=False)
+
+mcmc.run(jax.random.PRNGKey(42))
+\end{minted}
+More details can be found in the repository of this paper at this link \nblink{VectorizedNumPyro_block}.
 %% If you have bibdatabase file and want bibtex to generate the
@@
 %%
-\section*{Bibliography}
+%\section*{Bibliography}
 \bibliographystyle{elsarticle-harv} 

### 2022-05-26 paper.tex
fix reference error

@@
     \item second for the warm-up phase we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
-This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_svi_nuts_4}.
+This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
 \end{itemize}

### 2022-05-27 paper.tex
typo

@@
 Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing.
-We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming with \texttt{NumPy} that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
+We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming language that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
 
@@
 We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (aka NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI).
-All  the methods have been implemented using the \numpyro Probabilistic Programming Language (PPL).
+All  the methods have been implemented using the \numpyro probabilistic programming language (PPL).
 %

### 2022-05-30 paper.tex
CCIN2P3 acknowlegments
Tentative to write a short version of the Modeling & implementation section. The guideline is not to duplicate CCL article but to make clear what is in  the present version, what is not there that someone can think about (eg. neutrino contrib)...

@@
 \label{sec-jaxcosmo-design}
-In this section, we describe the cosmological modeling provided by jax-cosmo, and its implementation in JAX.
+In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX.
+The general development follows the Core Cosmology Library structure \citep{2019ApJS..242....2C} although it does not include all the options as the present \jaxcosmo\ version is more to be viewed as a proof of concept.
+
+The cosmological parameters are organised in a JAX \textit{pyTree} container\footnote{\url{https://jax.readthedocs.io/en/latest/jax-101/05.1-pytrees.html}} which allows one to use \textit{autodiff} and \texttt{vmap} transparently.
+The present choice of parameters are: $\Omega_{cdm}$ the cold dark matter density fraction, $\Omega_b$ the Byronic matter density fraction, $h$ the Hubble constant divided by $100$~km/s/Mpc, $n_s$ the primordial scalar perturbation spectral index, $\sigma_8$ the variance of matter density perturbations at an $8$~Mpc/h scale, $\Omega_k$ the curvature density fraction, $w_0$ the first order term of dark energy equation and $w_a$ the second order term of dark energy equation of state.
+
+The traditional computation of the different cosmological background follows \cite{2005A&A...443..819P}  with namely the Friedmann equation
+\begin{equation}
+    E^2(a) = \frac{H^2(a)}{H^2_0} = \Omega_m a^{-3} + \Omega_k a^{-2} + \Omega_{de} e^{f(a)}
+\end{equation}
+with $a=1/(1+z)$ the scale factor related to the redshift $z$, $H(a)=\dot{a}/a$ the Hubble parameter with $H_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and
+\begin{equation}
+    f(a) = -3 (1 + w_0 + w_a) \ln(a) + 3 w_a (a - 1)
+\end{equation}
+Noticeably, the relativist contributions of massless neutrinos and photon radiations, as well as the massive neutrino contribution are neglected.
+
+Currently, \jaxcosmo\  implements the \cite{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} (w/o the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}).
+It is worth to mention that to perform \textit{autodiff} we have implemented a series of tools as differentiable ODE (Romberg algorithm) as well as integration quadrature (eg.
+Simpson).
+
+Future version of the library would offer the possibility to call a JAX emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.
+
+To make the bridge with observations, two pytrees are introduced, namely \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing  respectively,  with for each probe a bunch of redshift bins.
+A long this line, the angular power spectra $C_\ell^{pq}(z,z^\prime)$ for the probes $(p,q)$ and for redshift bin window selections $(N(z), N(z^\prime))$ are computed in the Limber approximation.
+Remarkably, all the integral involved in the computation can be easily differentiated with respect to the cosmological parameters thanks to \textit{autodif} and the implementation of the above mentioned integration quadrature.
+An example is given in the context of DES Y1 3x2pts analysis (\ref{app-DESY1}).
 
@@
 
-\subsection{Background cosmology}
+%\subsection{Background cosmology}
 % Usual stuff
 
-\subsection{Growth of perturbations}
+%\subsection{Growth of perturbations}
 % The interesting point to mention here is the differentiable growth ODE
 
-\subsection{Matter power spectrum}
+%\subsection{Matter power spectrum}
 % Here we describe the EH power spectrum, mention possible extensions with emulators
 
-\subsection{Tracers}
+%\subsection{Tracers}
 % Here we describe the kernels, like in CCL
 
-\subsection{Angular power-spectra}
+%\subsection{Angular power-spectra}
 % Limber integration, this can be an opportunity to talk about differentiation through integrals as well.
@@
 %
-\subsection{Validation against the Core Cosmology Library (CCL)}
+%\subsection{Validation against the Core Cosmology Library (CCL)}
 To illustrate the different features available with the present version of the library, we have written a  notebook to make  comparison with the well established  Core Cosmology Library \citep{2019ApJS..242....2C}.
@@
 %
+%\Acknowledgements
+\section*{Acknowledgements}
+Some of the numerical experiments have been conducted at the IN2P3 Computing Center (CC-IN2P3 - Lyon/Villeurbanne - France) funded by the Centre National de la Recherche Scientifique.
 
@@
 \begin{minted}[fontsize=\footnotesize]{python}
+# Define the lensing and number counts probe
+    probes = [jc.probes.WeakLensing(nzs_s_sys, 
+                                    ia_bias=b_ia,
+                       multiplicative_bias=m),
+             jc.probes.NumberCounts(nzs_l, b)]
 cl, C = gaussian_cl_covariance_and_mean(cosmo, ell, probes, 

### 2022-05-31 paper.tex
avoid double citation of CCL and halofit model

@@
 %\subsection{Validation against the Core Cosmology Library (CCL)}
-To illustrate the different features available with the present version of the library, we have written a  notebook to make  comparison with the well established  Core Cosmology Library \citep{2019ApJS..242....2C}.
-As an example, the Figure \ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function \citep{2012ApJ...761..152T, 2003MNRAS.341.1311S}.
+To illustrate the different features available with the present version of the library, we have written a  notebook to make  comparison with the well established  Core Cosmology Library.
+As an example, the Figure \ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function.
 %

### 2022-06-01 paper.tex
add refence to 2018PhRvD..98d3526A: Dark Energy Survey Year 1 Results: Cosmological Constraints from Galaxy Clustering and Weak Lensing

@@
 
-Then, using the \numpyro\ PPL, we have set up a \textit{forward model} using the prior description  Table \ref{tab-DESY1} as for instance the $\Omega_c$ parameter
+Then, using the \numpyro\ PPL, we have set up a \textit{forward model} following the reference \citep{2018PhRvD..98d3526A}.
+The prior descriptions of the different model parameters are described in Table \ref{tab-DESY1}.
+For instance the $\Omega_c$ parameter is treated as random variable as followed
 \begin{minted}[fontsize=\footnotesize]{python}

### 2022-06-01 paper.tex
minor update

@@
 
-Currently, \jaxcosmo\  implements the \cite{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} (w/o the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}).
-It is worth to mention that to perform \textit{autodiff} we have implemented a series of tools as differentiable ODE (Romberg algorithm) as well as integration quadrature (eg.
-Simpson).
+Currently, \jaxcosmo\  implements the \cite{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}.
+No baryonic feedback is considered yet.
+It is worth to mention that to perform \textit{autodiff} we have implemented a series of tools as differentiable ODE via Runge-Kutta as well as integration methods as Simpson quadrature and Romberg algorithm.
 

### 2022-06-03 paper.tex
minor
typo Cell

@@
 Currently, \jaxcosmo\  implements the \cite{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}.
-No baryonic feedback is considered yet.
+No Baryon feedback modeling is considered yet.
 It is worth to mention that to perform \textit{autodiff} we have implemented a series of tools as differentiable ODE via Runge-Kutta as well as integration methods as Simpson quadrature and Romberg algorithm.
@@
 \end{minted} 
-with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} a series of 50 angular modes.
+with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} (ie.
+$\ell$) a series of 50 angular modes.
 The generation of the mock data yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
@@
 \end{minted}
-The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generates the model $C_\ell$ from the priors
+The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generates the model $C_{\ell}$ from the priors
 \begin{minted}[fontsize=\footnotesize]{python}
@@
 \end{minted}
-These model $C_\ell$ are then conditioned by the mock $C_ell$ as followed
+These theoretical $C_{\ell}$ are in turn conditioned by the mock $C_\ell$ as followed
 \begin{minted}[fontsize=\footnotesize]{python}

### 2022-06-28 paper.tex
minor
- Abstract : first draft - cite PyMC as new PPL with JAX backend - New sections: CRediT authorship contribution statement & Declaration of Competing Interest following the model of https://www.sciencedirect.com/science/article/pii/S2213133722000373#d1e10839

@@
 
+\newcommand{\FL}[1]{{\color{cyan}FL: #1}}
+
 \journal{Astronomy and Computing}
@@
 \begin{abstract}
-With \jaxcosmo, Fisher forecasts are accurate and instantaneous, inference is lightning-fast, survey optimization is trivial. \github
+The recent JAX library creates a new  ecosystem in which the developments of probabilistic programming software take benefit of the attractive speed up as batch acceleration, Just-In-Time compilation and optimization of code for different hardware (CPU, GPU, TPU).
+The JAX automatic differentiation mechanism makes more robust algorithms used in probability density sampling, eg.
+Hamiltonian like Monte Carlo as well as Variotional Inference.
+The current development of Bayesian Inference (with of without likelihood) with large data sets in Cosmology rises the opportunity to develop the \texttt{jax-cosmo} library \github.
+The current version allows to compute linear and non linear power spectra (\texttt{halofit}) with Eisenstein and Hu transfer function, as well as angular power spectra ($C_\ell$) in Limber approximation for galaxy and weak lensing probes.
+The comparison with the Core Cosmological Library is exposed.
+Then, we have conducted in the context of DES Y1 3x2pts analysis a series of cosmological parameters as well as nuisance parameters estimations using different MCMC methods using \numpyro\ as PPL framework.
+The mathematical background of the different methods is detailed.
+As result of the numerical experiments, We show a clear improvement using HMC compared to Metreoplosis-Hasting generation and that the Normalizing Flows using the Neutral Transport is very effective in the early phases of Bayesian model development.
 \end{abstract}
@@
 Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing.
-We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming language that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
+We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library,  \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming language (PPL) that is used in this paper.
+Notice also that other PPL softwares have switched to JAX backend as for instance the latest \texttt{PyMC} v4.0.1\footnote{\url{https://www.pymc.io/projects/docs/en/stable/installation.html}}.
 
@@
 % illustrates why autodiff is interesting from several points of view
+
+\FL{Add description of problem and small demos}
 
@@
 
+\FL{mention something about scaling with parallel MCMC}
+
 \subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
@@
 %
+
+%%%
+
+{\color{red} the next sections are written along the model of \url{https://www.sciencedirect.com/science/article/pii/S2213133722000373} in Astronomy and Computing}
+%%%
+\section*{CRediT authorship contribution statement}
+\textbf{F. Lanusse}, \textbf{J. Zuntz}...
+\textbf{J.E Campagne} Conceptualization, Methodology, Software, Validation, Writing.
+
+\section*{Declaration of Competing Interest}
+The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
+
 %\Acknowledgements

### 2022-07-31 paper_OJA.tex
Open Journal of AstroPhysics test

--- 2022-06-03 paper.tex
@@
-%% 
-%% Copyright 2007-2019 Elsevier Ltd
-%% 
-%% This file is part of the 'Elsarticle Bundle'.
-%% ---------------------------------------------
-%% 
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%% 
-%% The list of all files belonging to the 'Elsarticle Bundle' is
-%% given in the file `manifest.txt'.
-%% 
-%% Template article for Elsevier's document class `elsarticle'
-%% with harvard style bibliographic references
+%\documentclass[10pt,onecolumn,preprintnumbers,amsmath,amssymb,
+%nofootinbib,superscriptaddress,aps,prd,numberedappendix]{openjournal}
 
-%%\documentclass[preprint,12pt,authoryear]{elsarticle}
+\RequirePackage{rotating}
+\PassOptionsToPackage{usenames,dvipsnames}{xcolour}
+\documentclass[twocolumn,twocolappendix,nofootinbib]{openjournal}
 
-%% Use the option review to obtain double line spacing
-%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
+% Add your own macros here:
+\pdfoutput=1 %for arXiv submission
+%\usepackage{amsmath,amssymb,amstext}
+\usepackage{amsmath,amstext}
+\usepackage[T1]{fontenc}
+\usepackage{apjfonts}
+\usepackage{ae,aecompl}
+\usepackage[utf8]{inputenc}
+\usepackage{hyperref}
+\usepackage[figure,figure*]{hypcap}
+\usepackage{natbib}
+\usepackage{url}
+\usepackage{mdwlist}
+%\usepackage{listings}
+%\usepackage{rotating}
+\usepackage{multirow}
+\urlstyle{same}
 
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times,authoryear]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,authoryear]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
-%% \documentclass[final,5p,times,authoryear]{elsarticle}
-\documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
+\usepackage{lineno}
 
-%% For including figures, graphicx.sty has been loaded in
-%% elsarticle.cls. If you prefer to use the old commands
-%% please give \usepackage{epsfig}
+%\documentclass[twocolumn,iop]{openjournal}
+%\usepackage{newtxtext,newtxmath}
+%\usepackage{amsmath}
+%\usepackage{amssymb}
+%\usepackage{graphicx}
+%\usepackage{dcolumn}
+%\usepackage{bm}
+\usepackage[dvipsnames]{xcolor}
+%\usepackage{xspace}
+%\usepackage{ulem}
+%% The amssymb package provides various useful mathematical symbols
 
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{minted}
-\usepackage{amssymb}
-\usepackage{amsmath}
-%% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
-\usepackage{aas_macros}
-\usepackage{fontawesome}
+%\usepackage{fontawesome}
 %% The lineno packages adds line numbers. Start line numbering with
@@
 
-\usepackage{xcolor}
 \definecolor{xlinkcolor}{cmyk}{1,1,0,0}
@@
 
+%%%% sepackage{minted}
+
 \newcommand{\nblink}[1]{\href{https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/blob/master/notebooks/#1.ipynb}{\faFileCodeO}}
@@
 
-\journal{Astronomy and Computing}
+\newcommand{\FLan}[1]{{\color{cyan}FLan: #1}}
 
@@
 \begin{document}
+\journalinfo{The Open Journal of Astrophysics}
+\submitted{submitted xxx, revised yyy, accepted zzz}
 
-\begin{frontmatter}
-
-%% Title, authors and addresses
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for theassociated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for theassociated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for theassociated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-\title{\jaxcosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library}
+\title{\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library}
 
@@
 
-\author{\jaxcosmo\ contributors}
+\author{
+J.~E. Campagne$^{1,\ast}$, 
+F. Lanusse$^2$, 
+J. Zuntz$^3$,\\
+\jaxcosmo\ contributors}
+\thanks{$^\ast$jean-eric.campagne@ijclab.in2p3.fr}
 
-\address{}
+\affiliation{$^1$Universit́e Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
+
+\date{\today}
 
 \begin{abstract}
-With \jaxcosmo, Fisher forecasts are accurate and instantaneous, inference is lightning-fast, survey optimization is trivial. \github
+The recent JAX library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems.
+In particular, JAX exposes an \textit{automatic differentiation} mechanism, which can take a python function that uses JAX and return its (vector) derivative.
+This gives us the opportunity to apply a range of powerful but otherwise unfeasible algorithms used in Bayesian inference, such as Hamiltonian Monte Carlo (HMC) and Variational Inference.
+
+To take advantage of these possibilities within cosmology we have developed the \texttt{jax-cosmo} library, which implements background evolution, linear and non linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes.
+We discuss algorithms made possible by this library, and present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical usage.
+
+We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that the Normalizing Flows using the Neural Transport is a promising methodology.
+;in the early phases of Bayesian model development JAZ not sure what this means so I left it out
+
 \end{abstract}
 
-%%Graphical abstract
-%\begin{graphicalabstract}
-%\includegraphics{grabs}
-%\end{graphicalabstract}
-
-%%Research highlights
-%\begin{highlights}
-%\item Research highlight 1
-%\item Research highlight 2
-%\end{highlights}
-
-\begin{keyword}
-%% keywords here, in the form: keyword \sep keyword
-
-%% PACS codes here, in the form: \PACS code \sep code
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
-\end{keyword}
-
-\end{frontmatter}
+%\keywords{Suggested keywords}%Use showkeys class option if keyword
+                              %display desired
+\maketitle
 
@@
 \section{Introduction}
-Monte Carlo Markov Chain Bayesian inference has been widely used in Cosmology since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for past and current analysis thanks to packages as  \texttt{CosmoMC}\footnote{\url{https://cosmologist.info/cosmomc/readme.html}} \citep{2002PhRvD..66j3511L}, \texttt{CosmoSIS}\footnote{\url{http://bitbucket.org/joezuntz/cosmosis}} \citep{2015A&C....12...45Z}, \texttt{MontePython}\footnote{\url{https://github.com/brinckmann/montepython_public}} \citep{2019PDU....24..260B} as well as   \texttt{Cobaya}\footnote{\url{https://cobaya.readthedocs.io}} \citep{2019ascl.soft10019T,2021JCAP...05..057T}.
-See for instance the list of citations of these popular packages.
-Note that some direct optimization of likelihood function has also been successfully undertaken for instance for some Planck analysis \citep{2014A&A...566A..54P},
+Bayesian inference has been widely used in cosmology in the form of Monte Carlo Markov Chains (MCMC) since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for past and current analysis thanks partly to packages as  \texttt{CosmoMC} \citep{2002PhRvD..66j3511L}, \texttt{CosmoSIS} \citep{2015A&C....12...45Z}, \texttt{MontePython} \citep{2019PDU....24..260B} as well as   \texttt{Cobaya} \citep{2019ascl.soft10019T,2021JCAP...05..057T}; see, for instance, the list of citations to these popular packages for an idea of the wide usage in the community.
 
-Even if the above mentioned MCMC packages are widely used and are very valuable, they do not take benefit of recent advance in automatic differentiation (aka \textit{autodiff}) \citep{2015arXiv150205767G, 2018arXiv181105031M} which is widely used in neural network optimization and more generally for machine learning and computational statistics as for instance Hamiltonian Monte Carlo sampling and Variational Inference, two subjects covered in this paper.
-The \textit{autodiff} has been implemented in widely used libraries as for instance \texttt{Stan}\footnote{\url{https://mc-stan.org/users/documentation}} \citep{JSSv076i01} and \texttt{TensorFlow}\footnote{\url{https://www.tensorflow.org/}}  \citep{tensorflow2015-whitepaper} or \texttt{PyTorch}\footnote{\url{https://pytorch.org/docs/}} \citep{NEURIPS2019_9015}.
+% \footnote{\url{https://cosmologist.info/cosmomc/readme.html}}
+% \footnote{\url{http://bitbucket.org/joezuntz/cosmosis}}
+% \footnote{\url{https://cobaya.readthedocs.io}}
+% \footnote{\url{https://github.com/brinckmann/montepython_public}}
 
-Recently, the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} brings a rapid rise as it can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions which is quite attractive to speed up the learning process.
-\texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), the Just-In-Time compilation (\texttt{jit}) and optimization code for CPU, GPU and TPU hardware thanks to \texttt{XLA}\footnote{\url{https://www.tensorflow.org/xla}} which overall make this new library very attractive.
-Besides the fact that it contains dedicated packages as \texttt{jax.numpy} and \texttt{jax.scipy} which are not simple wrapping, as well as example libraries as \texttt{Stax} for simple but flexible neural network development, it has been the seed for an ecosystem\footnote{A more exhaustive list can be consulted at the following Web site \url{https://project-awesome.org/n2cholas/awesome-jax}.} which is rapidly growing.
-We can list some libraries that the authors have already used as \texttt{Flax}\footnote{\url{https://flax.readthedocs.io/}} \citep{flax2020github} a high-performance neural network library, \numpyro\footnote{\url{https://num.pyro.ai}} \citep{phan2019composable,bingham2019pyro} a probabilistic programming language that is used in this paper, \texttt{JAXopt}\footnote{\url{https://jaxopt.github.io}} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable optimizers, \texttt{Optax}\footnote{\url{https://optax.readthedocs.io}} \citep{optax2020github} a gradient processing and optimization library.
+% Note that some direct optimization of likelihood function has also been successfully undertaken for instance for some Planck analysis \citep{2014A&A...566A..54P}, 
+Since the development of these MCMC packages, major advances have been made in automatic differentiation (\textit{autodiff}) \citep{2015arXiv150205767G, 2018arXiv181105031M}, a set of technologies for transforming pieces of code into their derivatives.
 
-In this context we have developed the \texttt{jax-cosmo} library available on GitHub repository \github\footnote{ \url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}} described in section \ref{sec-jaxcosmo-design}.
-We show how to use this library in different use-cases as rapid and numerically stable Fisher forecasts (Sec.~\ref{sec-fisher-forecast}), optimisation strategy as in the LSST-DESC 3x2pt tomography optimization challenge \citep{2021OJAp....4E..13Z} (Sec.~\ref{sec-FOM-optimisation}), and more advanced gradient-based cosmological inference with Stochastic Variational Inference (Sec.~\ref{sec-SVI}) as well as Hamiltonian Monte carlo and its variant No-U-Turn (Sec.\ref{sec-NUTS}) and a mixture of these methods detailed in section \ref{sec-Neural-Reparametrisation}.
+While these tools have especially been applied to neural network optimization and machine learning in general, they can also enable classical statistical methods that require the derivatives of (e.g.
+likelihood) functions to operate: we consider such methods in this paper.
+\textit{Autodiff} has been implemented in widely used libraries like \texttt{Stan} \citep{JSSv076i01}, \texttt{TensorFlow}  \citep{tensorflow2015-whitepaper}, and  \texttt{PyTorch} \citep{NEURIPS2019_9015}.
+
+A recent entrant to this field is the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} which has undergone rapid development and can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions, offering a speed up to the development process and indeed code runtimes.
+\texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), and just-in-time compilation (\texttt{jit}) and optimization targetting CPU, GPU and TPU hardware thanks to the \texttt{XLA} library that powers TensorFlow .
+These attractive features have driven wide adoption of \texttt{JAX} in computational research, and motivate us to consider its usage in cosmology.
+
+\texttt{JAX} contains bespoke reimplementation of packages such as \texttt{jax.numpy} and \texttt{jax.scipy}, as well as example libraries as \texttt{Stax} for simple but flexible neural network development.
+It has formed the seed for a wider ecosystem of packages, including, for example:
+\texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper.
+Other PPL packages such as\texttt{PyMC} \citep{Salvatier2016} have switched to JAX backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}
+.
+
+% I don't think we need all these URLs, it's getting excessive, especially when they have papers:
+%\footnote{\url{https://flax.readthedocs.io/}}
+% \footnote{\url{https://jaxopt.github.io}}
+% \footnote{\url{https://optax.readthedocs.io}}
+% \footnote{\url{https://num.pyro.ai}}
+% \footnote{\url{https://www.pymc.io/projects/docs/en/stable/installation.html}}
+% \footnote{\url{https://www.tensorflow.org/xla}}
+% \footnote{\url{https://www.tensorflow.org/}}
+% \footnote{\url{https://pytorch.org/docs/}}
+% \footnote{\url{https://mc-stan.org/users/documentation}}
+
+In this context we have developed the open source \texttt{jax-cosmo} library \footnote{ \url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
+The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable JAX functions.
+
+We give an overview of the code's design and contents in Section \ref{sec-jaxcosmo-design}.
+We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more advanced gradient-based cosmological inference with variants of Hamiltonian Monte Carlo and its variant No-U-Turn in Section~\ref{sec:chmc}, and in ML-accelerated Stochastic Variational Inference in Section~\ref{sec-SVI}.
+We discuss and compare these methods in Section~\ref{sec-discussion} and conclude in Section~\ref{sec-conclusion}
 
@@
 
-\section{Example of applications}
+\section{Application: Fisher Forecasting}
+\label{sec-fisher-forecast}
 % illustrates why autodiff is interesting from several points of view
+
+\FLa{Add description of problem and small demos}
 
@@
 \subsection{Instantaneous Fisher Forecasts}
-\label{sec-fisher-forecast}
+
 \begin{minted}[fontsize=\footnotesize]{python}
@@
 
+Fisher forecasts are also commonly used in survey and analysis strategy strategy, where running a full posterior analysis for each possible choice would be unfeasible.
+The inverse area of a projection of a Fisher matrix in parameters of interest can be used as a metric for survey constraining power, such as in the Dark Energy Task Force report \citep{detf}.
+
+\texttt{jax-cosmo} was used in a recent example of such a process, for the  the LSST-DESC 3x2pt tomography optimization challenge \citep{2021OJAp....4E..13Z}, where the best methodology for assignment of galaxies to tomographic bins was assessed using several such figures of merit and related calculations.
+The \texttt{jax-cosmo} metric proved to be stable and fast.
+
 \subsection{Massive Optimal Compression in 3 Lines}
@@
 
-\section{Gradient-based fast cosmological inference}
+\section{Application: Classical Hamiltonian Monte Carlo}
+\label{sec:chmc}
 %
 In the following sections we review some statistical methods used to make the cosmological parameter inference with \jaxcosmo.
-We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (aka NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI).
+We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI).
 All  the methods have been implemented using the \numpyro probabilistic programming language (PPL).
@@
 %
-\subsection{Hamiltonian Monte Carlo}
 %
@@
 \end{figure*}
-\subsection{Stochastic Variational Inference}
+
+\section{Application: Stochastic Variational Inference}
 \label{sec-SVI}
@@
 %
-\subsection{Discussion}
+\section{Discussion}
+\label{sec-discussion}
 %
@@
 
-\section{Conclusion}
+\section{Conclusions \& Prospects}
+\label{sec-conclusion}
+
+\FLa{mention something about scaling with parallel MCMC}
 

### 2022-08-01 paper_OJA.tex
Pass with listing for Python code due to interaction between minted & openjournal author/affiliation

@@
 \usepackage{mdwlist}
-%\usepackage{listings}
 %\usepackage{rotating}
@@
 %\usepackage{bm}
-\usepackage[dvipsnames]{xcolor}
+%\usepackage[dvipsnames]{xcolor}
 %\usepackage{xspace}
@@
 
-\definecolor{xlinkcolor}{cmyk}{1,1,0,0}
-\usepackage[colorlinks = true,
-            linkcolor = xlinkcolor,
-            urlcolor  = xlinkcolor,
-            citecolor = xlinkcolor,
-            anchorcolor = xlinkcolor]{hyperref}
+%\definecolor{xlinkcolor}{cmyk}{1,1,0,0}
+%\usepackage[colorlinks = true,
+%            linkcolor = xlinkcolor,
+%            urlcolor  = xlinkcolor,
+%            citecolor = xlinkcolor,
+%            anchorcolor = xlinkcolor]{hyperref}
 
-%%%% sepackage{minted}
+%%%% usepackage{minted}
+\usepackage{xcolor}
+\definecolor{maroon}{cmyk}{0, 0.87, 0.68, 0.32}
+\definecolor{halfgray}{gray}{0.55}
+\definecolor{ipython_frame}{RGB}{207, 207, 207}
+\definecolor{ipython_bg}{RGB}{247, 247, 247}
+\definecolor{ipython_red}{RGB}{186, 33, 33}
+\definecolor{ipython_green}{RGB}{0, 128, 0}
+\definecolor{ipython_cyan}{RGB}{64, 128, 128}
+\definecolor{ipython_purple}{RGB}{170, 34, 255}
+
+\usepackage{listings}
+
+\lstset{
+    breaklines=true,
+    %
+    extendedchars=true,
+    literate=
+    {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
+    {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
+    {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
+    {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
+    {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
+    {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
+    {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
+    {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
+    {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
+    {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
+    {€}{{\EUR}}1 {£}{{\pounds}}1
+}
+
+%%
+%% Python definition (c) 1998 Michael Weber
+%% Additional definitions (2013) Alexis Dimitriadis
+%% modified by me (should not have empty lines)
+%%
+\lstdefinelanguage{iPython}{
+    morekeywords={access,and,break,class,continue,def,del,elif,else,except,exec,finally,for,from,global,if,import,in,is,lambda,not,or,pass,print,raise,return,try,while},%
+    %
+    % Built-ins
+    morekeywords=[2]{abs,all,any,basestring,bin,bool,bytearray,callable,chr,classmethod,cmp,compile,complex,delattr,dict,dir,divmod,enumerate,eval,execfile,file,filter,float,format,frozenset,getattr,globals,hasattr,hash,help,hex,id,input,int,isinstance,issubclass,iter,len,list,locals,long,map,max,memoryview,min,next,object,oct,open,ord,pow,property,range,raw_input,reduce,reload,repr,reversed,round,set,setattr,slice,sorted,staticmethod,str,sum,super,tuple,type,unichr,unicode,vars,xrange,zip,apply,buffer,coerce,intern},%
+    %
+    sensitive=true,%
+    morecomment=[l]\#,%
+    morestring=[b]',%
+    morestring=[b]",%
+    %
+    morestring=[s]{'''}{'''},% used for documentation text (mulitiline strings)
+    morestring=[s]{"""}{"""},% added by Philipp Matthias Hahn
+    %
+    morestring=[s]{r'}{'},% `raw' strings
+    morestring=[s]{r"}{"},%
+    morestring=[s]{r'''}{'''},%
+    morestring=[s]{r"""}{"""},%
+    morestring=[s]{u'}{'},% unicode strings
+    morestring=[s]{u"}{"},%
+    morestring=[s]{u'''}{'''},%
+    morestring=[s]{u"""}{"""},%
+    %
+    % {replace}{replacement}{lenght of replace}
+    % *{-}{-}{1} will not replace in comments and so on
+    literate=
+    {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
+    {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
+    {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
+    {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
+    {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
+    {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
+    {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
+    {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
+    {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
+    {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
+    {€}{{\EUR}}1 {£}{{\pounds}}1
+    %
+    {^}{{{\color{ipython_purple}\^{}}}}1
+    {=}{{{\color{ipython_purple}=}}}1
+    %
+    {+}{{{\color{ipython_purple}+}}}1
+    {*}{{{\color{ipython_purple}$^\ast$}}}1
+    {/}{{{\color{ipython_purple}/}}}1
+    %
+    {+=}{{{+=}}}1
+    {-=}{{{-=}}}1
+    {*=}{{{$^\ast$=}}}1
+    {/=}{{{/=}}}1,
+    literate=
+    *{-}{{{\color{ipython_purple}-}}}1
+     {?}{{{\color{ipython_purple}?}}}1,
+    %
+    identifierstyle=\color{black}\ttfamily,
+    commentstyle=\color{ipython_cyan}\ttfamily,
+    stringstyle=\color{ipython_red}\ttfamily,
+    keepspaces=true,
+    showspaces=false,
+    showstringspaces=false,
+    %
+    rulecolor=\color{ipython_frame},
+    frame=none,
+    frameround={t}{t}{t}{t},
+    framexleftmargin=6mm,
+    numbers=none,
+    numberstyle=\tiny\color{halfgray},
+    %
+    %
+    backgroundcolor=\color{ipython_bg},
+    %   extendedchars=true,
+    basicstyle=\scriptsize,
+    keywordstyle=\color{ipython_green}\ttfamily,
+}
 
@@
 
-\begin{minted}[fontsize=\footnotesize]{python}
+%\begin{minted}[fontsize=\footnotesize]{python}
+\begin{lstlisting}[language=iPython]
 def likelihood(cosmo):
@@
 F = - jax.hessian(likelihood)(cosmo)
-\end{minted}
+\end{lstlisting}
+%\end{minted}
 
@@
 %%%
+\section*{CRediT authorship contribution statement}
+\textbf{J.E Campagne} Conceptualization, Methodology, Software, Validation, Writing.
+\textbf{F. Lanusse}, \textbf{J. Zuntz}...
 
-{\color{red} the next sections are written along the model of \url{https://www.sciencedirect.com/science/article/pii/S2213133722000373} in Astronomy and Computing}
-%%%
-\section*{CRediT authorship contribution statement}
-\textbf{F. Lanusse}, \textbf{J. Zuntz}...
-\textbf{J.E Campagne} Conceptualization, Methodology, Software, Validation, Writing.
-
-\section*{Declaration of Competing Interest}
-The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
+%\section*{Declaration of Competing Interest}
+%The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
 
@@
 For instance the $\Omega_c$ parameter is treated as random variable as followed
-\begin{minted}[fontsize=\footnotesize]{python}
+%\begin{minted}[fontsize=\footnotesize]{python}
+\begin{lstlisting}[language=iPython]
 Omega_c = numpyro.sample('Omega_c', Uniform(0.1, 0.9))
-\end{minted}
+\end{lstlisting}
+%\end{minted}
+
 Then, we generate mock angular power spectra $C_\ell$ from the auto \& cross correlations of the Number Counts and Weak Lensing probes computed with the \texttt{gaussian\_cl\_covariance\_and\_mean} function in the \texttt{jc.angular\_cl} module.
 One reads
-\begin{minted}[fontsize=\footnotesize]{python}
+%\begin{minted}[fontsize=\footnotesize]{python}
+\begin{lstlisting}[language=iPython]
 # Define the lensing and number counts probe
@@
                                f_sky=0.25, sparse=True)
-\end{minted} 
+\end{lstlisting}
+%\end{minted} 
 with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} (ie.
@@
 The generation of the mock data yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
-\begin{minted}[fontsize=\footnotesize]{python}
+%\begin{minted}[fontsize=\footnotesize]{python}
+\begin{lstlisting}[language=iPython]
 fiducial_model = condition(model,
@@
     data, P, C = fiducial_model()
-\end{minted}
+%\end{minted}
+\end{lstlisting}
+
 The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generates the model $C_{\ell}$ from the priors
-\begin{minted}[fontsize=\footnotesize]{python}
+%\begin{minted}[fontsize=\footnotesize]{python}
+\begin{lstlisting}[language=iPython]
 cl = jc.angular_cl.angular_cl(cosmo, ell, probes).flatten()
@@
                             covariance_matrix=C))
-\end{minted}
+\end{lstlisting}
+%\end{minted}
+
 These theoretical $C_{\ell}$ are in turn conditioned by the mock $C_\ell$ as followed
-\begin{minted}[fontsize=\footnotesize]{python}
+%\begin{minted}[fontsize=\footnotesize]{python}
+\begin{lstlisting}[language=iPython]
 observed_model = condition(model, {'cl': data})
-\end{minted}
+\end{lstlisting}
+%\end{minted}
 Then, as an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed
-\begin{minted}[fontsize=\footnotesize]{python}
+%\begin{minted}[fontsize=\footnotesize]{python}
+\begin{lstlisting}[language=iPython]
 nuts_kernel = numpyro.infer.NUTS(observed_model,
@@
 mcmc.run(jax.random.PRNGKey(42))
-\end{minted}
+\end{lstlisting}
+%\end{minted}
 More details can be found in the repository of this paper at this link \nblink{VectorizedNumPyro_block}.

### 2022-08-01 paper_OJA.tex
tweak the rendering. correct a typo on the SVI paragraph

@@
-%\documentclass[10pt,onecolumn,preprintnumbers,amsmath,amssymb,
-%nofootinbib,superscriptaddress,aps,prd,numberedappendix]{openjournal}
-
-\RequirePackage{rotating}
 \PassOptionsToPackage{usenames,dvipsnames}{xcolour}
-\documentclass[twocolumn,twocolappendix,nofootinbib]{openjournal}
+\documentclass[twocolumn,twocolappendix,nofootinbib,iop]{openjournal}
 
@@
 \pdfoutput=1 %for arXiv submission
-%\usepackage{amsmath,amssymb,amstext}
-\usepackage{amsmath,amstext}
+\usepackage{amsmath,amssymb,amstext}
 \usepackage[T1]{fontenc}
@@
 \usepackage[utf8]{inputenc}
-\usepackage{hyperref}
+%\usepackage{hyperref}
+\usepackage[colorlinks,allcolors=blue]{hyperref}
 \usepackage[figure,figure*]{hypcap}
@@
 \usepackage{mdwlist}
-%\usepackage{rotating}
 \usepackage{multirow}
 \urlstyle{same}
-
 \usepackage{lineno}
-
-%\documentclass[twocolumn,iop]{openjournal}
-%\usepackage{newtxtext,newtxmath}
-%\usepackage{amsmath}
-%\usepackage{amssymb}
-%\usepackage{graphicx}
-%\usepackage{dcolumn}
-%\usepackage{bm}
-%\usepackage[dvipsnames]{xcolor}
-%\usepackage{xspace}
-%\usepackage{ulem}
-%% The amssymb package provides various useful mathematical symbols
-
-%\usepackage{fontawesome}
-%% The lineno packages adds line numbers. Start line numbering with
-%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
-%% for the whole article with \linenumbers.
-%% \usepackage{lineno}
-
-%\definecolor{xlinkcolor}{cmyk}{1,1,0,0}
-%\usepackage[colorlinks = true,
-%            linkcolor = xlinkcolor,
-%            urlcolor  = xlinkcolor,
-%            citecolor = xlinkcolor,
-%            anchorcolor = xlinkcolor]{hyperref}
+\usepackage{fontawesome}
 
@@
     {+}{{{\color{ipython_purple}+}}}1
+    {-}{{{\color{ipython_purple}-}}}1
     {*}{{{\color{ipython_purple}$^\ast$}}}1
@@
     rulecolor=\color{ipython_frame},
-    frame=none,
     frameround={t}{t}{t}{t},
-    framexleftmargin=6mm,
     numbers=none,
@@
     %   extendedchars=true,
-    basicstyle=\scriptsize,
+    %basicstyle=\scriptsize,
+    basicstyle=\ttfamily\footnotesize,
+    columns=fullflexible,
     keywordstyle=\color{ipython_green}\ttfamily,
@@
 
-\newcommand{\FLan}[1]{{\color{cyan}FLan: #1}}
+%%% color Comments 
+\newcommand{\FrL}[1]{{\color{cyan}FL: #1}}
 
@@
 \journalinfo{The Open Journal of Astrophysics}
-\submitted{submitted xxx, revised yyy, accepted zzz}
+%\submitted{submitted xxx, revised yyy, accepted zzz}
 
@@
 
-\affiliation{$^1$Universit́e Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
+\affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
 
@@
 We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that the Normalizing Flows using the Neural Transport is a promising methodology.
-;in the early phases of Bayesian model development JAZ not sure what this means so I left it out
+{\color{red}
+;in the early phases of Bayesian model development JAZ not sure what this means so I left it out}
 
@@
 
-In this context we have developed the open source \texttt{jax-cosmo} library \footnote{ \url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
+In this context we have developed the open source \texttt{jax-cosmo} library \footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
 The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable JAX functions.
@@
 
-\FLa{Add description of problem and small demos}
+\FrL{Add description of problem and small demos}
 
@@
 # parameters
-F = - jax.hessian(likelihood)(cosmo)
+F = -jax.hessian(likelihood)(cosmo)
 \end{lstlisting}
@@
 \begin{multline}
--\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
+-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
 \label{eq-loss-svi-3}
@@
 where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model. Then,
-to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
+to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to\footnote{To simplify the notation, $T^{-1}(S_\lambda^{-1}(\zeta))$ has been replaced by $z$.}
 \begin{multline}
@@
 
-Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to use these schema to ease the model development in its early phase.
+Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to ease the model development in its early phase.
 %
@@
 
-\FLa{mention something about scaling with parallel MCMC}
+\FrL{mention something about scaling with parallel MCMC}
 
@@
           jc.probes.NumberCounts(nzs_l, b)]
-cl, C = gaussian_cl_covariance_and_mean(cosmo, ell, probes, 
+cl, C = gaussian_cl_covariance_and_mean(cosmo, 
+                 ell, probes, 
                  f_sky=0.25, sparse=True)
@@
     data, P, C = fiducial_model()
+\end{lstlisting}
 %\end{minted}
-\end{lstlisting}
 
@@
 \begin{lstlisting}[language=iPython]
-cl = jc.angular_cl.angular_cl(cosmo, ell, probes).flatten()
+cl = jc.angular_cl.angular_cl(cosmo, ell, 
+                              probes).flatten()
 return numpyro.sample('cl', MultivariateNormal(cl, 
@@
 %\section*{Bibliography}
-\bibliographystyle{elsarticle-harv} 
+%\bibliographystyle{elsarticle-harv} 
+\bibliographystyle{aa} 
 \typeout{}

### 2022-08-01 paper_OJA.tex
fix Dark Energy Task force reference

@@
 
-\date{\today}
+%\date{\today}
 
@@
 Fisher forecasts are also commonly used in survey and analysis strategy strategy, where running a full posterior analysis for each possible choice would be unfeasible.
-The inverse area of a projection of a Fisher matrix in parameters of interest can be used as a metric for survey constraining power, such as in the Dark Energy Task Force report \citep{detf}.
+The inverse area of a projection of a Fisher matrix in parameters of interest can be used as a metric for survey constraining power, such as in the Dark Energy Task Force report \citep{2006astro.ph..9591A}.
 

### 2022-08-01 paper_deprecated.tex
fix Dark Energy Task force reference

--- 2022-08-01 paper_OJA.tex
@@
-%\documentclass[10pt,onecolumn,preprintnumbers,amsmath,amssymb,
-%nofootinbib,superscriptaddress,aps,prd,numberedappendix]{openjournal}
+%% 
+%% Copyright 2007-2019 Elsevier Ltd
+%% 
+%% This file is part of the 'Elsarticle Bundle'.
+%% ---------------------------------------------
+%% 
+%% It may be distributed under the conditions of the LaTeX Project Public
+%% License, either version 1.2 of this license or (at your option) any
+%% later version.  The latest version of this license is in
+%%    http://www.latex-project.org/lppl.txt
+%% and version 1.2 or later is part of all distributions of LaTeX
+%% version 1999/12/01 or later.
+%% 
+%% The list of all files belonging to the 'Elsarticle Bundle' is
+%% given in the file `manifest.txt'.
+%% 
+%% Template article for Elsevier's document class `elsarticle'
+%% with harvard style bibliographic references
 
-\RequirePackage{rotating}
-\PassOptionsToPackage{usenames,dvipsnames}{xcolour}
-\documentclass[twocolumn,twocolappendix,nofootinbib]{openjournal}
+%%\documentclass[preprint,12pt,authoryear]{elsarticle}
 
-% Add your own macros here:
-\pdfoutput=1 %for arXiv submission
-%\usepackage{amsmath,amssymb,amstext}
-\usepackage{amsmath,amstext}
-\usepackage[T1]{fontenc}
-\usepackage{apjfonts}
-\usepackage{ae,aecompl}
-\usepackage[utf8]{inputenc}
-\usepackage{hyperref}
-\usepackage[figure,figure*]{hypcap}
-\usepackage{natbib}
-\usepackage{url}
-\usepackage{mdwlist}
-%\usepackage{rotating}
-\usepackage{multirow}
-\urlstyle{same}
+%% Use the option review to obtain double line spacing
+%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}
 
-\usepackage{lineno}
+%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
+%% for a journal layout:
+%% \documentclass[final,1p,times,authoryear]{elsarticle}
+%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
+%% \documentclass[final,3p,times,authoryear]{elsarticle}
+%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
+%% \documentclass[final,5p,times,authoryear]{elsarticle}
+\documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}
 
-%\documentclass[twocolumn,iop]{openjournal}
-%\usepackage{newtxtext,newtxmath}
-%\usepackage{amsmath}
-%\usepackage{amssymb}
-%\usepackage{graphicx}
-%\usepackage{dcolumn}
-%\usepackage{bm}
-%\usepackage[dvipsnames]{xcolor}
-%\usepackage{xspace}
-%\usepackage{ulem}
+%% For including figures, graphicx.sty has been loaded in
+%% elsarticle.cls. If you prefer to use the old commands
+%% please give \usepackage{epsfig}
+
 %% The amssymb package provides various useful mathematical symbols
-
-%\usepackage{fontawesome}
+\usepackage{minted}
+\usepackage{amssymb}
+\usepackage{amsmath}
+%% The amsthm package provides extended theorem environments
+%% \usepackage{amsthm}
+\usepackage{aas_macros}
+\usepackage{fontawesome}
 %% The lineno packages adds line numbers. Start line numbering with
@@
 
-%\definecolor{xlinkcolor}{cmyk}{1,1,0,0}
-%\usepackage[colorlinks = true,
-%            linkcolor = xlinkcolor,
-%            urlcolor  = xlinkcolor,
-%            citecolor = xlinkcolor,
-%            anchorcolor = xlinkcolor]{hyperref}
-
-%%%% usepackage{minted}
 \usepackage{xcolor}
-\definecolor{maroon}{cmyk}{0, 0.87, 0.68, 0.32}
-\definecolor{halfgray}{gray}{0.55}
-\definecolor{ipython_frame}{RGB}{207, 207, 207}
-\definecolor{ipython_bg}{RGB}{247, 247, 247}
-\definecolor{ipython_red}{RGB}{186, 33, 33}
-\definecolor{ipython_green}{RGB}{0, 128, 0}
-\definecolor{ipython_cyan}{RGB}{64, 128, 128}
-\definecolor{ipython_purple}{RGB}{170, 34, 255}
-
-\usepackage{listings}
-
-\lstset{
-    breaklines=true,
-    %
-    extendedchars=true,
-    literate=
-    {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
-    {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
-    {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
-    {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
-    {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
-    {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
-    {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
-    {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
-    {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
-    {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
-    {€}{{\EUR}}1 {£}{{\pounds}}1
-}
-
-%%
-%% Python definition (c) 1998 Michael Weber
-%% Additional definitions (2013) Alexis Dimitriadis
-%% modified by me (should not have empty lines)
-%%
-\lstdefinelanguage{iPython}{
-    morekeywords={access,and,break,class,continue,def,del,elif,else,except,exec,finally,for,from,global,if,import,in,is,lambda,not,or,pass,print,raise,return,try,while},%
-    %
-    % Built-ins
-    morekeywords=[2]{abs,all,any,basestring,bin,bool,bytearray,callable,chr,classmethod,cmp,compile,complex,delattr,dict,dir,divmod,enumerate,eval,execfile,file,filter,float,format,frozenset,getattr,globals,hasattr,hash,help,hex,id,input,int,isinstance,issubclass,iter,len,list,locals,long,map,max,memoryview,min,next,object,oct,open,ord,pow,property,range,raw_input,reduce,reload,repr,reversed,round,set,setattr,slice,sorted,staticmethod,str,sum,super,tuple,type,unichr,unicode,vars,xrange,zip,apply,buffer,coerce,intern},%
-    %
-    sensitive=true,%
-    morecomment=[l]\#,%
-    morestring=[b]',%
-    morestring=[b]",%
-    %
-    morestring=[s]{'''}{'''},% used for documentation text (mulitiline strings)
-    morestring=[s]{"""}{"""},% added by Philipp Matthias Hahn
-    %
-    morestring=[s]{r'}{'},% `raw' strings
-    morestring=[s]{r"}{"},%
-    morestring=[s]{r'''}{'''},%
-    morestring=[s]{r"""}{"""},%
-    morestring=[s]{u'}{'},% unicode strings
-    morestring=[s]{u"}{"},%
-    morestring=[s]{u'''}{'''},%
-    morestring=[s]{u"""}{"""},%
-    %
-    % {replace}{replacement}{lenght of replace}
-    % *{-}{-}{1} will not replace in comments and so on
-    literate=
-    {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
-    {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
-    {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
-    {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
-    {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
-    {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
-    {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
-    {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
-    {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
-    {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
-    {€}{{\EUR}}1 {£}{{\pounds}}1
-    %
-    {^}{{{\color{ipython_purple}\^{}}}}1
-    {=}{{{\color{ipython_purple}=}}}1
-    %
-    {+}{{{\color{ipython_purple}+}}}1
-    {*}{{{\color{ipython_purple}$^\ast$}}}1
-    {/}{{{\color{ipython_purple}/}}}1
-    %
-    {+=}{{{+=}}}1
-    {-=}{{{-=}}}1
-    {*=}{{{$^\ast$=}}}1
-    {/=}{{{/=}}}1,
-    literate=
-    *{-}{{{\color{ipython_purple}-}}}1
-     {?}{{{\color{ipython_purple}?}}}1,
-    %
-    identifierstyle=\color{black}\ttfamily,
-    commentstyle=\color{ipython_cyan}\ttfamily,
-    stringstyle=\color{ipython_red}\ttfamily,
-    keepspaces=true,
-    showspaces=false,
-    showstringspaces=false,
-    %
-    rulecolor=\color{ipython_frame},
-    frame=none,
-    frameround={t}{t}{t}{t},
-    framexleftmargin=6mm,
-    numbers=none,
-    numberstyle=\tiny\color{halfgray},
-    %
-    %
-    backgroundcolor=\color{ipython_bg},
-    %   extendedchars=true,
-    basicstyle=\scriptsize,
-    keywordstyle=\color{ipython_green}\ttfamily,
-}
+\definecolor{xlinkcolor}{cmyk}{1,1,0,0}
+\usepackage[colorlinks = true,
+            linkcolor = xlinkcolor,
+            urlcolor  = xlinkcolor,
+            citecolor = xlinkcolor,
+            anchorcolor = xlinkcolor]{hyperref}
 
@@
 
-\newcommand{\FLan}[1]{{\color{cyan}FLan: #1}}
+\newcommand{\FL}[1]{{\color{cyan}FL: #1}}
+
+\journal{Astronomy and Computing}
 
@@
 \begin{document}
-\journalinfo{The Open Journal of Astrophysics}
-\submitted{submitted xxx, revised yyy, accepted zzz}
+
+\begin{frontmatter}
+
+%% Title, authors and addresses
+
+%% use the tnoteref command within \title for footnotes;
+%% use the tnotetext command for theassociated footnote;
+%% use the fnref command within \author or \address for footnotes;
+%% use the fntext command for theassociated footnote;
+%% use the corref command within \author for corresponding author footnotes;
+%% use the cortext command for theassociated footnote;
+%% use the ead command for the email address,
+%% and the form \ead[url] for the home page:
+%% \title{Title\tnoteref{label1}}
+%% \tnotetext[label1]{}
+%% \author{Name\corref{cor1}\fnref{label2}}
+%% \ead{email address}
+%% \ead[url]{home page}
+%% \fntext[label2]{}
+%% \cortext[cor1]{}
+%% \address{Address\fnref{label3}}
+%% \fntext[label3]{}
 
@@
 
-\author{
-J.~E. Campagne$^{1,\ast}$, 
-F. Lanusse$^2$, 
-J. Zuntz$^3$,\\
-\jaxcosmo\ contributors}
-\thanks{$^\ast$jean-eric.campagne@ijclab.in2p3.fr}
+\author{\jaxcosmo\ contributors}
 
-\affiliation{$^1$Universit́e Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
-
-\date{\today}
+\address{}
 
@@
 
-%\keywords{Suggested keywords}%Use showkeys class option if keyword
-                              %display desired
-\maketitle
+%%Graphical abstract
+%\begin{graphicalabstract}
+%\includegraphics{grabs}
+%\end{graphicalabstract}
+
+%%Research highlights
+%\begin{highlights}
+%\item Research highlight 1
+%\item Research highlight 2
+%\end{highlights}
+
+\begin{keyword}
+%% keywords here, in the form: keyword \sep keyword
+
+%% PACS codes here, in the form: \PACS code \sep code
+
+%% MSC codes here, in the form: \MSC code \sep code
+%% or \MSC[2008] code \sep code (2000 is the default)
+
+\end{keyword}
+
+\end{frontmatter}
 
@@
 
-\FLa{Add description of problem and small demos}
+\FL{Add description of problem and small demos}
 
@@
 
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
+\begin{minted}[fontsize=\footnotesize]{python}
 def likelihood(cosmo):
@@
 F = - jax.hessian(likelihood)(cosmo)
-\end{lstlisting}
-%\end{minted}
+\end{minted}
 
@@
 
-\FLa{mention something about scaling with parallel MCMC}
+\FL{mention something about scaling with parallel MCMC}
 
@@
 %%%
+
+{\color{red} the next sections are written along the model of \url{https://www.sciencedirect.com/science/article/pii/S2213133722000373} in Astronomy and Computing}
+%%%
 \section*{CRediT authorship contribution statement}
+\textbf{F. Lanusse}, \textbf{J. Zuntz}...
 \textbf{J.E Campagne} Conceptualization, Methodology, Software, Validation, Writing.
-\textbf{F. Lanusse}, \textbf{J. Zuntz}...
 
-%\section*{Declaration of Competing Interest}
-%The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
+\section*{Declaration of Competing Interest}
+The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
 
@@
 For instance the $\Omega_c$ parameter is treated as random variable as followed
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
+\begin{minted}[fontsize=\footnotesize]{python}
 Omega_c = numpyro.sample('Omega_c', Uniform(0.1, 0.9))
-\end{lstlisting}
-%\end{minted}
-
+\end{minted}
 Then, we generate mock angular power spectra $C_\ell$ from the auto \& cross correlations of the Number Counts and Weak Lensing probes computed with the \texttt{gaussian\_cl\_covariance\_and\_mean} function in the \texttt{jc.angular\_cl} module.
 One reads
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
+\begin{minted}[fontsize=\footnotesize]{python}
 # Define the lensing and number counts probe
@@
                                f_sky=0.25, sparse=True)
-\end{lstlisting}
-%\end{minted} 
+\end{minted} 
 with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} (ie.
@@
 The generation of the mock data yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
+\begin{minted}[fontsize=\footnotesize]{python}
 fiducial_model = condition(model,
@@
     data, P, C = fiducial_model()
-%\end{minted}
-\end{lstlisting}
-
+\end{minted}
 The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generates the model $C_{\ell}$ from the priors
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
+\begin{minted}[fontsize=\footnotesize]{python}
 cl = jc.angular_cl.angular_cl(cosmo, ell, probes).flatten()
@@
                             covariance_matrix=C))
-\end{lstlisting}
-%\end{minted}
-
+\end{minted}
 These theoretical $C_{\ell}$ are in turn conditioned by the mock $C_\ell$ as followed
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
+\begin{minted}[fontsize=\footnotesize]{python}
 observed_model = condition(model, {'cl': data})
-\end{lstlisting}
-%\end{minted}
+\end{minted}
 Then, as an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
+\begin{minted}[fontsize=\footnotesize]{python}
 nuts_kernel = numpyro.infer.NUTS(observed_model,
@@
 mcmc.run(jax.random.PRNGKey(42))
-\end{lstlisting}
-%\end{minted}
+\end{minted}
 More details can be found in the repository of this paper at this link \nblink{VectorizedNumPyro_block}.

### 2022-08-05 paper_OJA.tex
Fix numbering of subsections
add CCL_comparison nb

@@
 We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that the Normalizing Flows using the Neural Transport is a promising methodology.
-{\color{red}
-;in the early phases of Bayesian model development JAZ not sure what this means so I left it out}
+%{\color{red}
+%;in the early phases of Bayesian model development JAZ not sure what this means so I left it out}
 
@@
 A recent entrant to this field is the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} which has undergone rapid development and can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions, offering a speed up to the development process and indeed code runtimes.
-\texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), and just-in-time compilation (\texttt{jit}) and optimization targetting CPU, GPU and TPU hardware thanks to the \texttt{XLA} library that powers TensorFlow .
+\texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), and just-in-time compilation (\texttt{jit}) and optimization targetting CPU, GPU and TPU hardware thanks to the \texttt{XLA} library that powers TensorFlow.
 These attractive features have driven wide adoption of \texttt{JAX} in computational research, and motivate us to consider its usage in cosmology.
@@
 \texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper.
-Other PPL packages such as\texttt{PyMC} \citep{Salvatier2016} have switched to JAX backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}
-.
+Other PPL packages such as \texttt{PyMC} \citep{Salvatier2016} have switched to JAX backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}.
 
@@
 
-In this context we have developed the open source \texttt{jax-cosmo} library \footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
+In this context we have developed the open source \texttt{jax-cosmo} library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
 The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable JAX functions.
@@
 We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more advanced gradient-based cosmological inference with variants of Hamiltonian Monte Carlo and its variant No-U-Turn in Section~\ref{sec:chmc}, and in ML-accelerated Stochastic Variational Inference in Section~\ref{sec-SVI}.
-We discuss and compare these methods in Section~\ref{sec-discussion} and conclude in Section~\ref{sec-conclusion}
+We discuss and compare these methods in Section~\ref{sec-discussion} and conclude in Section~\ref{sec-conclusion}.
 
@@
 
-To make the bridge with observations, two pytrees are introduced, namely \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing  respectively,  with for each probe a bunch of redshift bins.
+To make the bridge with observations, two \textit{pyTrees} are introduced, namely \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing  respectively,  with for each probe a bunch of redshift bins.
 A long this line, the angular power spectra $C_\ell^{pq}(z,z^\prime)$ for the probes $(p,q)$ and for redshift bin window selections $(N(z), N(z^\prime))$ are computed in the Limber approximation.
-Remarkably, all the integral involved in the computation can be easily differentiated with respect to the cosmological parameters thanks to \textit{autodif} and the implementation of the above mentioned integration quadrature.
-An example is given in the context of DES Y1 3x2pts analysis (\ref{app-DESY1}).
+Remarkably, all the integral involved in the computation can be easily differentiated with respect to the cosmological parameters thanks to \textit{autodiff} and the implementation of the above mentioned integration quadrature.
+An example is given in the context of DES Y1 3x2pts analysis (Appendix~\ref{app-DESY1}).
 
@@
 %
-\subsubsection{Vanilla HMC}
+\subsection{Vanilla HMC}
 %
@@
 
-\subsubsection{NUTS}
+\subsection{NUTS}
 \label{sec-NUTS}
@@
 %
-\subsubsection{Neural Transport}
+\subsection{Neural Transport}
 \label{sec-Neural-Reparametrisation}

### 2022-08-05 paper_OJA.tex
add nb on SVI/NUTS

@@
 %%% color Comments 
+%% JEC : \FL produces a clash we another definition (but coannot figure out which one)  
 \newcommand{\FrL}[1]{{\color{cyan}FL: #1}}
@@
 
-To take advantage of these possibilities within cosmology we have developed the \texttt{jax-cosmo} library, which implements background evolution, linear and non linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes.
+To take advantage of these possibilities within cosmology we have developed the \jaxcosmo\ library, which implements background evolution, linear and non linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes.
 We discuss algorithms made possible by this library, and present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical usage.
@@
 
-In this context we have developed the open source \texttt{jax-cosmo} library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
+In this context we have developed the open source \jaxcosmo\ library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
 The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable JAX functions.
 
-We give an overview of the code's design and contents in Section \ref{sec-jaxcosmo-design}.
+We give an overview of the code's design and contents in Section~\ref{sec-jaxcosmo-design}.
 We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more advanced gradient-based cosmological inference with variants of Hamiltonian Monte Carlo and its variant No-U-Turn in Section~\ref{sec:chmc}, and in ML-accelerated Stochastic Variational Inference in Section~\ref{sec-SVI}.
@@
 %\subsection{Validation against the Core Cosmology Library (CCL)}
-To illustrate the different features available with the present version of the library, we have written a  notebook to make  comparison with the well established  Core Cosmology Library.
-As an example, the Figure \ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function.
+To illustrate the different features available with the present version of the library, we have written a  companion notebook \nblink{CCL_comparison} to make  comparison with the well established  Core Cosmology Library.
+As an example, the Figure~\ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function.
 %
@@
 
-\subsection{DES Y1 setting}
+%JEC 5 August 22: not sure that should be given here as we have an Appendix for that and the main application described in the following sections concern DES Y1. 
+%\subsection{DES Y1 setting}
 % Here we describe the data and 3x2pt model for DES Y1 
-
-In this section, we adopt as a setting a DES Y1 3x2pt-like analysis.
+% In this section, we adopt as a setting a DES Y1 3x2pt-like analysis.
 
@@
 
-\texttt{jax-cosmo} was used in a recent example of such a process, for the  the LSST-DESC 3x2pt tomography optimization challenge \citep{2021OJAp....4E..13Z}, where the best methodology for assignment of galaxies to tomographic bins was assessed using several such figures of merit and related calculations.
-The \texttt{jax-cosmo} metric proved to be stable and fast.
+\jaxcosmo\ was used in a recent example of such a process, for the  the LSST-DESC 3x2pt tomography optimization challenge \citep{2021OJAp....4E..13Z}, where the best methodology for assignment of galaxies to tomographic bins was assessed using several such figures of merit and related calculations.
+The \jaxcosmo\ metric proved to be stable and fast.
 
@@
 
-% Here we can implement MOPED for the compression of
+\FrL{Here we can implement MOPED for the compression}
 
@@
 We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI).
-All  the methods have been implemented using the \numpyro probabilistic programming language (PPL).
+All  the methods have been implemented using the \numpyro\ probabilistic programming language (PPL).
 %
@@
 So, in a automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
-The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls a each generation.
+The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls at each generation.
 
 Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
-After the forward model creation for the DES-Y1 3x2pts exercise described in \ref{app-DESY1}, we have performed
+After the forward model creation for the DES-Y1 3x2pts exercise described in Appendix~\ref{app-DESY1}, we have performed
 \begin{itemize}
-    \item first a reparametrisation of the latent variable uniform priors (Tab.~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$;
+    \item first a reparametrisation of the latent variable uniform priors (Table~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$;
     \item second for the warm-up phase we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
@@
 
-The results in terms of relative effective sample sizes (aka "ESS") are detailed in Tab.~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
+The results in terms of relative effective sample sizes (aka "ESS") are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
 One notices the dramatic improvement of the ESS by about a factor 10 using the NUTS sampler compared to the highly-optimized \texttt{Cobaya}  Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T} which has last about 40 hours on CPU to generate 70,000 samples.
@@
 It is worth to point out that the mass matrix structuring operation is a very effective as it increases the sampling efficiency by about a factor 2.
-Anticipating the section \ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
-This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Tab.~\ref{tab-ESS-NUTS_SVI-1}), then the user is highly encouraged to tune this critical parameter.
+Anticipating the Section~\ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
+This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Table~\ref{tab-ESS-NUTS_SVI-1}), then the user is highly encouraged to tune this critical parameter.
 %
@@
 Naturally, more samples lead to better contour precision.
-But what is illustrative is the "low" schema results as shown in the figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
+But what is illustrative is the "low" schema results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
 Notice the good results even for the highly correlated lens bias parameters.
@@
 
-Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to ease the model development in its early phase.
+Results in terms of sampling efficiency are shown on Table~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to ease the model development in its early phase.
 %
@@
 %
-The key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
+To a large extend looking at results of Table~\ref{tab-ESS-NUTS_SVI-1}, the key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
 In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
@@
 \end{equation}
-with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation.
+with $N_s$ the total number of samples, $\varepsilon$ the effective sampler efficiency  and $n_{steps}$ the number of steps (calls) per sample generation.
 Then, for \texttt{Cobaya} one roughly gets $\varepsilon\approx 3\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more.

### 2022-08-05 paper_OJA.tex
fix some typos and rephrase some sentences. Add link to notebook at the end of the appendix.

@@
 (a): results from Cobaya 70,000 samples;
-(b): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Sec.~\ref{sec-NUTS}); 
-(c) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 1 chain of 200 samples and 200 warm-up (Sec.~\ref{sec-Neural-Reparametrisation}).}
+(b): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Section~\ref{sec-NUTS}); 
+(c) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 1 chain of 200 samples and 200 warm-up (Section~\ref{sec-Neural-Reparametrisation}).}
 \label{tab-ESS-NUTS_SVI-1}
@@
 %\end{minted}
-More details can be found in the repository of this paper at this link \nblink{VectorizedNumPyro_block}.
+More details can be found in the repository of this paper to perform a \numpyro\ modeling using \jaxcosmo : a structured NUTS sampling described in Section~\ref{sec-NUTS} \nblink{VectorizedNumPyro_block}, and a SVI optimisation followed by NUTS sampling of Section~\ref{sec-Neural-Reparametrisation} \nblink{DESY_Y1_shear_cluster_ModelReparam_SVI-MVN_Neutra_NUTS}.
+Notice however that the notebooks are only illustrative and for a real sample production the authors invite the reader to run their own script.
 %% If you have bibdatabase file and want bibtex to generate the

### 2022-08-18 paper_OJA.tex
More details in the description of the angular Cl implementation.

@@
 \newcommand{\FrL}[1]{{\color{cyan}FL: #1}}
+\newcommand{\JZ}[1]{{\color{purple}JZ: #1}}
 
@@
 In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX.
-The general development follows the Core Cosmology Library structure \citep{2019ApJS..242....2C} although it does not include all the options as the present \jaxcosmo\ version is more to be viewed as a proof of concept.
+The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though the present proof-of-concept \jaxcosmo\ does not include as many features and options as CCL.
 
 The cosmological parameters are organised in a JAX \textit{pyTree} container\footnote{\url{https://jax.readthedocs.io/en/latest/jax-101/05.1-pytrees.html}} which allows one to use \textit{autodiff} and \texttt{vmap} transparently.
-The present choice of parameters are: $\Omega_{cdm}$ the cold dark matter density fraction, $\Omega_b$ the Byronic matter density fraction, $h$ the Hubble constant divided by $100$~km/s/Mpc, $n_s$ the primordial scalar perturbation spectral index, $\sigma_8$ the variance of matter density perturbations at an $8$~Mpc/h scale, $\Omega_k$ the curvature density fraction, $w_0$ the first order term of dark energy equation and $w_a$ the second order term of dark energy equation of state.
+The present set of exposed parameters is: $\Omega_{cdm}$ the cold dark matter density fraction, $\Omega_b$ the Baryonic matter density fraction, $h$ the Hubble constant divided by $100$~km/s/Mpc, $n_s$ the primordial scalar perturbation spectral index, $\sigma_8$ the variance of matter density perturbations at an $8$~Mpc/h scale, $\Omega_k$ the curvature density fraction, $w_0$ the first order term of dark energy equation and $w_a$ the second order term of dark energy equation of state.
 
-The traditional computation of the different cosmological background follows \cite{2005A&A...443..819P}  with namely the Friedmann equation
+\subsection{Background cosmology}
+
+The computation of the evolution of the cosmological background follows a typical implementation of a Friedmann equation (\citealp[see e.g.][]{2005A&A...443..819P}):
 \begin{equation}
-    E^2(a) = \frac{H^2(a)}{H^2_0} = \Omega_m a^{-3} + \Omega_k a^{-2} + \Omega_{de} e^{f(a)}
+    E^2(a) = \frac{\mathrm{H}^2(a)}{\mathrm{H}^2_0} = \Omega_m a^{-3} + \Omega_k a^{-2} + \Omega_{de} e^{f(a)}
 \end{equation}
-with $a=1/(1+z)$ the scale factor related to the redshift $z$, $H(a)=\dot{a}/a$ the Hubble parameter with $H_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and
+with $a=1/(1+z)$ the scale factor related to the redshift $z$, $H(a)=\dot{a}/a$ the Hubble parameter with $\mathrm{H}_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and
 \begin{equation}
@@
 \end{equation}
-Noticeably, the relativist contributions of massless neutrinos and photon radiations, as well as the massive neutrino contribution are neglected.
+Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are presently neglected.
+From these expressions, in \texttt{jc.background} where \texttt{jc} hereafter stands for \jaxcosmo, are computed the different cosmological distance functions.
+Notably, the radial comoving distance
+\begin{equation}
+     \chi(a) =  R_H \int_a^1 \frac{da^\prime}{{a^\prime}^2 E(a^\prime)}
+\end{equation}
+with $R_H$ the Hubble radius. Using the scale factor ($a$) redshift ($z$) relationship, $\chi$ can be viewed as a function of $z$. 
+%
+\subsection{Growth of perturbations}
+%
+Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}.
+No Baryon feedback modeling is considered yet.
+To perform \textit{autodiff} we have implemented various tools as differentiable ODEs using the Runge-Kutta algorithm, as well as Simpson quadrature and the Romberg algorithm for numerical integration.
 
-Currently, \jaxcosmo\  implements the \cite{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}.
-No Baryon feedback modeling is considered yet.
-It is worth to mention that to perform \textit{autodiff} we have implemented a series of tools as differentiable ODE via Runge-Kutta as well as integration methods as Simpson quadrature and Romberg algorithm.
+%Future version of the library would offer the possibility to call a JAX emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.
+\JZ{How close is this to being usable?
+If it's just an idea or an early prototype then skip this paragraph.}
+%
+\subsection{Angular power spectra}
+%
+\jaxcosmo\ is currently focused on predicting projected 2D Fourier-space 2pt galaxy lensing, clustering, and cross correlations, the $C_\ell$ angular power spectra that are a primary target of upcoming photometric surveys.
+The details of the implementation is in \texttt{jc.angular\_cl} which deals with the mean and Gaussian covariance matrix computations.
 
-Future version of the library would offer the possibility to call a JAX emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.
+To make a bridge to these observables, two \textit{pyTree} objects are implemented: \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing samples, with each instance of each representing a single tomographic redshift bin's number density.
+These tracers are implemented according to the following kernel functions:
+\begin{itemize}
+    \item \textbf{NumberCounts}: 
+    \begin{equation}
+        K_i(z) = n_i(z)\ b(z)\ H(z)
+    \end{equation}
+where $n_i(s)$ is the redshift distribution of the sample (eg. \texttt{jc.redshift.kde\_nz} function), $b(z)$ a galaxy bias function (see  \texttt{jc.bias.constant\_linear\_bias}). 
+ \item \textbf{WeakLensing}:
+    \begin{multline}
+ K_i(z) = \left( \frac{3 H_0^2\Omega_m}{2 c} \right) \left(\frac{(\ell+2)!}{(\ell-2)!} \right)^{1/2}\times 
+ \\  (1+z)\ \chi(z) \int_z^\infty p_i(z^\prime)\ \frac{\chi(z^\prime)-\chi(z)}{\chi(z^\prime)}\ dz^\prime + K_{IA}(z)
+    \end{multline}
+where $K_{IA}(z)$ is an optional kernel function to deal with the Intrinsic Alignment. The current implementation follows the reference \citep{2011A&A...527A..26} according to
+\begin{equation}
+    K_{IA}(z) =   \left(\frac{(\ell+2)!}{(\ell-2)!}\right)^{1/2}\ p_i(z) b(z) H(z)\ \frac{C \Omega_m}{D(z)}
+\end{equation}
+with $C\approx 0.0134$ a constant without unit and $D(z)$ the growth factor of perturbations.
+\end{itemize}
+Notice that all described kernel functions can be also viewed as function of the scale factor $a$ or as function of the radial comoving distance $\chi$.
 
-To make the bridge with observations, two \textit{pyTrees} are introduced, namely \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing  respectively,  with for each probe a bunch of redshift bins.
-A long this line, the angular power spectra $C_\ell^{pq}(z,z^\prime)$ for the probes $(p,q)$ and for redshift bin window selections $(N(z), N(z^\prime))$ are computed in the Limber approximation.
-Remarkably, all the integral involved in the computation can be easily differentiated with respect to the cosmological parameters thanks to \textit{autodiff} and the implementation of the above mentioned integration quadrature.
+From these tracer objects, the angular power spectra $C_\ell^{ij}$ for the probes $(i,j)$ and for redshift bin window selections are computed in the first order Limber approximation \citep{PhysRevD.78.123506}:
+\JZ{Discuss the implementation of the Limber integral here. Maybe move down the bit in the growth of perturbations section?}
+\begin{align}
+    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{d\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{limber_gen}
+\end{align}
+with $P(k,z)$ the power spectra computed using the growth of perturbation models described above and a primordial power spectra scaling according to
+\begin{equation}
+    P(k) = A k^{n_s}
+\end{equation}
+The normalisation $A$ is computed using the $\sigma_8$ parameter as 
+\begin{equation}
+    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k,z) W^2(kR_8) \right)^{-1}
+\end{equation}
+with $R_8 = 8$ in unit of Mpc and $W(x)$ is related to the $j_1$ sperical Bessel function as
+\begin{equation}
+    W(x) = \frac{3j_1(x)}{x}
+\end{equation}
+Finally, the factors $m_i$ are equal to $(0,-2)$ for the \textbf{NumberCounts}, \textbf{WeakLensing} related kernels, respectively.
+
+Remarkably, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters thanks to \textit{autodiff} and the implementation of the above mentioned integration quadrature.
 An example is given in the context of DES Y1 3x2pts analysis (Appendix~\ref{app-DESY1}).
-
-% Explain how the code is structured, mention some tricks of autodiff if we have some
-% Follows a bit the structure of the CCL library
-
-%\subsection{Background cosmology}
-% Usual stuff
-
-%\subsection{Growth of perturbations}
-% The interesting point to mention here is the differentiable growth ODE
-
-%\subsection{Matter power spectrum}
-% Here we describe the EH power spectrum, mention possible extensions with emulators
-
-%\subsection{Tracers}
-% Here we describe the kernels, like in CCL
-
-%\subsection{Angular power-spectra}
-% Limber integration, this can be an opportunity to talk about differentiation through integrals as well.
-
 %
-%\subsection{Validation against the Core Cosmology Library (CCL)}
-To illustrate the different features available with the present version of the library, we have written a  companion notebook \nblink{CCL_comparison} to make  comparison with the well established  Core Cosmology Library.
+\subsection{Validation against the Core Cosmology Library (CCL)}
+To illustrate the different features available with the present version of the library, we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library.
 As an example, the Figure~\ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function.
+\JZ{Show more plots from the notebook here.
+Also, show a relative error plot between the two.
+As Tim Eifler once said to me, ``Any codes can agree when you plot them over four orders of magnitude''.}
 %

### 2022-08-18 paper_deprecated.tex
More details in the description of the angular Cl implementation.

@@
 \newcommand{\FL}[1]{{\color{cyan}FL: #1}}
+\newcommand{\JZ}[1]{{\color{blue}JZ: #1}}
 
@@
 \end{equation}
-Noticeably, the relativist contributions of massless neutrinos and photon radiations, as well as the massive neutrino contribution are neglected.
+Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are presently neglected.
 
-Currently, \jaxcosmo\  implements the \cite{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}.
+\subsection{Growth of perturbations}
+
+Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}.
 No Baryon feedback modeling is considered yet.
-It is worth to mention that to perform \textit{autodiff} we have implemented a series of tools as differentiable ODE via Runge-Kutta as well as integration methods as Simpson quadrature and Romberg algorithm.
+To perform \textit{autodiff} we have implemented various tools as differentiable ODEs using the Runge-Kutta algorithm, as well as Simpson quadrature and the Romberg algorithm for numerical integration.
 
 Future version of the library would offer the possibility to call a JAX emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.
+\JZ{How close is this to readiness?
+If it's just an idea or an early prototype then skip this paragraph}
+
+\subsection{Angular power spectra}
+
+\jaxcosmo\ is currently focused on predicting projected 2D Fourier-space 2pt galaxy lensing, clustering, and cross correlations, the $C_\ell$ values that are a primary object of
 
@@
 A long this line, the angular power spectra $C_\ell^{pq}(z,z^\prime)$ for the probes $(p,q)$ and for redshift bin window selections $(N(z), N(z^\prime))$ are computed in the Limber approximation.
+
+\JZ{Discuss the Limber implementation here}
+
 Remarkably, all the integral involved in the computation can be easily differentiated with respect to the cosmological parameters thanks to \textit{autodif} and the implementation of the above mentioned integration quadrature.
@@
 % Explain how the code is structured, mention some tricks of autodiff if we have some
-% Follows a bit the structure of the CCL library
 
-%\subsection{Background cosmology}
 % Usual stuff
 
-%\subsection{Growth of perturbations}
 % The interesting point to mention here is the differentiable growth ODE
@@
 % Here we describe the EH power spectrum, mention possible extensions with emulators
-
-%\subsection{Tracers}
-% Here we describe the kernels, like in CCL
 

### 2022-08-23 paper_OJA.tex
Updates from Overleaf

@@
     \end{equation}
-where $n_i(s)$ is the redshift distribution of the sample (eg. \texttt{jc.redshift.kde\_nz} function), $b(z)$ a galaxy bias function (see  \texttt{jc.bias.constant\_linear\_bias}). 
+where $n_i(s)$ is the redshift distribution of the sample (eg.
+\texttt{jc.redshift.kde\_nz} function), $b(z)$ a galaxy bias function (see  \texttt{jc.bias.constant\_linear\_bias}).
+No redshift space distorsions are taken into account.
  \item \textbf{WeakLensing}:
@@
 To illustrate the different features available with the present version of the library, we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library.
-As an example, the Figure~\ref{fig:halofit_comparison} shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function.
+As examples, the Figures \ref{fig:chi_comparison},
+\ref{fig:halofit_comparison}
+shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function.
 \JZ{Show more plots from the notebook here.
@@
 %
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\columnwidth]{figures/comp_radial_com_dist.png}
+    \caption{Comparison of the radial comoving distance between CCL and \jaxcosmo.} 
+    \label{fig:chi_comparison}
+\end{figure}
 \begin{figure}[ht]

### 2022-08-23 paper_OJA.tex
Update of the CCL/jax-cosmo validation sectiion

@@
      \chi(a) =  R_H \int_a^1 \frac{da^\prime}{{a^\prime}^2 E(a^\prime)}
+     \label{eq:radial_comoving}
 \end{equation}
@@
 From these tracer objects, the angular power spectra $C_\ell^{ij}$ for the probes $(i,j)$ and for redshift bin window selections are computed in the first order Limber approximation \citep{PhysRevD.78.123506}:
-\JZ{Discuss the implementation of the Limber integral here. Maybe move down the bit in the growth of perturbations section?}
+%\JZ{Discuss the implementation of the Limber integral here. Maybe move down the bit in the growth of perturbations section?}
 \begin{align}
-    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{d\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{limber_gen}
+    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{d\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
 \end{align}
@@
 \end{equation}
-with $R_8 = 8$ in unit of Mpc and $W(x)$ is related to the $j_1$ sperical Bessel function as
+with $R_8 = 8$ in unit of Mpc and $W(x)$ is related to the $j_1$ spherical Bessel function as
 \begin{equation}
@@
 \end{equation}
-Finally, the factors $m_i$ are equal to $(0,-2)$ for the \textbf{NumberCounts}, \textbf{WeakLensing} related kernels, respectively.
+Finally, the factors $m_i$ are equal to $(0,-2)$ for the \texttt{NumberCounts}, \texttt{WeakLensing} related kernels, respectively.
 
@@
 \subsection{Validation against the Core Cosmology Library (CCL)}
-To illustrate the different features available with the present version of the library, we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library.
+%
+To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}), we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library \citep{2019ApJS..242....2C} \footnote{https://ccl.readthedocs.io, version \texttt{2.5.1}.}.
 As examples, the Figures \ref{fig:chi_comparison},
-\ref{fig:halofit_comparison}
-shows an example of the non linear matter power spectrum computation using the \texttt{halofit} fitting function.
-\JZ{Show more plots from the notebook here.
-Also, show a relative error plot between the two.
-As Tim Eifler once said to me, ``Any codes can agree when you plot them over four orders of magnitude''.}
+\ref{fig:halofit_comparison} and \ref{fig:Cell_comparison}
+show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non linear matter power spectrum computation using the \texttt{halofit} fitting function and the angular power spectrum (Eq.~\ref{eq:Cell_limber}) using the correlation between the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions.
+The relative differences between the two libraries are at the level of few $10^{-3}$ or better.
+ These observed numerical differences are mostly due to different choices of integration methods and accuracy parameters (eg.
+quadrature number of points).
+Increasing these parameters lead to performance degradation as for instance for \jaxcosmo\ this increases the XLA compilation memory storage especially for the angular power spectra computation.
+So, the authors have made compromise and the numerical applications detailed in the following sections are done with the \jaxcosmo\ version available via \texttt{PyPi}\footnote{https://github.com/pypa/pip}.
+%\JZ{Show more plots from the notebook here. Also, show a relative error plot between the two. As Tim Eifler once said to me, ``Any codes can agree when you plot them over four orders of magnitude''.}
 %
-\begin{figure}[ht]
+\begin{figure}
     \centering
     \includegraphics[width=\columnwidth]{figures/comp_radial_com_dist.png}
-    \caption{Comparison of the radial comoving distance between CCL and \jaxcosmo.} 
+    \caption{Comparison of the radial comoving distance between CCL and \jaxcosmo. More plots are available in the  companion notebook \nblink{CCL_comparison}.} 
     \label{fig:chi_comparison}
 \end{figure}
-\begin{figure}[ht]
+\begin{figure}
     \centering
     \includegraphics[width=\columnwidth]{figures/halofit_pk.png}
-    \caption{Comparison of matter power spectrum between CCL and jax-cosmo. \textbf{Example of validation plot} \nblink{CCL_comparison}}
-    \label{fig:halofit_comparison}
+    \caption{Comparison of the non-linear matter power spectrum (\textit{halofit} function) between CCL and \jaxcosmo. Is also shown the linear power spectrum.}     \label{fig:halofit_comparison}
+\end{figure}
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/comp_Cell_shape_pos.png}
+    \caption{Comparison of the shape-position angular power spectrum between CCL and \jaxcosmo.} 
+    \label{fig:Cell_comparison}
 \end{figure}

### 2022-08-23 paper_OJA.tex
minor corrections

@@
 
-The cosmological parameters are organised in a JAX \textit{pyTree} container\footnote{\url{https://jax.readthedocs.io/en/latest/jax-101/05.1-pytrees.html}} which allows one to use \textit{autodiff} and \texttt{vmap} transparently.
+The cosmological parameters are organised in a JAX \textit{pyTree} container\footnote{\url{https://jax.readthedocs.io/en/latest/jax-101/05.1-pytrees.html}} \texttt{jc.Cosmology} where \texttt{jc} hereafter stands for \jaxcosmo.
+This container allows one to use \textit{autodiff} and \texttt{vmap} transparently.
 The present set of exposed parameters is: $\Omega_{cdm}$ the cold dark matter density fraction, $\Omega_b$ the Baryonic matter density fraction, $h$ the Hubble constant divided by $100$~km/s/Mpc, $n_s$ the primordial scalar perturbation spectral index, $\sigma_8$ the variance of matter density perturbations at an $8$~Mpc/h scale, $\Omega_k$ the curvature density fraction, $w_0$ the first order term of dark energy equation and $w_a$ the second order term of dark energy equation of state.
+Notably, \texttt{jc.Planck15} instantiates a set of cosmological parameters inspired from the Planck 2015 results \citep{2016A&A...594A..13P}.
 
@@
 Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are presently neglected.
-From these expressions, in \texttt{jc.background} where \texttt{jc} hereafter stands for \jaxcosmo, are computed the different cosmological distance functions.
+From these expressions, in \texttt{jc.background}, are computed the different cosmological distance functions.
 Notably, the radial comoving distance
@@
 To make a bridge to these observables, two \textit{pyTree} objects are implemented: \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing samples, with each instance of each representing a single tomographic redshift bin's number density.
-These tracers are implemented according to the following kernel functions:
-\begin{itemize}
-    \item \textbf{NumberCounts}: 
+These tracers are implemented as two kernel functions:
+
+\begin{description}
+\item[\texttt{NumberCounts}]
     \begin{equation}
@@
 \texttt{jc.redshift.kde\_nz} function), $b(z)$ a galaxy bias function (see  \texttt{jc.bias.constant\_linear\_bias}).
-No redshift space distorsions are taken into account.
- \item \textbf{WeakLensing}:
+No redshift space distortions are taken into account.
+
+\item [\texttt{WeakLensing}]
     \begin{multline}
@@
     \end{multline}
-where $K_{IA}(z)$ is an optional kernel function to deal with the Intrinsic Alignment. The current implementation follows the reference \citep{2011A&A...527A..26} according to
+where $K_{IA}(z)$ is an optional kernel function to deal with the Intrinsic Alignment. The implementation of this term currently follows \citet{2011A&A...527A..26J}, and is given by:
 \begin{equation}
-    K_{IA}(z) =   \left(\frac{(\ell+2)!}{(\ell-2)!}\right)^{1/2}\ p_i(z) b(z) H(z)\ \frac{C \Omega_m}{D(z)}
+    K_{IA}(z) =   \left(\frac{(\ell+2)!}{(\ell-2)!}\right)^{1/2}\ p_i(z)\ b(z)\  H(z)\ \frac{C\  \Omega_m}{D(z)}
 \end{equation}
-with $C\approx 0.0134$ a constant without unit and $D(z)$ the growth factor of perturbations.
-\end{itemize}
+with $C\approx 0.0134$ a dimensionless constant and $D(z)$ the growth factor of linear perturbations.
+\end{description}
+
 Notice that all described kernel functions can be also viewed as function of the scale factor $a$ or as function of the radial comoving distance $\chi$.
@@
 %
-To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}), we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library \citep{2019ApJS..242....2C} \footnote{https://ccl.readthedocs.io, version \texttt{2.5.1}.}.
+To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{https://ccl.readthedocs.io, version \texttt{2.5.1}.}.
 As examples, the Figures \ref{fig:chi_comparison},
 \ref{fig:halofit_comparison} and \ref{fig:Cell_comparison}
-show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non linear matter power spectrum computation using the \texttt{halofit} fitting function and the angular power spectrum (Eq.~\ref{eq:Cell_limber}) using the correlation between the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions.
+show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non linear matter power spectrum computation using the \texttt{halofit} fitting function and the angular power spectrum for galaxy-galaxy lensing (Eq.~\ref{eq:Cell_limber}) using the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions.
 The relative differences between the two libraries are at the level of few $10^{-3}$ or better.
- These observed numerical differences are mostly due to different choices of integration methods and accuracy parameters (eg.
+
+These numerical differences are mostly due to different choices of integration methods and accuracy parameters (eg.
 quadrature number of points).
-Increasing these parameters lead to performance degradation as for instance for \jaxcosmo\ this increases the XLA compilation memory storage especially for the angular power spectra computation.
-So, the authors have made compromise and the numerical applications detailed in the following sections are done with the \jaxcosmo\ version available via \texttt{PyPi}\footnote{https://github.com/pypa/pip}.
-%\JZ{Show more plots from the notebook here. Also, show a relative error plot between the two. As Tim Eifler once said to me, ``Any codes can agree when you plot them over four orders of magnitude''.}
-%
+Increasing these parameters lead to performance degradation for \jaxcosmo\ as it increases the XLA compilation memory requirements significantly, especially for the angular power spectra computation.
+Since these differences are likely to be within the tolerance of the current generation of cosmological surveys, this trade-off is an acceptable one.
+
 \begin{figure}

### 2022-08-24 paper_OJA.tex

@@
 %
-To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{https://ccl.readthedocs.io, version \texttt{2.5.1}.}.
+To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{\url{https://ccl.readthedocs.io}, version \texttt{2.5.1}.}.
 As examples, the Figures \ref{fig:chi_comparison},
@@
 \subsection{Instantaneous Fisher Forecasts}
-
+In this section, we illustrate with a purely fictional case study, the computation of the Fisher matrix by two methods thanks to the \textit{autodiff} ability of JAX.
+For the detailed implementation, the reader is invited to have a look at the following companion notebook \nblink{Simple-Fisher}.
+In this use-case we use four tracers redshift distributions: two to define \texttt{WeakLensing} kernels and two for \texttt{NumberCounts} kernels.
+Then, the $10$ angular power spectra $C_\ell^{i,j}$ ($i,j:1,\dots,4$) are computed for $50$  logarithmically spaced angular moments between $10$ and $1000$ using Equation \ref{eq:Cell_limber}.
+The Gaussian covariance matrix is computed simultaneously.
+A dataset is obtained from the computation of the $C_\ell^{i,j}$ with a fiducial cosmology.
+Then, the following snippet shows the log likelihood function $\mathcal{L}(\theta)$ implementation ($\theta$ stands for the set of cosmological parameters).
+The \texttt{jax.jit} decorator is there to build a compiled version of the function at the first use.
 %\begin{minted}[fontsize=\footnotesize]{python}
 \begin{lstlisting}[language=iPython]
-def likelihood(cosmo):
-  # Compute mean and covariance of angular Cls, 
-  # for specific probes
-  mu, cov = gaussian_cl_covariance_and_mean(cosmo, 
-                    ell, probes)
-  # Return likelihood value
-  return gaussian_log_likelihood(data, mu, cov)
-
-# Compute derivatives of the likelihood with 
-# respect to cosmological parameters
-g = jax.grad(likelihood)(cosmo)
-
-# Compute Fisher matrix of cosmological 
-# parameters
-F = -jax.hessian(likelihood)(cosmo)
+@jax.jit
+def likelihood(p):
+    # Create a new cosmology at these parameters
+    cosmo = jc.Planck15(Omega_c=p[0], sigma8=p[1])
+    # Compute mean and covariance of angular Cls
+    mu, C = jc.angular_cl.gaussian_cl_covariance_and_mean(cosmo, ell, tracers, sparse=True)
+    # Return likelihood value assuming constant covariance, so we stop the gradient
+    # at the level of the precision matrix, and we will not include the logdet term
+    # in the likelihood
+    P = jc.sparse.inv(jax.lax.stop_gradient(C))
+    r = data - mu
+    return -0.5 * r.T @ jc.sparse.sparse_dot_vec(P, r)
 \end{lstlisting}
 %\end{minted}
+The goal is to obtain the 1-sigma contour forecast of the two parameters ($\Omega_c, \sigma_8$).
+To do so, we use the Fisher matrix computation.
+One way is to use the Hessian of the log-likelihood as follows
+\begin{equation}
+F_{\alpha\beta} = -\big\langle \frac{\partial^2\mathcal{L}(\theta)}{\partial \theta_\alpha \partial \theta_\beta}\big\rangle
+\qquad (\theta_\alpha=\Omega_c, \theta_\beta=\sigma_8)
+\end{equation}
+which is mostly  1-line JAX code
+\begin{lstlisting}[language=iPython]
+hessian_loglik = jax.jit(jax.hessian(likelihood))
+F = - hessian_loglik(params)
+\end{lstlisting}
+An other way uses the computation of the Fisher matrix when the likelihood is a Gaussian distribution as ii is the case above.
+First, we define a function that computes the mean  $\mu(\ell; \theta)$, and it follows that the Fisher matrix element reads
+\begin{equation}
+    F_{\alpha\beta} = \sum_\ell \frac{\partial \mu(\ell)}{\partial \theta_\alpha}^T C^{-1}(\ell)\frac{\partial \mu(\ell)}{\partial \theta_\beta}
+\end{equation}
+where $C^{-1}(\ell)$ is the covariance matrix computed with the fiduccial cosmology.
 

### 2022-08-24 paper_OJA.tex

@@
 A dataset is obtained from the computation of the $C_\ell^{i,j}$ with a fiducial cosmology.
-Then, the following snippet shows the log likelihood function $\mathcal{L}(\theta)$ implementation ($\theta$ stands for the set of cosmological parameters).
-The \texttt{jax.jit} decorator is there to build a compiled version of the function at the first use.
+Then, the following snippet shows the log likelihood function $\mathcal{L}(\theta)$ implementation considering a constant covariance matrix ($\theta$ stands for the set of cosmological parameters).
 %\begin{minted}[fontsize=\footnotesize]{python}
@@
 %\end{minted}
+The \texttt{jc.sparse} functions are implementation of block matrix computations: sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_diag)$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.
+The \texttt{jax.jit} decorator is there to build a compiled version of the function at the first use.
 The goal is to obtain the 1-sigma contour forecast of the two parameters ($\Omega_c, \sigma_8$).
@@
 \begin{equation}
-F_{\alpha\beta} = -\big\langle \frac{\partial^2\mathcal{L}(\theta)}{\partial \theta_\alpha \partial \theta_\beta}\big\rangle
+F_{\alpha\beta} = - \frac{\partial^2\mathcal{L}(\theta)}{\partial \theta_\alpha \partial \theta_\beta}
 \qquad (\theta_\alpha=\Omega_c, \theta_\beta=\sigma_8)
+\label{eq:fisher_way1}
 \end{equation}
@@
     F_{\alpha\beta} = \sum_\ell \frac{\partial \mu(\ell)}{\partial \theta_\alpha}^T C^{-1}(\ell)\frac{\partial \mu(\ell)}{\partial \theta_\beta}
+    \label{eq:fisher_way2}
 \end{equation}
-where $C^{-1}(\ell)$ is the covariance matrix computed with the fiduccial cosmology.
+where $C^{-1}(\ell)$ is the covariance matrix computed with the fiducial cosmology. In this case one needs also just 1-line of JAX code 
+\begin{lstlisting}[language=iPython]
+F = jc.sparse.dot(dmu.T, jc.sparse.inv(cov), dmu)
+\end{lstlisting}
 
-% Plot showing error in Jacobian computation by finite diff
+The two methods agree perfectly as shown on Figure \ref{fig:simple_fisher_1}.
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/simple_fisher_1.png}
+    \caption{Comparison of the two methods to compute the Fisher matrix (Eqs.\ref{eq:fisher_way1},\ref{eq:fisher_way2}).} 
+    \label{fig:simple_fisher_1}
+\end{figure}
+%
+It is worth to notice that in the two above methods of Fisher matrix computations, the user does not need to take care of particular parameter to compute the 1st or 2nd order derivatives.
+This is in contrast to the cases where one deals with finite difference methods.
+AS an illustration, we have used the CCL library to compute the Fisher matrix via Equation \ref{eq:fisher_way2}.
+To do so, the Jacobian ($\partial \mu/\partial\theta_\alpha$) is computed with centered  order 4 finite differences available in the \texttt{Numdifftools} Python package\footnote{\url{https://numdifftools.readthedocs.io}}.
+Using the spacing parameter values ($10^{-6}$, $10^{-2}$, $10^{-1}$) one gets the 1-sigma contours
 
-% Plot showing error on some parameters
-
+%
 \subsection{Survey Optimization by FoM Maximization}

### 2022-08-24 paper_OJA.tex
Section Fisher matrix computation update

@@
     \centering
-    \includegraphics[width=\columnwidth]{figures/simple_fisher_1.png}
+    \includegraphics[width=0.7\columnwidth]{figures/simple_fisher_1.png}
     \caption{Comparison of the two methods to compute the Fisher matrix (Eqs.\ref{eq:fisher_way1},\ref{eq:fisher_way2}).} 
@@
 To do so, the Jacobian ($\partial \mu/\partial\theta_\alpha$) is computed with centered  order 4 finite differences available in the \texttt{Numdifftools} Python package\footnote{\url{https://numdifftools.readthedocs.io}}.
-Using the spacing parameter values ($10^{-6}$, $10^{-2}$, $10^{-1}$) one gets the 1-sigma contours
-
+Using the spacing parameter values ($10^{-6}$, $10^{-2}$, $10^{-1}$) one obtains three different approximation of the 1-sigma contours as shown on Figure \ref{fig:simple_fisher_2}.
+One notices that the contour that agrees quite well with the \jaxcosmo\ one, is obtained with the spacing parameter equals to $10^{-2}$ which implies that the user has to  tune this parameter carefully.
+Although very simple, this case study exhibits the intrinsic limitation of finite difference methods for computing the Fisher matrix as it as been shown for instance in a more advanced case study in reference \citet{2021arXiv210100298B}.
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/simple_fisher_2.png}
+    \caption{Computations of the Fisher matrix (Eq.\ref{eq:fisher_way1}) using the CCL library for the angular power spectra estimation and finite difference method to get the Jacobian  with different spacing parameters ($10^{-6}$, $10^{-2}$, $10^{-1}$).
+For comparison, the contour obtained with \jaxcosmo\ is reproduced from Figure \ref{fig:simple_fisher_1} in black.}
+    \label{fig:simple_fisher_2}
+\end{figure}
 %

### 2022-08-25 paper_OJA.tex
MOPED (begin an introduction)

@@
 \subsection{Instantaneous Fisher Forecasts}
-In this section, we illustrate with a purely fictional case study, the computation of the Fisher matrix by two methods thanks to the \textit{autodiff} ability of JAX.
+In this section, we illustrate with a purely fictional case study, the computation of the Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} by two methods thanks to the \textit{autodiff} ability of JAX.
 For the detailed implementation, the reader is invited to have a look at the following companion notebook \nblink{Simple-Fisher}.
 In this use-case we use four tracers redshift distributions: two to define \texttt{WeakLensing} kernels and two for \texttt{NumberCounts} kernels.
-Then, the $10$ angular power spectra $C_\ell^{i,j}$ ($i,j:1,\dots,4$) are computed for $50$  logarithmically spaced angular moments between $10$ and $1000$ using Equation \ref{eq:Cell_limber}.
+Then, the $10$ angular power spectra $C_\ell^{p,q}$ ($p,q:1,\dots,4$) are computed for $50$  logarithmically spaced angular moments between $10$ and $1000$ using Equation \ref{eq:Cell_limber}.
 The Gaussian covariance matrix is computed simultaneously.
-A dataset is obtained from the computation of the $C_\ell^{i,j}$ with a fiducial cosmology.
+A dataset is obtained from the computation of the $C_\ell^{p,q}$ with a fiducial cosmology.
 Then, the following snippet shows the log likelihood function $\mathcal{L}(\theta)$ implementation considering a constant covariance matrix ($\theta$ stands for the set of cosmological parameters).
@@
 %\end{minted}
-The \texttt{jc.sparse} functions are implementation of block matrix computations: sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_diag)$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.
+The \texttt{jc.sparse} functions are implementation of block matrix computations: sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_{diag})$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.
 The \texttt{jax.jit} decorator is there to build a compiled version of the function at the first use.
@@
 \begin{equation}
-F_{\alpha\beta} = - \frac{\partial^2\mathcal{L}(\theta)}{\partial \theta_\alpha \partial \theta_\beta}
-\qquad (\theta_\alpha=\Omega_c, \theta_\beta=\sigma_8)
+F_{i,j} = - \frac{\partial^2\mathcal{L}(\theta)}{\partial \theta_i \partial \theta_j}
+\qquad (\theta_1=\Omega_c, \theta_2=\sigma_8)
 \label{eq:fisher_way1}
@@
 \begin{equation}
-    F_{\alpha\beta} = \sum_\ell \frac{\partial \mu(\ell)}{\partial \theta_\alpha}^T C^{-1}(\ell)\frac{\partial \mu(\ell)}{\partial \theta_\beta}
+    F_{i,j} = \sum_\ell \frac{\partial \mu(\ell)}{\partial \theta_i}^T C^{-1}(\ell)\frac{\partial \mu(\ell)}{\partial \theta_j}
     \label{eq:fisher_way2}
@@
 
+Once the Fisher matrix can be safely estimated, one can use the MOPED\footnote{Massively Optimised Parameter Estimation and Data compres-
+sion} algorithm \citep{2000MNRAS.317..965H,2016PhRvD..93h3525Z, 2017MNRAS.472.4244H}.
+In the case of the constant covariance matrix the algorithm leads to a  lossless compression which reduces possibly large data vector $x$ (eg.
+the $N=500$ $C_\ell^{p,q}$ values in the previous section) to $M$ numbers, where $M$ is the number of parameters $\theta_\alpha$ considered (eg.
+$M=2$ in the previous section).
+The algorithm computes by iteration $M$ orthogonal vectors of size $N$ such that (taking the notation of the previous section)
+\begin{equation}
+    b_i = \frac{C^{-1}\mu_{,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)b_j}{\sqrt{F_{i,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)^2}}
+\end{equation}
+where $\mu_{,i}=\partial \mu/\partial \theta_i$ ($i=1,\dots,M$). The vectors $b_i$, $b_j$ satisfy the following orthogonality constraint
+\begin{equation}
+    b_i^T C b_j = \delta_{i,j}
+\end{equation}
+and in turn from the data vector $x=(C_\ell^{p,q})_{p,q,\ell}$ one defines the compressed data set composed of $M$ numbers $(y_i)_{i\leq M}$ according to
+\begin{equation}
+    y_i = b_i^T x
+\end{equation}
+These numbers are uncorrelated and of unit variance and $(y_i)_{i\leq M}$ is a set of sufficient statistic for the set of parameters $(\theta_i)_{i\leq M}$.
+The case where the covariant matrix depends upon the parameters is discussed by  \citet{2017MNRAS.472.4244H}
+%
 \section{Application: Classical Hamiltonian Monte Carlo}

### 2022-08-26 paper_OJA.tex
Updates from Overleaf

@@
 sion} algorithm \citep{2000MNRAS.317..965H,2016PhRvD..93h3525Z, 2017MNRAS.472.4244H}.
-In the case of the constant covariance matrix the algorithm leads to a  lossless compression which reduces possibly large data vector $x$ (eg.
-the $N=500$ $C_\ell^{p,q}$ values in the previous section) to $M$ numbers, where $M$ is the number of parameters $\theta_\alpha$ considered (eg.
-$M=2$ in the previous section).
-The algorithm computes by iteration $M$ orthogonal vectors of size $N$ such that (taking the notation of the previous section)
+In the case of the constant covariance matrix the algorithm leads to a  lossless compression which reduces possibly large data vector $x$ of size $N$ to $M$ numbers, where $M$ is the number of parameters $\theta_i$ considered.
+For instance, in the previous section, $N=500$ as $x=(C_\ell^{p,q})$ and $M=2$ for $(\Omega_c,\sigma_8)$.
+The algorithm computes by iteration $M$ vectors of size $N$ such that (taking the notation of the previous section)
 \begin{equation}
@@
 \end{equation}
-where $\mu_{,i}=\partial \mu/\partial \theta_i$ ($i=1,\dots,M$). The vectors $b_i$, $b_j$ satisfy the following orthogonality constraint
+where $\mu_{,i}=\partial \mu/\partial \theta_i$ ($i=1,\dots,M$). The vectors $(b_i)_{i\leq M}$ satisfy the following orthogonality constraint
 \begin{equation}
@@
 \end{equation}
-and in turn from the data vector $x=(C_\ell^{p,q})_{p,q,\ell}$ one defines the compressed data set composed of $M$ numbers $(y_i)_{i\leq M}$ according to
+The algorithm is similar to the Gram-Schmidt process with a constant covariance matrix $C$to define the scalar product $\langle b_i, b_j\rangle$.
+Then, the original data set $x=(C_\ell^{p,q})_{p,q,\ell}$ is compressed in a data set composed of $M$ numbers $(y_i)_{i\leq M}$ according to
 \begin{equation}
@@
 \end{equation}
-These numbers are uncorrelated and of unit variance and $(y_i)_{i\leq M}$ is a set of sufficient statistic for the set of parameters $(\theta_i)_{i\leq M}$.
-The case where the covariant matrix depends upon the parameters is discussed by  \citet{2017MNRAS.472.4244H}
+These numbers are uncorrelated and of unit variance and $(y_i)_{i\leq M}$ is a set of sufficient statistic for the set of parameters $(\theta_i)_{i\leq M}$, $y_i$ contains the maximum information about the parameter $\theta_i$.The case where the covariant matrix depends upon the parameters is discussed by  \citet{2017MNRAS.472.4244H}.
+
+The net gain of MOPED algorithm is to eliminate the need of large covariance matrix inversion of size $N\times N$ requiring $O(N^3)$ operations.
+This inversion takes place not only for the Fisher matrix computation (Eq.~\ref{eq:fisher_way2}, but more importantly in the log-likelihood computation (see the snippet of the previous section).
+The MOPED algorithm reduces the complexity to $O(M)$ operations.
+
+To give an illustration here is a snippet that use the mock data set and the results of the Fisher matrix computation as in the previous section, to get the MOPED compressed data set composed of two numbers $(y_0,y_1)$ with maximal information on $(Omega_c, \sigma_8)$ respectively:
+\begin{lstlisting}[language=iPython]
+# Orthogonal vectors
+b0 = jc.sparse.dot(C_inv,dmu[:,0])/jax.numpy.sqrt(F[0,0])
+a = dmu[:,1].T @ b0
+b1 = (jc.sparse.dot(C_inv,dmu[:,1]) - a * b0)/jax.numpy.sqrt(F[1,1]-a*a)
+# MOPED vector
+y0 = b0.T @ data
+y1 = b1.T @ data
+\end{lstlisting}
+Then, the log-likelihood can easily be implemented as followed
+\begin{lstlisting}[language=iPython]
+@jax.jit
+def compress_likelihood(p):
+    # Create a new cosmology at these parameters
+    cosmo = jc.Planck15(Omega_c=p[0], sigma8=p[1])
+    # Compute mean Cl
+    mu = jc.angular_cl.angular_cl(cosmo, ell, tracers).flatten()
+    # likelihood using the MOPED vector
+    return -0.5 * ((y0 - b0.T @ mu)**2 + (y1 - b1.T @ mu)**2)
+\end{lstlisting}
 %

### 2022-09-13 paper_OJA.tex
Joe modif

@@
 \subsection{Instantaneous Fisher Forecasts}
-In this section, we illustrate with a purely fictional case study, the computation of the Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} by two methods thanks to the \textit{autodiff} ability of JAX.
-For the detailed implementation, the reader is invited to have a look at the following companion notebook \nblink{Simple-Fisher}.
-In this use-case we use four tracers redshift distributions: two to define \texttt{WeakLensing} kernels and two for \texttt{NumberCounts} kernels.
-Then, the $10$ angular power spectra $C_\ell^{p,q}$ ($p,q:1,\dots,4$) are computed for $50$  logarithmically spaced angular moments between $10$ and $1000$ using Equation \ref{eq:Cell_limber}.
+In this section, we illustrate, with a purely fictional case study, the computation of a Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} using two methods with the \textit{autodiff} ability of JAX.
+For the detailed implementation, the reader is invited to look at the following companion notebook \nblink{Simple-Fisher}.
+In this example we use four tracer redshift distributions: two to define \texttt{WeakLensing} kernels and two for \texttt{NumberCounts} kernels.
+Then, the $10$ angular power spectra $C_\ell^{p,q}$ ($p,q:1,\dots,4$) are computed for $50$  logarithmically-spaced angular moments between $\ell=10$ and $\ell=1000$ using Equation \ref{eq:Cell_limber}.
 The Gaussian covariance matrix is computed simultaneously.
@@
 %\end{minted}
-The \texttt{jc.sparse} functions are implementation of block matrix computations: sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_{diag})$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.
-The \texttt{jax.jit} decorator is there to build a compiled version of the function at the first use.
-The goal is to obtain the 1-sigma contour forecast of the two parameters ($\Omega_c, \sigma_8$).
-To do so, we use the Fisher matrix computation.
-One way is to use the Hessian of the log-likelihood as follows
+The \texttt{jc.sparse} functions are implementation of block matrix computations: a sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_{diag})$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.
+The \texttt{jax.jit} decorator builds a compiled version of the function at the first use.
+
+The first approach to obtaining approximate 1-sigma contours of the two parameters ($\Omega_c, \sigma_8$) with aa Fisher matrix uses the Hessian of the log-likelihood as follows:
 \begin{equation}
@@
 \end{equation}
-which is mostly  1-line JAX code
+which is accomplished in two lines of JAX code:
 \begin{lstlisting}[language=iPython]
-hessian_loglik = jax.jit(jax.hessian(likelihood))
-F = - hessian_loglik(params)
+hessian_loglike = jax.jit(jax.hessian(likelihood))
+F = - hessian_loglike(params)
 \end{lstlisting}
-An other way uses the computation of the Fisher matrix when the likelihood is a Gaussian distribution as ii is the case above.
-First, we define a function that computes the mean  $\mu(\ell; \theta)$, and it follows that the Fisher matrix element reads
+
+The second approach to computing the Fisher matric, which is restricted to Gaussian likelihoods but is more commonly used in the field because of its numerical stability, is to define a function that computes the observable mean  $\mu(\ell; \theta)$; it follows that the Fisher matrix elements are
 \begin{equation}
@@
 \end{equation}
-where $C^{-1}(\ell)$ is the covariance matrix computed with the fiducial cosmology. In this case one needs also just 1-line of JAX code 
+where $C^{-1}(\ell)$ is the covariance matrix computed with the fiducial cosmology. This can also be simply computed in JAX: 
 \begin{lstlisting}[language=iPython]
@@
 \end{lstlisting}
+\JZ{dmu in the above is not defined anywhere}
 
-The two methods agree perfectly as shown on Figure \ref{fig:simple_fisher_1}.
+JAX implementations of the two methods agree to near perfect accuracy, as shown in Figure \ref{fig:simple_fisher_1}.
 \begin{figure}
@@
 %
-It is worth to notice that in the two above methods of Fisher matrix computations, the user does not need to take care of particular parameter to compute the 1st or 2nd order derivatives.
-This is in contrast to the cases where one deals with finite difference methods.
-AS an illustration, we have used the CCL library to compute the Fisher matrix via Equation \ref{eq:fisher_way2}.
+It is worth noting that in the two methods for Fisher matrix computations described above, the user does not need to vary individual parameter values to compute the 1st or 2nd order derivatives; this is in contrast to the usual finite difference methods.
+As an illustration, we have used the CCL library to compute the Fisher matrix via Equation \ref{eq:fisher_way2}.
 To do so, the Jacobian ($\partial \mu/\partial\theta_\alpha$) is computed with centered  order 4 finite differences available in the \texttt{Numdifftools} Python package\footnote{\url{https://numdifftools.readthedocs.io}}.
-Using the spacing parameter values ($10^{-6}$, $10^{-2}$, $10^{-1}$) one obtains three different approximation of the 1-sigma contours as shown on Figure \ref{fig:simple_fisher_2}.
-One notices that the contour that agrees quite well with the \jaxcosmo\ one, is obtained with the spacing parameter equals to $10^{-2}$ which implies that the user has to  tune this parameter carefully.
-Although very simple, this case study exhibits the intrinsic limitation of finite difference methods for computing the Fisher matrix as it as been shown for instance in a more advanced case study in reference \citet{2021arXiv210100298B}.
+Using the parameter values spaced by ($10^{-6}$, $10^{-2}$, $10^{-1}$) one obtains three different approximation of the 1-sigma contours as shown on Figure \ref{fig:simple_fisher_2}.
+The contour that agrees best with the \jaxcosmo\ method is obtained with the intermediate spacing parameter $10^{-2}$, implying that the user must tune this parameter carefully.
+Although very simple, this case study exhibits a significant challenge of finite difference methods for computing the Fisher matrix as has been shown for instance in a more advanced case study in \citet{2021arXiv210100298B}.
+\JZ{Should probably latex the step sizes $10^{-1}, 10^{-2}, 10^{-6}$ in the figure legend}
 \begin{figure}
@@
     \includegraphics[width=\columnwidth]{figures/simple_fisher_2.png}
-    \caption{Computations of the Fisher matrix (Eq.\ref{eq:fisher_way1}) using the CCL library for the angular power spectra estimation and finite difference method to get the Jacobian  with different spacing parameters ($10^{-6}$, $10^{-2}$, $10^{-1}$).
+    \caption{Fisher matrices (Eq.\ref{eq:fisher_way1}) estimated using the CCL for the angular power spectra estimation and finite difference method to get the Jacobian  with different spacing of parameters ($10^{-6}$, $10^{-2}$, $10^{-1}$).
 For comparison, the contour obtained with \jaxcosmo\ is reproduced from Figure \ref{fig:simple_fisher_1} in black.}
@@
 \end{figure}
+
 %
@@
 sion} algorithm \citep{2000MNRAS.317..965H,2016PhRvD..93h3525Z, 2017MNRAS.472.4244H}.
-In the case of the constant covariance matrix the algorithm leads to a  lossless compression which reduces possibly large data vector $x$ of size $N$ to $M$ numbers, where $M$ is the number of parameters $\theta_i$ considered.
+In the case of the constant covariance matrix the algorithm compresses data in a way that is lossless at the Fisher matrix level (i.e.
+Fisher matrices estimated using the compressed and full data are identical, by construction)  which reduces a possibly large data vector $x$ of size $N$ to $M$ numbers, where $M$ is the number of parameters $\theta_i$ considered.
 For instance, in the previous section, $N=500$ as $x=(C_\ell^{p,q})$ and $M=2$ for $(\Omega_c,\sigma_8)$.
@@
 \end{equation}
-The algorithm is similar to the Gram-Schmidt process with a constant covariance matrix $C$to define the scalar product $\langle b_i, b_j\rangle$.
+The algorithm is similar to the Gram-Schmidt process, using the constant covariance matrix $C$ to define the scalar product $\langle b_i, b_j\rangle$.
 Then, the original data set $x=(C_\ell^{p,q})_{p,q,\ell}$ is compressed in a data set composed of $M$ numbers $(y_i)_{i\leq M}$ according to
@@
 \end{equation}
-These numbers are uncorrelated and of unit variance and $(y_i)_{i\leq M}$ is a set of sufficient statistic for the set of parameters $(\theta_i)_{i\leq M}$, $y_i$ contains the maximum information about the parameter $\theta_i$.The case where the covariant matrix depends upon the parameters is discussed by  \citet{2017MNRAS.472.4244H}.
+These numbers are uncorrelated and of unit variance and $(y_i)_{i\leq M}$ is a set of sufficient statistics for the set of parameters $(\theta_i)_{i\leq M}$, $y_i$ contains the maximum information about the parameter $\theta_i$.
+\JZ{I don't think this is true beyond the quadratic approximation to the log posterior - in particular I don't think these are a sufficient statistic for the full data set.
+If they were SS's then the curves in fig 6 would be identical.}
 
@@
 The MOPED algorithm reduces the complexity to $O(M)$ operations.
+\JZ{The primary advantage in cosmology at least is making the estimation of the covariance simpler, because you need far fewer simulations.}
 
-To give an illustration here is a snippet that use the mock data set and the results of the Fisher matrix computation as in the previous section, to get the MOPED compressed data set composed of two numbers $(y_0,y_1)$ with maximal information on $(Omega_c, \sigma_8)$ respectively:
+To give an illustration, this snippet uses the mock data set and results of the Fisher matrix computation from the previous section to obtain the MOPED compressed data set composed of two parameters $(y_0,y_1)$ with maximal information on $(Omega_c, \sigma_8)$:
 \begin{lstlisting}[language=iPython]
@@
 b1 = (jc.sparse.dot(C_inv,dmu[:,1]) - a * b0)/jax.numpy.sqrt(F[1,1]-a*a)
-# MOPED vector
+# MOPED vectors
 y0 = b0.T @ data
@@
 \end{lstlisting}
-Then, the log-likelihood can easily be implemented as followed
+Then, the log-likelihood can easily be implemented as:
 \begin{lstlisting}[language=iPython]
@@
 \end{lstlisting}
+The comparison between contour lines obtained with the original likelihood (uncompressed data set) and the MOPED version are shown in Figure \ref{fig:moped} in the case study of the previous section.
+Clearly, close to the negative likelihood minimum the lines agree very well.
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figures/moded.png}
+    \caption{Illustration of the log-likelihood contours obtained with an uncompressed data set (plain lines) and a MOPED version (dashed lines).} 
+    \label{fig:moped}
+\end{figure}
+%
+The case where the covariant matrix depends upon the parameters is discussed by  \citet{2017MNRAS.472.4244H} and leads to similar advantages of the MOPED compression algorithm.
+
 %

### 2022-09-13 paper_OJA.tex
MOPED update by Joe

@@
 
-The second approach to computing the Fisher matric, which is restricted to Gaussian likelihoods but is more commonly used in the field because of its numerical stability, is to define a function that computes the observable mean  $\mu(\ell; \theta)$; it follows that the Fisher matrix elements are
+The second approach to computing the Fisher matrix, which is restricted to Gaussian likelihoods but is more commonly used in the field because of its numerical stability, is to define a function that computes the observable mean  $\mu(\ell; \theta)$; it follows that the Fisher matrix elements are
 \begin{equation}
@@
 \begin{lstlisting}[language=iPython]
+# We define a parameter dependent function that computes the mean
+@jax.jit
+def jc_mean(p):
+    cosmo = jc.Planck15(Omega_c=p[0], sigma8=p[1])
+    # Compute signal vector
+    mu = jc.angular_cl.angular_cl(cosmo, ell, tracers)
+    # We want mu in 1d to operate against the covariance matrix
+    return mu.flatten() 
+# We compute it's jacobian with JAX, and we JIT it for efficiency
+jac_mean = jax.jit(jax.jacfwd(jc_mean))
+# We can now evaluate the jacobian at the fiducial cosmology
+dmu = jac_mean(params)
+# Now we can compose the Fisher matrix
 F = jc.sparse.dot(dmu.T, jc.sparse.inv(cov), dmu)
 \end{lstlisting}
-\JZ{dmu in the above is not defined anywhere}
+% done by JEC 13Sept22 \JZ{dmu in the above is not defined anywhere}
 
@@
 Although very simple, this case study exhibits a significant challenge of finite difference methods for computing the Fisher matrix as has been shown for instance in a more advanced case study in \citet{2021arXiv210100298B}.
-\JZ{Should probably latex the step sizes $10^{-1}, 10^{-2}, 10^{-6}$ in the figure legend}
+% done by JEC 13Sept22 \JZ{Should probably latex the step sizes $10^{-1}, 10^{-2}, 10^{-6}$ in the figure legend}
 \begin{figure}
@@
 \end{equation}
-These numbers are uncorrelated and of unit variance and $(y_i)_{i\leq M}$ is a set of sufficient statistics for the set of parameters $(\theta_i)_{i\leq M}$, $y_i$ contains the maximum information about the parameter $\theta_i$.
-\JZ{I don't think this is true beyond the quadratic approximation to the log posterior - in particular I don't think these are a sufficient statistic for the full data set.
-If they were SS's then the curves in fig 6 would be identical.}
+These numbers are uncorrelated and of unit variance and this construction ensures that the log-likelihood of $(y_i)_{i\leq M}$  $(\theta_i)_{i\leq M}$, $y_i$ is identical to that of $x$ up to second order, meaning that the Fisher matrices derived from the two parameters should be identical, and in general the $y$ values should lose very little information compared to the full likelihood.
 
-The net gain of MOPED algorithm is to eliminate the need of large covariance matrix inversion of size $N\times N$ requiring $O(N^3)$ operations.
+In problems where a (constant) covariance matrix is estimated from simulations, the number of such simulations required for a given accuracy typically scales with the number of data points used. 
+MOPED therefore greatly reduces this number, often by a factor of hundreds.
+Since the uncertainty in the covariance due to a finite number of simulations must be accounted for \citep{2018MNRAS.473.2355S,2007A&A...464..399H}, this reduction can also offset any loss of information from the compression.
+
+Another key advantage of the MOPED algorithm is to eliminate the need of large covariance matrix inversion of size $N\times N$ requiring $O(N^3)$ operations.
 This inversion takes place not only for the Fisher matrix computation (Eq.~\ref{eq:fisher_way2}, but more importantly in the log-likelihood computation (see the snippet of the previous section).
 The MOPED algorithm reduces the complexity to $O(M)$ operations.
-\JZ{The primary advantage in cosmology at least is making the estimation of the covariance simpler, because you need far fewer simulations.}
 
-To give an illustration, this snippet uses the mock data set and results of the Fisher matrix computation from the previous section to obtain the MOPED compressed data set composed of two parameters $(y_0,y_1)$ with maximal information on $(Omega_c, \sigma_8)$:
+To give an illustration, the following snippet uses the mock data set and results on the Fisher matrix computation from the previous section, to obtain the MOPED compressed data set composed of two parameters $(y_0,y_1)$ with maximal information on $(Omega_c, \sigma_8)$:
 \begin{lstlisting}[language=iPython]

### 2022-09-16 paper_OJA.tex
rephrasing

@@
 It has formed the seed for a wider ecosystem of packages, including, for example:
-\texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper.
+\texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro\ \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper.
 Other PPL packages such as \texttt{PyMC} \citep{Salvatier2016} have switched to JAX backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}.
@@
 We give an overview of the code's design and contents in Section~\ref{sec-jaxcosmo-design}.
-We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more advanced gradient-based cosmological inference with variants of Hamiltonian Monte Carlo and its variant No-U-Turn in Section~\ref{sec:chmc}, and in ML-accelerated Stochastic Variational Inference in Section~\ref{sec-SVI}.
+We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more complete gradient-based cosmological inference with variants of Hamiltonian Monte Carlo and its variant No-U-Turn, and ML-accelerated Stochastic Variational Inference in Section~\ref{sec:chmc}.
 We discuss and compare these methods in Section~\ref{sec-discussion} and conclude in Section~\ref{sec-conclusion}.
@@
 
-\section{Application: Fisher Forecasting}
+\section{Fisher Forecasting}
 \label{sec-fisher-forecast}
@@
 
-To give an illustration, the following snippet uses the mock data set and results on the Fisher matrix computation from the previous section, to obtain the MOPED compressed data set composed of two parameters $(y_0,y_1)$ with maximal information on $(Omega_c, \sigma_8)$:
+To give an illustration, the following snippet uses the mock data set and results on the Fisher matrix computation from the previous section, to obtain the MOPED compressed data set composed of two parameters $(y_0,y_1)$ with maximal information on $(\Omega_c, \sigma_8)$:
 \begin{lstlisting}[language=iPython]
@@
 %
-\section{Application: Classical Hamiltonian Monte Carlo}
+\section{Full Inference Methods}
 \label{sec:chmc}
@@
 %
-\subsection{Vanilla HMC}
+\subsection{Vanilla Hamiltonian Monte Carlo}
 %
@@
 
-The results in terms of relative effective sample sizes (aka "ESS") are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
+The results in terms of relative effective sample sizes (aka ``ESS'') are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
 One notices the dramatic improvement of the ESS by about a factor 10 using the NUTS sampler compared to the highly-optimized \texttt{Cobaya}  Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T} which has last about 40 hours on CPU to generate 70,000 samples.
@@
 
-\section{Application: Stochastic Variational Inference}
+\subsection{Stochastic Variational Inference}
 \label{sec-SVI}
@@
 
-Using the auto-differentation tool, one can use "black-box" guides (aka \textit{automatic differentiation variational inference}).
+Using the auto-differentation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}).
 As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
@@
 First is defined a bijective differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
-One can develop "AutoGuides" (\numpyro\ terminology) that can be adapted to the user models.
+One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models.
 Then, the cost function reads
@@
 \end{align}
-The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the "S" of SVI.
+The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
 As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX.
@@
 %
-\subsection{Neural Transport}
+\subsubsection{Neural Transport}
 \label{sec-Neural-Reparametrisation}
@@
 
-Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler, the chains have run 200 warm samples, and then 200 samples ("low" schema), 1000 samples on a single chain, and finally we have run 10 chains  with 1000 samples each.
+Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler, the chains have run 200 warm samples, and then 200 samples (``low'' schema), 1000 samples on a single chain, and finally we have run 10 chains  with 1000 samples each.
 All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
@@
 Naturally, more samples lead to better contour precision.
-But what is illustrative is the "low" schema results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
+But what is illustrative is the ``low'' schema results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
 Notice the good results even for the highly correlated lens bias parameters.
@@
 %
-\section{Discussion}
+\section{Full Inference Methods: Results}
 \label{sec-discussion}
@@
 \begin{table}[htb]
-\caption{Results of the \textit{relative effective sample size} (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}. The ESS can be larger than 100\% in some cases. 
+\caption{Results of the \textit{relative effective sample size} (aka ESS, ``bulk'') in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
+The ESS can be larger than 100\% in some cases.
 %{\color{red} 
@@
 The lens distributions $N(z)$ for five redshift bins are also extracted.
-The corresponding shapes of the different distributions are performed thanks to \texttt{jc.redshift.kde\_nz} function ("\texttt{jc}" hereafter stands for \texttt{jax\_cosmo}) and are presented in Figure \ref{fig-DESY1-src-lens-redshift}.
+The corresponding shapes of the different distributions are performed thanks to \texttt{jc.redshift.kde\_nz} function (``\texttt{jc}'' hereafter stands for \texttt{jax\_cosmo}) and are presented in Figure \ref{fig-DESY1-src-lens-redshift}.
 \begin{figure}

### 2022-10-26 paper_OJA.tex

@@
 \newcommand{\jaxcosmo}{\texttt{jax-cosmo}}
+\newcommand{\autodiff}{\texttt{autodiff}}
 
@@
 We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more complete gradient-based cosmological inference with variants of Hamiltonian Monte Carlo and its variant No-U-Turn, and ML-accelerated Stochastic Variational Inference in Section~\ref{sec:chmc}.
-We discuss and compare these methods in Section~\ref{sec-discussion} and conclude in Section~\ref{sec-conclusion}.
-
-% \section{JAX: Autograd meets XLA}
-
-\section{Modeling and implementation}
+We discuss and compare these methods in Section~\ref{sec-results} and conclude in Section~\ref{sec-conclusion}.
+
+\section{JAX: Autograd meets XLA}
+\label{sec-primer}
+
+The aim of this session is to provide a brief technical primer on JAX, necessary to fully grasp the potential of a cosmology library implemented in this framework.
+
+% Maybe a short history explaining what autograd and XLA are.
+\FrL{Add a short history of XLA and Autograd.}
+
+\FrL{Is this too much information below?
+On the one hand I don't want to go too technical, on the other hand, we talk about stuff like jax.grad/hessian, JIT, and vmap in the rest of the paper, it feels like these concepts should be introduced somewhere first.}
+
+\paragraph{Automatic Differentiation} Traditionally, two different approaches have been used in cosmology to obtain derivatives of given numerical expressions.
+
+The first approach is to derive analytically the formula for the derivatives of interest \citep{BORG}, with or without the help of tools such as \texttt{Mathematica}.
+This is however only practical typically for simple analytical models.
+
+The second approach is to compute numerical derivatives by finite differences.
+This approach can be applied on black-box models of abitrary complexity (from Boltzmann codes \FrL{some citations} to cosmological simulations \FrL{some citations}).
+However it is notoriously difficult to obtain stable derivatives by finite differences \FrL{cite papers trying to do fisher}.
+In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them completely inpractical whenever derivatives are needed on the fly.\FrL{this last sentence is not great.}
+
+Automatic differentiation frameworks like JAX take a different approach.
+They trace the execution of a given model and decompose this trace into fundamental operations with known derivatives (e.g.
+the derivative of a multiplication operation is known).
+Then, by applying the chain rule formula, the compatutional graph for the derivatives (of any order) of the model can be built from the known derivatives of every elementary operations.
+A new function corresponding to the derivative of the original function is therefore built automatically for the user.
+We direct the interested reader to \FrL{a good reference} for an in-depth introduction to automatic differentiation.
+
+JAX provides in particular a number of operators (\texttt{jax.grad}, \texttt{jax.jacobian}, \texttt{jax.hessian}) which can compute derivatives of any function written in JAX.
+\begin{lstlisting}[language=iPython]
+# Let us define a simple function 
+def f(x):
+	return y = 5 * x + 2
+# Take the derivative
+df_dx = jax.grad(f)
+# df_dx is a new function that always return 5 
+\end{lstlisting}
+
+\textbf{Why is that interesting?} \autodiff\ makes it possible to obtain \textit{exact gradients of cosmological likelihoods} with respect to all input parameters at the cost of only 2 likelihood evaluations.
+
+\paragraph{Just In Time Compilation (JIT)} Despite its convenience and wide adoption in astrophysics, Python still suffers from slow execution times compared to fully compiled languages such as C/C++.
+One approach to mitigate these issues and make Python code fast is Just In Time compilation, which traces the execution of a given Python function the first time it is called, and compiles it into a fast executable (by-passing the Python interpreter), which can be transparently used in subsequent calls to this function.
+Compared to other strategies for speeding up Python code \FrL{cpython and such}, JIT allows the user to simply write plain Python code, and reap \FrL{spell check} the benefits of compiled code.
+
+A number of libraries allowing for JIT have already been used in astrophysics, in particular Numba (ngmix), and \FrL{The JIT code developped by our friends in Zurich}.
+JAX stands out compared to these other frameworks in that it relies on the highly optimized XLA library \FrL{citation} for executing the compiled expressions.
+XLA is continuously developed by Google as part of the TensorFlow project, for efficient training and inference of large scale deep learning applications, and as such supports computational backends such as GPU an TPU clusters.
+\FrL{Maybe we want to say why GPU is such a good idea for our use cases.}
+
+In JAX, jitting is achieving simply by transforming a function with \texttt{jax.jit} operation:
+\begin{lstlisting}[language=iPython]
+# Let us redefine our function
+def f(x):
+	return y = 5 * x + 2
+# And JIT it
+jitted_f = jax.jit(f)
+# The first call to jitted_f will be relatively slow
+# subsequent calls will be extremely fast
+# and run as a compiled code directly on GPU
+\end{lstlisting}
+
+\textbf{Why is that interesting?} JIT makes it possible to execute entire cosmological \textit{MCMC chains directly on GPUs as compiled code}, with orders of magnitude in speedup over Python code.
+
+\paragraph{Automatic Vectorization} Another extremely powerful feature of JAX is its ability to automatically vectorize or paralellize any function.
+Through the same tracing mechanism used for automatic differentiation, JAX can decompose a given computation into fundamental operations and add a new \textit{batch} dimension so that the computation can be applied on a batch of inputs as opposed to individual inputs.
+In doing so, note that the computation will not run sequentially over all entries of the batch, but trully in parallel making full use of the intrinsic parallel architecture of modern GPUs.
+
+In JAX automatic vectorization is achieved using the \textit{jax.vmap} operation:
+\begin{lstlisting}[language=iPython]
+# Our function f only applies to scalars
+def f(x):
+	return y = 5 * x + 2
+# Applying automatic vectorization
+batched_f = jax.vmap(f)
+# batched_f now applies to 1D arrays
+\end{lstlisting}
+Again, we stress that in this example, \texttt{batched\_f} will not be implemented in terms of a for loop, but with operations over vectors.
+The function above is trivial, but the same approach can be used to parallelize any function, from Limber integrals, to an entire likelihood.
+
+\textbf{Why is that interesting?} Automatic Vectorization makes it possible to trivially parallelize cosmological likelihood evaluations, to run many parallel MCMC chains on a single GPU.
+
+\paragraph{NumPy API compliance} Finally, the last point to note about JAX, is that it mirrors the NumPy API (with only a few exceptions).
+This means in practice that existing NumPy code can easily be converted to JAX.
+This is in contrast to other similar frameworks like TensorFlow, PyTorch, or Julia which all require the user to learn, and adapt their code to, a new API or even a new language.
+
+\textbf{Why is that interesting?} NumPy compliance implies improved maintainability and lower barrier to entry for new contributors.
+
+\section{Design of the \jaxcosmo\ library}
 \label{sec-jaxcosmo-design}
+\FrL{I'll do another pass over this section when I have internet, I want to compare to what we did in CCL.}
+
 In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX.
-The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though the present proof-of-concept \jaxcosmo\ does not include as many features and options as CCL.
+The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though \textbf{in its initial release, }
+ \jaxcosmo\ only implements a subset of CCL features and options.
 
 The cosmological parameters are organised in a JAX \textit{pyTree} container\footnote{\url{https://jax.readthedocs.io/en/latest/jax-101/05.1-pytrees.html}} \texttt{jc.Cosmology} where \texttt{jc} hereafter stands for \jaxcosmo.
-This container allows one to use \textit{autodiff} and \texttt{vmap} transparently.
+This container allows one to use \textit{autodiff} and \texttt{vmap} transparently \FrL{I don't think people know what autodiff and vmap are.}.
 The present set of exposed parameters is: $\Omega_{cdm}$ the cold dark matter density fraction, $\Omega_b$ the Baryonic matter density fraction, $h$ the Hubble constant divided by $100$~km/s/Mpc, $n_s$ the primordial scalar perturbation spectral index, $\sigma_8$ the variance of matter density perturbations at an $8$~Mpc/h scale, $\Omega_k$ the curvature density fraction, $w_0$ the first order term of dark energy equation and $w_a$ the second order term of dark energy equation of state.
 Notably, \texttt{jc.Planck15} instantiates a set of cosmological parameters inspired from the Planck 2015 results \citep{2016A&A...594A..13P}.
+\FrL{I'm not sure we want to keep this paragraph, it's not extremely useful. The only useful thing to discuss I guess might be the choice of parametrisation in terms of A or sigma8.}
 
@@
 To perform \textit{autodiff} we have implemented various tools as differentiable ODEs using the Runge-Kutta algorithm, as well as Simpson quadrature and the Romberg algorithm for numerical integration.
+\FrL{Hummm it's a bit strange to talk about the theory and implementation in the same paragraph...
+Especially because we don't specify which concrete implementation is used here.}
 
@@
 
-To make a bridge to these observables, two \textit{pyTree} objects are implemented: \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing samples, with each instance of each representing a single tomographic redshift bin's number density.
+To make a bridge to these observables, two \textit{pyTree} objects are implemented: \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing samples, with each instance of each representing a single tomographic redshift bin's number density \FrL{We probably should refer to those as Probes instead of Pytrees, in general we should separate the implementation details from the theory}.
 These tracers are implemented as two kernel functions:
@@
 
+\FrL{We probably actually want to start by the Limber formula, which explicits what a kernel is, otherwise people don't necessarily know where these kernels come from.}
 From these tracer objects, the angular power spectra $C_\ell^{ij}$ for the probes $(i,j)$ and for redshift bin window selections are computed in the first order Limber approximation \citep{PhysRevD.78.123506}:
@@
 \end{equation}
+\FrL{At which z ?}
 with $R_8 = 8$ in unit of Mpc and $W(x)$ is related to the $j_1$ spherical Bessel function as
@@
 
-Remarkably, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters thanks to \textit{autodiff} and the implementation of the above mentioned integration quadrature.
+Remarkably, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters thanks to \textit{autodiff} and the implementation of the above mentioned integration quadrature \FrL{Are we using quadratures currently?
+I think by default we are using simple Simpson no?}.
 An example is given in the context of DES Y1 3x2pts analysis (Appendix~\ref{app-DESY1}).
@@
 show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non linear matter power spectrum computation using the \texttt{halofit} fitting function and the angular power spectrum for galaxy-galaxy lensing (Eq.~\ref{eq:Cell_limber}) using the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions.
-The relative differences between the two libraries are at the level of few $10^{-3}$ or better.
+\FrL{In addition, \jaxcosmo features a suite of validation tests against CCL, automatically validating the precision of all computations to within the desired numberical accuracy.} The relative differences between the two libraries are at the level of few $10^{-3}$ or better.
 
@@
 
-\section{Fisher Forecasting}
+\section{Unleashing the Power of Fisher Forecasts with Automatic Differentiation}
 \label{sec-fisher-forecast}
-% illustrates why autodiff is interesting from several points of view
-
-\FrL{Add description of problem and small demos}
+
+As a first illustration of the interest of \textit{autodiff} in cosmological computations, we present in this section a few direct applications involving the computation of the Fisher information matrix, widely used in cosmology.
+\\
+Not only does the computation of the Fisher matrix become a trivial business, but the Fisher matrix itself becomes differentiable, allowing in particular for powerful survey optimization applications.
 
@@
 \subsection{Instantaneous Fisher Forecasts}
-In this section, we illustrate, with a purely fictional case study, the computation of a Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} using two methods with the \textit{autodiff} ability of JAX.
+\FrL{We should discuss why Fisher matrices are usually difficult first.}
+We first illustrate, with a purely fictional case study, the computation of a Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} using two methods with the \textit{autodiff} ability of JAX.
 For the detailed implementation, the reader is invited to look at the following companion notebook \nblink{Simple-Fisher}.
@@
 % Maybe this can be only mentioned in the discussion.
-
+\FrL{Insist here on the fact that the FoM itself is differentiable.}
 Fisher forecasts are also commonly used in survey and analysis strategy strategy, where running a full posterior analysis for each possible choice would be unfeasible.
@@
 Since the uncertainty in the covariance due to a finite number of simulations must be accounted for \citep{2018MNRAS.473.2355S,2007A&A...464..399H}, this reduction can also offset any loss of information from the compression.
+\FrL{hummm but we still do need a full covariance in the compression algorithm, right?} \FrL{And we still do need to invert it...
+is the idea rather that if we dont have a perfect matrix at that stage the worse that can happen is sub optimal compression?}
 
@@
 %
-The case where the covariant matrix depends upon the parameters is discussed by  \citet{2017MNRAS.472.4244H} and leads to similar advantages of the MOPED compression algorithm.
-
-%
-\section{Full Inference Methods}
+The case where the covariance matrix depends upon the parameters is discussed by  \citet{2017MNRAS.472.4244H} and leads to similar advantages of the MOPED compression algorithm.
+
+%
+\section{Posterior Inference made fast by Gradient-Based Inference Methods}
 \label{sec:chmc}
 %
-In the following sections we review some statistical methods used to make the cosmological parameter inference with \jaxcosmo.
-We have used a variant of Hamiltonian Monte Carlo knows as \textit{No-U-Turn} (NUTS) and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI).
+In the following sections we review some statistical methods which directly benefit from the automatic differentiablity of \jaxcosmo likelihoods.
+We demonstrate a range of gradient-based methods, from Hamiltonian Monte Carlo (HMC), and its \textit{No-U-Turn} (NUTS) variant, to Stochastic Variational Inference (SVI).
+We further report on their respective computational costs in a DES-Y1 like analysis.
+All  the methods have been implemented using the \numpyro\ probabilistic programming language (PPL).
+%
+% and NUTS after a \textit{Neural Transport} perform using a \textit{Stochastic Variational Inference} (aka SVI).
 All  the methods have been implemented using the \numpyro\ probabilistic programming language (PPL).
@@
 calculating a Hessian, as in \jaxcosmo, makes it feasible and efficient.
+\FrL{I don't think it uses a Hesssian?
+you only need the derivative of the log posterior wrt parameters.
+The hessian would be needed to get a good estimate of M.}
 
@@
 The next section addresses this problem thanks to the No-U-Turn HMC version.
+
+\FrL{Needs a little paragraph to explain with HMC is supposed to be better than plain Metropolis-Hastings.}
 
@@
 
-The No-U-Turn (aka NUTS))  variant of the traditional HMC sampler has been introduced in reference \citep{nuts}.
+The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler has been introduced in \citet{nuts}.
 It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$.
@@
 As stated by the authors, NUTS  uses a recursive algorithm to build a set of likely candidate samples that spans a wide swath of the target distribution, stopping automatically when it starts to double back and retrace its steps.
-So, in a automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
+\FrL{Parameters L an epsilon should be introduced at the level of HMC then.} So, in an automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
 The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls at each generation.
@@
 Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
-After the forward model creation for the DES-Y1 3x2pts exercise described in Appendix~\ref{app-DESY1}, we have performed
+\FrL{class inheritance is TMI.}  After the forward model creation for the DES-Y1 3x2pts exercise described in Appendix~\ref{app-DESY1} \FrL{We should introduce the model at the beginning of this section.}, we have performed
 \begin{itemize}
-    \item first a reparametrisation of the latent variable uniform priors (Table~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$;
-    \item second for the warm-up phase we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
+    \item first a reparametrisation of the latent variable uniform priors (Table~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$; \FrL{these are not latent variables, but model parameters.
+Also people don't know what this reparametrisation means.}
+    \item second for the warm-up phase \FrL{People don't know what the warmup phase is...} we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
 This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
 \end{itemize}
-The NUTS sampler have been running with 16 chains of 1,000 samples each (\texttt{vectorized} option) after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie.
+The NUTS sampler have been running with 16 chains of 1,000 samples each (\texttt{vectorized} option) \FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.} after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie.
 128 steps for each iteration).
-The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or a NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
+The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
 The elapse time for these experiments was 20 hours.
@@
 The CL contours are also in very good agreement.
-It is worth to point out that the mass matrix structuring operation is a very effective as it increases the sampling efficiency by about a factor 2.
+\FrL{we should rephrase this a bit to say that Cobaya is our reference.} It is worth to point out that the mass matrix structuring operation is a very effective as it increases the sampling efficiency by about a factor 2.
 Anticipating the Section~\ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
@@
 If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
-The VI aims to get an approximation of this distribution, ie.
+VI aims to get an approximation of this distribution, ie.
 $p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
@@
 \end{equation}
-which exhibit the Jacobian of the $T^{-1}$ transformation. The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
+which exhibit the Jacobian of the $T^{-1}$ transformation.
+The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
+\FrL{we should say that before.}
 
@@
 \end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model. Then,
+where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model.
+\FrL{Hummm this is a little bit too specific, in practice I would use a flow rather than an MVN....
+Maybe this can go to an appendix for people who want to understand how this works on a concrete example?} Then,
 to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to\footnote{To simplify the notation, $T^{-1}(S_\lambda^{-1}(\zeta))$ has been replaced by $z$.}
@@
 \end{align}
+\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.} \FrL{let's move the MVN details to an appendix.}
 The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
@@
 Now even if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to a NUTS sampler which is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}.
+\FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.}
+\FrL{we don't have plots of just the VI?}
 %
@@
 It takes 2 hours on the NVidia V100 GPU.
+\FrL{Arf...
+2 hours !}
 
@@
 %
-\section{Full Inference Methods: Results}
-\label{sec-discussion}
+\subsection{Results}
+\label{sec-results}
 %
@@
 
+\section{Discussion}
+\label{sec-discussion}
+
+Having demonstrated the interest of \jaxcosmo\ as a differentiable cosmology library library, we now discuss several limitations, and open questions.
+
+%Lack of autodiff Boltzmann code (emulator)
+The first essential barrier to a full-fledged automatically differentiable cosmology library is the need for a differentiable Boltzmann solver to compute the matter power spectrum.
+At this stage, \jaxcosmo\ relies on the analytic Einsenstein \& Hu fitting formula, which is not accurate enough for Stage IV requirements, nor does it allow to explore beyond $\Lambda$CDM models.
+Existing solvers such as CLASS or CAMB \FrL{citations} are large and complex codes which are not easily reimplemented in an \autodiff\ framework and therefore cannot be directly integrated in \jaxcosmo\ .
+
+A first option to resolve this issue is to implement from scratch a new Boltzmann code in a framework that will support automatic differentiation.
+This is the approach behind works such as \FrL{the Julia code} which is provides a simplified Boltzmann solver in Julia, or \FrL{the code from our friends in Zurich} which is based on a SymPy and could be relatively easily compatible with JAX.
+However, even if very promising, both of these options thus far remain limited.
+While we do believe a automatically differentiable Boltzmann code is the best option, it seems that the cost of developing such a code remains very high at this time.
+
+A second approach is to develop emulators of a full-fledged Boltzmann code.
+Emulators based on neural networks or Gaussian processes are themselves automatically differentiable with respect to cosmological parameters.
+In fact, the litterature is now rich in examples of such emulators \FrL{cite all emulators}.
+After validating their accuracy against the reference CAMB or CLASS implementation, they can be directly integrated as a plug-and-play replacement for the computation of the matter power spectrum.
+\FrL{need to talk about cosmopower somewhere.} At this time, it seems that using emulators will be the most straightforward approach to bring more accurate models to \jaxcosmo\ .
+\FrL{However, we want to emphasize that one of the reason for the wide diversity of emulators is a lack of standardization}.
+
+\bigskip
+
+% Reasoning behind not making a full emulator of the correlation functions
+Connected to this discussion about emulators, one point that could be discussed is whether one even needs a library like \jaxcosmo\ if anyway one can build emulators of a CCL likelihood for use inside gradient-based inference algorithms.
+While this could indeed be feasible, and of similar cost as just making an emulator of the matter power spectrum, the drawback of this approach is that many analysis parameters and choices become hardcoded in the emulator.
+While the model for the linear matter power spectrum is typically kept fixed in practical analyses, all the of the choices related in particular to systematics modeling (e.g.
+photometric redshift errors, galaxy bias, or intrinsic galaxy alignments) are very likely to vary significantly in the process of developing the analysis.
+Building an emulator for the likelihood would mean the emulator needs to be retrained everytime the likelihood is changed.
+
+\bigskip
+
+%  - Massively Parallel MCMC on GPU (linked to vmap and/or pmap)
+%  - Efficiency of gradient-based inference methods
+Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo\ .
+As we illustrated in the previous section, gradient-based inference techniques yield significantly less correllated MCMC chains.\FrL{Hummm here I guess we want to say also it scales better with dimensions, and that you can get very fast first results.} \FrL{we also want to say that the GPU parallelisation is going to change the game, all the more when you account for new massive GPU machines.}
+
+\bigskip
+
+%  - Respective role of CCL and jax-cosmo
+Finally, one question worth asking is how does \jaxcosmo\ position itself against classical codes such as CCL?
+Or in other terms, will it replace CCL?
+While we are convinced of the benefits of a JAX implementation, we expect CCL to remain extremely valuable as the reference cosmology implementation.
+\FrL{it allows to leverage the legacy codes, has a lot more features, etc.} Ultimately, a natural transition may occur towards more modern frameworks like \jaxcosmo\ when they reach the ability to run full-fledged Stage IV likelihoods.
+
 \section{Conclusions \& Prospects}
 \label{sec-conclusion}
-
-\FrL{mention something about scaling with parallel MCMC}
-
-\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
-
-\subsection{Lightning fast inference with BayesFast}
-
-%
-%
-\subsection{DES-Y1 Example}
+%- We have presented a library using automatic differentiation and and automatic GPU offload for a set of key cosmology theory calculations
+%- We currently cover only a small number of the many calculations needed in cosmo theory, and welcome contributions for other areas such as CMB.
+%- The calculation of Fisher matrices using this method, which is infamous for requiring extensive careful tuning, becomes very simple and robust using JC.
+%- We have demonstrated the efficiency in JC of the derivative-aware MCMC methods HMC and NUTS, which are regarded as the only way for samplers to evade the curse of dimensionality up to the hundreds of dimensions we are likely to need for the next generation of surveys.
+%- We have shown a proof-of-concept for the use of JC with machine learning methods, opening a whole new space of methods in cosmology.
+
+In this paper, we have presented \jaxcosmo\ a cosmology library implemented in the JAX framework, enabling the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, as well as greatly speeding up likelihood evaluations thanks to automatic parallelisation and just in time compilation on GPUs.
+Currently, \jaxcosmo\ only contains a small set of features limited corresponding to a DES-Y1 3x2pt analysis.
+Being an open source project, contribution of additional features is additional areas such as CMB or spectroscopic galaxy clustering are warmly welcome.
+
+To demonstrate the interest of an automatically differentiable library, we have illustrated on concrete examples how Fisher matrices, which are notoriously difficult to compute and require extensive and careful fine tuning, can now be computed robustly and at much lower cost.
+In addition, Fisher matrices becomes themselves differentiable, which allows for  Figure of Merit optimization by gradient descent, making survey optimization extremely simple.
+
+Going beyond Fisher forecasts, we have also compared to simple Metropolis-Hastings several gradient-based inference techniques (namely, Hamiltonian Monte-Carlo, No-U-Turn-Sampler, Stochastic Variational Inference).
+We show that the posterior samples with gradient-based methods have a much larger Effective Sample Size, meaning that \FrL{let's think a bit more about interpretation here....
+the thing I care about is not as much the ESS, but the number of likelihood evaluation per independent sample.
+I don't think we have quantified that too much}.
+These inference techniques can scale to hundreds of dimensions, and may become necessary in Stage IV analysis as the number of nuisance parameters is likely to become very large.
+
+\bigskip
+
+The next extensions of this framework will be the inclusion of additional cosmological probes, as well as integrating emulators for the matter power spectrum trained on CAMB or CLASS, as a means to go beyond the current analytic Eisenstein \& Hu model.
+
+\bigskip
+
+In the spirit of reproducible research, all results presented in this paper can be reproduced from the following GitHub repository:
+
+\FrL{add link}
+
+%
+%
+%
+%\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
+%
+%%
+%%
+%\subsection{DES-Y1 Example}
 %
@@
   \multicolumn{2}{c}{\textbf{Cosmology}} \\
-$\Omega_b$ & $\mathcal{U}[0.10, 0.9]$ \\
-$\Omega_c$ & $\mathcal{U}[0.40, 1.0]$ \\
-$\sigma_8$ & $\mathcal{U}[0.03, 0.07]$ \\
+$\Omega_c$ & $\mathcal{U}[0.10, 0.9]$ \\ %JEC 21/10/22 there was exchanges in Omega_c, sigma8, Omega_b intervalle (but ok in the notebooks)
+$\sigma_8$ & $\mathcal{U}[0.40, 1.0]$ \\
+$\Omega_b$ & $\mathcal{U}[0.03, 0.07]$ \\
 $w_0$ & $\mathcal{U}[-2.00, -0.33]$ \\
@@
 \multicolumn{2}{c}{\textbf{Source photo-$z$ shift}} \\
-$dz_1$ & $\mathcal{N}[0.001,0.016]$ \\
+$dz_1$ & $\mathcal{N}[0.001,0.016]$ \\ % JEC note the sign is the one we have used but is different from https://arxiv.org/pdf/1708.01530.pdf
 $dz_2$ & $\mathcal{N}[-0.019,0.013]$ \\

### 2022-11-10 paper_OJA.tex
Updates from Overleaf

@@
 
+%to cross out a sentence: \st{Hello World}
+\usepackage{soul}
+
 %%%% usepackage{minted}
@@
 \newcommand{\JZ}[1]{{\color{purple}JZ: #1}}
+\newcommand{\JEC}[1]{{\color{green}JEC: #1}}
 
@@
 the derivative of a multiplication operation is known).
-Then, by applying the chain rule formula, the compatutional graph for the derivatives (of any order) of the model can be built from the known derivatives of every elementary operations.
+Then, by applying the chain rule formula, the computational graph for the derivatives (of any order) of the model can be built from the known derivatives of every elementary operations.
 A new function corresponding to the derivative of the original function is therefore built automatically for the user.
@@
 \begin{equation}
-    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k,z) W^2(kR_8) \right)^{-1}
-\end{equation}
-\FrL{At which z ?}
+    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
+\end{equation}
+\JEC{\st{At which z ?}}
 with $R_8 = 8$ in unit of Mpc and $W(x)$ is related to the $j_1$ spherical Bessel function as
@@
 Remarkably, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters thanks to \textit{autodiff} and the implementation of the above mentioned integration quadrature \FrL{Are we using quadratures currently?
-I think by default we are using simple Simpson no?}.
+I think by default we are using simple Simpson no?}\JEC{Correct: the current pip version uses Simpson but it is a quadrature even if it is qualified "simple".}.
 An example is given in the context of DES Y1 3x2pts analysis (Appendix~\ref{app-DESY1}).
@@
 In the case of the constant covariance matrix the algorithm compresses data in a way that is lossless at the Fisher matrix level (i.e.
-Fisher matrices estimated using the compressed and full data are identical, by construction)  which reduces a possibly large data vector $x$ of size $N$ to $M$ numbers, where $M$ is the number of parameters $\theta_i$ considered.
-For instance, in the previous section, $N=500$ as $x=(C_\ell^{p,q})$ and $M=2$ for $(\Omega_c,\sigma_8)$.
+Fisher matrices estimated using the compressed and full data are identical, by construction)  which reduces a possibly large data vector $\mu$ of size $N$ to $M$ numbers, where $M$ is the number of parameters $\theta_i$ considered.
+For instance, in the previous section, $N=500$ as $\mu=(C_\ell^{p,q})$ and $M=2$ for $(\Omega_c,\sigma_8)$.
 The algorithm computes by iteration $M$ vectors of size $N$ such that (taking the notation of the previous section)
@@
 \FrL{hummm but we still do need a full covariance in the compression algorithm, right?} \FrL{And we still do need to invert it...
-is the idea rather that if we dont have a perfect matrix at that stage the worse that can happen is sub optimal compression?}
+is the idea rather that if we dont have a perfect matrix at that stage the worse that can happen is sub optimal compression?}\JEC{Well there are several ways that $C$ can be questioned: parameter dependence (ie.
+fiducial model/true model), non-gaussianities, errors...
+So not sure that all these effects induce only sub-optimal compression.
+But it may be out of the scope of this paragraph/paper.}
 

### 2022-11-10 paper_OJA.tex
leapfrog algo =>L & eps parameters introduced in HMC

@@
 \end{equation}
-At each sample, a trajectory is initialized with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration (so called \textit{leapfrog} algorithm):
+At each sample, a trajectory is initialized with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration :
 \begin{align}
-\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial \log{\cal P}}{\mathrm{d} q}\\
+\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial \log{\cal P}}{\mathrm{d} q} \\
 \frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
 \end{align}
-where $M$ is a mass matrix which should be set to approximate the covariance of the posterior. This is also used to set the scale of the random initial velocities.
+where $M$ is a mass matrix which should be set to approximate the covariance of the posterior.
+This is also used to set the scale of the random initial velocities.
+\JEC{These differential equations may be integrated iterating over $L$ steps the following \textit{leapfrog} algorithm
+\begin{align}
+    p_{n+\frac{1}{2}} &= p_n -\frac{\varepsilon}{2} \frac{\partial V}{\mathrm{d} q}(q_n) \\
+    q_{n+1} & = q_n +\varepsilon M^{-1} p_{n+\frac{1}{2}} \\
+    p_{n+1} &=  p_{n+\frac{1}{2}} -\frac{\varepsilon}{2} \frac{\partial V}{\mathrm{d} q}(q_{n+1})
+\end{align}
+where $\varepsilon$ is a step size parameter.}
 
@@
 you only need the derivative of the log posterior wrt parameters.
-The hessian would be needed to get a good estimate of M.}
+The hessian would be needed to get a good estimate of M.}\JEC{Yes.
+Why the Hessian?}
+
+\FrL{Needs a little paragraph to explain with HMC is supposed to be better than plain Metropolis-Hastings.}
+\JEC{Metropolis-Hastings suffers from random walk to explore the geometry of the target distribution to get the region of maximum probability (ie.
+typical set) which is particularly  penalising in large dimensions.
+The HMC dynamics allow to make large jumps to explore the region around a sample to guess the next good one.}
 
@@
 The next section addresses this problem thanks to the No-U-Turn HMC version.
-
-\FrL{Needs a little paragraph to explain with HMC is supposed to be better than plain Metropolis-Hastings.}
 
@@
 As stated by the authors, NUTS  uses a recursive algorithm to build a set of likely candidate samples that spans a wide swath of the target distribution, stopping automatically when it starts to double back and retrace its steps.
-\FrL{Parameters L an epsilon should be introduced at the level of HMC then.} So, in an automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
+\FrL{\st{Parameters L an epsilon should be introduced at the level of HMC then.}}\JEC{done} So, in an automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
 The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls at each generation.

### 2022-11-10 paper_OJA.tex
colorJEC
some modifs following FL

@@
 \newcommand{\JZ}[1]{{\color{purple}JZ: #1}}
-\newcommand{\JEC}[1]{{\color{green}JEC: #1}}
+\newcommand{\JEC}[1]{{\color{magenta}JEC: #1}}
 
@@
 However it is notoriously difficult to obtain stable derivatives by finite differences \FrL{cite papers trying to do fisher}.
-In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them completely inpractical whenever derivatives are needed on the fly.\FrL{this last sentence is not great.}
+In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them completely impractical whenever derivatives are needed on the fly.\FrL{this last sentence is not great.}
 
@@
 Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
-\FrL{class inheritance is TMI.}  After the forward model creation for the DES-Y1 3x2pts exercise described in Appendix~\ref{app-DESY1} \FrL{We should introduce the model at the beginning of this section.}, we have performed
+\FrL{class inheritance is TMI.}  After the forward model creation for the DES-Y1 3x2pts exercise described in Appendix~\ref{app-DESY1} \FrL{We should introduce the model at the beginning of this section.},
 \begin{itemize}
-    \item first a reparametrisation of the latent variable uniform priors (Table~\ref{tab-DESY1}) to use a unified uniform prior $\mathcal{U}[-5,5]$; \FrL{these are not latent variables, but model parameters.
+    \item \st{a reparametrisation of the latent variable uniform priors to use a unified uniform prior $\mathcal{U}[-5,5]$} \JEC{we have first applyied  an affine transformation on cosmological, intrinsic alignment and biais parameters to use an unique uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1})}; \FrL{these are not latent variables, but model parameters.
 Also people don't know what this reparametrisation means.}
-    \item second for the warm-up phase \FrL{People don't know what the warmup phase is...} we ask for mass matrix optimisation with essentially a diagonal form but with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
+    \item  \FrL{People don't know what the warmup phase is...}\JEC{we use a structured mass matrix consisting on essentially a diagonal form  with a block structure for the following sets of parameters} $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
 This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
@@
 \FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.}
-\FrL{we don't have plots of just the VI?}
+\FrL{we don't have plots of just the VI?}\JEC{SVI alone (ie. not followed by HMC) gives poor results.}
 %
@@
 Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo\ .
-As we illustrated in the previous section, gradient-based inference techniques yield significantly less correllated MCMC chains.\FrL{Hummm here I guess we want to say also it scales better with dimensions, and that you can get very fast first results.} \FrL{we also want to say that the GPU parallelisation is going to change the game, all the more when you account for new massive GPU machines.}
+As we illustrated in the previous section, gradient-based inference techniques yield significantly less correlated MCMC chains.\FrL{Hummm here I guess we want to say also it scales better with dimensions, and that you can get very fast first results.} \FrL{we also want to say that the GPU parallelisation is going to change the game, all the more when you account for new massive GPU machines.}
 

### 2022-11-10 paper_OJA.tex
more adds
Start mv MVN SVI (sec. 5.3) to Appendix

@@
 function, taking derivatives, and running a gradient-based or coordinate-ascent optimization.
-First is defined a bijective differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
+First is defined a invertible differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
+The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
 One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models.
@@
 which exhibit the Jacobian of the $T^{-1}$ transformation.
-The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
-\FrL{we should say that before.}
-
-Now, one is faced with the evaluation of the expectations during gradient descent.
-The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}).
-Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
-a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
-One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg.
-$\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$.
-The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
+\JEC{
+The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{2015arXiv150603431K}).
+Let us illustrated the method using an invertible transformation (aka \textit{normalizing flow})  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, such that $\zeta\sim \mathcal{N}(0,I)$.
+The Jacobian of this distribution is 1 by definition (volume conservation), so the lost function reads
 \begin{multline}
@@
 \end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model.
-\FrL{Hummm this is a little bit too specific, in practice I would use a flow rather than an MVN....
-Maybe this can go to an appendix for people who want to understand how this works on a concrete example?} Then,
+where $\mathbb{H}(q)$ is the Shannon entropy of the $q$ distribution, so its gradient can be computed once for all for a given $q$ distribution family and reused in any user model.  
+Then,
 to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to\footnote{To simplify the notation, $T^{-1}(S_\lambda^{-1}(\zeta))$ has been replaced by $z$.}
@@
 \end{multline}
-For the use case of the MVN (auto)guide, one gets
-\begin{align}
--\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
-+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
-\right\} \nn \\
--\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
-&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
-\right\}+ (L^{-1})^T
-\end{align}
-\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.} \FrL{let's move the MVN details to an appendix.}
-The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
-As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX.
-Now even if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to a NUTS sampler which is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}.
+An concrete example of the gradient formula is given in Appendix \ref{app-SVI_MVN} with the multivariate Gaussian distribution family used as the guide.
+}%JEC
+%
+
+One can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to a NUTS sampler which is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}.
 \FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.}
-\FrL{we don't have plots of just the VI?}\JEC{SVI alone (ie. not followed by HMC) gives poor results.}
+\FrL{we don't have plots of just the VI?}\JEC{Good point: I have SVI alone with both MVN or Block Neural Autoregressive Flow. The contours are narrower which would require to elaborate...}
 %
@@
 Notice however that the notebooks are only illustrative and for a real sample production the authors invite the reader to run their own script.
+
+\section{SVI example with MVN guide}
+\label{app-SVI_MVN}
+\JEC{To be cleaned}
+Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
+a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
+One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg.
+$\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$.
+The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
+\begin{multline}
+-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
+\label{eq-loss-svi-3}
+\end{multline}
+where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model.  
+For the use case of the MVN (auto)guide, one gets
+\begin{align}
+-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
++ \nabla_\xi \log|J_{T^{-1}}(\xi)|
+\right\} \nn \\
+-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
+&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
+\right\}+ (L^{-1})^T
+\end{align}
+\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.} \FrL{let's move the MVN details to an appendix.}
+The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
+As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX.
+
 %% If you have bibdatabase file and want bibtex to generate the

### 2022-11-11 paper_OJA.tex
SVI appendix, SVI alone plots, Neural Transport... show listing codes

@@
 \centering
-\includegraphics[width=1.5\columnwidth]{figures/Cobaya-NUTS-SVI200.png}
-\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue, 70,000 samples), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red; 16,000 samples), and a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green, 200 samples).}
+\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_v1.pdf}
+\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue, 70,000 samples), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red; 16,000 samples), and a SVI MVN optimisation followed by a Neural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green, 200 samples).}
 \label{fig_cobaya_NUTS_SVI}
@@
 The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{2015arXiv150603431K}).
-Let us illustrated the method using an invertible transformation (aka \textit{normalizing flow})  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, such that $\zeta\sim \mathcal{N}(0,I)$.
+Let us illustrated the method using an invertible transformation (aka \textit{normalizing flow})  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, such that $\zeta\sim \mathcal{N}(0,I)$\footnote{Notice that $z=T^{-1}(\xi)=(T^{-1} \circ S_\lambda^{-1})(\zeta)=(S_\lambda \circ T)^{-1}(\zeta)=F_\lambda(\zeta)$.}.
 The Jacobian of this distribution is 1 by definition (volume conservation), so the lost function reads
@@
 %
-
-One can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to a NUTS sampler which is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}.
 \FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.} 
 \FrL{we don't have plots of just the VI?}\JEC{Good point: I have SVI alone with both MVN or Block Neural Autoregressive Flow. The contours are narrower which would require to elaborate...}
+
+Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{app-DESY1}) where the \texttt{model} is described and the \texttt{cl_obs} are the observed $C_\ell$.}
+\begin{lstlisting}[language=iPython]
+import numpyro.infer.autoguide as autoguide
+from numpyro.infer import SVI, Trace_ELBO
+from numpyro.optim import Adam
+# choose a SVI guide 
+guide = autoguide.AutoMultivariateNormal(model,
+        init_loc_fn=numpyro.infer.init_to_median())
+# SVI optimisation 
+optimizer = numpyro.optim.Adam(1e-3)
+svi = SVI(model_spl, guide,optimizer,loss=Trace_ELBO(num_particles=10))
+svi_result = svi.run(jax.random.PRNGKey(0),20_000, cl_obs)
+#To get samples from SVI approximated posterior 
+samples = guide.sample_posterior(jax.random.PRNGKey(1), svi_result.params, sample_shape=(100_000,))
+\end{lstlisting}
+%
+\JEC{Once the optimisation is done on gets $z$ i.i.d.
+samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
+Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (BNAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
+The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$.
+We have stopped the optimization after 20,000 steps to ensure a good convergence of the ELBO loss, but no particular tuning of the optimizer has been undertaken to decrease the number of steps.
+It takes about 2 hours on the NVidia V100 GPU}.
+\FrL{Arf...
+2 hours !}
+
+\JEC{
+In the figure \ref{fig_cobaya_SVIs} we compare the contours obtained with \texttt{Cobaya} (Fig.~\ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and BNAF guides for the SVI}.
+\JEC{\textit{Add comments on the fact that BNAF looks ok, while MVN is much narrower}}
+
+%
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya_SVIs_v1.pdf}
+\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated perameter posteriorf using SVI with MVN and BNAF guides compared to \texttt{Cobaya} sampling.}
+\label{fig_cobaya_SVIs}
+\end{figure*}
+%
+
+One benefit of SVI is to get a rather quick estimate of an approximated posterior which can serve as front-end to a NUTS sampler.
+This techniques  is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}.
 %
@@
 If the SVI method can be used as is to get $z$ i.i.d.
-samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
-From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
+samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
+From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ (eg.
+$F_\lambda=T^{-1}\circ S^{-1}_\lambda$) is such that
 \begin{equation}
@@
 \end{equation}
-which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg.
-$F_\lambda=T^{-1}\circ S^{-1}_\lambda$).
+which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO.
 So, one can use a HMC sampler (eg.
@@
 Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples.
-The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.
-
-%\begin{figure}
-%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200.pdf}
-%\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 22 parameters of a simulated DES-Y1 3x2pt likelihood, in blue the NUTS sampling described in section \ref{sec-NUTS}, and in orange a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
-%\label{fig_svi_nuts_1}
-%\end{figure}
-
-%\begin{figure}
-%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x1000.pdf}
-%\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 1 chain of 200 warm followed by 1000 samples for the NUTS sampler after SVI optimization.}
-%\label{fig_svi_nuts_2}
-%\end{figure}
-%
-%\begin{figure}
-%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_10x200x1000.pdf}
-%\caption{Same as Fig.~\ref{fig_svi_nuts_1} but with 10 chains of 200 warm followed by 1000 samples for the NUTS sampler (\texttt{max\_tree\_depth=5}) after SVI optimization.}
-%\label{fig_svi_nuts_3}
-%\end{figure}
-
-%\begin{figure}
-%\includegraphics[width=\columnwidth]{figures/SVI_NUTS_1x200x200_bis.pdf}
-%\caption{
-%Constraints (30\%, 68\%, 90\%) on the five lens galaxy bias.
-In blue the NUTS sampling described in section \ref{sec-NUTS}, and in red a SVI MVN optimisation followed by a Naural Transport parametrisation to run a NUTS sampling (1 chain of 200 warm followed by 200 samples; a dense mass matrix and \texttt{max\_tree\_depth} parameter set to 5).}
-%\label{fig_svi_nuts_4}
-%\end{figure}
-
-\begin{figure}
+\begin{figure*}
 \centering
-\includegraphics[width=\columnwidth]{figures/Cobaya-NUTS-SVI200_bis.png}
+\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_bis_v1.pdf}
 \caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but for the highly correlated five lens galaxy bias.}
 \label{fig_cobaya_NUTS_SVI_bis}
-\end{figure}
-
-Using the same DES-Y1 simulation as in previous section, we use a MVN distribution as \textit{guide} to approximate the true posterior (\textit{model}) and feed a SVI.
-The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$.
-We have stopped the optimization after 20,000 steps to ensure a good convergence of the ELBO loss, but no particular tuning of the optimizer has been undertaken to decrease the number of steps.
-It takes 2 hours on the NVidia V100 GPU.
-\FrL{Arf...
-2 hours !}
-
-Then, we have used the Neural Transport method to generate Markov Chains of different sizes by the NUTS sampler, the chains have run 200 warm samples, and then 200 samples (``low'' schema), 1000 samples on a single chain, and finally we have run 10 chains  with 1000 samples each.
+\end{figure*}
+
+\JEC{The SVI optim is described in the SVI section, so the following show be rephrased.}
+
+Using the MVN guided SVI described in the previous section applying the \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed:
+\begin{lstlisting}[language=iPython]
+from numpyro.infer.reparam import NeuTraReparam
+# Reparametrisation of model to feed NUTS sampler
+neutra = NeuTraReparam(guide, svi_result.params)
+neutra_model = neutra.reparam(model_spl) 
+\end{lstlisting}
+%
+The NUTS sampler has run 1 chain of 200 samples (``low'' schema), 1 chain with 1000 samples and 10 chains  with 1000 samples each (each setup starts all chains with 200 samples for initialisation).
 All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
-The elapse time for each of these 3 runs is 50 minutes, 150 minutes and 5 hours, respectively.
+The elapse time for each of the 3 setups is 50 minutes, 150 minutes and 5 hours, respectively.
 Naturally, more samples lead to better contour precision.
@@
 %
-\subsection{Results}
+\subsection{\JEC{Sampling efficiency}}
 \label{sec-results}
@@
 \end{table}
-
-\section{Discussion}
+%
+\section{General discussion}
 \label{sec-discussion}
-
+%
 Having demonstrated the interest of \jaxcosmo\ as a differentiable cosmology library library, we now discuss several limitations, and open questions.
@@
 $\ell$) a series of 50 angular modes.
-The generation of the mock data yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
+The generation of the mock data (\texttt{cl_obs}) yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
 %\begin{minted}[fontsize=\footnotesize]{python}
@@
 with seed(rng_seed=42):
-    data, P, C = fiducial_model()
+    cl_obs, P, C = fiducial_model()
 \end{lstlisting}
@@
 
-\section{SVI example with MVN guide}
+\section{\JEC{SVI example with MVN guide}}
 \label{app-SVI_MVN}
-\JEC{To be cleaned}
-Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
+Let us illustrate the \texit{elliptical standardisation} described in section \label{ec-SVI} in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
 a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
@@
 -\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
-+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
-\right\} \nn \\
+\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \nn \\
+& \left. \qquad \qquad \qquad \qquad \qquad \qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right\} \nn \\
 -\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
@@
 \end{align}
-\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.} \FrL{let's move the MVN details to an appendix.}
-The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
+\FrL{Yeah.
+this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.} The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
 As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX.
-
 %% If you have bibdatabase file and want bibtex to generate the

### 2022-11-12 paper_OJA.tex
SVI contours comment

@@
 \JEC{
-In the figure \ref{fig_cobaya_SVIs} we compare the contours obtained with \texttt{Cobaya} (Fig.~\ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and BNAF guides for the SVI}.
-\JEC{\textit{Add comments on the fact that BNAF looks ok, while MVN is much narrower}}
+In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and BNAF guides for the SVI.
+As mentioned in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution.
+It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that with SVI, one can get a quick estimate of the posterior which can serve as front-end to a NUTS sampler.
+This techniques is described in the next section.}
 
@@
 %
-
-One benefit of SVI is to get a rather quick estimate of an approximated posterior which can serve as front-end to a NUTS sampler.
-This techniques  is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}.
 %
@@
 \end{figure*}
-
-\JEC{The SVI optim is described in the SVI section, so the following show be rephrased.}
-
-Using the MVN guided SVI described in the previous section applying the \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed:
+The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed:
 \begin{lstlisting}[language=iPython]
@@
 %
+We have used the MVN guided SVI described in the previous section even if it may be not the best choice.
 The NUTS sampler has run 1 chain of 200 samples (``low'' schema), 1 chain with 1000 samples and 10 chains  with 1000 samples each (each setup starts all chains with 200 samples for initialisation).

### 2022-11-12 paper_OJA.tex
fix some label typo

@@
 
-Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{app-DESY1}) where the \texttt{model} is described and the \texttt{cl_obs} are the observed $C_\ell$.}
+Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{app-DESY1}) where the \texttt{model} is described and the \texttt{cl\_obs} are the observed $C_\ell$.}
 \begin{lstlisting}[language=iPython]
@@
 $\ell$) a series of 50 angular modes.
-The generation of the mock data (\texttt{cl_obs}) yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
+The generation of the mock data (\texttt{cl\_obs}) yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
 %\begin{minted}[fontsize=\footnotesize]{python}
@@
 \label{app-SVI_MVN}
-Let us illustrate the \texit{elliptical standardisation} described in section \label{ec-SVI} in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
+Let us illustrate the \textit{elliptical standardisation} described in section \label{ec-SVI} in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
 a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
@@
 -\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
-\label{eq-loss-svi-3}
+\label{eq-loss-svi-3b}
 \end{multline}

### 2022-11-13 paper_OJA.tex
table caption left aligned

@@
 
+\usepackage[
+singlelinecheck=false % <-- important
+]{caption}
+
 %to cross out a sentence: \st{Hello World}
@@
 
-The first approach is to derive analytically the formula for the derivatives of interest \citep{BORG}, with or without the help of tools such as \texttt{Mathematica}.
+The first approach is to derive analytically the formula for the derivatives of interest \citep{2019A&A...625A..64J} \JEC{This reference should be validated by F.L}, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}.
 This is however only practical typically for simple analytical models.

### 2022-11-14 paper_OJA.tex
Review SVI alone result according to new MVN convergence.

@@
 The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$.
-We have stopped the optimization after 20,000 steps to ensure a good convergence of the ELBO loss, but no particular tuning of the optimizer has been undertaken to decrease the number of steps.
-It takes about 2 hours on the NVidia V100 GPU}.
+We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the BNAF (MVN) guides, and no particular tuning of the optimizer has been undertaken as for instance a learning rate scheduling.
+It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps}.
 \FrL{Arf...
@@
 \JEC{
-In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and BNAF guides for the SVI.
+In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and BNAF guided SVI.
 As mentioned in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution.
-It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that with SVI, one can get a quick estimate of the posterior which can serve as front-end to a NUTS sampler.
-This techniques is described in the next section.}
-
+It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling in a short time scale.
+This SVI posterior estimates can serve as front-end to a NUTS sampler as described in the next section.}
 %
@@
 \includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya_SVIs_v1.pdf}
-\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated perameter posteriorf using SVI with MVN and BNAF guides compared to \texttt{Cobaya} sampling.}
+\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated posterior using SVI with MVN and BNAF guides (see text) compared to \texttt{Cobaya} sampling.}
 \label{fig_cobaya_SVIs}
 \end{figure*}
-%
 %
@@
 %
-We have used the MVN guided SVI described in the previous section even if it may be not the best choice.
+We have used the MVN guided SVI described in the previous section.
 The NUTS sampler has run 1 chain of 200 samples (``low'' schema), 1 chain with 1000 samples and 10 chains  with 1000 samples each (each setup starts all chains with 200 samples for initialisation).

### 2022-11-14 paper_OJA.tex

@@
 samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
-Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (BNAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
+Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
+Concerning the B-NAF architecture it is composed of  a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
 The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$.
-We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the BNAF (MVN) guides, and no particular tuning of the optimizer has been undertaken as for instance a learning rate scheduling.
+We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the B-NAF (MVN) guides, and no tuning of the optimizer has been undertaken as for instance a learning rate scheduling.
 It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps}.
@@
 \JEC{
-In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and BNAF guided SVI.
+In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and B-NAF guided SVI.
 As mentioned in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution.
-It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling in a short time scale.
-This SVI posterior estimates can serve as front-end to a NUTS sampler as described in the next section.}
+It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling.
+The difference is that these SVI approximate posteriors have been obtained in a short time and these posterior estimates can serve as front-end to a NUTS sampler as described in the next section.}
 %
@@
 \includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya_SVIs_v1.pdf}
-\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated posterior using SVI with MVN and BNAF guides (see text) compared to \texttt{Cobaya} sampling.}
+\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated posterior using SVI with MVN and B-NAF guides (see text) compared to \texttt{Cobaya} sampling.}
 \label{fig_cobaya_SVIs}

### 2022-11-14 paper_OJA.tex
Link to the jax-cosmo-paper repo.

@@
 Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
+The reader can have a glance to the following notebook \nblink{DESY_Y1_SVI_and_NeutraReparam}.
 Concerning the B-NAF architecture it is composed of  a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
@@
 \end{figure*}
-The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed:
+The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed (See the notebook \nblink{DESY_Y1_SVI_and_NeutraReparam}):
 \begin{lstlisting}[language=iPython]
@@
 
-\FrL{add link}
+\FrL{add link}\JEC{Verif if all the nb listed are there, and if some have to be withdrawn. }
+
+\url{https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/}
 

### 2022-11-15 paper_OJA.tex

@@
 Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
-The reader can have a glance to the following notebook \nblink{DESY_Y1_SVI_and_NeutraReparam}.
+The reader can have a glance to the following notebook \nblink{DES_Y1_SVI_and_NeutraReparam}.
 Concerning the B-NAF architecture it is composed of  a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
@@
 \end{figure*}
-The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed (See the notebook \nblink{DESY_Y1_SVI_and_NeutraReparam}):
+The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed (See the notebook \nblink{DES_Y1_SVI_and_NeutraReparam}):
 \begin{lstlisting}[language=iPython]
@@
 %\end{minted}
-More details can be found in the repository of this paper to perform a \numpyro\ modeling using \jaxcosmo : a structured NUTS sampling described in Section~\ref{sec-NUTS} \nblink{VectorizedNumPyro_block}, and a SVI optimisation followed by NUTS sampling of Section~\ref{sec-Neural-Reparametrisation} \nblink{DESY_Y1_shear_cluster_ModelReparam_SVI-MVN_Neutra_NUTS}.
+More details can be found in the repository of this paper to perform a \numpyro\ modeling using \jaxcosmo : a structured NUTS sampling described in Section~\ref{sec-NUTS} \nblink{VectorizedNumPyro_block}, and a SVI optimisation followed by NUTS sampling of Section~\ref{sec-Neural-Reparametrisation} \nblink{DESY_Y1_SVI_and_NeutraReparam}.
 Notice however that the notebooks are only illustrative and for a real sample production the authors invite the reader to run their own script.

### 2022-11-15 paper_OJA.tex
vectorized/parallel chain_method MCMC a bit explained although it can certainly be more accurate if needed.

@@
 I think by default we are using simple Simpson no?}\JEC{Correct: the current pip version uses Simpson but it is a quadrature even if it is qualified "simple".}.
-An example is given in the context of DES Y1 3x2pts analysis (Appendix~\ref{app-DESY1}).
+An example is given in the context of DES Y1 3x2pts analysis (Sec~\ref{sec-DESY1}).
 %
@@
 %
-%
-\subsection{Vanilla Hamiltonian Monte Carlo}
-%
-% Show Joe's vanilla HMC results against Cobaya
-Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
-samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
-and developed extensively since.  See \citet{betancourt} for a full review; we describe
-very basic features here.
-
-HMC samples spaces by generating particle trajectories through the space, using the log-posterior as the negative potential energy of a particle at each point $q$ in the space.
-Associated to $q$, we introduce an auxiliary $p$ variable in the spirit of Hamiltonian momentum such that
-\begin{equation}
-- \log{\cal P}(q) = V(q) + cte; \quad H(q,p) = V(q) + U(p)
-\end{equation}
-where $U(p)$ is a sort of \textit{kinetic} term defined by 
-\begin{equation}
-U(p) = p^T M^{-1} p
-\end{equation}
-At each sample, a trajectory is initialized with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration :
-\begin{align}
-\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial \log{\cal P}}{\mathrm{d} q} \\
-\frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
-\end{align}
-where $M$ is a mass matrix which should be set to approximate the covariance of the posterior.
-This is also used to set the scale of the random initial velocities.
-\JEC{These differential equations may be integrated iterating over $L$ steps the following \textit{leapfrog} algorithm
-\begin{align}
-    p_{n+\frac{1}{2}} &= p_n -\frac{\varepsilon}{2} \frac{\partial V}{\mathrm{d} q}(q_n) \\
-    q_{n+1} & = q_n +\varepsilon M^{-1} p_{n+\frac{1}{2}} \\
-    p_{n+1} &=  p_{n+\frac{1}{2}} -\frac{\varepsilon}{2} \frac{\partial V}{\mathrm{d} q}(q_{n+1})
-\end{align}
-where $\varepsilon$ is a step size parameter.}
-
-Formally, the set of $n_\mathrm{dim}$ moment are treated as new parameters, and after 
-some number of integration steps a final point in the trajectory is compared to the initial one,
-and a Metropolis-Hastings acceptance criterion on the total energy $H(q,p)$ is applied.
-If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
-a relaxation to the integration accuracy.
-
-That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
-but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
-in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
-calculating a Hessian, as in \jaxcosmo, makes it feasible and efficient.
-\FrL{I don't think it uses a Hesssian?
-you only need the derivative of the log posterior wrt parameters.
-The hessian would be needed to get a good estimate of M.}\JEC{Yes.
-Why the Hessian?}
-
-\FrL{Needs a little paragraph to explain with HMC is supposed to be better than plain Metropolis-Hastings.}
-\JEC{Metropolis-Hastings suffers from random walk to explore the geometry of the target distribution to get the region of maximum probability (ie.
-typical set) which is particularly  penalising in large dimensions.
-The HMC dynamics allow to make large jumps to explore the region around a sample to guess the next good one.}
-
-The tricky point is that the \textit{leapfrog} algorithm needs tuning to set the number of steps as well as the step size of integration.
-The next section addresses this problem thanks to the No-U-Turn HMC version.
-
-%Below we show an example set of HMC constraints using \jaxcosmo\ to constrain a simulated
-%Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
-%five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
-%decomposition
-%of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
-%edge of the parameter space, as recommended in \citep{NIPS2015_8303a79b}.  We use a fixed number (25) of integration steps and 
-%a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
-%far from optimal.  We compare to the highly-optimized \texttt{Cobaya} 
-%Metropolis-Hastings implementation  %\citep{cobaya2,cobaya1} JEC it may preferable to have a single up to date reference in JCAP
-%\citep{2019ascl.soft10019T,2021JCAP...05..057T}, initializing it with the Fisher matrix as a proposal and
-%tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.
-%
-%The Hamiltonian sampler required around half the number of samples as the Metropolis (70,000 vs 140,000) to reach approximately the same
-%convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
-%on a CPU.
-This proof-of-principle example can certainly be hugely improved using various extensions to HMC such as No U-Turn Sampling \citep{nuts} or sampling from the entire trajectories \cite{betancourt}.
-On next generation GPU high-performance systems, the advantage will be large.
-%
-%\begin{figure}
-%\includegraphics[width=0.95\columnwidth]{figures/hmc.pdf}
-%\caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
-%Monte Carlo sampler with \jaxcosmo\ (blue), compared to the \texttt{Cobaya} Metropolis-Hastings sampler (orange).}
-%\end{figure}
-%
-%
-%We note that a full Boltzmann code accurate enough to calculate the CMB spectra to Planck \citep{planck18}
-%or future \citep{s4,simons} accuracy is currently beyond the scope of this project.  
-%If this is needed then HMC could use a fast approximation to a CMB theory spectrum, and importance sampling applied to the complete chain.
-
-\subsection{NUTS}
-\label{sec-NUTS}
-%
-% Show the advantage of using NUTS 
-% look at difference in efficiency in terms of how many times we need to call the model.
-
-The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler has been introduced in \citet{nuts}.
-It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$.
-Namely, the root algorithm of HMC deals with the \textit{leapfrog} integration which requires to set two main parameters: $\varepsilon$ the step size and $L$ the number of steps.
-As stated by the authors, NUTS  uses a recursive algorithm to build a set of likely candidate samples that spans a wide swath of the target distribution, stopping automatically when it starts to double back and retrace its steps.
-\FrL{\st{Parameters L an epsilon should be introduced at the level of HMC then.}}\JEC{done} So, in an automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
-The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls at each generation.
-
-Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
-\FrL{class inheritance is TMI.}  After the forward model creation for the DES-Y1 3x2pts exercise described in Appendix~\ref{app-DESY1} \FrL{We should introduce the model at the beginning of this section.},
-\begin{itemize}
-    \item \st{a reparametrisation of the latent variable uniform priors to use a unified uniform prior $\mathcal{U}[-5,5]$} \JEC{we have first applyied  an affine transformation on cosmological, intrinsic alignment and biais parameters to use an unique uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1})}; \FrL{these are not latent variables, but model parameters.
-Also people don't know what this reparametrisation means.}
-    \item  \FrL{People don't know what the warmup phase is...}\JEC{we use a structured mass matrix consisting on essentially a diagonal form  with a block structure for the following sets of parameters} $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
-This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
-\end{itemize}
-The NUTS sampler have been running with 16 chains of 1,000 samples each (\texttt{vectorized} option) \FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.} after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie.
-128 steps for each iteration).
-The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
-The elapse time for these experiments was 20 hours.
-
-The results in terms of relative effective sample sizes (aka ``ESS'') are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
-One notices the dramatic improvement of the ESS by about a factor 10 using the NUTS sampler compared to the highly-optimized \texttt{Cobaya} Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T} which has last about 40 hours on CPU to generate 70,000 samples.
-The CL contours are also in very good agreement.
-\FrL{we should rephrase this a bit to say that Cobaya is our reference.} It is worth to point out that the mass matrix structuring operation is a very effective as it increases the sampling efficiency by about a factor 2.
-Anticipating the Section~\ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
-This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Table~\ref{tab-ESS-NUTS_SVI-1}), then the user is highly encouraged to tune this critical parameter.
-%
-\begin{figure*}
-\centering
-\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_v1.pdf}
-\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue, 70,000 samples), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red; 16,000 samples), and a SVI MVN optimisation followed by a Neural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green, 200 samples).}
-\label{fig_cobaya_NUTS_SVI}
-\end{figure*}
-
-\subsection{Stochastic Variational Inference}
-\label{sec-SVI}
-%
-Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps.
-If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
-VI aims to get an approximation of this distribution, ie.
-$p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
-To do so, one uses the Kullback-Leibler divergence of the two distributions $KL(q(z;\lambda)||p(z|\mathcal{D}))$ leading to the following relation
-\begin{align}
-\log p(\mathcal{D}) &= \mathtt{ELBO} +  KL(q(z;\lambda)||p(z|\mathcal{D})) \label{eq-ELBO} \\
-\mathrm{with} \ \mathtt{ELBO} &\equiv -\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right] + \mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right] 
-\end{align}
-which defines the \textit{evidence lower bound} (aka ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
-\begin{equation}
-q(z;\lambda^\ast) = \underset{q(z;\lambda)}{\mathrm{argmax}}\  \mathtt{ELBO} = 
-\underset{\lambda}{\mathrm{argmin}}\ \mathcal{L}(\lambda)
-\end{equation}
-The function $\mathcal{L}(\lambda)$ is the cost function used in practice. It is composed of two parts:
-\begin{equation}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right]}_{model}
-\label{eq-loss-svi-1}
-\end{equation}
-where the \textit{guide} in the \numpyro\ library may be a multi-variate Gaussian distribution (MVN) for instance.
-
-Using the auto-differentation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}).
-As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
-function, taking derivatives, and running a gradient-based or coordinate-ascent optimization.
-First is defined a invertible differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
-The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
-One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models.
-Then, the cost function reads
-\begin{equation}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log p(\xi,\mathcal{D}) \right]}_{model}
-\label{eq-loss-svi-2}
-\end{equation}
-with
-\begin{equation}
-p(\xi,\mathcal{D}) \bydef p(T^{-1}(\xi),\mathcal{D}) |J_{T^{-1}}(\xi)|
-\end{equation}
-which exhibit the Jacobian of the $T^{-1}$ transformation.
-\JEC{
-The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{2015arXiv150603431K}).
-Let us illustrated the method using an invertible transformation (aka \textit{normalizing flow})  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, such that $\zeta\sim \mathcal{N}(0,I)$\footnote{Notice that $z=T^{-1}(\xi)=(T^{-1} \circ S_\lambda^{-1})(\zeta)=(S_\lambda \circ T)^{-1}(\zeta)=F_\lambda(\zeta)$.}.
-The Jacobian of this distribution is 1 by definition (volume conservation), so the lost function reads
-\begin{multline}
--\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
-\label{eq-loss-svi-3}
-\end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the $q$ distribution, so its gradient can be computed once for all for a given $q$ distribution family and reused in any user model.  
-Then,
-to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to\footnote{To simplify the notation, $T^{-1}(S_\lambda^{-1}(\zeta))$ has been replaced by $z$.}
-\begin{multline}
--\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
-+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
-\right\}
-+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
-\label{eq-loss-svi-4}
-\end{multline}
-An concrete example of the gradient formula is given in Appendix \ref{app-SVI_MVN} with the multivariate Gaussian distribution family used as the guide.
-}%JEC
-%
-\FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.} 
-\FrL{we don't have plots of just the VI?}\JEC{Good point: I have SVI alone with both MVN or Block Neural Autoregressive Flow. The contours are narrower which would require to elaborate...}
-
-Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{app-DESY1}) where the \texttt{model} is described and the \texttt{cl\_obs} are the observed $C_\ell$.}
-\begin{lstlisting}[language=iPython]
-import numpyro.infer.autoguide as autoguide
-from numpyro.infer import SVI, Trace_ELBO
-from numpyro.optim import Adam
-# choose a SVI guide 
-guide = autoguide.AutoMultivariateNormal(model,
-        init_loc_fn=numpyro.infer.init_to_median())
-# SVI optimisation 
-optimizer = numpyro.optim.Adam(1e-3)
-svi = SVI(model_spl, guide,optimizer,loss=Trace_ELBO(num_particles=10))
-svi_result = svi.run(jax.random.PRNGKey(0),20_000, cl_obs)
-#To get samples from SVI approximated posterior 
-samples = guide.sample_posterior(jax.random.PRNGKey(1), svi_result.params, sample_shape=(100_000,))
-\end{lstlisting}
-%
-\JEC{Once the optimisation is done on gets $z$ i.i.d.
-samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
-Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
-The reader can have a glance to the following notebook \nblink{DES_Y1_SVI_and_NeutraReparam}.
-Concerning the B-NAF architecture it is composed of  a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
-The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$.
-We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the B-NAF (MVN) guides, and no tuning of the optimizer has been undertaken as for instance a learning rate scheduling.
-It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps}.
-\FrL{Arf...
-2 hours !}
-
-\JEC{
-In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and B-NAF guided SVI.
-As mentioned in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution.
-It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling.
-The difference is that these SVI approximate posteriors have been obtained in a short time and these posterior estimates can serve as front-end to a NUTS sampler as described in the next section.}
-%
-\begin{figure*}
-\centering
-\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya_SVIs_v1.pdf}
-\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated posterior using SVI with MVN and B-NAF guides (see text) compared to \texttt{Cobaya} sampling.}
-\label{fig_cobaya_SVIs}
-\end{figure*}
-%
-\subsubsection{Neural Transport}
-\label{sec-Neural-Reparametrisation}
-%
-If the SVI method can be used as is to get $z$ i.i.d.
-samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
-From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ (eg.
-$F_\lambda=T^{-1}\circ S^{-1}_\lambda$) is such that
-\begin{equation}
-q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
-\end{equation}
-which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO.
-So, one can use a HMC sampler (eg.
-NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$.
-Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples.
-\begin{figure*}
-\centering
-\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_bis_v1.pdf}
-\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but for the highly correlated five lens galaxy bias.}
-\label{fig_cobaya_NUTS_SVI_bis}
-\end{figure*}
-The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed (See the notebook \nblink{DES_Y1_SVI_and_NeutraReparam}):
-\begin{lstlisting}[language=iPython]
-from numpyro.infer.reparam import NeuTraReparam
-# Reparametrisation of model to feed NUTS sampler
-neutra = NeuTraReparam(guide, svi_result.params)
-neutra_model = neutra.reparam(model_spl) 
-\end{lstlisting}
-%
-We have used the MVN guided SVI described in the previous section.
-The NUTS sampler has run 1 chain of 200 samples (``low'' schema), 1 chain with 1000 samples and 10 chains  with 1000 samples each (each setup starts all chains with 200 samples for initialisation).
-All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
-The elapse time for each of the 3 setups is 50 minutes, 150 minutes and 5 hours, respectively.
-Naturally, more samples lead to better contour precision.
-But what is illustrative is the ``low'' schema results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
-Notice the good results even for the highly correlated lens bias parameters.
-It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a rather poor behaviour with only 200 samples.
-
-Results in terms of sampling efficiency are shown on Table~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to ease the model development in its early phase.
-%
-\subsection{\JEC{Sampling efficiency}}
-\label{sec-results}
-%
-To a large extend looking at results of Table~\ref{tab-ESS-NUTS_SVI-1}, the key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
-In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
-\begin{equation}
-    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
-\end{equation}
-with $N_s$ the total number of samples, $\varepsilon$ the effective sampler efficiency  and $n_{steps}$ the number of steps (calls) per sample generation.
-Then, for \texttt{Cobaya} one roughly gets $\varepsilon\approx 3\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more.
-At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie.
-about $1\%$).
-So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.
-
-On the other hand, the above metric is may be too crude to cover all aspects of the sample generation.
-One should probably have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
-several days on dedicated infrastructure.
-Moreover, the user is faced to the dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, 2015arXiv151007727O}.
-
-So, the authors think that it is worth considering methods like HMC/NUTS that by construction are more efficient.
-Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of the model with rather modest sample set (eg.
-starting with 200 samples) producing good enough marginal contours in few hours.
-In practice, this validation phase can save time before producing sizeable batch for final analysis.
-The authors have not investigated very high dimensional ($O(10^2)$ parameters) as well as multi-modal analysis, but the key argument in favour of HMC/NUTS sampling is that it exploit the geometry of the typical set of the posterior distribution contrary to the standard random walk of Metropolis-Hasting sampling.
-Furthermore, using reparametrisation  one can adapt to poor geometry cases (eg.
-\cite{2019arXiv190303704H}).
-%
-\begin{table}[htb]
-\caption{Results of the \textit{relative effective sample size} (aka ESS, ``bulk'') in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
-The ESS can be larger than 100\% in some cases.
-%{\color{red} 
-%(0) "F.L run" are computed from NUTS sampler with 16 %parallel chains of 1,000 samples each and 200 warm-up %and \texttt{max\_tree\_depth=7} (see text) The numbers %are from \texttt{n\_eff/16,000};
-%}
-(a): results from Cobaya 70,000 samples;
-(b): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Section~\ref{sec-NUTS}); 
-(c) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 1 chain of 200 samples and 200 warm-up (Section~\ref{sec-Neural-Reparametrisation}).}
-\label{tab-ESS-NUTS_SVI-1}
- \centering
-\begin{tabular}{ccccccccccc}
-\hline
-    & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$ & $n_s$ & $A$ & $\eta$\\
-\hline
-(a) &  $3.1$ & $2.5$       & $2.9$      & $2.9$  & $2.6$  & $3.1$  & $3.1$ & $2.8$ \\  
-(b) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\
-(c) & $84.0$ &  $28.0$     & $26.0$     & $20.5$ & $80.0$ & $110.5$ & $58.5$ & $29.5$\\
-\hline
-\end{tabular}
-\end{table}
-%
-\section{General discussion}
-\label{sec-discussion}
-%
-Having demonstrated the interest of \jaxcosmo\ as a differentiable cosmology library library, we now discuss several limitations, and open questions.
-
-%Lack of autodiff Boltzmann code (emulator)
-The first essential barrier to a full-fledged automatically differentiable cosmology library is the need for a differentiable Boltzmann solver to compute the matter power spectrum.
-At this stage, \jaxcosmo\ relies on the analytic Einsenstein \& Hu fitting formula, which is not accurate enough for Stage IV requirements, nor does it allow to explore beyond $\Lambda$CDM models.
-Existing solvers such as CLASS or CAMB \FrL{citations} are large and complex codes which are not easily reimplemented in an \autodiff\ framework and therefore cannot be directly integrated in \jaxcosmo\ .
-
-A first option to resolve this issue is to implement from scratch a new Boltzmann code in a framework that will support automatic differentiation.
-This is the approach behind works such as \FrL{the Julia code} which is provides a simplified Boltzmann solver in Julia, or \FrL{the code from our friends in Zurich} which is based on a SymPy and could be relatively easily compatible with JAX.
-However, even if very promising, both of these options thus far remain limited.
-While we do believe a automatically differentiable Boltzmann code is the best option, it seems that the cost of developing such a code remains very high at this time.
-
-A second approach is to develop emulators of a full-fledged Boltzmann code.
-Emulators based on neural networks or Gaussian processes are themselves automatically differentiable with respect to cosmological parameters.
-In fact, the litterature is now rich in examples of such emulators \FrL{cite all emulators}.
-After validating their accuracy against the reference CAMB or CLASS implementation, they can be directly integrated as a plug-and-play replacement for the computation of the matter power spectrum.
-\FrL{need to talk about cosmopower somewhere.} At this time, it seems that using emulators will be the most straightforward approach to bring more accurate models to \jaxcosmo\ .
-\FrL{However, we want to emphasize that one of the reason for the wide diversity of emulators is a lack of standardization}.
-
-\bigskip
-
-% Reasoning behind not making a full emulator of the correlation functions
-Connected to this discussion about emulators, one point that could be discussed is whether one even needs a library like \jaxcosmo\ if anyway one can build emulators of a CCL likelihood for use inside gradient-based inference algorithms.
-While this could indeed be feasible, and of similar cost as just making an emulator of the matter power spectrum, the drawback of this approach is that many analysis parameters and choices become hardcoded in the emulator.
-While the model for the linear matter power spectrum is typically kept fixed in practical analyses, all the of the choices related in particular to systematics modeling (e.g.
-photometric redshift errors, galaxy bias, or intrinsic galaxy alignments) are very likely to vary significantly in the process of developing the analysis.
-Building an emulator for the likelihood would mean the emulator needs to be retrained everytime the likelihood is changed.
-
-\bigskip
-
-%  - Massively Parallel MCMC on GPU (linked to vmap and/or pmap)
-%  - Efficiency of gradient-based inference methods
-Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo\ .
-As we illustrated in the previous section, gradient-based inference techniques yield significantly less correlated MCMC chains.\FrL{Hummm here I guess we want to say also it scales better with dimensions, and that you can get very fast first results.} \FrL{we also want to say that the GPU parallelisation is going to change the game, all the more when you account for new massive GPU machines.}
-
-\bigskip
-
-%  - Respective role of CCL and jax-cosmo
-Finally, one question worth asking is how does \jaxcosmo\ position itself against classical codes such as CCL?
-Or in other terms, will it replace CCL?
-While we are convinced of the benefits of a JAX implementation, we expect CCL to remain extremely valuable as the reference cosmology implementation.
-\FrL{it allows to leverage the legacy codes, has a lot more features, etc.} Ultimately, a natural transition may occur towards more modern frameworks like \jaxcosmo\ when they reach the ability to run full-fledged Stage IV likelihoods.
-
-\section{Conclusions \& Prospects}
-\label{sec-conclusion}
-%- We have presented a library using automatic differentiation and and automatic GPU offload for a set of key cosmology theory calculations
-%- We currently cover only a small number of the many calculations needed in cosmo theory, and welcome contributions for other areas such as CMB.
-%- The calculation of Fisher matrices using this method, which is infamous for requiring extensive careful tuning, becomes very simple and robust using JC.
-%- We have demonstrated the efficiency in JC of the derivative-aware MCMC methods HMC and NUTS, which are regarded as the only way for samplers to evade the curse of dimensionality up to the hundreds of dimensions we are likely to need for the next generation of surveys.
-%- We have shown a proof-of-concept for the use of JC with machine learning methods, opening a whole new space of methods in cosmology.
-
-In this paper, we have presented \jaxcosmo\ a cosmology library implemented in the JAX framework, enabling the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, as well as greatly speeding up likelihood evaluations thanks to automatic parallelisation and just in time compilation on GPUs.
-Currently, \jaxcosmo\ only contains a small set of features limited corresponding to a DES-Y1 3x2pt analysis.
-Being an open source project, contribution of additional features is additional areas such as CMB or spectroscopic galaxy clustering are warmly welcome.
-
-To demonstrate the interest of an automatically differentiable library, we have illustrated on concrete examples how Fisher matrices, which are notoriously difficult to compute and require extensive and careful fine tuning, can now be computed robustly and at much lower cost.
-In addition, Fisher matrices becomes themselves differentiable, which allows for  Figure of Merit optimization by gradient descent, making survey optimization extremely simple.
-
-Going beyond Fisher forecasts, we have also compared to simple Metropolis-Hastings several gradient-based inference techniques (namely, Hamiltonian Monte-Carlo, No-U-Turn-Sampler, Stochastic Variational Inference).
-We show that the posterior samples with gradient-based methods have a much larger Effective Sample Size, meaning that \FrL{let's think a bit more about interpretation here....
-the thing I care about is not as much the ESS, but the number of likelihood evaluation per independent sample.
-I don't think we have quantified that too much}.
-These inference techniques can scale to hundreds of dimensions, and may become necessary in Stage IV analysis as the number of nuisance parameters is likely to become very large.
-
-\bigskip
-
-The next extensions of this framework will be the inclusion of additional cosmological probes, as well as integrating emulators for the matter power spectrum trained on CAMB or CLASS, as a means to go beyond the current analytic Eisenstein \& Hu model.
-
-\bigskip
-
-In the spirit of reproducible research, all results presented in this paper can be reproduced from the following GitHub repository:
-
-\FrL{add link}\JEC{Verif if all the nb listed are there, and if some have to be withdrawn. }
-
-\url{https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/}
-
-%
-%
-%
-%\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
-%
-%%
-%%
-%\subsection{DES-Y1 Example}
-%
-
-%%%
-\section*{CRediT authorship contribution statement}
-\textbf{J.E Campagne} Conceptualization, Methodology, Software, Validation, Writing.
-\textbf{F. Lanusse}, \textbf{J. Zuntz}...
-
-%\section*{Declaration of Competing Interest}
-%The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
-
-%\Acknowledgements
-\section*{Acknowledgements}
-Some of the numerical experiments have been conducted at the IN2P3 Computing Center (CC-IN2P3 - Lyon/Villeurbanne - France) funded by the Centre National de la Recherche Scientifique.
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-\appendix
-
-\section{Description of the DES-Y1 exercise}
-\label{app-DESY1}
+
+\subsection{Description of the DES-Y1 exercise}
+\label{sec-DESY1}
+\JEC{Transfer of the model context from apprndix to here to ease the reading.}
+
 From the DES server\footnote{\url{http://desdr-server.ncsa.illinois.edu/despublic/y1a1_files/chains/2pt_NG_mcal_1110.fit}} we have extracted the $N(z)$ distributions of the sources and normalize them to $[1.47, 1.46, 1.50, 0.73]$ effective number of sources per $\mathrm{arcmin}^2$ for each of the four redshift bins, respectively.
@@
 %\end{minted}
-Then, as an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed
+%Then, as an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed
 %\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
-nuts_kernel = numpyro.infer.NUTS(observed_model,
+%\begin{lstlisting}[language=iPython]
+%nuts_kernel = numpyro.infer.NUTS(observed_model,
+%              step_size=1e-1, 
+%              init_strategy=numpyro.infer.init_to_median,
+%              dense_mass=True,
+%              max_tree_depth=7)
+%
+%mcmc = numpyro.infer.MCMC(nuts_kernel, 
+%                          num_warmup=200, 
+%                          num_samples=1000,
+%                          num_chains=16,
+%                          chain_method='vectorized',
+%                          progress_bar=False)
+%
+%mcmc.run(jax.random.PRNGKey(42))
+%\end{lstlisting}
+%%\end{minted}
+%More details can be found in the repository of this paper to perform a \numpyro\ modeling using \jaxcosmo: a structured NUTS sampling described in Section~\ref{sec-NUTS} \nblink{VectorizedNumPyro_block}, and a SVI optimisation followed by NUTS sampling of Section~\ref{sec-Neural-Reparametrisation} \nblink{DESY_Y1_SVI_and_NeutraReparam}.
+How to use this model to perform inference is described in the sections \ref{sec-NUTS} and \ref{sec-SVI}.
+%
+\subsection{Vanilla Hamiltonian Monte Carlo}
+%
+% Show Joe's vanilla HMC results against Cobaya
+Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
+samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
+and developed extensively since.  See \citet{betancourt} for a full review; we describe
+very basic features here.
+
+HMC samples spaces by generating particle trajectories through the space, using the log-posterior as the negative potential energy of a particle at each point $q$ in the space.
+Associated to $q$, we introduce an auxiliary $p$ variable in the spirit of Hamiltonian momentum such that
+\begin{equation}
+- \log{\cal P}(q) = V(q) + cte; \quad H(q,p) = V(q) + U(p)
+\end{equation}
+where $U(p)$ is a sort of \textit{kinetic} term defined by 
+\begin{equation}
+U(p) = p^T M^{-1} p
+\end{equation}
+At each sample, a trajectory is initialized with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration :
+\begin{align}
+\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial \log{\cal P}}{\mathrm{d} q} \\
+\frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
+\end{align}
+where $M$ is a mass matrix which should be set to approximate the covariance of the posterior.
+This is also used to set the scale of the random initial velocities.
+\JEC{These differential equations may be integrated iterating over $L$ steps the following \textit{leapfrog} algorithm
+\begin{align}
+    p_{n+\frac{1}{2}} &= p_n -\frac{\varepsilon}{2} \frac{\partial V}{\mathrm{d} q}(q_n) \\
+    q_{n+1} & = q_n +\varepsilon M^{-1} p_{n+\frac{1}{2}} \\
+    p_{n+1} &=  p_{n+\frac{1}{2}} -\frac{\varepsilon}{2} \frac{\partial V}{\mathrm{d} q}(q_{n+1})
+\end{align}
+where $\varepsilon$ is a step size parameter.}
+
+Formally, the set of $n_\mathrm{dim}$ moment are treated as new parameters, and after 
+some number of integration steps a final point in the trajectory is compared to the initial one,
+and a Metropolis-Hastings acceptance criterion on the total energy $H(q,p)$ is applied.
+If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
+a relaxation to the integration accuracy.
+
+That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
+but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
+in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
+calculating a Hessian, as in \jaxcosmo, makes it feasible and efficient.
+\FrL{I don't think it uses a Hesssian?
+you only need the derivative of the log posterior wrt parameters.
+The hessian would be needed to get a good estimate of M.}\JEC{Yes.
+Why the Hessian?}
+
+\FrL{Needs a little paragraph to explain with HMC is supposed to be better than plain Metropolis-Hastings.}
+\JEC{Metropolis-Hastings suffers from random walk to explore the geometry of the target distribution to get the region of maximum probability (ie.
+typical set) which is particularly  penalising in large dimensions.
+The HMC dynamics allow to make large jumps to explore the region around a sample to guess the next good one.}
+
+The tricky point is that the \textit{leapfrog} algorithm needs tuning to set the number of steps as well as the step size of integration.
+The next section addresses this problem thanks to the No-U-Turn HMC version.
+
+%Below we show an example set of HMC constraints using \jaxcosmo\ to constrain a simulated
+%Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
+%five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
+%decomposition
+%of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
+%edge of the parameter space, as recommended in \citep{NIPS2015_8303a79b}.  We use a fixed number (25) of integration steps and 
+%a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
+%far from optimal.  We compare to the highly-optimized \texttt{Cobaya} 
+%Metropolis-Hastings implementation  %\citep{cobaya2,cobaya1} JEC it may preferable to have a single up to date reference in JCAP
+%\citep{2019ascl.soft10019T,2021JCAP...05..057T}, initializing it with the Fisher matrix as a proposal and
+%tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.
+%
+%The Hamiltonian sampler required around half the number of samples as the Metropolis (70,000 vs 140,000) to reach approximately the same
+%convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
+%on a CPU.
+This proof-of-principle example can certainly be hugely improved using various extensions to HMC such as No U-Turn Sampling \citep{nuts} or sampling from the entire trajectories \cite{betancourt}.
+On next generation GPU high-performance systems, the advantage will be large.
+%
+%\begin{figure}
+%\includegraphics[width=0.95\columnwidth]{figures/hmc.pdf}
+%\caption{Constraints on three of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using a simple Hamiltonian
+%Monte Carlo sampler with \jaxcosmo\ (blue), compared to the \texttt{Cobaya} Metropolis-Hastings sampler (orange).}
+%\end{figure}
+%
+%
+%We note that a full Boltzmann code accurate enough to calculate the CMB spectra to Planck \citep{planck18}
+%or future \citep{s4,simons} accuracy is currently beyond the scope of this project.  
+%If this is needed then HMC could use a fast approximation to a CMB theory spectrum, and importance sampling applied to the complete chain.
+
+\subsection{NUTS}
+\label{sec-NUTS}
+%
+% Show the advantage of using NUTS 
+% look at difference in efficiency in terms of how many times we need to call the model.
+
+The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler has been introduced in \citet{nuts}.
+It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$.
+Namely, the root algorithm of HMC deals with the \textit{leapfrog} integration which requires to set two main parameters: $\varepsilon$ the step size and $L$ the number of steps.
+As stated by the authors, NUTS  uses a recursive algorithm to build a set of likely candidate samples that spans a wide swath of the target distribution, stopping automatically when it starts to double back and retrace its steps.
+\FrL{\st{Parameters L an epsilon should be introduced at the level of HMC then.}}\JEC{done} So, in an automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
+The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls at each generation.
+
+Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
+\FrL{class inheritance is TMI.}  After the forward model creation for the DES-Y1 3x2pts exercise described in section \ref{sec-DESY1} \FrL{We should introduce the model at the beginning of this section.}\JEC{Done}
+\begin{itemize}
+    \item \st{a reparametrisation of the latent variable uniform priors to use a unified uniform prior $\mathcal{U}[-5,5]$} \JEC{we have first applyied  an affine transformation on cosmological, intrinsic alignment and biais parameters to use an unique uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1})}; \FrL{these are not latent variables, but model parameters.
+Also people don't know what this reparametrisation means.}
+    \item  \FrL{People don't know what the warmup phase is...}\JEC{we use a structured mass matrix consisting on essentially a diagonal form  with a block structure for the following sets of parameters} $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
+This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
+\end{itemize}
+
+\JEC{As an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed with \numpyro:}
+\begin{lstlisting}[language=iPython]
+%nuts_kernel = numpyro.infer.NUTS(observed_model,
               step_size=1e-1,
@@
 \end{lstlisting}
-%\end{minted}
-More details can be found in the repository of this paper to perform a \numpyro\ modeling using \jaxcosmo : a structured NUTS sampling described in Section~\ref{sec-NUTS} \nblink{VectorizedNumPyro_block}, and a SVI optimisation followed by NUTS sampling of Section~\ref{sec-Neural-Reparametrisation} \nblink{DESY_Y1_SVI_and_NeutraReparam}.
-Notice however that the notebooks are only illustrative and for a real sample production the authors invite the reader to run their own script.
+
+The NUTS sampler have been running with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie.
+128 steps for each iteration).
+\FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
+\JEC{The "vectorized" option runs multi chains in a single GPU thanks to the \jax \textit{vmap} mechanism.
+If one dispose of several GPUs then the "parallel" option using \textit{pmap}  mechanism runs one chain per GPU.
+As noted in \numpyro\ MCMC documentation, it is possible to mix parallel and vectorized sampling\footnote{Since version 0.9.0.}.}
+
+The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
+The elapse time for these experiments was 20 hours.
+
+The results in terms of relative effective sample sizes (aka ``ESS'') are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
+One notices the dramatic improvement of the ESS by about a factor 10 using the NUTS sampler compared to the highly-optimized \texttt{Cobaya} Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T} which has last about 40 hours on CPU to generate 70,000 samples.
+The CL contours are also in very good agreement.
+\FrL{we should rephrase this a bit to say that Cobaya is our reference.} It is worth to point out that the mass matrix structuring operation is a very effective as it increases the sampling efficiency by about a factor 2.
+Anticipating the Section~\ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
+This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Table~\ref{tab-ESS-NUTS_SVI-1}), then the user is highly encouraged to tune this critical parameter.
+%
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_v1.pdf}
+\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue, 70,000 samples), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red; 16,000 samples), and a SVI MVN optimisation followed by a Neural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green, 200 samples).}
+\label{fig_cobaya_NUTS_SVI}
+\end{figure*}
+
+\subsection{Stochastic Variational Inference}
+\label{sec-SVI}
+%
+Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps.
+If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
+VI aims to get an approximation of this distribution, ie.
+$p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
+To do so, one uses the Kullback-Leibler divergence of the two distributions $KL(q(z;\lambda)||p(z|\mathcal{D}))$ leading to the following relation
+\begin{align}
+\log p(\mathcal{D}) &= \mathtt{ELBO} +  KL(q(z;\lambda)||p(z|\mathcal{D})) \label{eq-ELBO} \\
+\mathrm{with} \ \mathtt{ELBO} &\equiv -\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right] + \mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right] 
+\end{align}
+which defines the \textit{evidence lower bound} (aka ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
+\begin{equation}
+q(z;\lambda^\ast) = \underset{q(z;\lambda)}{\mathrm{argmax}}\  \mathtt{ELBO} = 
+\underset{\lambda}{\mathrm{argmin}}\ \mathcal{L}(\lambda)
+\end{equation}
+The function $\mathcal{L}(\lambda)$ is the cost function used in practice. It is composed of two parts:
+\begin{equation}
+\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right]}_{model}
+\label{eq-loss-svi-1}
+\end{equation}
+where the \textit{guide} in the \numpyro\ library may be a multi-variate Gaussian distribution (MVN) for instance.
+
+Using the auto-differentation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}).
+As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
+function, taking derivatives, and running a gradient-based or coordinate-ascent optimization.
+First is defined a invertible differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
+The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
+One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models.
+Then, the cost function reads
+\begin{equation}
+\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log p(\xi,\mathcal{D}) \right]}_{model}
+\label{eq-loss-svi-2}
+\end{equation}
+with
+\begin{equation}
+p(\xi,\mathcal{D}) \bydef p(T^{-1}(\xi),\mathcal{D}) |J_{T^{-1}}(\xi)|
+\end{equation}
+which exhibit the Jacobian of the $T^{-1}$ transformation.
+\JEC{
+The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{2015arXiv150603431K}).
+Let us illustrated the method using an invertible transformation (aka \textit{normalizing flow})  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, such that $\zeta\sim \mathcal{N}(0,I)$\footnote{Notice that $z=T^{-1}(\xi)=(T^{-1} \circ S_\lambda^{-1})(\zeta)=(S_\lambda \circ T)^{-1}(\zeta)=F_\lambda(\zeta)$.}.
+The Jacobian of this distribution is 1 by definition (volume conservation), so the lost function reads
+\begin{multline}
+-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
+\label{eq-loss-svi-3}
+\end{multline}
+where $\mathbb{H}(q)$ is the Shannon entropy of the $q$ distribution, so its gradient can be computed once for all for a given $q$ distribution family and reused in any user model.  
+Then,
+to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to\footnote{To simplify the notation, $T^{-1}(S_\lambda^{-1}(\zeta))$ has been replaced by $z$.}
+\begin{multline}
+-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
+\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
++ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
+\right\}
++ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
+\label{eq-loss-svi-4}
+\end{multline}
+An concrete example of the gradient formula is given in Appendix \ref{app-SVI_MVN} with the multivariate Gaussian distribution family used as the guide.
+}%JEC
+%
+\FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.} 
+\FrL{we don't have plots of just the VI?}\JEC{Good point: I have SVI alone with both MVN or Block Neural Autoregressive Flow. The contours are narrower which would require to elaborate...}
+
+Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{sec-DESY1}) where the \texttt{model} is described and the \texttt{cl\_obs} are the observed $C_\ell$.}
+\begin{lstlisting}[language=iPython]
+import numpyro.infer.autoguide as autoguide
+from numpyro.infer import SVI, Trace_ELBO
+from numpyro.optim import Adam
+# choose a SVI guide 
+guide = autoguide.AutoMultivariateNormal(model,
+        init_loc_fn=numpyro.infer.init_to_median())
+# SVI optimisation 
+optimizer = numpyro.optim.Adam(1e-3)
+svi = SVI(model_spl, guide,optimizer,loss=Trace_ELBO(num_particles=10))
+svi_result = svi.run(jax.random.PRNGKey(0),20_000, cl_obs)
+#To get samples from SVI approximated posterior 
+samples = guide.sample_posterior(jax.random.PRNGKey(1), svi_result.params, sample_shape=(100_000,))
+\end{lstlisting}
+%
+\JEC{Once the optimisation is done on gets $z$ i.i.d.
+samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
+Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
+The reader can have a glance to the following notebook \nblink{DES_Y1_SVI_and_NeutraReparam}.
+Concerning the B-NAF architecture it is composed of  a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
+The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$.
+We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the B-NAF (MVN) guides, and no tuning of the optimizer has been undertaken as for instance a learning rate scheduling.
+It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps}.
+\FrL{Arf...
+2 hours !}
+
+\JEC{
+In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and B-NAF guided SVI.
+As mentioned in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution.
+It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling.
+The difference is that these SVI approximate posteriors have been obtained in a short time and these posterior estimates can serve as front-end to a NUTS sampler as described in the next section.}
+%
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya_SVIs_v1.pdf}
+\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but using sampling of approximated posterior using SVI with MVN and B-NAF guides (see text) compared to \texttt{Cobaya} sampling.}
+\label{fig_cobaya_SVIs}
+\end{figure*}
+%
+\subsubsection{Neural Transport}
+\label{sec-Neural-Reparametrisation}
+%
+If the SVI method can be used as is to get $z$ i.i.d.
+samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
+From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ (eg.
+$F_\lambda=T^{-1}\circ S^{-1}_\lambda$) is such that
+\begin{equation}
+q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
+\end{equation}
+which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO.
+So, one can use a HMC sampler (eg.
+NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$.
+Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples.
+\begin{figure*}
+\centering
+\includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_bis_v1.pdf}
+\caption{Same configuration as in Figure \ref{fig_cobaya_NUTS_SVI} but for the highly correlated five lens galaxy bias.}
+\label{fig_cobaya_NUTS_SVI_bis}
+\end{figure*}
+The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed (See the notebook \nblink{DES_Y1_SVI_and_NeutraReparam}):
+\begin{lstlisting}[language=iPython]
+from numpyro.infer.reparam import NeuTraReparam
+# Reparametrisation of model to feed NUTS sampler
+neutra = NeuTraReparam(guide, svi_result.params)
+neutra_model = neutra.reparam(model_spl) 
+\end{lstlisting}
+%
+We have used the MVN guided SVI described in the previous section.
+The NUTS sampler has run 1 chain of 200 samples (``low'' schema), 1 chain with 1000 samples and 10 chains  with 1000 samples each (each setup starts all chains with 200 samples for initialisation).
+All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
+The elapse time for each of the 3 setups is 50 minutes, 150 minutes and 5 hours, respectively.
+Naturally, more samples lead to better contour precision.
+But what is illustrative is the ``low'' schema results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
+Notice the good results even for the highly correlated lens bias parameters.
+It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a rather poor behaviour with only 200 samples.
+
+Results in terms of sampling efficiency are shown on Table~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to ease the model development in its early phase.
+%
+\subsection{\JEC{Sampling efficiency}}
+\label{sec-results}
+%
+To a large extend looking at results of Table~\ref{tab-ESS-NUTS_SVI-1}, the key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
+In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
+\begin{equation}
+    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
+\end{equation}
+with $N_s$ the total number of samples, $\varepsilon$ the effective sampler efficiency  and $n_{steps}$ the number of steps (calls) per sample generation.
+Then, for \texttt{Cobaya} one roughly gets $\varepsilon\approx 3\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more.
+At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie.
+about $1\%$).
+So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.
+
+On the other hand, the above metric is may be too crude to cover all aspects of the sample generation.
+One should probably have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
+several days on dedicated infrastructure.
+Moreover, the user is faced to the dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, 2015arXiv151007727O}.
+
+So, the authors think that it is worth considering methods like HMC/NUTS that by construction are more efficient.
+Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of the model with rather modest sample set (eg.
+starting with 200 samples) producing good enough marginal contours in few hours.
+In practice, this validation phase can save time before producing sizeable batch for final analysis.
+The authors have not investigated very high dimensional ($O(10^2)$ parameters) as well as multi-modal analysis, but the key argument in favour of HMC/NUTS sampling is that it exploit the geometry of the typical set of the posterior distribution contrary to the standard random walk of Metropolis-Hasting sampling.
+Furthermore, using reparametrisation  one can adapt to poor geometry cases (eg.
+\cite{2019arXiv190303704H}).
+%
+\begin{table}[htb]
+\caption{Results of the \textit{relative effective sample size} (aka ESS, ``bulk'') in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
+The ESS can be larger than 100\% in some cases.
+%{\color{red} 
+%(0) "F.L run" are computed from NUTS sampler with 16 %parallel chains of 1,000 samples each and 200 warm-up %and \texttt{max\_tree\_depth=7} (see text) The numbers %are from \texttt{n\_eff/16,000};
+%}
+(a): results from Cobaya 70,000 samples;
+(b): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Section~\ref{sec-NUTS}); 
+(c) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 1 chain of 200 samples and 200 warm-up (Section~\ref{sec-Neural-Reparametrisation}).}
+\label{tab-ESS-NUTS_SVI-1}
+ \centering
+\begin{tabular}{ccccccccccc}
+\hline
+    & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$ & $n_s$ & $A$ & $\eta$\\
+\hline
+(a) &  $3.1$ & $2.5$       & $2.9$      & $2.9$  & $2.6$  & $3.1$  & $3.1$ & $2.8$ \\  
+(b) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\
+(c) & $84.0$ &  $28.0$     & $26.0$     & $20.5$ & $80.0$ & $110.5$ & $58.5$ & $29.5$\\
+\hline
+\end{tabular}
+\end{table}
+%
+\section{General discussion}
+\label{sec-discussion}
+%
+Having demonstrated the interest of \jaxcosmo\ as a differentiable cosmology library library, we now discuss several limitations, and open questions.
+
+%Lack of autodiff Boltzmann code (emulator)
+The first essential barrier to a full-fledged automatically differentiable cosmology library is the need for a differentiable Boltzmann solver to compute the matter power spectrum.
+At this stage, \jaxcosmo\ relies on the analytic Einsenstein \& Hu fitting formula, which is not accurate enough for Stage IV requirements, nor does it allow to explore beyond $\Lambda$CDM models.
+Existing solvers such as CLASS or CAMB \FrL{citations} are large and complex codes which are not easily reimplemented in an \autodiff\ framework and therefore cannot be directly integrated in \jaxcosmo\ .
+
+A first option to resolve this issue is to implement from scratch a new Boltzmann code in a framework that will support automatic differentiation.
+This is the approach behind works such as \FrL{the Julia code} which is provides a simplified Boltzmann solver in Julia, or \FrL{the code from our friends in Zurich} which is based on a SymPy and could be relatively easily compatible with JAX.
+However, even if very promising, both of these options thus far remain limited.
+While we do believe a automatically differentiable Boltzmann code is the best option, it seems that the cost of developing such a code remains very high at this time.
+
+A second approach is to develop emulators of a full-fledged Boltzmann code.
+Emulators based on neural networks or Gaussian processes are themselves automatically differentiable with respect to cosmological parameters.
+In fact, the litterature is now rich in examples of such emulators \FrL{cite all emulators}.
+After validating their accuracy against the reference CAMB or CLASS implementation, they can be directly integrated as a plug-and-play replacement for the computation of the matter power spectrum.
+\FrL{need to talk about cosmopower somewhere.} At this time, it seems that using emulators will be the most straightforward approach to bring more accurate models to \jaxcosmo\ .
+\FrL{However, we want to emphasize that one of the reason for the wide diversity of emulators is a lack of standardization}.
+
+\bigskip
+
+% Reasoning behind not making a full emulator of the correlation functions
+Connected to this discussion about emulators, one point that could be discussed is whether one even needs a library like \jaxcosmo\ if anyway one can build emulators of a CCL likelihood for use inside gradient-based inference algorithms.
+While this could indeed be feasible, and of similar cost as just making an emulator of the matter power spectrum, the drawback of this approach is that many analysis parameters and choices become hardcoded in the emulator.
+While the model for the linear matter power spectrum is typically kept fixed in practical analyses, all the of the choices related in particular to systematics modeling (e.g.
+photometric redshift errors, galaxy bias, or intrinsic galaxy alignments) are very likely to vary significantly in the process of developing the analysis.
+Building an emulator for the likelihood would mean the emulator needs to be retrained everytime the likelihood is changed.
+
+\bigskip
+
+%  - Massively Parallel MCMC on GPU (linked to vmap and/or pmap)
+%  - Efficiency of gradient-based inference methods
+Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo\ .
+As we illustrated in the previous section, gradient-based inference techniques yield significantly less correlated MCMC chains.\FrL{Hummm here I guess we want to say also it scales better with dimensions, and that you can get very fast first results.} \FrL{we also want to say that the GPU parallelisation is going to change the game, all the more when you account for new massive GPU machines.}
+
+\bigskip
+
+%  - Respective role of CCL and jax-cosmo
+Finally, one question worth asking is how does \jaxcosmo\ position itself against classical codes such as CCL?
+Or in other terms, will it replace CCL?
+While we are convinced of the benefits of a JAX implementation, we expect CCL to remain extremely valuable as the reference cosmology implementation.
+\FrL{it allows to leverage the legacy codes, has a lot more features, etc.} Ultimately, a natural transition may occur towards more modern frameworks like \jaxcosmo\ when they reach the ability to run full-fledged Stage IV likelihoods.
+
+\section{Conclusions \& Prospects}
+\label{sec-conclusion}
+%- We have presented a library using automatic differentiation and and automatic GPU offload for a set of key cosmology theory calculations
+%- We currently cover only a small number of the many calculations needed in cosmo theory, and welcome contributions for other areas such as CMB.
+%- The calculation of Fisher matrices using this method, which is infamous for requiring extensive careful tuning, becomes very simple and robust using JC.
+%- We have demonstrated the efficiency in JC of the derivative-aware MCMC methods HMC and NUTS, which are regarded as the only way for samplers to evade the curse of dimensionality up to the hundreds of dimensions we are likely to need for the next generation of surveys.
+%- We have shown a proof-of-concept for the use of JC with machine learning methods, opening a whole new space of methods in cosmology.
+
+In this paper, we have presented \jaxcosmo\ a cosmology library implemented in the JAX framework, enabling the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, as well as greatly speeding up likelihood evaluations thanks to automatic parallelisation and just in time compilation on GPUs.
+Currently, \jaxcosmo\ only contains a small set of features limited corresponding to a DES-Y1 3x2pt analysis.
+Being an open source project, contribution of additional features is additional areas such as CMB or spectroscopic galaxy clustering are warmly welcome.
+
+To demonstrate the interest of an automatically differentiable library, we have illustrated on concrete examples how Fisher matrices, which are notoriously difficult to compute and require extensive and careful fine tuning, can now be computed robustly and at much lower cost.
+In addition, Fisher matrices becomes themselves differentiable, which allows for  Figure of Merit optimization by gradient descent, making survey optimization extremely simple.
+
+Going beyond Fisher forecasts, we have also compared to simple Metropolis-Hastings several gradient-based inference techniques (namely, Hamiltonian Monte-Carlo, No-U-Turn-Sampler, Stochastic Variational Inference).
+We show that the posterior samples with gradient-based methods have a much larger Effective Sample Size, meaning that \FrL{let's think a bit more about interpretation here....
+the thing I care about is not as much the ESS, but the number of likelihood evaluation per independent sample.
+I don't think we have quantified that too much}.
+These inference techniques can scale to hundreds of dimensions, and may become necessary in Stage IV analysis as the number of nuisance parameters is likely to become very large.
+
+\bigskip
+
+The next extensions of this framework will be the inclusion of additional cosmological probes, as well as integrating emulators for the matter power spectrum trained on CAMB or CLASS, as a means to go beyond the current analytic Eisenstein \& Hu model.
+
+\bigskip
+
+In the spirit of reproducible research, all results presented in this paper can be reproduced from the following GitHub repository:
+
+\FrL{add link}\JEC{Verif if all the nb listed are there, and if some have to be withdrawn. }
+
+\url{https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/}
+
+%
+%
+%
+%\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
+%
+%%
+%%
+%\subsection{DES-Y1 Example}
+%
+
+%%%
+\section*{CRediT authorship contribution statement}
+\textbf{J.E Campagne} Conceptualization, Methodology, Software, Validation, Writing.
+\textbf{F. Lanusse}, \textbf{J. Zuntz}...
+
+%\section*{Declaration of Competing Interest}
+%The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
+
+%\Acknowledgements
+\section*{Acknowledgements}
+Some of the numerical experiments have been conducted at the IN2P3 Computing Center (CC-IN2P3 - Lyon/Villeurbanne - France) funded by the Centre National de la Recherche Scientifique.
+
+%% The Appendices part is started with the command \appendix;
+%% appendix sections are then done as normal sections
+\appendix
 

### 2022-11-15 paper_OJA.tex
Switch arxiv references to published versions when possible

@@
 % Note that some direct optimization of likelihood function has also been successfully undertaken for instance for some Planck analysis \citep{2014A&A...566A..54P}, 
-Since the development of these MCMC packages, major advances have been made in automatic differentiation (\textit{autodiff}) \citep{2015arXiv150205767G, 2018arXiv181105031M}, a set of technologies for transforming pieces of code into their derivatives.
+Since the development of these MCMC packages, major advances have been made in automatic differentiation (\textit{autodiff}) \citep{10.5555/3122009.3242010, Margossian2019}, a set of technologies for transforming pieces of code into their derivatives.
 
@@
 
-The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler has been introduced in \citet{nuts}.
+The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler has been introduced in \citet{10.5555/2627435.2638586}.
 It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$.
@@
 %
-Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps.
+Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{10.5555/2567709.2502622, 8588399} and show where auto-differentiation helps.
 If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
@@
 Using the auto-differentation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}).
-As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
+As stated by the authors of \citep{10.5555/3122009.3122023} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
 function, taking derivatives, and running a gradient-based or coordinate-ascent optimization.
@@
 \JEC{
-The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{2015arXiv150603431K}).
+The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{10.5555/2969239.2969303}).
 Let us illustrated the method using an invertible transformation (aka \textit{normalizing flow})  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, such that $\zeta\sim \mathcal{N}(0,I)$\footnote{Notice that $z=T^{-1}(\xi)=(T^{-1} \circ S_\lambda^{-1})(\zeta)=(S_\lambda \circ T)^{-1}(\zeta)=F_\lambda(\zeta)$.}.
@@
 Concerning the B-NAF architecture it is composed of  a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
-The SVI optimization has been performed with the Adam optimizer \citep{2014arXiv1412.6980K} and a learning rate set to $10^{-3}$.
+The SVI optimization has been performed with the Adam optimizer \citep{KingmaB14} and a learning rate set to $10^{-3}$.
 We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the B-NAF (MVN) guides, and no tuning of the optimizer has been undertaken as for instance a learning rate scheduling.
@@
 If the SVI method can be used as is to get $z$ i.i.d.
-samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
+samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{Parno2018,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
 From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ (eg.
@@
 several days on dedicated infrastructure.
-Moreover, the user is faced to the dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, 2015arXiv151007727O}.
+Moreover, the user is faced to the dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, Owen2017}.
 

### 2022-11-17 paper_OJA.tex
remove \centering in caption openjournal.cls (before \@table@type@size...   at line 952)

@@
 
-\usepackage[
-singlelinecheck=false % <-- important
-]{caption}
+% jZ had to remove this to avoid clash
+% \usepackage[
+% singlelinecheck=false % <-- important
+% ]{caption}
 
@@
 \FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
-\JEC{The "vectorized" option runs multi chains in a single GPU thanks to the \jax \textit{vmap} mechanism.
+\JEC{The "vectorized" option runs multi chains in a single GPU thanks to the JAX \textit{vmap} mechanism.
 If one dispose of several GPUs then the "parallel" option using \textit{pmap}  mechanism runs one chain per GPU.
@@
 Connected to this discussion about emulators, one point that could be discussed is whether one even needs a library like \jaxcosmo\ if anyway one can build emulators of a CCL likelihood for use inside gradient-based inference algorithms.
-While this could indeed be feasible, and of similar cost as just making an emulator of the matter power spectrum, the drawback of this approach is that many analysis parameters and choices become hardcoded in the emulator.
+While this could indeed be feasible, and of similar cost as just making an emulator of the matter power spectrum, the drawback of this approach is that many analysis parameters and choices become hard coded in the emulator.
 While the model for the linear matter power spectrum is typically kept fixed in practical analyses, all the of the choices related in particular to systematics modeling (e.g.

### 2022-11-21 paper_OJA.tex
Introducing in one-sentence the pmap and multi-mesh devel.

@@
 The function above is trivial, but the same approach can be used to parallelize any function, from Limber integrals, to an entire likelihood.
+\JEC{Considering the case of multi devices (eg.
+GPU or TPU), JAX provides \texttt{pmap} which compile and execute in parallel replicates of the same code on each device, and recent experimental developments deal with parallelization of function over supercomputer-scale hardware meshes.
+In the examples detailed in this article, we have only relied on \texttt{vmap} functionality.}
 
@@
 \label{sec-jaxcosmo-design}
-\FrL{I'll do another pass over this section when I have internet, I want to compare to what we did in CCL.}
 
 In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX.
-The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though \textbf{in its initial release, }
+The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though, in its initial release,
  \jaxcosmo\ only implements a subset of CCL features and options.
 
-The cosmological parameters are organised in a JAX \textit{pyTree} container\footnote{\url{https://jax.readthedocs.io/en/latest/jax-101/05.1-pytrees.html}} \texttt{jc.Cosmology} where \texttt{jc} hereafter stands for \jaxcosmo.
-This container allows one to use \textit{autodiff} and \texttt{vmap} transparently \FrL{I don't think people know what autodiff and vmap are.}.
-The present set of exposed parameters is: $\Omega_{cdm}$ the cold dark matter density fraction, $\Omega_b$ the Baryonic matter density fraction, $h$ the Hubble constant divided by $100$~km/s/Mpc, $n_s$ the primordial scalar perturbation spectral index, $\sigma_8$ the variance of matter density perturbations at an $8$~Mpc/h scale, $\Omega_k$ the curvature density fraction, $w_0$ the first order term of dark energy equation and $w_a$ the second order term of dark energy equation of state.
-Notably, \texttt{jc.Planck15} instantiates a set of cosmological parameters inspired from the Planck 2015 results \citep{2016A&A...594A..13P}.
-\FrL{I'm not sure we want to keep this paragraph, it's not extremely useful. The only useful thing to discuss I guess might be the choice of parametrisation in terms of A or sigma8.}
+All \jaxcosmo\ data structures are organized as JAX \textit{pyTree} container objects, ensuring that two key JAX features are available to them: \textit{autodiff} and \textit{vmap}.
+The vmap feature enables any operation defined in JAX (including complicated composite operations) to be applied efficiently as a vector operation.
+The autodiff feature further makes it possible to take the derivative of any operation, automatically transforming a function that takes $n$ inputs and produces $m$ outputs into a new function that generates an $m \times n$ matrix of partial derivatives.
+
+\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, and Simpson \& Romberg integration routines through this framework, so that we can automatically compute the derivative of their solutions with respect to their inputs.
+This includes the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also the other input quantities such as redshift distributions described below.
+
+In the rest of this section we describe the cosmological calculations that are implemented using these facilities.
 
@@
 \end{equation}
-with $a=1/(1+z)$ the scale factor related to the redshift $z$, $H(a)=\dot{a}/a$ the Hubble parameter with $\mathrm{H}_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and
+with $a=1/(1+z)$ the scale factor related to the redshift $z$, $\mathrm{H}(a)=\dot{a}/a$ the Hubble parameter with $\mathrm{H}_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and
 \begin{equation}
@@
 \end{equation}
-Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are presently neglected.
-From these expressions, in \texttt{jc.background}, are computed the different cosmological distance functions.
-Notably, the radial comoving distance
+Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are currently neglected.
+From these expressions, in \texttt{jc.background}, are computed the different cosmological distance functions, such as the radial comoving distance
 \begin{equation}
@@
 \end{equation}
-with $R_H$ the Hubble radius. Using the scale factor ($a$) redshift ($z$) relationship, $\chi$ can be viewed as a function of $z$. 
+with $R_H$ the Hubble radius. 
+%Using the scale factor ($a$) redshift ($z$) relationship, $\chi$ can be viewed as a function of $z$. 
 %
@@
 %
-Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function and non-linear features through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}.
+Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function $T$ which transforms the primordial matter power spectrum to its late-time non-linear value:
+\begin{equation}
+    P(k, z) = P(k, z=\infty) \cdot T(k, z; \Omega_m, \Omega_b, ...),
+\end{equation}
+through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}.
 No Baryon feedback modeling is considered yet.
-To perform \textit{autodiff} we have implemented various tools as differentiable ODEs using the Runge-Kutta algorithm, as well as Simpson quadrature and the Romberg algorithm for numerical integration.
-\FrL{Hummm it's a bit strange to talk about the theory and implementation in the same paragraph...
-Especially because we don't specify which concrete implementation is used here.}
+
+The primordial power spectrum is modelled with the standard form:
+\begin{equation}
+    P(k) = A k^{n_s}.
+\end{equation}
+
+The normalisation $A$ is parametrized via $\sigma_8$ at $z=0$ as 
+\begin{equation}
+    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
+\end{equation}
+with $R_8 = 8 \mathrm{Mpc}/h$ and $W(x)$ related to the $j_1$ spherical Bessel function as
+\begin{equation}
+    W(x) = \frac{3j_1(x)}{x}
+\end{equation}
 
@@
 
-To make a bridge to these observables, two \textit{pyTree} objects are implemented: \texttt{NumberCounts} and \texttt{WeakLensing} tracers for galaxy clustering and weak lensing samples, with each instance of each representing a single tomographic redshift bin's number density \FrL{We probably should refer to those as Probes instead of Pytrees, in general we should separate the implementation details from the theory}.
+The angular power spectra $C_\ell^{ij}$ for the probes $(i,j)$ and for redshift bin window selections are computed in the first order Limber approximation \citep{PhysRevD.78.123506}:
+\begin{align}
+    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{d\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
+\end{align}
+The factors $m_i$ are equal to $(0,-2)$ for the galaxy clustering and weak lensing respectively, and the functions $K(z)$ each represent a single tomographic redshift bin's number density.
 These tracers are implemented as two kernel functions:
@@
 
-Notice that all described kernel functions can be also viewed as function of the scale factor $a$ or as function of the radial comoving distance $\chi$.
-
-\FrL{We probably actually want to start by the Limber formula, which explicits what a kernel is, otherwise people don't necessarily know where these kernels come from.}
-From these tracer objects, the angular power spectra $C_\ell^{ij}$ for the probes $(i,j)$ and for redshift bin window selections are computed in the first order Limber approximation \citep{PhysRevD.78.123506}:
-%\JZ{Discuss the implementation of the Limber integral here. Maybe move down the bit in the growth of perturbations section?}
-\begin{align}
-    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{d\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
-\end{align}
-with $P(k,z)$ the power spectra computed using the growth of perturbation models described above and a primordial power spectra scaling according to
-\begin{equation}
-    P(k) = A k^{n_s}
-\end{equation}
-The normalisation $A$ is computed using the $\sigma_8$ parameter as 
-\begin{equation}
-    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
-\end{equation}
-\JEC{\st{At which z ?}}
-with $R_8 = 8$ in unit of Mpc and $W(x)$ is related to the $j_1$ spherical Bessel function as
-\begin{equation}
-    W(x) = \frac{3j_1(x)}{x}
-\end{equation}
-Finally, the factors $m_i$ are equal to $(0,-2)$ for the \texttt{NumberCounts}, \texttt{WeakLensing} related kernels, respectively.
-
-Remarkably, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters thanks to \textit{autodiff} and the implementation of the above mentioned integration quadrature \FrL{Are we using quadratures currently?
-I think by default we are using simple Simpson no?}\JEC{Correct: the current pip version uses Simpson but it is a quadrature even if it is qualified "simple".}.
+% Notice that all described kernel functions can be also viewed as function of the scale factor $a$ or as function of the radial comoving distance $\chi$.
+
+% \FrL{We probably actually want to start by the Limber formula, which explicits what a kernel is, otherwise people don't necessarily know where these kernels come from.}
+
+Because, like the other ingredients, the kernels are implemented as pyTree objects, \texttt{NumberCounts} and \texttt{WeakLensing}, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters  or the number densities using \textit{autodiff} and \jaxcosmo's implementation of integration quadrature.
 An example is given in the context of DES Y1 3x2pts analysis (Sec~\ref{sec-DESY1}).
@@
 %
-To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to make a comparison with the well-established  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{\url{https://ccl.readthedocs.io}, version \texttt{2.5.1}.}.
-As examples, the Figures \ref{fig:chi_comparison},
+To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to compare it to the well-validated  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{\url{https://ccl.readthedocs.io}, version \texttt{2.5.1}.}.
+As examples, Figures \ref{fig:chi_comparison},
 \ref{fig:halofit_comparison} and \ref{fig:Cell_comparison}
-show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non linear matter power spectrum computation using the \texttt{halofit} fitting function and the angular power spectrum for galaxy-galaxy lensing (Eq.~\ref{eq:Cell_limber}) using the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions.
-\FrL{In addition, \jaxcosmo features a suite of validation tests against CCL, automatically validating the precision of all computations to within the desired numberical accuracy.} The relative differences between the two libraries are at the level of few $10^{-3}$ or better.
+show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non-linear matter power spectrum computation, and the angular power spectrum for galaxy-galaxy lensing (Eq.~\ref{eq:Cell_limber}) using the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions.
+\jaxcosmo\ features a suite of validation tests against CCL, automatically validating the precision of all computations to within the desired numberical accuracy; the relative differences between the two libraries are at the level of few $10^{-3}$ or better.
 
@@
 \subsection{Instantaneous Fisher Forecasts}
-\FrL{We should discuss why Fisher matrices are usually difficult first.}
-We first illustrate, with a purely fictional case study, the computation of a Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} using two methods with the \textit{autodiff} ability of JAX.
+Fisher matrices are a key tool in cosmology forecasting and experimental planning.
+By computing the Hessian matrix of a likelihood with respect to its parameters, we can find a Gaussian approximation to a posterior, which is usually sufficient for comparing the constraining power of different experimental configurations.
+As noted above, computing Fisher matrices is notoriously error-prone, since finite difference approximations to likelihoods must be carefully tuned for convergence, and observable calculations can be numerically unstable.
+Autodiff can help avoid this challenge.
+
+We first illustrate, with an artificial case study, the computation of a Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} using two methods with the \textit{autodiff} ability of JAX.
 For the detailed implementation, the reader is invited to look at the following companion notebook \nblink{Simple-Fisher}.
@@
 \JEC{The "vectorized" option runs multi chains in a single GPU thanks to the JAX \textit{vmap} mechanism.
-If one dispose of several GPUs then the "parallel" option using \textit{pmap}  mechanism runs one chain per GPU.
-As noted in \numpyro\ MCMC documentation, it is possible to mix parallel and vectorized sampling\footnote{Since version 0.9.0.}.}
-
-The experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
+If one dispose of several GPU devices, then the using the JAX paralellization mechanism (\textit{pmap}), it is possible to launch on each GPU device the "vectorized" sampling as done in the previous snippet, and get back all the MCMC chains.}
+
+Although, experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
 The elapse time for these experiments was 20 hours.

### 2022-11-30 paper_OJA.tex
polish JEC comments/adds

@@
 
-The first approach is to derive analytically the formula for the derivatives of interest \citep{2019A&A...625A..64J} \JEC{This reference should be validated by F.L}, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}.
-This is however only practical typically for simple analytical models.
+The first approach is to derive analytically the formula for the derivatives of interest \citep{2019A&A...625A..64J}%JEC: This reference should be validated by F.L
+, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}. This is however only practical typically for simple analytical models.
 
@@
 The function above is trivial, but the same approach can be used to parallelize any function, from Limber integrals, to an entire likelihood.
-\JEC{Considering the case of multi devices (eg.
-GPU or TPU), JAX provides \texttt{pmap} which compile and execute in parallel replicates of the same code on each device, and recent experimental developments deal with parallelization of function over supercomputer-scale hardware meshes.
-In the examples detailed in this article, we have only relied on \texttt{vmap} functionality.}
+When considering multi-devices use-case (eg.
+GPU or TPU), JAX provides \texttt{pmap} which compile and execute in parallel replicates of the same code on each device.
+Moreover, recent experimental developments deal with parallelization of function over supercomputer-scale hardware meshes.
+Notice that in the examples detailed in this article, we have only relied on \texttt{vmap} functionality.
 
@@
 \FrL{hummm but we still do need a full covariance in the compression algorithm, right?} \FrL{And we still do need to invert it...
-is the idea rather that if we dont have a perfect matrix at that stage the worse that can happen is sub optimal compression?}\JEC{Well there are several ways that $C$ can be questioned: parameter dependence (ie.
+is the idea rather that if we dont have a perfect matrix at that stage the worse that can happen is sub optimal compression?}
+%JEC{Well there are several ways that $C$ can be questioned: parameter dependence (ie.
 fiducial model/true model), non-gaussianities, errors...
@@
 \label{sec-DESY1}
-\JEC{Transfer of the model context from apprndix to here to ease the reading.}
-
+%JEC{Transfer of the model context from appendix here to ease the reading.}
 From the DES server\footnote{\url{http://desdr-server.ncsa.illinois.edu/despublic/y1a1_files/chains/2pt_NG_mcal_1110.fit}} we have extracted the $N(z)$ distributions of the sources and normalize them to $[1.47, 1.46, 1.50, 0.73]$ effective number of sources per $\mathrm{arcmin}^2$ for each of the four redshift bins, respectively.
@@
 This is also used to set the scale of the random initial velocities.
-\JEC{These differential equations may be integrated iterating over $L$ steps the following \textit{leapfrog} algorithm
+These differential equations may be integrated iterating over $L$ steps the following \textit{leapfrog} algorithm
 \begin{align}
@@
 \end{align}
-where $\varepsilon$ is a step size parameter.}
+where $\varepsilon$ is a step size parameter.
 
@@
 you only need the derivative of the log posterior wrt parameters.
-The hessian would be needed to get a good estimate of M.}\JEC{Yes.
-Why the Hessian?}
+The hessian would be needed to get a good estimate of M.}
 
 \FrL{Needs a little paragraph to explain with HMC is supposed to be better than plain Metropolis-Hastings.}
-\JEC{Metropolis-Hastings suffers from random walk to explore the geometry of the target distribution to get the region of maximum probability (ie.
+Metropolis-Hastings suffers from random walk to explore the geometry of the target distribution to get the region of maximum probability (ie.
 typical set) which is particularly  penalising in large dimensions.
-The HMC dynamics allow to make large jumps to explore the region around a sample to guess the next good one.}
+The HMC dynamics allow to make large jumps to explore the region around a sample to guess the next good one.
 
@@
 As stated by the authors, NUTS  uses a recursive algorithm to build a set of likely candidate samples that spans a wide swath of the target distribution, stopping automatically when it starts to double back and retrace its steps.
-\FrL{\st{Parameters L an epsilon should be introduced at the level of HMC then.}}\JEC{done} So, in an automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
+So, in an automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
 The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls at each generation.
 
-Both sampler HMC and NUTS are available in the \numpyro\ library: \texttt{NUTS} is inherited from \texttt{HMC} which in turns is inherited from the generic \texttt{MCMC} class.
-\FrL{class inheritance is TMI.}  After the forward model creation for the DES-Y1 3x2pts exercise described in section \ref{sec-DESY1} \FrL{We should introduce the model at the beginning of this section.}\JEC{Done}
-\begin{itemize}
-    \item \st{a reparametrisation of the latent variable uniform priors to use a unified uniform prior $\mathcal{U}[-5,5]$} \JEC{we have first applyied  an affine transformation on cosmological, intrinsic alignment and biais parameters to use an unique uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1})}; \FrL{these are not latent variables, but model parameters.
+Both sampler HMC and NUTS are available in the \numpyro\ library. After the forward model creation for the DES-Y1 3x2pts exercise described in section \ref{sec-DESY1}: \begin{itemize}
+    \item we have first applyied  an affine transformation on cosmological, intrinsic alignment and biais parameters to use an unique uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1}); %\FrL{these are not latent variables, but model parameters.
 Also people don't know what this reparametrisation means.}
-    \item  \FrL{People don't know what the warmup phase is...}\JEC{we use a structured mass matrix consisting on essentially a diagonal form  with a block structure for the following sets of parameters} $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
+    \item  %\FrL{People don't know what the warmup phase is...}
+    we use a structured mass matrix consisting on essentially a diagonal form  with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
 This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
@@
 
-\JEC{As an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed with \numpyro:}
+As an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed with \numpyro:
 \begin{lstlisting}[language=iPython]
@@
 128 steps for each iteration).
-\FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
-\JEC{The "vectorized" option runs multi chains in a single GPU thanks to the JAX \textit{vmap} mechanism.
-If one dispose of several GPU devices, then the using the JAX paralellization mechanism (\textit{pmap}), it is possible to launch on each GPU device the "vectorized" sampling as done in the previous snippet, and get back all the MCMC chains.}
-
-Although, experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
+%\FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
+The "vectorized" option runs multi chains in a single GPU thanks to the JAX \textit{vmap} mechanism.
+If one dispose of several GPU devices, then the using the JAX paralellization mechanism (\textit{pmap}), it is possible to launch on each GPU device the "vectorized" sampling as done in the previous snippet, and get back all the MCMC chains.
+However, experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
 The elapse time for these experiments was 20 hours.
@@
 which exhibit the Jacobian of the $T^{-1}$ transformation.
-\JEC{
 The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{10.5555/2969239.2969303}).
@@
 An concrete example of the gradient formula is given in Appendix \ref{app-SVI_MVN} with the multivariate Gaussian distribution family used as the guide.
-}%JEC
-%
-\FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.} 
-\FrL{we don't have plots of just the VI?}\JEC{Good point: I have SVI alone with both MVN or Block Neural Autoregressive Flow. The contours are narrower which would require to elaborate...}
+%
+%\FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.} 
+%\FrL{we don't have plots of just the VI?}\JEC{Good point: I have SVI alone with both MVN or Block Neural Autoregressive Flow. See new figure.}
 
@@
 %
-\JEC{Once the optimisation is done on gets $z$ i.i.d.
+Once the optimisation is done on gets $z$ i.i.d.
 samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
@@
 We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the B-NAF (MVN) guides, and no tuning of the optimizer has been undertaken as for instance a learning rate scheduling.
-It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps}.
+It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps.
 \FrL{Arf...
@@
 
-\JEC{
 In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and B-NAF guided SVI.
@@
 It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling.
-The difference is that these SVI approximate posteriors have been obtained in a short time and these posterior estimates can serve as front-end to a NUTS sampler as described in the next section.}
+The difference is that these SVI approximate posteriors have been obtained in a short time and these posterior estimates can serve as front-end to a NUTS sampler as described in the next section.
 %
@@
 %
-\subsection{\JEC{Sampling efficiency}}
+\subsection{Sampling efficiency}
 \label{sec-results}
@@
 
-\FrL{add link}\JEC{Verif if all the nb listed are there, and if some have to be withdrawn. }
+%\FrL{add link}\JEC{Verif if all the nb listed are there, and if some have to be withdrawn. }
 
@@
 
-\section{\JEC{SVI example with MVN guide}}
+\section{SVI example with MVN guide}
 \label{app-SVI_MVN}

### 2022-11-30 paper_OJA.tex
correct a section ref

@@
 \label{app-SVI_MVN}
-Let us illustrate the \textit{elliptical standardisation} described in section \label{ec-SVI} in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
+Let us illustrate the \textit{elliptical standardisation} described in section \ref{sec-SVI} in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
 a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).

### 2022-11-30 paper_OJA.tex
reduce Appendix

@@
 \end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model.  
-For the use case of the MVN (auto)guide, one gets
-\begin{align}
--\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \nn \\
-& \left. \qquad \qquad \qquad \qquad \qquad \qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right\} \nn \\
--\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
-\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
-&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
-\right\}+ (L^{-1})^T
-\end{align}
-\FrL{Yeah.
-this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.} The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
+where $\mathbb{H}(q)$ is the Shannon entropy of the $q$ distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model.
+In case of the MVN for completeness  $\nabla_\mu \mathbb{H}(q)=0$, $\nabla_L \mathbb{H}(q)=(L^{-1})^T$ while the other gradients are computed by the JAX \texttt{autodiff} mechanism.
+%\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.}
+
+The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
 As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX.

### 2022-12-01 paper_OJA.tex
Joe comments/rephrasing

@@
 
-\section{Unleashing the Power of Fisher Forecasts with Automatic Differentiation}
+% \section{Unleashing the Power of Fisher Forecasts with Automatic Differentiation}
+\section{Fisher Forecasts \& Data Compression}
 \label{sec-fisher-forecast}
 
-As a first illustration of the interest of \textit{autodiff} in cosmological computations, we present in this section a few direct applications involving the computation of the Fisher information matrix, widely used in cosmology.
+As a first illustration of the value of \textit{autodiff} in cosmological computations, we present in this section a few direct applications involving the computation of the Fisher information matrix, widely used in cosmology.
 \\
-Not only does the computation of the Fisher matrix become a trivial business, but the Fisher matrix itself becomes differentiable, allowing in particular for powerful survey optimization applications.
+Not only does the computation of the Fisher matrix become trivial, but the Fisher matrix itself becomes differentiable, allowing in particular for powerful survey optimization applications.
 
@@
 
-The first approach to obtaining approximate 1-sigma contours of the two parameters ($\Omega_c, \sigma_8$) with aa Fisher matrix uses the Hessian of the log-likelihood as follows:
+The first approach to obtaining approximate 1-sigma contours of the two parameters ($\Omega_c, \sigma_8$) with a Fisher matrix uses the Hessian of the log-likelihood as follows:
 \begin{equation}
@@
 % Maybe this can be only mentioned in the discussion.
-\FrL{Insist here on the fact that the FoM itself is differentiable.}
+% \FrL{Insist here on the fact that the FoM itself is differentiable.}
 Fisher forecasts are also commonly used in survey and analysis strategy strategy, where running a full posterior analysis for each possible choice would be unfeasible.
@@
 
+Because JAX functions are differentiable with respect to all their inputs, including survey configuration parameters (e.g.
+depth, area, etc), we can even compute the derivative of an FoM with respect to these inputs, allowing for rapid and complete survey optimization.
+
 \subsection{Massive Optimal Compression in 3 Lines}
 
-\FrL{Here we can implement MOPED for the compression}
-
-Once the Fisher matrix can be safely estimated, one can use the MOPED\footnote{Massively Optimised Parameter Estimation and Data compres-
-sion} algorithm \citep{2000MNRAS.317..965H,2016PhRvD..93h3525Z, 2017MNRAS.472.4244H}.
+% \FrL{Here we can implement MOPED for the compression}
+
+Once the Fisher matrix has been accurately estimated, the MOPED\footnote{Massively Optimised Parameter Estimation and Data compression} algorithm can be used to compress data sets with minimal information loss \citep{2000MNRAS.317..965H,2016PhRvD..93h3525Z, 2017MNRAS.472.4244H}.
 In the case of the constant covariance matrix the algorithm compresses data in a way that is lossless at the Fisher matrix level (i.e.
@@
 For instance, in the previous section, $N=500$ as $\mu=(C_\ell^{p,q})$ and $M=2$ for $(\Omega_c,\sigma_8)$.
+
 The algorithm computes by iteration $M$ vectors of size $N$ such that (taking the notation of the previous section)
@@
     b_i = \frac{C^{-1}\mu_{,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)b_j}{\sqrt{F_{i,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)^2}}
+\label{eq:moped}
 \end{equation}
@@
 The algorithm is similar to the Gram-Schmidt process, using the constant covariance matrix $C$ to define the scalar product $\langle b_i, b_j\rangle$.
-Then, the original data set $x=(C_\ell^{p,q})_{p,q,\ell}$ is compressed in a data set composed of $M$ numbers $(y_i)_{i\leq M}$ according to
+Then, the original data set $x=C_\ell^{p,q}$ is compressed in a data set composed of $M$ numbers $y_i$ according to
 \begin{equation}
@@
 \end{equation}
-These numbers are uncorrelated and of unit variance and this construction ensures that the log-likelihood of $(y_i)_{i\leq M}$  $(\theta_i)_{i\leq M}$, $y_i$ is identical to that of $x$ up to second order, meaning that the Fisher matrices derived from the two parameters should be identical, and in general the $y$ values should lose very little information compared to the full likelihood.
+These numbers are uncorrelated and of unit variance and this construction ensures that the log-likelihood of $y_i$ given  $\theta_i$ is identical to that of $x$ up to second order, meaning that the Fisher matrices derived from the two parameters should be identical, and in general the $y$ values should lose very little information compared to the full likelihood.
 
@@
 Since the uncertainty in the covariance due to a finite number of simulations must be accounted for \citep{2018MNRAS.473.2355S,2007A&A...464..399H}, this reduction can also offset any loss of information from the compression.
-\FrL{hummm but we still do need a full covariance in the compression algorithm, right?} \FrL{And we still do need to invert it...
+Inaccuracies in the full covariance matrix used in the data compression result only in information loss, not bias, as does mis-specification of the fiducial parameter set $\theta_i$ where equation \ref{eq:moped}.
+
+% \FrL{hummm but we still do need a full covariance in the compression algorithm, right?} \FrL{And we still do need to invert it...
 is the idea rather that if we dont have a perfect matrix at that stage the worse that can happen is sub optimal compression?}
@@
 @jax.jit
-def compress_likelihood(p):
+def compressed_likelihood(p):
     # Create a new cosmology at these parameters
@@
 \end{lstlisting}
-The comparison between contour lines obtained with the original likelihood (uncompressed data set) and the MOPED version are shown in Figure \ref{fig:moped} in the case study of the previous section.
+The comparison between contour lines obtained with the original likelihood (uncompressed data set) and the MOPED version are shown in Figure \ref{fig:moped} for the case study of the previous section.
 Clearly, close to the negative likelihood minimum the lines agree very well.
@@
 %
-In the following sections we review some statistical methods which directly benefit from the automatic differentiablity of \jaxcosmo likelihoods.
+In the following sections we review some statistical methods which directly benefit from the automatic differentiablity of \jaxcosmo\ likelihoods.
 We demonstrate a range of gradient-based methods, from Hamiltonian Monte Carlo (HMC), and its \textit{No-U-Turn} (NUTS) variant, to Stochastic Variational Inference (SVI).
@@
 %JEC{Transfer of the model context from appendix here to ease the reading.}
-From the DES server\footnote{\url{http://desdr-server.ncsa.illinois.edu/despublic/y1a1_files/chains/2pt_NG_mcal_1110.fit}} we have extracted the $N(z)$ distributions of the sources and normalize them to $[1.47, 1.46, 1.50, 0.73]$ effective number of sources per $\mathrm{arcmin}^2$ for each of the four redshift bins, respectively.
-The lens distributions $N(z)$ for five redshift bins are also extracted.
-The corresponding shapes of the different distributions are performed thanks to \texttt{jc.redshift.kde\_nz} function (``\texttt{jc}'' hereafter stands for \texttt{jax\_cosmo}) and are presented in Figure \ref{fig-DESY1-src-lens-redshift}.
+From the DES Year 1 lensing and clustering data release \footnote{\url{http://desdr-server.ncsa.illinois.edu/despublic/y1a1_files/chains/2pt_NG_mcal_1110.fit}} we have extracted the $N(z)$ distributions of four source and five lens samples.
+We normalize the sources to $[1.47, 1.46, 1.50, 0.73]$ effective number of sources per $\mathrm{arcmin}^2$.
+These distributions are modelled in \jaxcosmo\ using a kernel density estimation in the \texttt{jc.redshift.kde\_nz} function and are presented in Figure \ref{fig-DESY1-src-lens-redshift}.
 \begin{figure}
@@
 
-Then, using the \numpyro\ PPL, we have set up a \textit{forward model} following the reference \citep{2018PhRvD..98d3526A}.
-The prior descriptions of the different model parameters are described in Table \ref{tab-DESY1}.
-For instance the $\Omega_c$ parameter is treated as random variable as followed
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
-Omega_c = numpyro.sample('Omega_c', Uniform(0.1, 0.9))
-\end{lstlisting}
-%\end{minted}
+Using the \numpyro\ PPL, we then set up a forward model following the DES Y1 Pipeline \citep{2018PhRvD..98d3526A}.
+The model parameters and their prior distributions are shown in Table \ref{tab-DESY1}.
+For instance the $\Omega_c$ parameter is treated as random variable as follows
+
+\begin{lstlisting}[language=iPython]
+Omega_c = numpyro.sample("Omega_c", Uniform(0.1, 0.9))
+\end{lstlisting}
 
 Then, we generate mock angular power spectra $C_\ell$ from the auto \& cross correlations of the Number Counts and Weak Lensing probes computed with the \texttt{gaussian\_cl\_covariance\_and\_mean} function in the \texttt{jc.angular\_cl} module.
-One reads
-%\begin{minted}[fontsize=\footnotesize]{python}
+The code reads:
 \begin{lstlisting}[language=iPython]
@@
 \end{lstlisting}
-%\end{minted} 
+
 with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} (ie.
 $\ell$) a series of 50 angular modes.
-The generation of the mock data (\texttt{cl\_obs}) yields from the possibility to execute a conditioned model with a fiducial cosmology, leaving \numpyro\ dealing with the random number generation:
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
-fiducial_model = condition(model,
-    {'Omega_c':0.2545, 'sigma8':0.801, 
-    'h':0.682, 'Omega_b':0.0485, 'w0':-1.,'n_s':0.971,
-     'A':0.5,'eta':0.,
-     'm1':0.0,'m2':0.0,'m3':0.0,'m4':0.0,
-     'dz1':0.0,'dz2':0.0,'dz3':0.0,'dz4':0.0,
-     'b1':1.2,'b2':1.4,'b3':1.6,'b4':1.8,'b5':2.0
+Our mock data (\texttt{cl\_obs}) comes from this model evaluated at a fiducial cosmology, with \numpyro\ dealing with the random number generation that generates added noise:
+
+\begin{lstlisting}[language=iPython]
+fiducial_model = numpyro.condition(model,
+    {"Omega_c":0.2545, "sigma8":0.801, 
+    "h":0.682, "Omega_b":0.0485, "w0":-1.,"n_s":0.971,
+     "A":0.5,"eta":0.,
+     "m1":0.0,"m2":0.0,"m3":0.0,"m4":0.0,
+     "dz1":0.0,"dz2":0.0,"dz3":0.0,"dz4":0.0,
+     "b1":1.2,"b2":1.4,"b3":1.6,"b4":1.8,"b5":2.0
       })
@@
 \end{lstlisting}
-%\end{minted}
-
-The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generates the model $C_{\ell}$ from the priors
-%\begin{minted}[fontsize=\footnotesize]{python}
+\JZ{We do not describe the model function above here. I guess it just adds noise to the cl?}
+
+The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generates the model $C_{\ell}$ from the priors:
 \begin{lstlisting}[language=iPython]
@@
                               probes).flatten()
-return numpyro.sample('cl', MultivariateNormal(cl, 
+return numpyro.sample("cl", MultivariateNormal(cl, 
                             precision_matrix=P,
@@
 \end{lstlisting}
-%\end{minted}
-
-These theoretical $C_{\ell}$ are in turn conditioned by the mock $C_\ell$ as followed
-%\begin{minted}[fontsize=\footnotesize]{python}
-\begin{lstlisting}[language=iPython]
-observed_model = condition(model, {'cl': data})
-\end{lstlisting}
-%\end{minted}
-%Then, as an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed
+
+These theoretical $C_{\ell}$ are in turn conditioned on the mock $C_\ell$:
+
+\begin{lstlisting}[language=iPython]
+observed_model = numpyro.condition(model, {"cl": data})
+\end{lstlisting}
+
+%Then, as an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as follows
 %\begin{minted}[fontsize=\footnotesize]{python}
@@
 HMC samples spaces by generating particle trajectories through the space, using the log-posterior as the negative potential energy of a particle at each point $q$ in the space.
-Associated to $q$, we introduce an auxiliary $p$ variable in the spirit of Hamiltonian momentum such that
-\begin{equation}
-- \log{\cal P}(q) = V(q) + cte; \quad H(q,p) = V(q) + U(p)
-\end{equation}
-where $U(p)$ is a sort of \textit{kinetic} term defined by 
+Associated with $q$, we introduce an auxiliary $p$ variable as Hamiltonian momentum such that
+\begin{equation}
+- \log{\cal P}(q) = V(q) \quad H(q,p) = V(q) + U(p)
+\end{equation}
+where $U(p)$ is a kinetic energy-like term defined by 
 \begin{equation}
@@
 \end{equation}
-At each sample, a trajectory is initialized with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration :
+At each sample, a trajectory is initialized with a random momentum $p$, and then Hamilton's equations are integrated:
 \begin{align}
@@
 This is also used to set the scale of the random initial velocities.
-These differential equations may be integrated iterating over $L$ steps the following \textit{leapfrog} algorithm
+These differential equations may be integrated numerically, taking $L$ small steps of the \textit{leapfrog} algorithm:
 \begin{align}
@@
 
-Formally, the set of $n_\mathrm{dim}$ moment are treated as new parameters, and after 
+Formally, the set of $n_\mathrm{dim}$ momenta are treated as new parameters, and after 
 some number of integration steps a final point in the trajectory is compared to the initial one,
 and a Metropolis-Hastings acceptance criterion on the total energy $H(q,p)$ is applied.
-If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
-a relaxation to the integration accuracy.
-
-That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
+If the trajectory is perfectly simulated then this acceptance is unity, since energy is conserved; applying it allows
+a relaxation of the integration accuracy.
+
+The gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
 but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
 in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
-calculating a Hessian, as in \jaxcosmo, makes it feasible and efficient.
-\FrL{I don't think it uses a Hesssian?
-you only need the derivative of the log posterior wrt parameters.
-The hessian would be needed to get a good estimate of M.}
-
-\FrL{Needs a little paragraph to explain with HMC is supposed to be better than plain Metropolis-Hastings.}
-Metropolis-Hastings suffers from random walk to explore the geometry of the target distribution to get the region of maximum probability (ie.
-typical set) which is particularly  penalising in large dimensions.
-The HMC dynamics allow to make large jumps to explore the region around a sample to guess the next good one.
-
-The tricky point is that the \textit{leapfrog} algorithm needs tuning to set the number of steps as well as the step size of integration.
+calculating the derivative, as in \jaxcosmo, makes it feasible and efficient.
+
+Metropolis-Hastings, and related methods like \texttt{emcee} \citep{goodman-weare,emcee},  suffer as dimensionality increases,  as the region of high probability mass (the \textit{typical set}) becomes a very small fraction of the total parameter space volume.
+At high enough dimension they become a slow random walk around the space and cannot remain in typical set regions.
+The dynamics of HMC allows it to make large jumps that nonetheless stay within the region of high posterior.
+
+The tricky part of HMC is that the \textit{leapfrog} algorithm needs tuning to set the number of steps as well as the step size of integration.
 The next section addresses this problem thanks to the No-U-Turn HMC version.
@@
 
-The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler has been introduced in \citet{10.5555/2627435.2638586}.
-It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$.
-Namely, the root algorithm of HMC deals with the \textit{leapfrog} integration which requires to set two main parameters: $\varepsilon$ the step size and $L$ the number of steps.
-As stated by the authors, NUTS  uses a recursive algorithm to build a set of likely candidate samples that spans a wide swath of the target distribution, stopping automatically when it starts to double back and retrace its steps.
-So, in an automatic manner the parameter $L$ is adjusted to produce statistically  a new point $x_{i+1}$ independent from $x_i$, at the expense of the increase the number of model calls.
-The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives in power of $2$ the maximal number of model calls at each generation.
-
-Both sampler HMC and NUTS are available in the \numpyro\ library. After the forward model creation for the DES-Y1 3x2pts exercise described in section \ref{sec-DESY1}: \begin{itemize}
-    \item we have first applyied  an affine transformation on cosmological, intrinsic alignment and biais parameters to use an unique uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1}); %\FrL{these are not latent variables, but model parameters.
-Also people don't know what this reparametrisation means.}
-    \item  %\FrL{People don't know what the warmup phase is...}
-    we use a structured mass matrix consisting on essentially a diagonal form  with a block structure for the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
-This matrix structuring is motivated by the expected degree of correlation of the concerned parameters as can be shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
+The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler was introduced in \citet{10.5555/2627435.2638586}.
+It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$ by finding good and dynamic choices for the leapfrog integration parameters in the root HMC algorithms, the step size $\varepsilon$ and the number of steps $L$.
+
+NUTS sets iterates the leapfrog algorithm not for a fixed $L$, but until the trajectory starts to ``double back'' and return to previously visited region, at the cost of increasing the number of model evaluations.
+The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives as a power of $2$ the maximum number of model calls at each generation.
+
+Both sampler HMC and NUTS are available in the \numpyro\ library. After the forward model creation for the DES-Y1 3x2pt exercise described in section \ref{sec-DESY1}: \begin{itemize}
+    \item we apply an affine transformation on the cosmological, intrinsic alignment and bias parameters to use a consistent uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1});
+    \item
+    we use a structured mass matrix $M$ in a block diagonal form with the blocks as the following sets of parameters $(\Omega_b,\Omega_c,\sigma_8,w_0,h)$ and $(b_i)_{i:1\dots5}$.
+The remaining parameters have uncorrelated masses.
+This matrix structure is motivated by the expected degree of parameter correlation as shown for instance in Figure \ref{fig_cobaya_NUTS_SVI_bis}.
 \end{itemize}
 
-As an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as followed with \numpyro:
-\begin{lstlisting}[language=iPython]
-%nuts_kernel = numpyro.infer.NUTS(observed_model,
-              step_size=1e-1,
-              init_strategy=numpyro.infer.init_to_median,
-              dense_mass=True,
-              max_tree_depth=7)
-
-mcmc = numpyro.infer.MCMC(nuts_kernel, 
-                          num_warmup=200, 
-                          num_samples=1000,
-                          num_chains=16,
-                          chain_method='vectorized',
-                          progress_bar=False)
-
-mcmc.run(jax.random.PRNGKey(42))
-\end{lstlisting}
-
-The NUTS sampler have been running with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, and the \texttt{max\_tree\_depth} has been set to 7 by default (ie.
+% As an example the optimisation of the cosmological parameters using for instance the NUTS flavour of MCMC can be conducted as follows with \numpyro:
+% \begin{lstlisting}[language=iPython]
+% nuts_kernel = numpyro.infer.NUTS(observed_model,
+%               step_size=1e-1,
+%               init_strategy=numpyro.infer.init_to_median,
+%               dense_mass=True,
+%               max_tree_depth=7)
+
+% mcmc = numpyro.infer.MCMC(nuts_kernel, 
+%                           num_warmup=200, 
+%                           num_samples=1000,
+%                           num_chains=16,
+%                           chain_method='vectorized',
+%                           progress_bar=False)
+
+% mcmc.run(jax.random.PRNGKey(42))
+% \end{lstlisting}
+
+We have run the NUTS sampler using \texttt{numpyro.infer.NUTS} on the DES Y1 likelihood,  with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, with the \texttt{max\_tree\_depth} set to seven (ie.
 128 steps for each iteration).
 %\FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
-The "vectorized" option runs multi chains in a single GPU thanks to the JAX \textit{vmap} mechanism.
-If one dispose of several GPU devices, then the using the JAX paralellization mechanism (\textit{pmap}), it is possible to launch on each GPU device the "vectorized" sampling as done in the previous snippet, and get back all the MCMC chains.
-However, experiments have been undertaken on a single GPU either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
-The elapse time for these experiments was 20 hours.
-
-The results in terms of relative effective sample sizes (aka ``ESS'') are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
-One notices the dramatic improvement of the ESS by about a factor 10 using the NUTS sampler compared to the highly-optimized \texttt{Cobaya} Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T} which has last about 40 hours on CPU to generate 70,000 samples.
-The CL contours are also in very good agreement.
-\FrL{we should rephrase this a bit to say that Cobaya is our reference.} It is worth to point out that the mass matrix structuring operation is a very effective as it increases the sampling efficiency by about a factor 2.
-Anticipating the Section~\ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
+Using the ``vectorized'' \texttt{numpyro} option we ran all 16 chains simultaneously on a single GPU, made possible by the JAX \textit{vmap} mechanism.
+If one has several GPU devices available, then the using the JAX paralellization mechanism (\textit{pmap}), it is further possible to launch the vectorized sampling across the devices, and get back all the MCMC chains.
+However, the experiments have all been undertaken on a single GPU, either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
+The elapsed time for these experiments was 20 hours.
+
+The results in terms of relative effective sample sizes (ESS) are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
+We compare to a reference sample from the highly-optimized \texttt{Cobaya} Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T}, which is widely used in cosmology and which we ran for around 40 hours on CPU to obtain the set of contours shown.
+
+There is a dramatic improvement ofinthe ESS by about a factor of 10 using the NUTS sampler compared to Cobaya, with very good agreement between the CL contours.
+It is worth noting that the mass matrix structure described above increases the sampling efficiency by about a factor of two.
+
+The speed of the sampling could be improved: we have tested using the parameter \texttt{max\_tree\_depth=5} and found convergence in five hours, showing a linear scaling in this parameter while keeping the sampling efficiencies at a high level; the user is highly encouraged to tune this critical parameter.
+\JZ{I don't think the anticipation of the next section was needed here to make the point so I removed it, but please let me know if you're unhappy about this.
+Otherwise if you're happy please delete this comment.}
+
+% Anticipating the Section~\ref{sec-Neural-Reparametrisation} devoted to the Neural Transport which is an effective solution to validate the model with fewer samples, 
+% we have used a NUTS sampling with \texttt{max\_tree\_depth=5} those elapse time is 5 hours.
 This clearly shows a linear scaling with respect to the number of steps per iteration keeping the sampling efficiencies at a high level (see Table~\ref{tab-ESS-NUTS_SVI-1}), then the user is highly encouraged to tune this critical parameter.
@@
 %
-Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{10.5555/2567709.2502622, 8588399} and show where auto-differentiation helps.
-If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
-VI aims to get an approximation of this distribution, ie.
+
+We now explore \textit{Stochastic Variational Inference} \citep{10.5555/2567709.2502622, 8588399}, another inference algorithm enabled by auto-differentiation.
+If we  write $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to Bayes theorem we have $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ as the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
+Variational Inference (VI) aims to find an approximation to this distribution, ie.
 $p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
@@
 \end{equation}
-where the \textit{guide} in the \numpyro\ library may be a multi-variate Gaussian distribution (MVN) for instance.
+where the \textit{guide} in the \numpyro\ library (i.e. the parametrized function $q$) may be a multi-variate Gaussian distribution (MVN) for instance.
 
 Using the auto-differentation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}).
-As stated by the authors of \citep{10.5555/3122009.3122023} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
-function, taking derivatives, and running a gradient-based or coordinate-ascent optimization.
-First is defined a invertible differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$.
-The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.
-One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models.
-Then, the cost function reads
+As stated by the authors of \citep{10.5555/3122009.3122023} ADVI specifies a variational family appropriate to the model, computes the corresponding objective
+function, takes derivatives, and runs a gradient-based or coordinate-ascent optimization.
+First we define a invertible differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$, where the new $\xi$ parameters are unbounded, $xi_i in (\-infty, \infty)$ and so the subsequent minimization step can be performed with no bound constraints.
+% One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models.
+The cost function then reads
 \begin{equation}
@@
 \end{equation}
-which exhibit the Jacobian of the $T^{-1}$ transformation.
-The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{10.5555/2969239.2969303}).
-Let us illustrated the method using an invertible transformation (aka \textit{normalizing flow})  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, such that $\zeta\sim \mathcal{N}(0,I)$\footnote{Notice that $z=T^{-1}(\xi)=(T^{-1} \circ S_\lambda^{-1})(\zeta)=(S_\lambda \circ T)^{-1}(\zeta)=F_\lambda(\zeta)$.}.
-The Jacobian of this distribution is 1 by definition (volume conservation), so the lost function reads
+which includes the Jacobian of the $T^{-1}$ transformation.
+The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citealt{10.5555/2969239.2969303}).
+Let us illustrate the method using an invertible transformation  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, where $\zeta\sim \mathcal{N}(0,I)$\footnote{Notice that $z=T^{-1}(\xi)=(T^{-1} \circ S_\lambda^{-1})(\zeta)=(S_\lambda \circ T)^{-1}(\zeta)=F_\lambda(\zeta)$.}.
+The Jacobian of this distribution is 1 by definition (volume conservation), so the loss function reads
 \begin{multline}
@@
 \end{figure*}
-The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as followed (See the notebook \nblink{DES_Y1_SVI_and_NeutraReparam}):
+The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as follows (See the notebook \nblink{DES_Y1_SVI_and_NeutraReparam}):
 \begin{lstlisting}[language=iPython]

### 2022-12-02 paper_OJA.tex
Some review by Joe & JE

@@
 We discuss algorithms made possible by this library, and present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical usage.
-
 We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that the Normalizing Flows using the Neural Transport is a promising methodology.
@@
 \end{lstlisting}
-% done by JEC 13Sept22 \JZ{dmu in the above is not defined anywhere}
 
@@
 Although very simple, this case study exhibits a significant challenge of finite difference methods for computing the Fisher matrix as has been shown for instance in a more advanced case study in \citet{2021arXiv210100298B}.
-% done by JEC 13Sept22 \JZ{Should probably latex the step sizes $10^{-1}, 10^{-2}, 10^{-6}$ in the figure legend}
+
 \begin{figure}
@@
 Using the \numpyro\ PPL, we then set up a forward model following the DES Y1 Pipeline \citep{2018PhRvD..98d3526A}.
+Here we show some key elements of the implementation; the details and inference examples may be found in the following notebook \nblink{VectorizedNumPyro_block}.
 The model parameters and their prior distributions are shown in Table \ref{tab-DESY1}.
@@
 $\ell$) a series of 50 angular modes.
-Our mock data (\texttt{cl\_obs}) comes from this model evaluated at a fiducial cosmology, with \numpyro\ dealing with the random number generation that generates added noise:
+After encapsulating the code above with input sampled parameters (using \numpyro\ distribution classes) in a function \texttt{model}, we can generate our mock data (\texttt{cl\_obs}).
+This comes comes from this model function evaluated at a fiducial cosmology, with \numpyro\ dealing with the random number generation that generates added noise:
 
@@
 \end{lstlisting}
-\JZ{We do not describe the model function above here. I guess it just adds noise to the cl?}
 
@@
 How to use this model to perform inference is described in the sections \ref{sec-NUTS} and \ref{sec-SVI}.
+%For a complete example of this functionality see the attached notebook: \nblink{VectorizedNumPyro_block}.
 %
@@
 The speed of the sampling could be improved: we have tested using the parameter \texttt{max\_tree\_depth=5} and found convergence in five hours, showing a linear scaling in this parameter while keeping the sampling efficiencies at a high level; the user is highly encouraged to tune this critical parameter.
-\JZ{I don't think the anticipation of the next section was needed here to make the point so I removed it, but please let me know if you're unhappy about this.
+
+%\JZ{I don't think the anticipation of the next section was needed here to make the point so I removed it, but please let me know if you're unhappy about this.
 Otherwise if you're happy please delete this comment.}
@@
 \end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the $q$ distribution, so its gradient can be computed once for all for a given $q$ distribution family and reused in any user model.  
+where $\mathbb{H}(q)\equiv \mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]$ is the Shannon entropy of the $q$ distribution; its gradient can be computed once for all for a given $q$ distribution family and reused in any user model.
 Then,
@@
 \end{multline}
-An concrete example of the gradient formula is given in Appendix \ref{app-SVI_MVN} with the multivariate Gaussian distribution family used as the guide.
+%JEC: I think now it is not usefull. A concrete example of the gradient formula is given in Appendix \ref{app-SVI_MVN} with the multivariate Gaussian distribution family used as the guide.
+
 %
@@
 %\FrL{we don't have plots of just the VI?}\JEC{Good point: I have SVI alone with both MVN or Block Neural Autoregressive Flow. See new figure.}
-
-Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{sec-DESY1}) where the \texttt{model} is described and the \texttt{cl\_obs} are the observed $C_\ell$.}
-\begin{lstlisting}[language=iPython]
-import numpyro.infer.autoguide as autoguide
-from numpyro.infer import SVI, Trace_ELBO
-from numpyro.optim import Adam
-# choose a SVI guide 
-guide = autoguide.AutoMultivariateNormal(model,
-        init_loc_fn=numpyro.infer.init_to_median())
-# SVI optimisation 
-optimizer = numpyro.optim.Adam(1e-3)
-svi = SVI(model_spl, guide,optimizer,loss=Trace_ELBO(num_particles=10))
-svi_result = svi.run(jax.random.PRNGKey(0),20_000, cl_obs)
-#To get samples from SVI approximated posterior 
-samples = guide.sample_posterior(jax.random.PRNGKey(1), svi_result.params, sample_shape=(100_000,))
-\end{lstlisting}
-%
-Once the optimisation is done on gets $z$ i.i.d.
-samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
+% \JZ{I removed a code snippet here since it was really about numpyro instead of jax-cosmo and felt slightly out of place. Uncomment it if you prefer!}
+An implementation example using the \numpyro\ library may be found in this companion notebook: \nblink{DES_Y1_SVI_and_NeutraReparam}.
+% Here is the simple way \numpyro\ deals with SVI initialisation and optimisation\footnote{See also listing in Appendix \ref{sec-DESY1}) where the \texttt{model} is described and the \texttt{cl\_obs} are the observed $C_\ell$.}
+% \begin{lstlisting}[language=iPython]
+% import numpyro.infer.autoguide as autoguide
+% from numpyro.infer import SVI, Trace_ELBO
+% from numpyro.optim import Adam
+% # choose a SVI guide 
+% guide = autoguide.AutoMultivariateNormal(model,
+%         init_loc_fn=numpyro.infer.init_to_median())
+% # SVI optimisation 
+% optimizer = numpyro.optim.Adam(1e-3)
+% svi = SVI(model_spl, guide,optimizer,loss=Trace_ELBO(num_particles=10))
+% svi_result = svi.run(jax.random.PRNGKey(0),20_000, cl_obs)
+% #To get samples from SVI approximated posterior 
+% samples = guide.sample_posterior(jax.random.PRNGKey(1), svi_result.params, sample_shape=(100_000,))
+% \end{lstlisting}
+%
+Once the optimisation is done one can obtain  i.i.d.
+$z$ samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
 Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
-The reader can have a glance to the following notebook \nblink{DES_Y1_SVI_and_NeutraReparam}.
-Concerning the B-NAF architecture it is composed of  a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
+The B-NAF architecture is composed of a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
 The SVI optimization has been performed with the Adam optimizer \citep{KingmaB14} and a learning rate set to $10^{-3}$.
@@
 It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps.
-\FrL{Arf...
-2 hours !}
-
-In the figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and B-NAF guided SVI.
-As mentioned in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution.
-It is not in the scope of this article to elaborate a statistical diagnostics, rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling.
-The difference is that these SVI approximate posteriors have been obtained in a short time and these posterior estimates can serve as front-end to a NUTS sampler as described in the next section.
+
+In figure \ref{fig_cobaya_SVIs}, we compare the contours obtained with \texttt{Cobaya} (as in figure \ref{fig_cobaya_NUTS_SVI}) and those obtained with the MVN and B-NAF guided SVI.
+As noted in \cite{NEURIPS2020_7cac11e2} one challenge with variational inference is assessing how close the variational approximation $q(z,\lambda^\ast)$ is to the true posterior distribution.
+It is not in the scope of this article to elaborate a statistical diagnosis; rather we show that both guided SVI exhibit rather similar contours and both estimate are close to the Cobaya posterior sampling.
+The difference is that these SVI approximate posteriors have been obtained in a much shorter time, and can serve as starting point for a NUTS sampler as described in the next section.
 %
@@
 If the SVI method can be used as is to get $z$ i.i.d.
-samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{Parno2018,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
-From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ (eg.
+samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{Parno2018,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution with unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
+From SVI, one obtains a first approximation of the target distribution, and this approximation is used to choose a better transform $T$ to map the parameter space to a more convenient one $z=F_\lambda(\zeta)$ (eg.
 $F_\lambda=T^{-1}\circ S^{-1}_\lambda$) is such that
@@
 \end{equation}
-which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO.
+where $F_\lambda$ with the optimal $\lambda^\ast$ maps the best-fitting $q(z;\lambda^\ast)$ to a geometrically simple function like a unit multivariate normal distribution.
 So, one can use a HMC sampler (eg.
 NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$.
-Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples.
+Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Markov Chain with $(z_i)_{i<N}$ samples.
+% We provide a notebook illustrating this process: \nblink{DES_Y1_SVI_and_NeutraReparam}.
 \begin{figure*}
@@
 \end{figure*}
-The \textit{Neural Transport} is accessible throw the  \verb|NeuTraReparam| in \numpyro\ library as follows (See the notebook \nblink{DES_Y1_SVI_and_NeutraReparam}):
-\begin{lstlisting}[language=iPython]
-from numpyro.infer.reparam import NeuTraReparam
-# Reparametrisation of model to feed NUTS sampler
-neutra = NeuTraReparam(guide, svi_result.params)
-neutra_model = neutra.reparam(model_spl) 
-\end{lstlisting}
-%
-We have used the MVN guided SVI described in the previous section.
-The NUTS sampler has run 1 chain of 200 samples (``low'' schema), 1 chain with 1000 samples and 10 chains  with 1000 samples each (each setup starts all chains with 200 samples for initialisation).
-All NUTS sampling as been performed with dense mass matrix optimisation without special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
-The elapse time for each of the 3 setups is 50 minutes, 150 minutes and 5 hours, respectively.
+
+% \JZ{Deleted code snippet again, but please reinstate if you prefer}.
+% The core process is straightforwardly implemented in \numpyro\:
+% \begin{lstlisting}[language=iPython]
+% from numpyro.infer.reparam import NeuTraReparam
+% # Reparametrisation of model to feed NUTS sampler
+% neutra = NeuTraReparam(guide, svi_result.params)
+% neutra_model = neutra.reparam(model_spl) 
+% \end{lstlisting}
+%
+We have used the MVN guided SVI described in the previous section on the same DES-Y1 analysis described above.
+The NUTS sampler was run using 1 chain of 200 samples (a fast configuration), one chain with 1000 samples and a set of ten chains with 1000 samples each combined into a single run of 10,000 samples (each setup began all chains with 200 samples for initialisation).
+All NUTS sampling was performed with dense mass matrix optimisation without the special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
+The elapsed time for each of the 3 setups was 50 minutes, 150 minutes and 5 hours, respectively.
 Naturally, more samples lead to better contour precision.
-But what is illustrative is the ``low'' schema results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}  to be compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
-Notice the good results even for the highly correlated lens bias parameters.
+But what is illustrative is the fast configuration results as shown in the Figures \ref{fig_cobaya_NUTS_SVI} and \ref{fig_cobaya_NUTS_SVI_bis}, compared to \texttt{Cobaya} and NUTS results presented in section \ref{sec-NUTS}.
+The results are good even for the highly correlated lens bias parameters.
 It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a rather poor behaviour with only 200 samples.
 
-Results in terms of sampling efficiency are shown on Table~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency at low number of samples which is quite interesting to ease the model development in its early phase.
+Results in terms of sampling efficiency are shown in Table~\ref{tab-ESS-NUTS_SVI-1}.
+SVI followed by Neutral Transport gives high efficiency at low number of samples which may be particularly useful during early phases of model development.
 %
@@
 \label{sec-results}
-%
-To a large extend looking at results of Table~\ref{tab-ESS-NUTS_SVI-1}, the key question is: what is the HMC/NUTS gain compared to the well known highly optimized \texttt{Cobaya} sampler? 
-In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
-\begin{equation}
-    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
-\end{equation}
-with $N_s$ the total number of samples, $\varepsilon$ the effective sampler efficiency  and $n_{steps}$ the number of steps (calls) per sample generation.
-Then, for \texttt{Cobaya} one roughly gets $\varepsilon\approx 3\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more.
-At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie.
-about $1\%$).
-So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.
-
-On the other hand, the above metric is may be too crude to cover all aspects of the sample generation.
-One should probably have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
-several days on dedicated infrastructure.
-Moreover, the user is faced to the dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, Owen2017}.
-
-So, the authors think that it is worth considering methods like HMC/NUTS that by construction are more efficient.
-Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of the model with rather modest sample set (eg.
-starting with 200 samples) producing good enough marginal contours in few hours.
-In practice, this validation phase can save time before producing sizeable batch for final analysis.
-The authors have not investigated very high dimensional ($O(10^2)$ parameters) as well as multi-modal analysis, but the key argument in favour of HMC/NUTS sampling is that it exploit the geometry of the typical set of the posterior distribution contrary to the standard random walk of Metropolis-Hasting sampling.
-Furthermore, using reparametrisation  one can adapt to poor geometry cases (eg.
-\cite{2019arXiv190303704H}).
-%
 \begin{table}[htb]
-\caption{Results of the \textit{relative effective sample size} (aka ESS, ``bulk'') in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
-The ESS can be larger than 100\% in some cases.
+\caption{The \textit{relative effective sample size} (ESS) in percent computed by the \texttt{Arviz} library \citep{arviz_2019} from: 
 %{\color{red} 
@@
 %}
-(a): results from Cobaya 70,000 samples;
-(b): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Section~\ref{sec-NUTS}); 
-(c) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 1 chain of 200 samples and 200 warm-up (Section~\ref{sec-Neural-Reparametrisation}).}
+(a): Cobaya 70,000 samples;
+(b): NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Section~\ref{sec-NUTS}); 
+(c) SVI Multivariate Normal followed by NUTS and Neural Transform  with one chain of 200 samples and 200 warm-up.
+The ESS can be larger than 100\% in some cases.
+(Section~\ref{sec-Neural-Reparametrisation}).}
 \label{tab-ESS-NUTS_SVI-1}
@@
 %
+Looking at the results of Table~\ref{tab-ESS-NUTS_SVI-1}, a key question is: what is the HMC/NUTS gain compared to the highly optimized \texttt{Cobaya} sampler? 
+One useful metric is the number of effective samples per model evaluation:
+\begin{equation}
+    \eta = \frac{n_\mathrm{eff}}{n_\mathrm{eval}} = \frac{N_s \times \varepsilon}{N_s \times n_\mathrm{step}} = \frac{\varepsilon}{n_\mathrm{step}}
+\end{equation}
+with $N_s$ the total number of samples, $\varepsilon$ the effective sampler efficiency  and $n_\mathrm{step}$ the number of steps (calls) per generated sample.
+For \texttt{Cobaya} we find $\varepsilon\approx 3\%$ with $n_\mathrm{step}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_\mathrm{step}=2^5$ or more.
+With better tuning of the sampling parameters we would expect the $\eta$ values for both methods would to become more comparable, but at this intermediate dimensionality the gain from HMC/NUTS compared to standard MCMC sampler.
+The power of these approaches will become most evident at higher dimensionality still, such has when marginalizing over increasingly complex systematic models.
+NUTS also makes post-processing simpler, since samples are nearly uncorrelated, removing the need for a \textit{thinning} step, which is a rather delicate procedure, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, Owen2017}.
+
+% The $\eta$ metric is may be too crude, though, to cover all aspects of the sample generation.
+\JZ{I don't understand the next point.
+Doesn't the HMC/NUTS take a long time too on dedicated resources?} One should probably have in mind that the low sampling efficiency of a standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
+several days on dedicated infrastructure.
+%\JZ{I think you need to remove the burn-in when using NUTS too?}.
+
+As the dimensionality of cosmological models increases, methods like HMC/NUTS that by construction are more efficient will become increasingly important.
+Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of a model with rather modest sample set (eg.
+starting with 200 samples) producing good enough marginal contours in few hours.
+In practice, this validation phase can save time before producing sizeable batch for final analysis.
+The authors have not investigated higher dimensional ($O(10^2)$ parameters) or multi-modal problems, but the key argument in favour of HMC/NUTS sampling is that it exploits the geometry of the typical set of the posterior distribution automatically, unlike the standard random walk of Metropolis-Hasting sampling.
+Furthermore, using reparametrisation  one can adapt to poor geometry cases (eg.
+\cite{2019arXiv190303704H}).
+%
+
+%
 \section{General discussion}
@@
 %
-Having demonstrated the interest of \jaxcosmo\ as a differentiable cosmology library library, we now discuss several limitations, and open questions.
+Having demonstrated the utility of \jaxcosmo\ as a differentiable cosmology library library, we now discuss several limitations, and open questions.
 
 %Lack of autodiff Boltzmann code (emulator)
-The first essential barrier to a full-fledged automatically differentiable cosmology library is the need for a differentiable Boltzmann solver to compute the matter power spectrum.
-At this stage, \jaxcosmo\ relies on the analytic Einsenstein \& Hu fitting formula, which is not accurate enough for Stage IV requirements, nor does it allow to explore beyond $\Lambda$CDM models.
-Existing solvers such as CLASS or CAMB \FrL{citations} are large and complex codes which are not easily reimplemented in an \autodiff\ framework and therefore cannot be directly integrated in \jaxcosmo\ .
-
-A first option to resolve this issue is to implement from scratch a new Boltzmann code in a framework that will support automatic differentiation.
-This is the approach behind works such as \FrL{the Julia code} which is provides a simplified Boltzmann solver in Julia, or \FrL{the code from our friends in Zurich} which is based on a SymPy and could be relatively easily compatible with JAX.
+The first essential barrier to a fully-fledged automatically differentiable cosmology library is the need for a differentiable Boltzmann solver to compute the CMB or matter power spectra.
+At this stage, \jaxcosmo\ relies on the analytic Einsenstein \& Hu fitting formula for the latter, which is not accurate enough for Stage IV requirements, and it does not include models beyond $\Lambda$CDM.
+Existing solvers such as CLASS or CAMB \citep{2011JCAP...07..034B,camb} are large and complex codes which are not easily reimplemented in an \autodiff\ framework and therefore cannot be directly integrated in \jaxcosmo\ .
+
+A first option to resolve this issue would be to implement from scratch a new Boltzmann code in a framework that supports automatic differentiation.
+This is the approach behind works such as \texttt{Bolt.jl}\footnote{\url{https://github.com/xzackli/Bolt.jl}} which is provides a simplified Boltzmann solver in Julia, or PyCosmo \citep{pycosmo} which is based on a the SymPy symbolic mathematics library and could be relatively easily compatible with JAX.
 However, even if very promising, both of these options thus far remain limited.
@@
 
-A second approach is to develop emulators of a full-fledged Boltzmann code.
+A second approach would be to develop emulators of a fully-fledged Boltzmann code.
 Emulators based on neural networks or Gaussian processes are themselves automatically differentiable with respect to cosmological parameters.
-In fact, the litterature is now rich in examples of such emulators \FrL{cite all emulators}.
-After validating their accuracy against the reference CAMB or CLASS implementation, they can be directly integrated as a plug-and-play replacement for the computation of the matter power spectrum.
-\FrL{need to talk about cosmopower somewhere.} At this time, it seems that using emulators will be the most straightforward approach to bring more accurate models to \jaxcosmo\ .
-\FrL{However, we want to emphasize that one of the reason for the wide diversity of emulators is a lack of standardization}.
+In fact, the literature is now rich in examples of such emulators \citep[e.g.][and references therein]{emucmb, nygaard,cosmopower,cosmicnet}.
+After validating their accuracy against a reference CAMB or CLASS implementation, they could be directly integrated as a plug-and-play replacement for the computation of the matter power spectrum.
+At this time, it seems that using emulators will be the most straightforward approach to bring more accurate models to \jaxcosmo.
+We believe, though,  that one of the reason for the wide diversity in this is a lack of standardization - a unified interface and validation suite for such methods would provide a much simpler comparison between them and enable wider usage.
+\JZ{Francois - is this what you meant here?
+If not please edit.
+If Francois hasn't done this soon then just delete this comment.}
 
@@
 While the model for the linear matter power spectrum is typically kept fixed in practical analyses, all the of the choices related in particular to systematics modeling (e.g.
-photometric redshift errors, galaxy bias, or intrinsic galaxy alignments) are very likely to vary significantly in the process of developing the analysis.
-Building an emulator for the likelihood would mean the emulator needs to be retrained everytime the likelihood is changed.
+photometric redshift errors, galaxy bias, or intrinsic galaxy alignments) will vary significantly in the process of developing the analysis.
+Building an emulator for the likelihood would mean the emulator needs to be retrained every time the likelihood is changed.
 
@@
 %  - Efficiency of gradient-based inference methods
-Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo\ .
-As we illustrated in the previous section, gradient-based inference techniques yield significantly less correlated MCMC chains.\FrL{Hummm here I guess we want to say also it scales better with dimensions, and that you can get very fast first results.} \FrL{we also want to say that the GPU parallelisation is going to change the game, all the more when you account for new massive GPU machines.}
+Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo.
+As we illustrated in the previous section, gradient-based inference techniques yield significantly less correlated MCMC chains, scale better than any other known sampling method as dimension increases, and can provide very fast approximate posteriors if needed.
+\jaxcosmo\  is also well suited to aid in the parallelization of likelihoods and algorithms, especially on multiple GPUs, which will become increasingly important as the high-performance computing landscape evolves.
 
@@
 Finally, one question worth asking is how does \jaxcosmo\ position itself against classical codes such as CCL?
-Or in other terms, will it replace CCL?
-While we are convinced of the benefits of a JAX implementation, we expect CCL to remain extremely valuable as the reference cosmology implementation.
-\FrL{it allows to leverage the legacy codes, has a lot more features, etc.} Ultimately, a natural transition may occur towards more modern frameworks like \jaxcosmo\ when they reach the ability to run full-fledged Stage IV likelihoods.
+While we are convinced of the benefits of a JAX implementation, we expect CCL and other key codes to remain critical as standard cosmology implementations.
+Ultimately, a natural transition may occur towards differentiable frameworks like \jaxcosmo\ when they reach the ability to run full-fledged Stage IV likelihoods.
 
@@
 
-In this paper, we have presented \jaxcosmo\ a cosmology library implemented in the JAX framework, enabling the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, as well as greatly speeding up likelihood evaluations thanks to automatic parallelisation and just in time compilation on GPUs.
-Currently, \jaxcosmo\ only contains a small set of features limited corresponding to a DES-Y1 3x2pt analysis.
-Being an open source project, contribution of additional features is additional areas such as CMB or spectroscopic galaxy clustering are warmly welcome.
-
-To demonstrate the interest of an automatically differentiable library, we have illustrated on concrete examples how Fisher matrices, which are notoriously difficult to compute and require extensive and careful fine tuning, can now be computed robustly and at much lower cost.
+In this paper, we have presented \jaxcosmo, a cosmology library implemented in the JAX framework that enables the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, and greatly speeds up likelihood evaluations thanks to automatic parallelisation and just-in-time compilation on GPUs.
+Currently, \jaxcosmo\ only contains a small set of features corresponding to a DES-Y1 3x2pt analysis.
+Being an open source project, contributions of additional features is additional areas such as CMB or spectroscopic galaxy clustering are warmly welcome.
+
+To demonstrate the value of an automatically differentiable library, we have illustrated with concrete examples how Fisher matrices, which are notoriously unstable and require extensive and careful fine tuning, can now be computed robustly and at much lower cost.
 In addition, Fisher matrices becomes themselves differentiable, which allows for  Figure of Merit optimization by gradient descent, making survey optimization extremely simple.
 
-Going beyond Fisher forecasts, we have also compared to simple Metropolis-Hastings several gradient-based inference techniques (namely, Hamiltonian Monte-Carlo, No-U-Turn-Sampler, Stochastic Variational Inference).
-We show that the posterior samples with gradient-based methods have a much larger Effective Sample Size, meaning that \FrL{let's think a bit more about interpretation here....
-the thing I care about is not as much the ESS, but the number of likelihood evaluation per independent sample.
-I don't think we have quantified that too much}.
+Going beyond Fisher forecasts, we have also compared simple Metropolis-Hastings to several gradient-based inference techniques (Hamiltonian Monte-Carlo, No-U-Turn-Sampler, and Stochastic Variational Inference).
+We show that the posterior samples with gradient-based methods can reproduce classic methods very efficiently, and can provide approximate posteriors very rapidly.
 These inference techniques can scale to hundreds of dimensions, and may become necessary in Stage IV analysis as the number of nuisance parameters is likely to become very large.
 
-\bigskip
-
-The next extensions of this framework will be the inclusion of additional cosmological probes, as well as integrating emulators for the matter power spectrum trained on CAMB or CLASS, as a means to go beyond the current analytic Eisenstein \& Hu model.
-
-\bigskip
+The next extensions of this framework will be the inclusion of additional cosmological probes, as well as the integration of emulators for the matter power spectrum trained on CAMB or CLASS, as a means to go beyond the current analytic Eisenstein \& Hu model.
 
@@
 
-%\FrL{add link}\JEC{Verif if all the nb listed are there, and if some have to be withdrawn. }
-
 \url{https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/}
 
-%
-%
-%
-%\subsection{Efficient sampling in high dimensions with Hamiltonian Monte Carlo}
-%
-%%
-%%
-%\subsection{DES-Y1 Example}
-%
-
 %%%
-\section*{CRediT authorship contribution statement}
-\textbf{J.E Campagne} Conceptualization, Methodology, Software, Validation, Writing.
-\textbf{F. Lanusse}, \textbf{J. Zuntz}...
+\section*{CRedit authorship contribution statement}
+\textbf{J.E Campagne:} Conceptualization, Methodology, Software, Validation, Writing, Visualization.
+\textbf{F. Lanusse:}, Conceptualization, Methodology, Software, Validation, Writing, Project administration \textbf{J. Zuntz:} Investigation, Writing.
 
@@
 %% appendix sections are then done as normal sections
-\appendix
-
-\section{SVI example with MVN guide}
-\label{app-SVI_MVN}
-Let us illustrate the \textit{elliptical standardisation} described in section \ref{sec-SVI} in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie.
-a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
-One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg.
-$\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$.
-The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
-\begin{multline}
--\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
-\label{eq-loss-svi-3b}
-\end{multline}
-where $\mathbb{H}(q)$ is the Shannon entropy of the $q$ distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model.
-In case of the MVN for completeness  $\nabla_\mu \mathbb{H}(q)=0$, $\nabla_L \mathbb{H}(q)=(L^{-1})^T$ while the other gradients are computed by the JAX \texttt{autodiff} mechanism.
+%\appendix
+% JEC no more useful
+%\section{SVI example with MVN guide}
+%\label{app-SVI_MVN}
+%Let us illustrate the \textit{elliptical standardisation} described in section \ref{sec-SVI} in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (i.e.
+an MVN distribution with the parameters $\lambda=(\mu, \Sigma)$).
+The  function $S_\lambda$ which transforms the parameters as $S_\lambda(\xi)=\zeta$ is given here by  $S_\lambda(\xi) = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$.
+The Jacobian of this distribution is 1 (by volume conservation), so the cost function reads
+%\begin{multline}
+%-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
+%\label{eq-loss-svi-3b}
+%\end{multline}
+
+%In case of the MVN  $\nabla_\mu \mathbb{H}(q)=0$, $\nabla_L \mathbb{H}(q)=(L^{-1})^T$ while the other gradients are computed by the JAX \texttt{autodiff} mechanism.
 %\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.}
 
-The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI.
+%The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution, hence the ``S'' of SVI.
 As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%
 %% If you have bibdatabase file and want bibtex to generate the

### 2022-12-05 paper_OJA.tex
Joe abstract revision

@@
 \begin{abstract}
-The recent JAX library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems.
+
+We present \jaxcosmo, a library for differentiable cosmological theory calculations.
+\jaxcosmo\ uses the recent JAX library, which has created a new coding ecosystem, especially in probabilistic programming.
+%
+As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), JAX exposes an \textit{automatic differentiation} (AD) mechanism.
+AD can take a Python function that uses JAX and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn.
+This gives us the opportunity to apply a range of powerful Bayesian inference algorithms, otherwise impractical in cosmology, such as Hamiltonian Monte Carlo and Variational Inference.
+%
+\jaxcosmo\ implements background evolution, linear and non-linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes, all differentiable with respect to the cosmological parameters and their other inputs.
+%
+We discuss algorithms made possible by this library, and present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical inference.
+%
+We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that Normalizing Flows using the Neural Transport are a promising methodology.
+
+% The recent JAX library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems.
 In particular, JAX exposes an \textit{automatic differentiation} mechanism, which can take a python function that uses JAX and return its (vector) derivative.
@@
 
-To take advantage of these possibilities within cosmology we have developed the \jaxcosmo\ library, which implements background evolution, linear and non linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes.
+% To take advantage of these possibilities within cosmology we have developed the \jaxcosmo\ library, which implements background evolution, linear and non linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes.
 We discuss algorithms made possible by this library, and present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical usage.
-We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that the Normalizing Flows using the Neural Transport is a promising methodology.
+% We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that the Normalizing Flows using the Neural Transport is a promising methodology.
 %{\color{red}

### 2022-12-06 paper_OJA.tex
some ref

@@
 A new function corresponding to the derivative of the original function is therefore built automatically for the user.
-We direct the interested reader to \FrL{a good reference} for an in-depth introduction to automatic differentiation.
+We direct the interested reader to \FrL{a good reference} \cite{Margossian2019} for an in-depth introduction to automatic differentiation.
 
@@
 Emulators based on neural networks or Gaussian processes are themselves automatically differentiable with respect to cosmological parameters.
-In fact, the literature is now rich in examples of such emulators \citep[e.g.][and references therein]{emucmb, nygaard,cosmopower,cosmicnet}.
+In fact, the literature is now rich in examples of such emulators \citep[e.g.][and references therein]{Gunther_2022, nygaard,cosmopower,cosmicnet, emucmb}.
 After validating their accuracy against a reference CAMB or CLASS implementation, they could be directly integrated as a plug-and-play replacement for the computation of the matter power spectrum.

### 2022-12-20 paper_OJA.tex
Updating main text

@@
 We present \jaxcosmo, a library for differentiable cosmological theory calculations.
-\jaxcosmo\ uses the recent JAX library, which has created a new coding ecosystem, especially in probabilistic programming.
-%
-As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), JAX exposes an \textit{automatic differentiation} (AD) mechanism.
-AD can take a Python function that uses JAX and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn.
+\jaxcosmo\ uses the recent \texttt{JAX} library, which has created a new coding ecosystem, especially in probabilistic programming.
+%
+As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), \texttt{JAX} exposes an \textit{automatic differentiation} (AD) mechanism.
+AD can take a Python function that uses \texttt{JAX} and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn.
 This gives us the opportunity to apply a range of powerful Bayesian inference algorithms, otherwise impractical in cosmology, such as Hamiltonian Monte Carlo and Variational Inference.
@@
 
-% The recent JAX library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems.
-In particular, JAX exposes an \textit{automatic differentiation} mechanism, which can take a python function that uses JAX and return its (vector) derivative.
+% The recent \texttt{JAX} library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems.
+In particular, \texttt{JAX} exposes an \textit{automatic differentiation} mechanism, which can take a python function that uses \texttt{JAX} and return its (vector) derivative.
 This gives us the opportunity to apply a range of powerful but otherwise unfeasible algorithms used in Bayesian inference, such as Hamiltonian Monte Carlo (HMC) and Variational Inference.
@@
 \texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro\ \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper.
-Other PPL packages such as \texttt{PyMC} \citep{Salvatier2016} have switched to JAX backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}.
+Other PPL packages such as \texttt{PyMC} \citep{Salvatier2016} have switched to \texttt{JAX} backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}.
 
@@
 In this context we have developed the open source \jaxcosmo\ library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
-The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable JAX functions.
+The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable \texttt{JAX} functions.
 
@@
 
-\section{JAX: Autograd meets XLA}
+\section{JAX: GPU Accelerated and Automatically Differentiable Python Programming}
 \label{sec-primer}
 
-The aim of this session is to provide a brief technical primer on JAX, necessary to fully grasp the potential of a cosmology library implemented in this framework.
+The aim of this section is to provide a brief technical primer on \texttt{JAX}, necessary to fully grasp the potential of a cosmology library implemented in this framework.
 
 % Maybe a short history explaining what autograd and XLA are.
-\FrL{Add a short history of XLA and Autograd.}
-
-\FrL{Is this too much information below?
-On the one hand I don't want to go too technical, on the other hand, we talk about stuff like jax.grad/hessian, JIT, and vmap in the rest of the paper, it feels like these concepts should be introduced somewhere first.}
-
-\paragraph{Automatic Differentiation} Traditionally, two different approaches have been used in cosmology to obtain derivatives of given numerical expressions.
-
-The first approach is to derive analytically the formula for the derivatives of interest \citep{2019A&A...625A..64J}%JEC: This reference should be validated by F.L
-, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}. This is however only practical typically for simple analytical models.
-
+%\FrL{Add a short history of XLA and Autograd.} Or maybe not
+
+\paragraph{\textbf{Automatic Differentiation}} Traditionally, two different approaches have been used in cosmology to obtain derivatives of given numerical expressions.
+The first approach is to derive analytically the formula for the derivatives of interest \citep[e.g.][]{2013MNRAS.432..894J}, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}.
+This is however only practical typically for simple analytical models.
 The second approach is to compute numerical derivatives by finite differences.
-This approach can be applied on black-box models of abitrary complexity (from Boltzmann codes \FrL{some citations} to cosmological simulations \FrL{some citations}).
-However it is notoriously difficult to obtain stable derivatives by finite differences \FrL{cite papers trying to do fisher}.
-In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them completely impractical whenever derivatives are needed on the fly.\FrL{this last sentence is not great.}
-
-Automatic differentiation frameworks like JAX take a different approach.
+This approach can be applied on black-box models of arbitrary complexity (from typical Boltzmann codes to cosmological simulations \citep{2020ApJS..250....2V}).
+However it is notoriously difficult to obtain stable derivatives by finite differences \citep[e.g.][]{2021arXiv210100298B, 2021A&A...649A..52Y}.
+In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them impractical whenever derivatives are needed as part of outer iterative algorithm.
+
+Automatic differentiation frameworks like \texttt{JAX} take a different approach.
 They trace the execution of a given model and decompose this trace into fundamental operations with known derivatives (e.g.
@@
 A new function corresponding to the derivative of the original function is therefore built automatically for the user.
-We direct the interested reader to \FrL{a good reference} \cite{Margossian2019} for an in-depth introduction to automatic differentiation.
-
-JAX provides in particular a number of operators (\texttt{jax.grad}, \texttt{jax.jacobian}, \texttt{jax.hessian}) which can compute derivatives of any function written in JAX.
+We direct the interested reader to \cite{baydin2018automatic, Margossian2019} for in-depth introductions to automatic differentiation.
+
+JAX provides in particular a number of operators (\texttt{jax.grad}, \texttt{jax.jacobian}, \texttt{jax.hessian}) which can compute derivatives of any function written in \texttt{JAX}.
 \begin{lstlisting}[language=iPython]
@@
 
-\paragraph{Just In Time Compilation (JIT)} Despite its convenience and wide adoption in astrophysics, Python still suffers from slow execution times compared to fully compiled languages such as C/C++.
+\paragraph{\textbf{Just In Time Compilation (JIT)}} Despite its convenience and wide adoption in astrophysics, Python still suffers from slow execution times compared to fully compiled languages such as C/C++.
 One approach to mitigate these issues and make Python code fast is Just In Time compilation, which traces the execution of a given Python function the first time it is called, and compiles it into a fast executable (by-passing the Python interpreter), which can be transparently used in subsequent calls to this function.
-Compared to other strategies for speeding up Python code \FrL{cpython and such}, JIT allows the user to simply write plain Python code, and reap \FrL{spell check} the benefits of compiled code.
-
-A number of libraries allowing for JIT have already been used in astrophysics, in particular Numba (ngmix), and \FrL{The JIT code developped by our friends in Zurich}.
-JAX stands out compared to these other frameworks in that it relies on the highly optimized XLA library \FrL{citation} for executing the compiled expressions.
-XLA is continuously developed by Google as part of the TensorFlow project, for efficient training and inference of large scale deep learning applications, and as such supports computational backends such as GPU an TPU clusters.
-\FrL{Maybe we want to say why GPU is such a good idea for our use cases.}
-
-In JAX, jitting is achieving simply by transforming a function with \texttt{jax.jit} operation:
+Compared to other strategies for speeding up Python code such as Cython, JIT allows the user to simply write plain Python code, and reap the benefits of compiled code.
+
+A number of libraries allowing for JIT have already been used in astrophysics, in particular Numba\footnote{\url{https://numba.pydata.org/}}, or the HOPE library \cite{2015A&C....10....1A} developed specifically for the need of astrophysics.
+\texttt{JAX} stands out compared to these other frameworks in that it relies on the highly optimized XLA library\footnote{\url{https://www.tensorflow.org/xla}} for executing the compiled expressions.
+XLA is continuously developed by Google as part of the TensorFlow project, for efficient training and inference of large scale deep learning applications, and as such supports computational backends such as GPU an Tensor Processing Units (TPU) clusters.
+The ability to perform computations directly on GPUs through XLA is one of the major benefits of \texttt{JAX}, as speed-ups of at least two orders of magnitudes can be expected for typical parallel linear algebra computations compared to CPU.
+
+In \texttt{JAX}, jitting is achieved simply by transforming a function with \texttt{jax.jit} operation:
 \begin{lstlisting}[language=iPython]
@@
 
-\paragraph{Automatic Vectorization} Another extremely powerful feature of JAX is its ability to automatically vectorize or paralellize any function.
-Through the same tracing mechanism used for automatic differentiation, JAX can decompose a given computation into fundamental operations and add a new \textit{batch} dimension so that the computation can be applied on a batch of inputs as opposed to individual inputs.
+\paragraph{\textbf{Automatic Vectorization}} Another extremely powerful feature of \texttt{JAX} is its ability to automatically vectorize or paralellize any function.
+Through the same tracing mechanism used for automatic differentiation, \texttt{JAX} can decompose a given computation into fundamental operations and add a new \textit{batch} dimension so that the computation can be applied on a batch of inputs as opposed to individual inputs.
 In doing so, note that the computation will not run sequentially over all entries of the batch, but trully in parallel making full use of the intrinsic parallel architecture of modern GPUs.
 
-In JAX automatic vectorization is achieved using the \textit{jax.vmap} operation:
+In \texttt{JAX} automatic vectorization is achieved using the \textit{jax.vmap} operation:
 \begin{lstlisting}[language=iPython]
@@
 When considering multi-devices use-case (eg.
-GPU or TPU), JAX provides \texttt{pmap} which compile and execute in parallel replicates of the same code on each device.
+GPU or TPU), \texttt{JAX} provides \texttt{pmap} which compile and execute in parallel replicates of the same code on each device.
 Moreover, recent experimental developments deal with parallelization of function over supercomputer-scale hardware meshes.
@@
 
-\paragraph{NumPy API compliance} Finally, the last point to note about JAX, is that it mirrors the NumPy API (with only a few exceptions).
-This means in practice that existing NumPy code can easily be converted to JAX.
+\paragraph{\textbf{NumPy API compliance}} Finally, the last point to note about \texttt{JAX}, is that it mirrors the NumPy API (with only a few exceptions).
+This means in practice that existing NumPy code can easily be converted to \texttt{JAX}.
 This is in contrast to other similar frameworks like TensorFlow, PyTorch, or Julia which all require the user to learn, and adapt their code to, a new API or even a new language.
@@
 
-All \jaxcosmo\ data structures are organized as JAX \textit{pyTree} container objects, ensuring that two key JAX features are available to them: \textit{autodiff} and \textit{vmap}.
-The vmap feature enables any operation defined in JAX (including complicated composite operations) to be applied efficiently as a vector operation.
+All \jaxcosmo\ data structures are organized as \texttt{JAX} \textit{pyTree} container objects, ensuring that two key \texttt{JAX} features are available to them: \textit{autodiff} and \textit{vmap}.
+The vmap feature enables any operation defined in \texttt{JAX} (including complicated composite operations) to be applied efficiently as a vector operation.
 The autodiff feature further makes it possible to take the derivative of any operation, automatically transforming a function that takes $n$ inputs and produces $m$ outputs into a new function that generates an $m \times n$ matrix of partial derivatives.
@@
 
-%Future version of the library would offer the possibility to call a JAX emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.
+%Future version of the library would offer the possibility to call a \texttt{JAX} emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.
 \JZ{How close is this to being usable?
@@
 \end{equation}
-which is accomplished in two lines of JAX code:
+which is accomplished in two lines of \texttt{JAX} code:
 \begin{lstlisting}[language=iPython]
@@
 
-Because JAX functions are differentiable with respect to all their inputs, including survey configuration parameters (e.g.
+Because \texttt{JAX} functions are differentiable with respect to all their inputs, including survey configuration parameters (e.g.
 depth, area, etc), we can even compute the derivative of an FoM with respect to these inputs, allowing for rapid and complete survey optimization.
@@
 %\FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
-Using the ``vectorized'' \texttt{numpyro} option we ran all 16 chains simultaneously on a single GPU, made possible by the JAX \textit{vmap} mechanism.
-If one has several GPU devices available, then the using the JAX paralellization mechanism (\textit{pmap}), it is further possible to launch the vectorized sampling across the devices, and get back all the MCMC chains.
+Using the ``vectorized'' \texttt{numpyro} option we ran all 16 chains simultaneously on a single GPU, made possible by the \texttt{JAX} \textit{vmap} mechanism.
+If one has several GPU devices available, then the using the \texttt{JAX} paralellization mechanism (\textit{pmap}), it is further possible to launch the vectorized sampling across the devices, and get back all the MCMC chains.
 However, the experiments have all been undertaken on a single GPU, either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
@@
 We believe, though,  that one of the reason for the wide diversity in this is a lack of standardization - a unified interface and validation suite for such methods would provide a much simpler comparison between them and enable wider usage.
-\JZ{Francois - is this what you meant here?
-If not please edit.
-If Francois hasn't done this soon then just delete this comment.}
 
@@
 Finally, one question worth asking is how does \jaxcosmo\ position itself against classical codes such as CCL?
-While we are convinced of the benefits of a JAX implementation, we expect CCL and other key codes to remain critical as standard cosmology implementations.
+While we are convinced of the benefits of a \texttt{JAX} implementation, we expect CCL and other key codes to remain critical as standard cosmology implementations.
 Ultimately, a natural transition may occur towards differentiable frameworks like \jaxcosmo\ when they reach the ability to run full-fledged Stage IV likelihoods.
@@
 
-In this paper, we have presented \jaxcosmo, a cosmology library implemented in the JAX framework that enables the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, and greatly speeds up likelihood evaluations thanks to automatic parallelisation and just-in-time compilation on GPUs.
+In this paper, we have presented \jaxcosmo, a cosmology library implemented in the \texttt{JAX} framework that enables the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, and greatly speeds up likelihood evaluations thanks to automatic parallelisation and just-in-time compilation on GPUs.
 Currently, \jaxcosmo\ only contains a small set of features corresponding to a DES-Y1 3x2pt analysis.
@@
 
-%In case of the MVN  $\nabla_\mu \mathbb{H}(q)=0$, $\nabla_L \mathbb{H}(q)=(L^{-1})^T$ while the other gradients are computed by the JAX \texttt{autodiff} mechanism.
+%In case of the MVN  $\nabla_\mu \mathbb{H}(q)=0$, $\nabla_L \mathbb{H}(q)=(L^{-1})^T$ while the other gradients are computed by the \texttt{JAX} \texttt{autodiff} mechanism.
 %\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.}

### 2023-01-11 paper_OJA.tex
Add author D. Kirkby
Add author contribution

@@
 F. Lanusse$^2$, 
-J. Zuntz$^3$,\\
+J. Zuntz$^3$,
+A. Peel$^4$\\
 \jaxcosmo\ contributors}
@@
 
-\affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
+\affiliation{
+$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France\\
+$^4$Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Observatoire de Sauverny, 1290 Versoix, Switzerland}
 
@@
 \textbf{J.E Campagne:} Conceptualization, Methodology, Software, Validation, Writing, Visualization.
-\textbf{F. Lanusse:}, Conceptualization, Methodology, Software, Validation, Writing, Project administration \textbf{J. Zuntz:} Investigation, Writing.
+\textbf{F. Lanusse:}, Conceptualization, Methodology, Software, Validation, Writing, Project administration.
+\textbf{J. Zuntz:} Investigation, Writing.
+\textbf{A. Peel:} Software, Validation.
 

### 2023-01-11 paper_OJA.tex
Add author contribution dlanzieri

--- 2022-12-20 paper_OJA.tex
@@
 J. Zuntz$^3$,\\
+D. Lanzieri$^4$\\
 \jaxcosmo\ contributors}
@@
 
-\affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
+\affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France \\
+ $^4$Université Paris Cité, Université Paris-Saclay, CEA, CNRS, AIM, F-91191, Gif-sur-Yvette, France}
 
@@
 \textbf{F. Lanusse:}, Conceptualization, Methodology, Software, Validation, Writing, Project administration \textbf{J. Zuntz:} Investigation, Writing.
+\textbf{D. Lanzieri:}, Software and validation for redshift distribution.
 

### 2023-01-13 paper_OJA.tex

--- 2022-12-20 paper_OJA.tex
@@
 J. Zuntz$^3$,\\
+Y. Li$^{4,5}$,\\
 \jaxcosmo\ contributors}
@@
 \affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
+
+\affiliation{$^4$Department of Mathematics and Theory, Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
+\affiliation{$^5$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
 
@@
 
-While these tools have especially been applied to neural network optimization and machine learning in general, they can also enable classical statistical methods that require the derivatives of (e.g.
+While these tools have especially been applied to neural network optimization and machine learning (ML) in general, they can also enable classical statistical methods that require the derivatives of (e.g.
 likelihood) functions to operate: we consider such methods in this paper.
@@
 We give an overview of the code's design and contents in Section~\ref{sec-jaxcosmo-design}.
-We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more complete gradient-based cosmological inference with variants of Hamiltonian Monte Carlo and its variant No-U-Turn, and ML-accelerated Stochastic Variational Inference in Section~\ref{sec:chmc}.
-We discuss and compare these methods in Section~\ref{sec-results} and conclude in Section~\ref{sec-conclusion}.
+We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more complete gradient-based cosmological inference with variants of Hamiltonian Monte Carlo including the No-U-Turn Sampler, and ML-accelerated Stochastic Variational Inference in Section~\ref{sec:chmc}.
+We discuss and compare these methods in Section~\ref{sec-discussion} and conclude in Section~\ref{sec-conclusion}.
 
@@
 
-\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, and Simpson \& Romberg integration routines through this framework, so that we can automatically compute the derivative of their solutions with respect to their inputs.
-This includes the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also the other input quantities such as redshift distributions described below.
+\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, and Simpson \& Romberg integration routines through this framework, so that we can automatically compute the derivatives of their solutions with respect to their inputs.
+This includes not only the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also other input quantities such as redshift distributions as described below.
 
@@
 \end{align}
-The factors $m_i$ are equal to $(0,-2)$ for the galaxy clustering and weak lensing respectively, and the functions $K(z)$ each represent a single tomographic redshift bin's number density.
+The $m_i$ factors are $(0,-2)$ for the galaxy clustering and weak lensing, respectively, and each $K(z)$ function represent a single tomographic redshift bin's number density.
 These tracers are implemented as two kernel functions:
@@
     \end{equation}
-where $n_i(s)$ is the redshift distribution of the sample (eg.
-\texttt{jc.redshift.kde\_nz} function), $b(z)$ a galaxy bias function (see  \texttt{jc.bias.constant\_linear\_bias}).
+where $n_i(s)$ is the redshift distribution of the sample (e.g., \texttt{jc.redshift.kde\_nz} function), and $b(z)$ is the galaxy bias function (see  \texttt{jc.bias.constant\_linear\_bias}).
 No redshift space distortions are taken into account.
@@
 \end{equation}
-with $C\approx 0.0134$ a dimensionless constant and $D(z)$ the growth factor of linear perturbations.
+with $C\approx 0.0134$ being a dimensionless constant and $D(z)$ the growth factor of linear perturbations.
 \end{description}
@@
 
-Because, like the other ingredients, the kernels are implemented as pyTree objects, \texttt{NumberCounts} and \texttt{WeakLensing}, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters  or the number densities using \textit{autodiff} and \jaxcosmo's implementation of integration quadrature.
+Because, like the other ingredients, the kernels are implemented as pyTree objects, namely \texttt{NumberCounts} and \texttt{WeakLensing}, all the integrals involved in these computations can be differentiated with respect to the cosmological parameters and the number densities, using \textit{autodiff} and \jaxcosmo's implementation of integration quadrature.
 An example is given in the context of DES Y1 3x2pts analysis (Sec~\ref{sec-DESY1}).
@@
 The algorithm computes by iteration $M$ vectors of size $N$ such that (taking the notation of the previous section)
-\begin{equation}
-    b_i = \frac{C^{-1}\mu_{,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)b_j}{\sqrt{F_{i,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)^2}}
+%JEC 13/1/23 Typo fixed
+\begin{equation}
+    b_i = \frac{C^{-1}\mu_{,i}-\sum_{j=1}^{i-1}(\mu_{,i}^T b_j)b_j}{\sqrt{F_{i,i}-\sum_{j=1}^{i-1}(\mu_{,i}^T b_j)^2}}
 \label{eq:moped}
@@
 Since the uncertainty in the covariance due to a finite number of simulations must be accounted for \citep{2018MNRAS.473.2355S,2007A&A...464..399H}, this reduction can also offset any loss of information from the compression.
-Inaccuracies in the full covariance matrix used in the data compression result only in information loss, not bias, as does mis-specification of the fiducial parameter set $\theta_i$ where equation \ref{eq:moped}.
+Inaccuracies in the full covariance matrix used in the data compression result only in information loss, not bias, as does mis-specification of the fiducial parameter set $\theta_i$ in equation \ref{eq:moped}.
 
@@
 In the following sections we review some statistical methods which directly benefit from the automatic differentiablity of \jaxcosmo\ likelihoods.
-We demonstrate a range of gradient-based methods, from Hamiltonian Monte Carlo (HMC), and its \textit{No-U-Turn} (NUTS) variant, to Stochastic Variational Inference (SVI).
+We demonstrate a range of gradient-based methods, from Hamiltonian Monte Carlo (HMC), and its \textit{No-U-Turn Sampler} (NUTS) variant, to Stochastic Variational Inference (SVI).
 We further report on their respective computational costs in a DES-Y1 like analysis.
@@
 
-The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler was introduced in \citet{10.5555/2627435.2638586}.
+The No-U-Turn Sampler (\textit{NUTS})  variant of the traditional HMC sampler was introduced in \citet{10.5555/2627435.2638586}.
 It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$ by finding good and dynamic choices for the leapfrog integration parameters in the root HMC algorithms, the step size $\varepsilon$ and the number of steps $L$.

### 2023-01-13 paper_OJA.tex

@@
 J. Zuntz$^3$,\\
-Y. Li$^{4,5}$,\\
+D.~Kirkby$^4$,\\
+Y.~Li$^{5,6}$,\\
 \jaxcosmo\ contributors}
@@
 
-\affiliation{$^4$Department of Mathematics and Theory, Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
-\affiliation{$^5$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
+\affiliation{$^5$Department of Mathematics and Theory, Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
+\affiliation{$^6$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
 
@@
 \textbf{J.E Campagne:} Conceptualization, Methodology, Software, Validation, Writing, Visualization.
-\textbf{F. Lanusse:}, Conceptualization, Methodology, Software, Validation, Writing, Project administration \textbf{J. Zuntz:} Investigation, Writing.
+\textbf{F. Lanusse:}, Conceptualization, Methodology, Software, Validation, Writing, Project administration.
+\textbf{J. Zuntz:} Investigation, Writing.
+\textbf{D.~Kirkby:}, Software and validation for sparse linear algebra.
 

### 2023-01-13 paper_OJA.tex

@@
 Y.~Li$^{5,6}$,\\
+A. Peel$^7$\\
 \jaxcosmo\ contributors}
@@
 
-\affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
+\affiliation{
+$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France\\
+}
 
@@
 \affiliation{$^6$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
+\affiliation{$^7$Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Observatoire de Sauverny, 1290 Versoix, Switzerland}
 
@@
 %%%
-\section*{CRedit authorship contribution statement}
+\section*{Credit authorship contribution statement}
 \textbf{J.E Campagne:} Conceptualization, Methodology, Software, Validation, Writing, Visualization.
@@
 \textbf{D.~Kirkby:}, Software and validation for sparse linear algebra.
+\textbf{A. Peel:} Software and validation for spline interpolations in JAX.
 

### 2023-01-13 paper_OJA.tex
Add contributor

--- 2023-01-13 paper_OJA.tex
@@
 J. Zuntz$^3$,\\
-Y. Li$^{4,5}$,\\
+A. Boucaud$^4$, 
 \jaxcosmo\ contributors}
@@
 
-\affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France}
-
-\affiliation{$^4$Department of Mathematics and Theory, Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
-\affiliation{$^5$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
+\affiliation{$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France
+$^4$Université de Paris, CNRS, Astroparticule et Cosmologie, F-75013 Paris, France}
 
@@
 
-While these tools have especially been applied to neural network optimization and machine learning (ML) in general, they can also enable classical statistical methods that require the derivatives of (e.g.
+While these tools have especially been applied to neural network optimization and machine learning in general, they can also enable classical statistical methods that require the derivatives of (e.g.
 likelihood) functions to operate: we consider such methods in this paper.
@@
 We give an overview of the code's design and contents in Section~\ref{sec-jaxcosmo-design}.
-We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more complete gradient-based cosmological inference with variants of Hamiltonian Monte Carlo including the No-U-Turn Sampler, and ML-accelerated Stochastic Variational Inference in Section~\ref{sec:chmc}.
-We discuss and compare these methods in Section~\ref{sec-discussion} and conclude in Section~\ref{sec-conclusion}.
+We show how to use it for rapid and numerically stable Fisher forecasts in Section~\ref{sec-fisher-forecast}, in more complete gradient-based cosmological inference with variants of Hamiltonian Monte Carlo and its variant No-U-Turn, and ML-accelerated Stochastic Variational Inference in Section~\ref{sec:chmc}.
+We discuss and compare these methods in Section~\ref{sec-results} and conclude in Section~\ref{sec-conclusion}.
 
@@
 
-\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, and Simpson \& Romberg integration routines through this framework, so that we can automatically compute the derivatives of their solutions with respect to their inputs.
-This includes not only the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also other input quantities such as redshift distributions as described below.
+\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, and Simpson \& Romberg integration routines through this framework, so that we can automatically compute the derivative of their solutions with respect to their inputs.
+This includes the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also the other input quantities such as redshift distributions described below.
 
@@
 \end{align}
-The $m_i$ factors are $(0,-2)$ for the galaxy clustering and weak lensing, respectively, and each $K(z)$ function represent a single tomographic redshift bin's number density.
+The factors $m_i$ are equal to $(0,-2)$ for the galaxy clustering and weak lensing respectively, and the functions $K(z)$ each represent a single tomographic redshift bin's number density.
 These tracers are implemented as two kernel functions:
@@
     \end{equation}
-where $n_i(s)$ is the redshift distribution of the sample (e.g., \texttt{jc.redshift.kde\_nz} function), and $b(z)$ is the galaxy bias function (see  \texttt{jc.bias.constant\_linear\_bias}).
+where $n_i(s)$ is the redshift distribution of the sample (eg.
+\texttt{jc.redshift.kde\_nz} function), $b(z)$ a galaxy bias function (see  \texttt{jc.bias.constant\_linear\_bias}).
 No redshift space distortions are taken into account.
@@
 \end{equation}
-with $C\approx 0.0134$ being a dimensionless constant and $D(z)$ the growth factor of linear perturbations.
+with $C\approx 0.0134$ a dimensionless constant and $D(z)$ the growth factor of linear perturbations.
 \end{description}
@@
 
-Because, like the other ingredients, the kernels are implemented as pyTree objects, namely \texttt{NumberCounts} and \texttt{WeakLensing}, all the integrals involved in these computations can be differentiated with respect to the cosmological parameters and the number densities, using \textit{autodiff} and \jaxcosmo's implementation of integration quadrature.
+Because, like the other ingredients, the kernels are implemented as pyTree objects, \texttt{NumberCounts} and \texttt{WeakLensing}, all the integrals involved in these computations can be  differentiated with respect to the cosmological parameters  or the number densities using \textit{autodiff} and \jaxcosmo's implementation of integration quadrature.
 An example is given in the context of DES Y1 3x2pts analysis (Sec~\ref{sec-DESY1}).
@@
 The algorithm computes by iteration $M$ vectors of size $N$ such that (taking the notation of the previous section)
-%JEC 13/1/23 Typo fixed
-\begin{equation}
-    b_i = \frac{C^{-1}\mu_{,i}-\sum_{j=1}^{i-1}(\mu_{,i}^T b_j)b_j}{\sqrt{F_{i,i}-\sum_{j=1}^{i-1}(\mu_{,i}^T b_j)^2}}
+\begin{equation}
+    b_i = \frac{C^{-1}\mu_{,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)b_j}{\sqrt{F_{i,i}-\sum_{j=1}^{i-1}(\mu_{,j}^T b_j)^2}}
 \label{eq:moped}
@@
 Since the uncertainty in the covariance due to a finite number of simulations must be accounted for \citep{2018MNRAS.473.2355S,2007A&A...464..399H}, this reduction can also offset any loss of information from the compression.
-Inaccuracies in the full covariance matrix used in the data compression result only in information loss, not bias, as does mis-specification of the fiducial parameter set $\theta_i$ in equation \ref{eq:moped}.
+Inaccuracies in the full covariance matrix used in the data compression result only in information loss, not bias, as does mis-specification of the fiducial parameter set $\theta_i$ where equation \ref{eq:moped}.
 
@@
 In the following sections we review some statistical methods which directly benefit from the automatic differentiablity of \jaxcosmo\ likelihoods.
-We demonstrate a range of gradient-based methods, from Hamiltonian Monte Carlo (HMC), and its \textit{No-U-Turn Sampler} (NUTS) variant, to Stochastic Variational Inference (SVI).
+We demonstrate a range of gradient-based methods, from Hamiltonian Monte Carlo (HMC), and its \textit{No-U-Turn} (NUTS) variant, to Stochastic Variational Inference (SVI).
 We further report on their respective computational costs in a DES-Y1 like analysis.
@@
 
-The No-U-Turn Sampler (\textit{NUTS})  variant of the traditional HMC sampler was introduced in \citet{10.5555/2627435.2638586}.
+The No-U-Turn (\textit{NUTS})  variant of the traditional HMC sampler was introduced in \citet{10.5555/2627435.2638586}.
 It aims to help finding a new point  $x_{i+1}$ from  the current $x_i$ by finding good and dynamic choices for the leapfrog integration parameters in the root HMC algorithms, the step size $\varepsilon$ and the number of steps $L$.

### 2023-01-13 paper_OJA.tex

--- 2023-01-13 paper_OJA.tex
@@
 J. Zuntz$^3$,\\
-D.~Kirkby$^4$,\\
-Y.~Li$^{5,6}$,\\
-A. Peel$^7$\\
+A. Boucaud$^8$,
+D.~Kirkby$^4$,
+D. Lanzieri$^2$,
+Y.~Li$^{5,6}$,
+A. Peel$^7$
 \jaxcosmo\ contributors}
@@
 \affiliation{$^7$Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Observatoire de Sauverny, 1290 Versoix, Switzerland}
+\affiliation{$^8$Université de Paris, CNRS, Astroparticule et Cosmologie, F-75013 Paris, France}
 
@@
 \section*{Credit authorship contribution statement}
+\textbf{A. Boucaud:} Software and comments.
 \textbf{J.E Campagne:} Conceptualization, Methodology, Software, Validation, Writing, Visualization.
-\textbf{F. Lanusse:}, Conceptualization, Methodology, Software, Validation, Writing, Project administration.
+\textbf{F. Lanusse:} Conceptualization, Methodology, Software, Validation, Writing, Project administration.
+\textbf{D. Lanzieri:} Software and validation for redshift distribution.
 \textbf{J. Zuntz:} Investigation, Writing.
-\textbf{D.~Kirkby:}, Software and validation for sparse linear algebra.
+\textbf{D.~Kirkby:} Software and validation for sparse linear algebra.
 \textbf{A. Peel:} Software and validation for spline interpolations in JAX.

### 2023-02-05 paper_OJA.tex
Update the paper

@@
 \newcommand{\autodiff}{\texttt{autodiff}}
+\newcommand{\jax}{\texttt{JAX}}
 
@@
 J. Zuntz$^3$,\\
-A. Boucaud$^8$,
-D.~Kirkby$^4$,
-D. Lanzieri$^2$,
-Y.~Li$^{5,6}$,
-A. Peel$^7$
-\jaxcosmo\ contributors}
+A. Boucaud$^4$,
+S. Casas$^{5}$,
+M.~Karamanis$^{6,7}$,
+D.~Kirkby$^8$,
+D. Lanzieri$^9$,
+Y.~Li$^{10,11}$,
+A. Peel$^{12}$
+}
 \thanks{$^\ast$jean-eric.campagne@ijclab.in2p3.fr}
@@
 \affiliation{
-$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France\\
+$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France
 }
-
-\affiliation{$^5$Department of Mathematics and Theory, Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
-\affiliation{$^6$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
-\affiliation{$^7$Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Observatoire de Sauverny, 1290 Versoix, Switzerland}
-\affiliation{$^8$Université de Paris, CNRS, Astroparticule et Cosmologie, F-75013 Paris, France}
+\affiliation{$^2$Université Paris-Saclay, Université Paris Cité, CEA, CNRS, AIM, 91191, Gif-sur-Yvette, France}
+\affiliation{$^3$Institute for Astronomy, University of Edinburgh, Edinburgh EH9 3HJ, United Kingdom}
+\affiliation{$^4$Université de Paris, CNRS, Astroparticule et Cosmologie, F-75013 Paris, France}
+\affiliation{$^{5}$Institute for Theoretical Particle Physics and Cosmology (TTK), RWTH Aachen University, 52056 Aachen, Germany.}
+\affiliation{$^{6}$Berkeley Center for Cosmological Physics, University of California, Berkeley, CA 94720, USA}
+\affiliation{$^{7}$Lawrence Berkeley National Laboratory, 1 Cyclotron Road, Berkeley, CA 94720, USA}
+\affiliation{$^8$Department of Physics and Astronomy, University of California, Irvine, CA 92697, USA}
+\affiliation{$^9$Université Paris Cité, Université Paris-Saclay, CEA, CNRS, AIM, F-91191, Gif-sur-Yvette, France}
+\affiliation{$^{10}$Department of Mathematics and Theory, Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
+\affiliation{$^{11}$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
+\affiliation{$^{12}$Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Observatoire de Sauverny, 1290 Versoix, Switzerland}
 
@@
 \begin{abstract}
-
-We present \jaxcosmo, a library for differentiable cosmological theory calculations.
-\jaxcosmo\ uses the recent \texttt{JAX} library, which has created a new coding ecosystem, especially in probabilistic programming.
-%
-As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), \texttt{JAX} exposes an \textit{automatic differentiation} (AD) mechanism.
-AD can take a Python function that uses \texttt{JAX} and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn.
-This gives us the opportunity to apply a range of powerful Bayesian inference algorithms, otherwise impractical in cosmology, such as Hamiltonian Monte Carlo and Variational Inference.
-%
-\jaxcosmo\ implements background evolution, linear and non-linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes, all differentiable with respect to the cosmological parameters and their other inputs.
-%
-We discuss algorithms made possible by this library, and present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical inference.
-%
-We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that Normalizing Flows using the Neural Transport are a promising methodology.
-
-% The recent \texttt{JAX} library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems.
-In particular, \texttt{JAX} exposes an \textit{automatic differentiation} mechanism, which can take a python function that uses \texttt{JAX} and return its (vector) derivative.
+We present \jaxcosmo, a library for automatically differentiable cosmological theory calculations.
+\jaxcosmo\ uses the \jax\ library, which has created a new coding ecosystem, especially in probabilistic programming.
+%
+As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), \jax\ exposes an \textit{automatic differentiation} (AD) mechanism.
+Thanks to AD, \jaxcosmo\ gives access to the derivatives of cosmological likelihoods with respect to any of their parameters, and thus enables a range of powerful %  AD can take a Python function that uses \jax\ and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn.
+This gives us the opportunity to apply a range of powerful
+Bayesian inference algorithms, otherwise impractical in cosmology, such as Hamiltonian Monte Carlo and Variational Inference. 
+%
+In its initial release, \jaxcosmo\ implements background evolution, linear and non-linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes, all differentiable with respect to the cosmological parameters and their other inputs.
+%
+We illustrate how automatic differentiation can be a game-changer for common tasks involving Fisher matrix computations, or full posterior inference with gradient-based techniques (e.g.
+Hamiltonian Monte Carlo).
+In particular, we show how Fisher matrices are now fast, exact, no longer require any fine tuning, and are themselves differentiable with respect to parameters of the likelihood enabling complex survey optimization by simple gradient descent.
+Finally, using a Dark Energy Survey Year 1 3x2pt analysis as a benchmark, we demonstrate how \jaxcosmo\ can be combined with Probabilistic Programming Languages such as \numpyro\ to perform posterior inference with state-of-the-art algorithms including a No U-Turn Sampler (NUTS), Automatic Differentiation Variational Inference (ADVI), and Neural Transport HMC (NeuTra).% We discuss algorithms made possible by this library, present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical inference.
+%
+We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that Normalizing Flows using the Neural Transport are a promising methodology.\FrL{we need a better statement than this, what are our conclusions?
+Are we faster/better than Cobaya?}
+
+% The recent \jax\ library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems.
+In particular, \jax\ exposes an \textit{automatic differentiation} mechanism, which can take a python function that uses \jax\ and return its (vector) derivative.
 This gives us the opportunity to apply a range of powerful but otherwise unfeasible algorithms used in Bayesian inference, such as Hamiltonian Monte Carlo (HMC) and Variational Inference.
@@
 \section{Introduction}
-Bayesian inference has been widely used in cosmology in the form of Monte Carlo Markov Chains (MCMC) since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for past and current analysis thanks partly to packages as  \texttt{CosmoMC} \citep{2002PhRvD..66j3511L}, \texttt{CosmoSIS} \citep{2015A&C....12...45Z}, \texttt{MontePython} \citep{2019PDU....24..260B} as well as   \texttt{Cobaya} \citep{2019ascl.soft10019T,2021JCAP...05..057T}; see, for instance, the list of citations to these popular packages for an idea of the wide usage in the community.
+Bayesian inference has been widely used in cosmology in the form of Monte Carlo Markov Chains (MCMC) since the work of \citet{2001ApJ...563L..95K} and \citet{2003MNRAS.341.1084R}, and has been the keystone for past and current analysis thanks partly to packages such as  \texttt{CosmoMC} \citep{2002PhRvD..66j3511L}, \texttt{CosmoSIS} \citep{2015A&C....12...45Z}, \texttt{MontePython} \citep{2019PDU....24..260B}, and \texttt{Cobaya} \citep{2019ascl.soft10019T,2021JCAP...05..057T}; see, for instance, the list of citations to these popular packages for an idea of the wide usage in the community.
 
@@
 likelihood) functions to operate: we consider such methods in this paper.
-\textit{Autodiff} has been implemented in widely used libraries like \texttt{Stan} \citep{JSSv076i01}, \texttt{TensorFlow}  \citep{tensorflow2015-whitepaper}, and  \texttt{PyTorch} \citep{NEURIPS2019_9015}.
-
-A recent entrant to this field is the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} which has undergone rapid development and can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions, offering a speed up to the development process and indeed code runtimes.
-\texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), and just-in-time compilation (\texttt{jit}) and optimization targetting CPU, GPU and TPU hardware thanks to the \texttt{XLA} library that powers TensorFlow.
-These attractive features have driven wide adoption of \texttt{JAX} in computational research, and motivate us to consider its usage in cosmology.
-
-\texttt{JAX} contains bespoke reimplementation of packages such as \texttt{jax.numpy} and \texttt{jax.scipy}, as well as example libraries as \texttt{Stax} for simple but flexible neural network development.
+\textit{Autodiff} has been implemented in widely used libraries like \texttt{Stan} \citep{JSSv076i01}, \texttt{TensorFlow} \citep{tensorflow2015-whitepaper}, \texttt{Julia} \citep{bezanson2017julia}, and \texttt{PyTorch} \citep{NEURIPS2019_9015}.
+
+A recent entrant to this field is the \jax\ library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} which has undergone rapid development and can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions, offering a speed up to the development process and indeed code runtimes.
+\jax\ offers an easy parallelization mechanism (\texttt{vmap}), just-in-time compilation (\texttt{jit}), and optimization targeting CPU, GPU, and TPU hardware thanks to the \texttt{XLA} library that powers TensorFlow.
+These attractive features have driven wide adoption of \jax\ in computational research, and motivate us to consider its usage in cosmology.
+
+\jax\ contains bespoke reimplementations of packages such as \texttt{jax.numpy} and \texttt{jax.scipy}, as well as example libraries such as \texttt{Stax} for simple but flexible neural network development.
 It has formed the seed for a wider ecosystem of packages, including, for example:
-\texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro\ \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper.
-Other PPL packages such as \texttt{PyMC} \citep{Salvatier2016} have switched to \texttt{JAX} backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}.
+\texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} a hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro\ \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper.
+Other PPL packages such as \texttt{PyMC} \citep{Salvatier2016} have switched to a \jax\ backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}.
 
@@
 In this context we have developed the open source \jaxcosmo\ library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
-The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable \texttt{JAX} functions.
+The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable \jax\ functions.
 
@@
 
-\section{JAX: GPU Accelerated and Automatically Differentiable Python Programming}
+\section{\jax: GPU Accelerated and Automatically Differentiable Python Programming}
 \label{sec-primer}
 
-The aim of this section is to provide a brief technical primer on \texttt{JAX}, necessary to fully grasp the potential of a cosmology library implemented in this framework.
+The aim of this section is to provide a brief technical primer on \jax, necessary to fully grasp the potential of a cosmology library implemented in this framework.
 
@@
 The first approach is to derive analytically the formula for the derivatives of interest \citep[e.g.][]{2013MNRAS.432..894J}, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}.
-This is however only practical typically for simple analytical models.
+This is typically only practical, however, for simple analytical models.
 The second approach is to compute numerical derivatives by finite differences.
-This approach can be applied on black-box models of arbitrary complexity (from typical Boltzmann codes to cosmological simulations \citep{2020ApJS..250....2V}).
+This approach can be applied on black-box models of arbitrary complexity (from typical Boltzmann codes to cosmological simulations; \citealp{2020ApJS..250....2V}).
 However it is notoriously difficult to obtain stable derivatives by finite differences \citep[e.g.][]{2021arXiv210100298B, 2021A&A...649A..52Y}.
-In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them impractical whenever derivatives are needed as part of outer iterative algorithm.
-
-Automatic differentiation frameworks like \texttt{JAX} take a different approach.
-They trace the execution of a given model and decompose this trace into fundamental operations with known derivatives (e.g.
-the derivative of a multiplication operation is known).
-Then, by applying the chain rule formula, the computational graph for the derivatives (of any order) of the model can be built from the known derivatives of every elementary operations.
+In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them impractical whenever derivatives are needed as part of an outer iterative algorithm.
+
+Automatic differentiation frameworks like \jax\ take a different approach.
+They trace the execution of a given model and decompose this trace into primitive operations with known derivatives (e.g.
+multiplication).
+Then, by applying the chain rule formula, the computational graph for the derivatives (of any order) of the model can be built from the known derivatives of every elementary operation.
 A new function corresponding to the derivative of the original function is therefore built automatically for the user.
-We direct the interested reader to \cite{baydin2018automatic, Margossian2019} for in-depth introductions to automatic differentiation.
-
-JAX provides in particular a number of operators (\texttt{jax.grad}, \texttt{jax.jacobian}, \texttt{jax.hessian}) which can compute derivatives of any function written in \texttt{JAX}.
+We direct the interested reader to \citet{baydin2018automatic} and \citet{Margossian2019} for in-depth introductions to automatic differentiation.
+
+\jax\ provides in particular a number of operators (\texttt{jax.grad}, \texttt{jax.jacobian}, \texttt{jax.hessian}) which can compute derivatives of any function written in \jax:
 \begin{lstlisting}[language=iPython]
-# Let us define a simple function 
+# Define a simple function 
 def f(x):
@@
 df_dx = jax.grad(f)
-# df_dx is a new function that always return 5 
+# df_dx is a new function that always returns 5 
 \end{lstlisting}
 
-\textbf{Why is that interesting?} \autodiff\ makes it possible to obtain \textit{exact gradients of cosmological likelihoods} with respect to all input parameters at the cost of only 2 likelihood evaluations.
+\textbf{Why is this interesting?} \autodiff\ makes it possible to obtain \textit{exact gradients of cosmological likelihoods} with respect to all input parameters at the cost of only two likelihood evaluations.
 
@@
 One approach to mitigate these issues and make Python code fast is Just In Time compilation, which traces the execution of a given Python function the first time it is called, and compiles it into a fast executable (by-passing the Python interpreter), which can be transparently used in subsequent calls to this function.
-Compared to other strategies for speeding up Python code such as Cython, JIT allows the user to simply write plain Python code, and reap the benefits of compiled code.
-
-A number of libraries allowing for JIT have already been used in astrophysics, in particular Numba\footnote{\url{https://numba.pydata.org/}}, or the HOPE library \cite{2015A&C....10....1A} developed specifically for the need of astrophysics.
-\texttt{JAX} stands out compared to these other frameworks in that it relies on the highly optimized XLA library\footnote{\url{https://www.tensorflow.org/xla}} for executing the compiled expressions.
-XLA is continuously developed by Google as part of the TensorFlow project, for efficient training and inference of large scale deep learning applications, and as such supports computational backends such as GPU an Tensor Processing Units (TPU) clusters.
-The ability to perform computations directly on GPUs through XLA is one of the major benefits of \texttt{JAX}, as speed-ups of at least two orders of magnitudes can be expected for typical parallel linear algebra computations compared to CPU.
-
-In \texttt{JAX}, jitting is achieved simply by transforming a function with \texttt{jax.jit} operation:
+Compared to other strategies for speeding up Python code such as Cython, JIT allows the user to write plain Python code, and reap the benefits of compiled code.
+
+A number of libraries allowing for JIT have already been used in astrophysics, in particular Numba\footnote{\url{https://numba.pydata.org/}}, or the HOPE library \cite{2015A&C....10....1A} developed specifically for the needs of astrophysics.
+\jax\ stands out compared to these other frameworks in that it relies on the highly optimized XLA library\footnote{\url{https://www.tensorflow.org/xla}} for executing the compiled expressions.
+XLA is continuously developed by Google as part of the TensorFlow project, for efficient training and inference of large scale deep learning applications, and as such supports computational backends such as GPU and Tensor Processing Units (TPU) clusters.
+The ability to perform computations directly on GPUs through XLA is one of the major benefits of \jax, as speed-ups of at least two orders of magnitudes can be expected for typical parallel linear algebra computations compared to CPU.
+
+In \jax, jitting is achieved by transforming a function with the \texttt{jax.jit} operation:
 \begin{lstlisting}[language=iPython]
-# Let us redefine our function
+# Redefine our function
 def f(x):
@@
 
-\textbf{Why is that interesting?} JIT makes it possible to execute entire cosmological \textit{MCMC chains directly on GPUs as compiled code}, with orders of magnitude in speedup over Python code.
-
-\paragraph{\textbf{Automatic Vectorization}} Another extremely powerful feature of \texttt{JAX} is its ability to automatically vectorize or paralellize any function.
-Through the same tracing mechanism used for automatic differentiation, \texttt{JAX} can decompose a given computation into fundamental operations and add a new \textit{batch} dimension so that the computation can be applied on a batch of inputs as opposed to individual inputs.
-In doing so, note that the computation will not run sequentially over all entries of the batch, but trully in parallel making full use of the intrinsic parallel architecture of modern GPUs.
-
-In \texttt{JAX} automatic vectorization is achieved using the \textit{jax.vmap} operation:
+\textbf{Why is this interesting?} JIT makes it possible to execute entire cosmological \textit{MCMC chains directly on GPUs as compiled code}, with orders of magnitude gain in speedup over Python code.
+
+\paragraph{\textbf{Automatic Vectorization}} Another extremely powerful feature of \jax\ is its ability to automatically vectorize or paralellize any function.
+Through the same tracing mechanism used for automatic differentiation, \jax\ can decompose a given computation into primitive operations and add a new \textit{batch} dimension so that the computation can be applied to a batch of inputs as opposed to individual ones.
+In doing so, the computation will not run sequentially over all entries of the batch, but truly in parallel making full use of the intrinsic parallel architecture of modern GPUs.
+
+In \jax\ automatic vectorization is achieved using the \textit{jax.vmap} operation:
 \begin{lstlisting}[language=iPython]
@@
 The function above is trivial, but the same approach can be used to parallelize any function, from Limber integrals, to an entire likelihood.
-When considering multi-devices use-case (eg.
-GPU or TPU), \texttt{JAX} provides \texttt{pmap} which compile and execute in parallel replicates of the same code on each device.
-Moreover, recent experimental developments deal with parallelization of function over supercomputer-scale hardware meshes.
-Notice that in the examples detailed in this article, we have only relied on \texttt{vmap} functionality.
-
-\textbf{Why is that interesting?} Automatic Vectorization makes it possible to trivially parallelize cosmological likelihood evaluations, to run many parallel MCMC chains on a single GPU.
-
-\paragraph{\textbf{NumPy API compliance}} Finally, the last point to note about \texttt{JAX}, is that it mirrors the NumPy API (with only a few exceptions).
-This means in practice that existing NumPy code can easily be converted to \texttt{JAX}.
+For multi-device use-cases (e.g., several GPUs or TPUs), \jax\ provides \texttt{pmap} which compiles and executes, in parallel, replicas of the same code on each device.
+Moreover, recent experimental developments deal with parallelization of functions over supercomputer-scale hardware meshes.
+In the examples detailed in this article, we have only relied on \texttt{vmap} functionality.
+
+\textbf{Why is this interesting?} Automatic Vectorization makes it possible to trivially parallelize cosmological likelihood evaluations, to run many parallel MCMC chains on a single GPU.
+
+\paragraph{\textbf{NumPy API compliance}} Finally, the last point to note about \jax, is that it mirrors the NumPy API (with only a few exceptions).
+This means in practice that existing NumPy code can easily be converted to \jax.
 This is in contrast to other similar frameworks like TensorFlow, PyTorch, or Julia which all require the user to learn, and adapt their code to, a new API or even a new language.
 
-\textbf{Why is that interesting?} NumPy compliance implies improved maintainability and lower barrier to entry for new contributors.
-
-\section{Design of the \jaxcosmo\ library}
+\textbf{Why is this interesting?} NumPy compliance implies improved maintainability and lower barrier to entry for new contributors.
+
+\section{Capabilities of the \jaxcosmo\ library}
 \label{sec-jaxcosmo-design}
 
-In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX.
-The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though, in its initial release,
+In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in \jax.
+The general design follows that of the Core Cosmology Library (CCL; \citealp{2019ApJS..242....2C}), though in its initial release
  \jaxcosmo\ only implements a subset of CCL features and options.
 
-All \jaxcosmo\ data structures are organized as \texttt{JAX} \textit{pyTree} container objects, ensuring that two key \texttt{JAX} features are available to them: \textit{autodiff} and \textit{vmap}.
-The vmap feature enables any operation defined in \texttt{JAX} (including complicated composite operations) to be applied efficiently as a vector operation.
+All \jaxcosmo\ data structures are organized as \jax\ container objects, which means that two key \jax\ features are automatically available to them: \textit{autodiff} and \textit{vmap}.
+The vmap feature enables any operation defined in \jax\ (including complicated composite operations) to be applied efficiently as a vector operation.
 The autodiff feature further makes it possible to take the derivative of any operation, automatically transforming a function that takes $n$ inputs and produces $m$ outputs into a new function that generates an $m \times n$ matrix of partial derivatives.
 
-\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, and Simpson \& Romberg integration routines through this framework, so that we can automatically compute the derivatives of their solutions with respect to their inputs.
+\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, as well as Simpson and Romberg integration routines through this framework, so that we can automatically compute the derivatives of their solutions with respect to their inputs.
 This includes not only the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also other input quantities such as redshift distributions as described below.
@@
 In the rest of this section we describe the cosmological calculations that are implemented using these facilities.
-
-\subsection{Background cosmology}
+\subsection{Formalism \& Implementation}
+
+\subsubsection{Background cosmology}
 
@@
 \begin{equation}
-     \chi(a) =  R_H \int_a^1 \frac{da^\prime}{{a^\prime}^2 E(a^\prime)}
+     \chi(a) =  R_H \int_a^1 \frac{\mathrm{d}a^\prime}{{a^\prime}^2 E(a^\prime)}
      \label{eq:radial_comoving}
@@
 %
-\subsection{Growth of perturbations}
+\subsubsection{Growth of perturbations}
 %
@@
 \begin{equation}
-    P(k, z) = P(k, z=\infty) \cdot T(k, z; \Omega_m, \Omega_b, ...),
+    P(k, z) = P(k, z=\infty) \cdot T^2(k, z; \Omega_m, \Omega_b, ...),
 \end{equation}
@@
 
-The normalisation $A$ is parametrized via $\sigma_8$ at $z=0$ as 
-\begin{equation}
-    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
+The normalisation $A$ is parameterised via $\sigma_8$ at $z=0$ as 
+\begin{equation}
+    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{\mathrm{d}k}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
 \end{equation}
@@
 
-%Future version of the library would offer the possibility to call a \texttt{JAX} emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.
+%Future version of the library would offer the possibility to call a \jax\ emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}.
 \JZ{How close is this to being usable?
@@
 %
-\subsection{Angular power spectra}
+\subsubsection{Angular power spectra}
 %
@@
 \begin{align}
-    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{d\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
+    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{\mathrm{d}\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
 \end{align}
-The $m_i$ factors are $(0,-2)$ for the galaxy clustering and weak lensing, respectively, and each $K(z)$ function represent a single tomographic redshift bin's number density.
+The $m_i$ factors are $(0,-2)$ for the galaxy clustering and weak lensing, respectively, and each $K(z)$ function represents a single tomographic redshift bin's number density.
 These tracers are implemented as two kernel functions:
@@
  K_i(z) = \left( \frac{3 H_0^2\Omega_m}{2 c} \right) \left(\frac{(\ell+2)!}{(\ell-2)!} \right)^{1/2}\times 
- \\  (1+z)\ \chi(z) \int_z^\infty p_i(z^\prime)\ \frac{\chi(z^\prime)-\chi(z)}{\chi(z^\prime)}\ dz^\prime + K_{IA}(z)
+ \\  (1+z)\ \chi(z) \int_z^\infty p_i(z^\prime)\ \frac{\chi(z^\prime)-\chi(z)}{\chi(z^\prime)}\ \mathrm{d}z^\prime + K_{IA}(z)
     \end{multline}
@@
 
-Because, like the other ingredients, the kernels are implemented as pyTree objects, namely \texttt{NumberCounts} and \texttt{WeakLensing}, all the integrals involved in these computations can be differentiated with respect to the cosmological parameters and the number densities, using \textit{autodiff} and \jaxcosmo's implementation of integration quadrature.
+Because, like the other ingredients, the \texttt{NumberCounts} and \texttt{WeakLensing} kernels are implemented as JAX objects, all the integrals involved in these computations can be differentiated with respect to the cosmological parameters and the number densities, using \textit{autodiff} and \jaxcosmo's implementation of integration quadrature.
 An example is given in the context of DES Y1 3x2pts analysis (Sec~\ref{sec-DESY1}).
@@
 %
-To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1rc9}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to compare it to the well-validated  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{\url{https://ccl.readthedocs.io}, version \texttt{2.5.1}.}.
+To illustrate the different features available with the present version of the library (\jaxcosmo\ \texttt{0.1}, which is available in the Python Package Index PyPI\footnote{\url{https://pypi.org/}}), we have written a  companion notebook \nblink{CCL_comparison} to compare it to the well-validated  Core Cosmology Library \citep{2019ApJS..242....2C}\footnote{\url{https://ccl.readthedocs.io}, version \texttt{2.5.1}.}.
 As examples, Figures \ref{fig:chi_comparison},
@@
 show the radial comoving distance (Eq.~\ref{eq:radial_comoving}), the non-linear matter power spectrum computation, and the angular power spectrum for galaxy-galaxy lensing (Eq.~\ref{eq:Cell_limber}) using the \texttt{NumberCounts} and \texttt{WeakLensing} kernel functions.
-\jaxcosmo\ features a suite of validation tests against CCL, automatically validating the precision of all computations to within the desired numberical accuracy; the relative differences between the two libraries are at the level of few $10^{-3}$ or better.
+\jaxcosmo\ features a suite of validation tests against CCL, automatically validating the precision of all computations to within the desired numerical accuracy; the relative differences between the two libraries are at the level of few $10^{-3}$ or better.
 
 These numerical differences are mostly due to different choices of integration methods and accuracy parameters (eg.
-quadrature number of points).
-Increasing these parameters lead to performance degradation for \jaxcosmo\ as it increases the XLA compilation memory requirements significantly, especially for the angular power spectra computation.
+number of quadrature points).
+Increasing these parameters leads to performance degradation for \jaxcosmo\, but increases the XLA compilation memory requirements significantly, especially for the angular power spectra computation.
 Since these differences are likely to be within the tolerance of the current generation of cosmological surveys, this trade-off is an acceptable one.
@@
 
-As a first illustration of the value of \textit{autodiff} in cosmological computations, we present in this section a few direct applications involving the computation of the Fisher information matrix, widely used in cosmology.
+As a first illustration of the value of \textit{autodiff} in cosmological computations, we present in this section a few direct applications involving the computation of the Fisher information matrix, widely used in cosmology \citep{1997ApJ...480...22T,Stuart1991}.
 \\
@@
 
-We first illustrate, with an artificial case study, the computation of a Fisher matrix \citep{1997ApJ...480...22T,Stuart1991} using two methods with the \textit{autodiff} ability of JAX.
+We first illustrate, with an artificial case study, the computation of a Fisher matrix using two methods with the \textit{autodiff} ability of \jax.
 For the detailed implementation, the reader is invited to look at the following companion notebook \nblink{Simple-Fisher}.
@@
 %\end{minted}
-The \texttt{jc.sparse} functions are implementation of block matrix computations: a sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_{diag})$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.
-The \texttt{jax.jit} decorator builds a compiled version of the function at the first use.
+The \texttt{jc.sparse} functions are implementations of block matrix computations: a sparse matrix is represented as a 3D array of shape $(n_y, n_x, n_{diag})$ composed of $n_y \times n_x$ square blocks of size $n_{diag} \times n_{diag}$.
+The \texttt{jax.jit} decorator builds a compiled version of the function on first use.
 
@@
 \end{equation}
-which is accomplished in two lines of \texttt{JAX} code:
+which is accomplished in two lines of \jax\ code:
 \begin{lstlisting}[language=iPython]
@@
 
-The second approach to computing the Fisher matrix, which is restricted to Gaussian likelihoods but is more commonly used in the field because of its numerical stability, is to define a function that computes the observable mean  $\mu(\ell; \theta)$; it follows that the Fisher matrix elements are
+The second approach to computing the Fisher matrix, which is restricted to Gaussian likelihoods but is more commonly used in the field because of its numerical stability, is to define a function that computes the summary statistic mean  $\mu(\ell; \theta)$; the Fisher matrix elements are then:
 \begin{equation}
@@
 \end{equation}
-where $C^{-1}(\ell)$ is the covariance matrix computed with the fiducial cosmology. This can also be simply computed in JAX: 
+where $C^{-1}(\ell)$ is the covariance matrix computed with the fiducial cosmology. This can be computed in \jax\ as: 
 \begin{lstlisting}[language=iPython]
-# We define a parameter dependent function that computes the mean
+# We define a parameter dependent mean function
 @jax.jit
@@
     mu = jc.angular_cl.angular_cl(cosmo, ell, tracers)
-    # We want mu in 1d to operate against the covariance matrix
+    # We want mu in 1d to match the covariance matrix
     return mu.flatten() 
-# We compute it's jacobian with JAX, and we JIT it for efficiency
+# We compute its jacobian with JAX, and we JIT it for efficiency
 jac_mean = jax.jit(jax.jacfwd(jc_mean))
@@
 
-JAX implementations of the two methods agree to near perfect accuracy, as shown in Figure \ref{fig:simple_fisher_1}.
+\jax\ implementations of the two methods agree to near perfect accuracy, as shown in Figure \ref{fig:simple_fisher_1}.
 \begin{figure}
@@
     \includegraphics[width=0.7\columnwidth]{figures/simple_fisher_1.png}
-    \caption{Comparison of the two methods to compute the Fisher matrix (Eqs.\ref{eq:fisher_way1},\ref{eq:fisher_way2}).} 
+    \caption{Comparison of the two methods to compute the Fisher matrix: method 1 (method 2) is using  Eq.~\ref{eq:fisher_way1} (Eq.~\ref{eq:fisher_way2}).} 
     \label{fig:simple_fisher_1}
@@
 It is worth noting that in the two methods for Fisher matrix computations described above, the user does not need to vary individual parameter values to compute the 1st or 2nd order derivatives; this is in contrast to the usual finite difference methods.
-As an illustration, we have used the CCL library to compute the Fisher matrix via Equation \ref{eq:fisher_way2}.
+As an illustration, we have used CCL to compute the Fisher matrix via Equation \ref{eq:fisher_way2}.
 To do so, the Jacobian ($\partial \mu/\partial\theta_\alpha$) is computed with centered  order 4 finite differences available in the \texttt{Numdifftools} Python package\footnote{\url{https://numdifftools.readthedocs.io}}.
@@
 The contour that agrees best with the \jaxcosmo\ method is obtained with the intermediate spacing parameter $10^{-2}$, implying that the user must tune this parameter carefully.
-Although very simple, this case study exhibits a significant challenge of finite difference methods for computing the Fisher matrix as has been shown for instance in a more advanced case study in \citet{2021arXiv210100298B}.
+Although very simple, this case study demonstrates the significant challenge of using finite difference methods for computing the Fisher matrix, as has been shown for instance in a more advanced case study in \citet{2021arXiv210100298B}.
 
@@
 % \FrL{Insist here on the fact that the FoM itself is differentiable.}
-Fisher forecasts are also commonly used in survey and analysis strategy strategy, where running a full posterior analysis for each possible choice would be unfeasible.
+Fisher forecasts are also commonly used in survey and analysis strategy, where running a full posterior analysis for each possible choice would be unfeasible.
 The inverse area of a projection of a Fisher matrix in parameters of interest can be used as a metric for survey constraining power, such as in the Dark Energy Task Force report \citep{2006astro.ph..9591A}.
@@
 
-Because \texttt{JAX} functions are differentiable with respect to all their inputs, including survey configuration parameters (e.g.
+Because \jax\ functions are differentiable with respect to all their inputs, including survey configuration parameters (e.g.
 depth, area, etc), we can even compute the derivative of an FoM with respect to these inputs, allowing for rapid and complete survey optimization.
@@
 The algorithm computes by iteration $M$ vectors of size $N$ such that (taking the notation of the previous section)
-%JEC 13/1/23 Typo fixed
+
 \begin{equation}
@@
 
-Another key advantage of the MOPED algorithm is to eliminate the need of large covariance matrix inversion of size $N\times N$ requiring $O(N^3)$ operations.
-This inversion takes place not only for the Fisher matrix computation (Eq.~\ref{eq:fisher_way2}, but more importantly in the log-likelihood computation (see the snippet of the previous section).
-The MOPED algorithm reduces the complexity to $O(M)$ operations.
+Another key advantage of the MOPED algorithm is to eliminate the need for large covariance matrix inversion of size $N\times N$ requiring $O(N^3)$ operations.
+This inversion takes place not only for the Fisher matrix computation (Eq.~\ref{eq:fisher_way2}), but more importantly in the log-likelihood computation (see the snippet in the previous section).
+The MOPED algorithm reduces the complexity to $O(M^3)$ operations.
 
@@
 \end{lstlisting}
-Then, the log-likelihood can easily be implemented as:
+Then, the log-likelihood can be implemented as:
 \begin{lstlisting}[language=iPython]
@@
 The comparison between contour lines obtained with the original likelihood (uncompressed data set) and the MOPED version are shown in Figure \ref{fig:moped} for the case study of the previous section.
-Clearly, close to the negative likelihood minimum the lines agree very well.
+Close to the negative likelihood minimum, the lines agree very well.
 \begin{figure}
     \centering
-    \includegraphics[width=\columnwidth]{figures/moded.png}
+    \includegraphics[width=\columnwidth]{figures/moped.png}
     \caption{Illustration of the log-likelihood contours obtained with an uncompressed data set (plain lines) and a MOPED version (dashed lines).} 
@@
 %
-In the following sections we review some statistical methods which directly benefit from the automatic differentiablity of \jaxcosmo\ likelihoods.
+In the following sections we review more statistical methods which directly benefit from the automatic differentiablity of \jaxcosmo\ likelihoods.
 We demonstrate a range of gradient-based methods, from Hamiltonian Monte Carlo (HMC), and its \textit{No-U-Turn Sampler} (NUTS) variant, to Stochastic Variational Inference (SVI).
@@
 %JEC{Transfer of the model context from appendix here to ease the reading.}
-From the DES Year 1 lensing and clustering data release \footnote{\url{http://desdr-server.ncsa.illinois.edu/despublic/y1a1_files/chains/2pt_NG_mcal_1110.fit}} we have extracted the $N(z)$ distributions of four source and five lens samples.
+From the DES Year 1 lensing and clustering data release\footnote{\url{http://desdr-server.ncsa.illinois.edu/despublic/y1a1_files/chains/2pt_NG_mcal_1110.fit}} we have extracted the $N(z)$ distributions of the four source and five lens samples.
 We normalize the sources to $[1.47, 1.46, 1.50, 0.73]$ effective number of sources per $\mathrm{arcmin}^2$.
@@
 \centering
-\includegraphics[height=3cm]{figures/DESY1-source-redshift.png}
-\includegraphics[height=3cm]{figures/DESY1-lens-redshift.png}
+\includegraphics[height=5cm]{figures/DESY1-source-redshift.png}\\
+\includegraphics[height=5cm]{figures/DESY1-lens-redshift.png}
 \caption{Distributions of the sources and lenses for the different redshift bins considered.}
@@
 \begin{table}[htb]
-\caption{Priors of the 21 variables of the DES-Y1 of the 3x2pt likelihood (number counts and shear).}
+\caption{Priors on the 21 variables of the DES-Y1 of the 3x2pt likelihood (number counts and shear).}
 \label{tab-DESY1}
@@
 After encapsulating the code above with input sampled parameters (using \numpyro\ distribution classes) in a function \texttt{model}, we can generate our mock data (\texttt{cl\_obs}).
-This comes comes from this model function evaluated at a fiducial cosmology, with \numpyro\ dealing with the random number generation that generates added noise:
+This comes comes from this model function evaluated at a fiducial cosmology, with \numpyro\ dealing with the random number generation that adds noise:
 
@@
 
-The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generates the model $C_{\ell}$ from the priors:
+% The \textit{inference model} looks very similar to the \textit{forward model} excepts that first we fix the angular power spectra covariance matrix $C$ (and its inverse $P$) and secondly we let \numpyro\ generate the model $C_{\ell}$ from the priors:
+% \begin{lstlisting}[language=iPython]
+% cl = jc.angular_cl.angular_cl(cosmo, ell, 
+%                               probes).flatten()
+% return numpyro.sample("cl", MultivariateNormal(cl, 
+%                             precision_matrix=P,
+%                             covariance_matrix=C))
+% \end{lstlisting}
+
+These theoretical $C_{\ell}$ are in turn conditioned on the mock $C_\ell$:
+
 \begin{lstlisting}[language=iPython]
-cl = jc.angular_cl.angular_cl(cosmo, ell, 
-                              probes).flatten()
-return numpyro.sample("cl", MultivariateNormal(cl, 
-                            precision_matrix=P,
-                            covariance_matrix=C))
-\end{lstlisting}
-
-These theoretical $C_{\ell}$ are in turn conditioned on the mock $C_\ell$:
-
-\begin{lstlisting}[language=iPython]
-observed_model = numpyro.condition(model, {"cl": data})
+observed_model = numpyro.condition(model, {"cl": cl_obs})
 \end{lstlisting}
@@
 Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
-samples from high dimensional parameter spaces.  It was introduced in \citep{1987PhLB..195..216D}
+samples from high dimensional parameter spaces.  It was introduced in \citet{1987PhLB..195..216D}
 and developed extensively since.  See \citet{betancourt} for a full review; we describe
@@
 
-HMC samples spaces by generating particle trajectories through the space, using the log-posterior as the negative potential energy of a particle at each point $q$ in the space.
+HMC samples a space by generating particle trajectories through it, using the log-posterior as the negative potential energy of a particle at each point $q$ in the space.
 Associated with $q$, we introduce an auxiliary $p$ variable as Hamiltonian momentum such that
@@
 \end{equation}
+where $M$ is a mass matrix which should be set to approximate the covariance of the posterior.
 At each sample, a trajectory is initialized with a random momentum $p$, and then Hamilton's equations are integrated:
@@
 \end{align}
-where $M$ is a mass matrix which should be set to approximate the covariance of the posterior.
-This is also used to set the scale of the random initial velocities.
-These differential equations may be integrated numerically, taking $L$ small steps of the \textit{leapfrog} algorithm:
+This is also used to set the scale of the random initial velocities. These differential equations may be integrated numerically, taking $L$ small steps of the \textit{leapfrog} algorithm:
 \begin{align}
@@
 
-The gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
-but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
+The gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differences,
+but this requires at least $2 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
 in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
@@
 The tricky part of HMC is that the \textit{leapfrog} algorithm needs tuning to set the number of steps as well as the step size of integration.
-The next section addresses this problem thanks to the No-U-Turn HMC version.
+The next section examines a solution to this problem: the No-U-Turn HMC version.
 
@@
 
-NUTS sets iterates the leapfrog algorithm not for a fixed $L$, but until the trajectory starts to ``double back'' and return to previously visited region, at the cost of increasing the number of model evaluations.
+NUTS iterates the leapfrog algorithm not for a fixed $L$, but until the trajectory starts to ``double back'' and return to previously visited region, at the cost of increasing the number of model evaluations.
 The user has to set a new  parameter  (\texttt{max\_tree\_depth}) which gives as a power of $2$ the maximum number of model calls at each generation.
@@
 Both sampler HMC and NUTS are available in the \numpyro\ library. After the forward model creation for the DES-Y1 3x2pt exercise described in section \ref{sec-DESY1}: \begin{itemize}
-    \item we apply an affine transformation on the cosmological, intrinsic alignment and bias parameters to use a consistent uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1});
+    \item we apply a transformation to the cosmological, intrinsic alignment and bias parameters to use a consistent uniform prior $\mathcal{U}[-5,5]$ (Table~\ref{tab-DESY1});
     \item
@@
 
-We have run the NUTS sampler using \texttt{numpyro.infer.NUTS} on the DES Y1 likelihood,  with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, with the \texttt{max\_tree\_depth} set to seven (ie.
+We ran the NUTS sampler using \texttt{numpyro.infer.NUTS} on the DES Y1 likelihood,  with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, with the \texttt{max\_tree\_depth} set to seven (ie.
 128 steps for each iteration).
 %\FrL{it should explain that vectorized means 16 chains in parallel on a single GPU.}
-Using the ``vectorized'' \texttt{numpyro} option we ran all 16 chains simultaneously on a single GPU, made possible by the \texttt{JAX} \textit{vmap} mechanism.
-If one has several GPU devices available, then the using the \texttt{JAX} paralellization mechanism (\textit{pmap}), it is further possible to launch the vectorized sampling across the devices, and get back all the MCMC chains.
-However, the experiments have all been undertaken on a single GPU, either a NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
+Using the ``vectorized'' \texttt{numpyro} option we ran all 16 chains simultaneously on a single GPU, made possible by the \jax\ \textit{vmap} mechanism.
+If one has several GPU devices available, then the using the \jax\ paralellization mechanism (\textit{pmap}), it is further possible to launch the vectorized sampling across the devices, and get back all the MCMC chains.
+However, these experiments have all been undertaken on single GPUs, either an NVidia Titan Xp (12GB RAM) on a desktop or an NVidia V100 (32GB RAM) at the IN2P3 Computing Centre\footnote{\url{https://cc.in2p3.fr/en/}}.
 The elapsed time for these experiments was 20 hours.
 
-The results in terms of relative effective sample sizes (ESS) are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidential level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
+The results in terms of relative effective sample sizes (ESS) are detailed in Table~\ref{tab-ESS-NUTS_SVI-1} while the confidence level (CL) contours are presented in Figure \ref{fig_cobaya_NUTS_SVI}.
 We compare to a reference sample from the highly-optimized \texttt{Cobaya} Metropolis-Hastings implementation \citep{2019ascl.soft10019T,2021JCAP...05..057T}, which is widely used in cosmology and which we ran for around 40 hours on CPU to obtain the set of contours shown.
 
-There is a dramatic improvement ofinthe ESS by about a factor of 10 using the NUTS sampler compared to Cobaya, with very good agreement between the CL contours.
+There is a dramatic improvement of the ESS by about a factor of 10 using the NUTS sampler compared to Cobaya, with very good agreement between the CL contours.
 It is worth noting that the mass matrix structure described above increases the sampling efficiency by about a factor of two.
@@
 \end{align}
-which defines the \textit{evidence lower bound} (aka ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
+which defines the \textit{evidence lower bound} (ELBO) that one aims to maximize to get the $\lambda$ values. So, the optimal variational distribution satisfies
 \begin{equation}
@@
 \end{equation}
-where the \textit{guide} in the \numpyro\ library (i.e. the parametrized function $q$) may be a multi-variate Gaussian distribution (MVN) for instance.
-
-Using the auto-differentation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}).
+where the \textit{guide} in the \numpyro\ library (i.e. the parameterised function $q$) may be a multi-variate Gaussian distribution (MVN) for instance.
+
+Using the auto-differentiation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}).
 As stated by the authors of \citep{10.5555/3122009.3122023} ADVI specifies a variational family appropriate to the model, computes the corresponding objective
 function, takes derivatives, and runs a gradient-based or coordinate-ascent optimization.
-First we define a invertible differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$, where the new $\xi$ parameters are unbounded, $xi_i in (\-infty, \infty)$ and so the subsequent minimization step can be performed with no bound constraints.
+First we define a invertible differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$, where the new $\xi$ parameters are unbounded, $\xi_i \in (-\infty, \infty)$ and so the subsequent minimization step can be performed with no bound constraints.
 % One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models.
@@
 $z$ samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
-Using the same DES-Y1 simulation as in previous section, we use both a MultiVariate Normal distribution (MVN) and a Block Nueral Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
+Using the same DES-Y1 simulation as in previous section, we use both a Multivariate Normal distribution (MVN) and a Block Neural Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
 The B-NAF architecture is composed of a single flow using a block autoregressive structure with 2 hidden layers of 8 units each.
 The SVI optimization has been performed with the Adam optimizer \citep{KingmaB14} and a learning rate set to $10^{-3}$.
-We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the B-NAF (MVN) guides, and no tuning of the optimizer has been undertaken as for instance a learning rate scheduling.
-It takes about 2 or 3 hours on the NVidia V100 GPU scaling according to the number of steps.
+We have stopped the optimization after 20,000 (30,000) steps to ensure a stable ELBO loss convergence when using the B-NAF (MVN) guides, and no tuning of the optimizer learning rate scheduling or other parameters was performed.
+This takes about 2 or 3 hours on the NVidia V100 GPU scaling, depending on the number of steps.
 
@@
 If the SVI method can be used as is to get $z$ i.i.d.
-samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{Parno2018,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution with unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables.
+samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{Parno2018,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution with unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed joint distributions for a subset of variables.
 From SVI, one obtains a first approximation of the target distribution, and this approximation is used to choose a better transform $T$ to map the parameter space to a more convenient one $z=F_\lambda(\zeta)$ (eg.
@@
 We have used the MVN guided SVI described in the previous section on the same DES-Y1 analysis described above.
-The NUTS sampler was run using 1 chain of 200 samples (a fast configuration), one chain with 1000 samples and a set of ten chains with 1000 samples each combined into a single run of 10,000 samples (each setup began all chains with 200 samples for initialisation).
+NUTS was run using 1 chain of 200 samples (a fast configuration), one chain with 1000 samples and a set of ten chains with 1000 samples each combined into a single run of 10,000 samples (each setup began all chains with 200 samples for initialisation).
 All NUTS sampling was performed with dense mass matrix optimisation without the special block structuring and with \texttt{max\_tree\_depth=5} which is different than the default NUTS setting described in section \ref{sec-NUTS}.
@@
 For \texttt{Cobaya} we find $\varepsilon\approx 3\%$ with $n_\mathrm{step}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_\mathrm{step}=2^5$ or more.
-With better tuning of the sampling parameters we would expect the $\eta$ values for both methods would to become more comparable, but at this intermediate dimensionality the gain from HMC/NUTS compared to standard MCMC sampler.
+With better tuning of the sampling parameters we would expect the $\eta$ values for both methods to become more comparable, but at this intermediate dimensionality the gain from HMC/NUTS compared to a standard MCMC sampler is small.
 The power of these approaches will become most evident at higher dimensionality still, such has when marginalizing over increasingly complex systematic models.
-NUTS also makes post-processing simpler, since samples are nearly uncorrelated, removing the need for a \textit{thinning} step, which is a rather delicate procedure, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, Owen2017}.
+NUTS also makes post-processing simpler, since samples are nearly uncorrelated, removing the need for a \textit{thinning} step, which is a rather delicate procedure, needing know-how to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, Owen2017}.
 
@@
 Furthermore, using reparametrisation  one can adapt to poor geometry cases (eg.
-\cite{2019arXiv190303704H}).
+\citealp{2019arXiv190303704H}).
 %
@@
 %
-Having demonstrated the utility of \jaxcosmo\ as a differentiable cosmology library library, we now discuss several limitations, and open questions.
+Having demonstrated the utility of \jaxcosmo\ as a differentiable cosmology library, we now discuss several limitations, and raise a few questions.
 
@@
 The first essential barrier to a fully-fledged automatically differentiable cosmology library is the need for a differentiable Boltzmann solver to compute the CMB or matter power spectra.
-At this stage, \jaxcosmo\ relies on the analytic Einsenstein \& Hu fitting formula for the latter, which is not accurate enough for Stage IV requirements, and it does not include models beyond $\Lambda$CDM.
-Existing solvers such as CLASS or CAMB \citep{2011JCAP...07..034B,camb} are large and complex codes which are not easily reimplemented in an \autodiff\ framework and therefore cannot be directly integrated in \jaxcosmo\ .
+At this stage, \jaxcosmo\ relies on the analytic Eisenstein \& Hu fitting formula for the latter, which is not accurate enough for Stage IV \citep{detf} requirements, and it does not include models beyond $\Lambda$CDM.
+Existing solvers such as CLASS \citep{2011JCAP...07..034B} or CAMB \citep{camb} are large and complex codes which are not easily reimplemented in an \autodiff\ framework and therefore cannot be directly integrated in \jaxcosmo\ .
 
 A first option to resolve this issue would be to implement from scratch a new Boltzmann code in a framework that supports automatic differentiation.
-This is the approach behind works such as \texttt{Bolt.jl}\footnote{\url{https://github.com/xzackli/Bolt.jl}} which is provides a simplified Boltzmann solver in Julia, or PyCosmo \citep{pycosmo} which is based on a the SymPy symbolic mathematics library and could be relatively easily compatible with JAX.
+This is the approach behind works such as \texttt{Bolt.jl}\footnote{\url{https://github.com/xzackli/Bolt.jl}} which provides a simplified Boltzmann solver in Julia, or PyCosmo \citep{pycosmo} which is based on the SymPy symbolic mathematics library and could be relatively compatible with \jax.
 However, even if very promising, both of these options thus far remain limited.
-While we do believe a automatically differentiable Boltzmann code is the best option, it seems that the cost of developing such a code remains very high at this time.
+While we do believe an automatically differentiable Boltzmann code is the best option, it seems that the cost of developing such a code remains very high at this time.
 
@@
 % Reasoning behind not making a full emulator of the correlation functions
-Connected to this discussion about emulators, one point that could be discussed is whether one even needs a library like \jaxcosmo\ if anyway one can build emulators of a CCL likelihood for use inside gradient-based inference algorithms.
+Connected to this discussion about emulators, a point that could be discussed is whether one even needs a library like \jaxcosmo\ if one can build emulators of a CCL likelihood for use inside gradient-based inference algorithms.
 While this could indeed be feasible, and of similar cost as just making an emulator of the matter power spectrum, the drawback of this approach is that many analysis parameters and choices become hard coded in the emulator.
-While the model for the linear matter power spectrum is typically kept fixed in practical analyses, all the of the choices related in particular to systematics modeling (e.g.
+Since the model for the linear matter power spectrum is typically kept fixed in practical analyses, all the of the choices related in particular to systematics modeling (e.g.
 photometric redshift errors, galaxy bias, or intrinsic galaxy alignments) will vary significantly in the process of developing the analysis.
-Building an emulator for the likelihood would mean the emulator needs to be retrained every time the likelihood is changed.
+Building an emulator for the likelihood would require the emulator to be retrained every time the likelihood is changed.
 
@@
 %  - Efficiency of gradient-based inference methods
-Another aspect worth discussing are the prospects for scaling up and speeding up cosmological inference in practice given tools such as \jaxcosmo.
+Another aspect worth discussing are the prospects for scaling and speeding up cosmological inference in practice, given tools such as \jaxcosmo.
 As we illustrated in the previous section, gradient-based inference techniques yield significantly less correlated MCMC chains, scale better than any other known sampling method as dimension increases, and can provide very fast approximate posteriors if needed.
-\jaxcosmo\  is also well suited to aid in the parallelization of likelihoods and algorithms, especially on multiple GPUs, which will become increasingly important as the high-performance computing landscape evolves.
+\jaxcosmo\  is also well suited to aid in the parallelisation of likelihoods and algorithms, especially on multiple GPUs, which will become increasingly important as the high-performance computing landscape evolves.
 
@@
 %  - Respective role of CCL and jax-cosmo
-Finally, one question worth asking is how does \jaxcosmo\ position itself against classical codes such as CCL?
-While we are convinced of the benefits of a \texttt{JAX} implementation, we expect CCL and other key codes to remain critical as standard cosmology implementations.
-Ultimately, a natural transition may occur towards differentiable frameworks like \jaxcosmo\ when they reach the ability to run full-fledged Stage IV likelihoods.
+Finally, how does \jaxcosmo\ position itself against classical codes such as CCL?
+While we are convinced of the benefits of a \jax\ implementation, we expect CCL and other key codes to remain critical as standard cosmology implementations.
+Ultimately, a natural transition may occur towards differentiable frameworks like \jaxcosmo\ when they reach the ability to run fully-fledged Stage IV likelihoods.
 
@@
 
-In this paper, we have presented \jaxcosmo, a cosmology library implemented in the \texttt{JAX} framework that enables the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, and greatly speeds up likelihood evaluations thanks to automatic parallelisation and just-in-time compilation on GPUs.
-Currently, \jaxcosmo\ only contains a small set of features corresponding to a DES-Y1 3x2pt analysis.
-Being an open source project, contributions of additional features is additional areas such as CMB or spectroscopic galaxy clustering are warmly welcome.
+In this paper, we have presented \jaxcosmo, a cosmology library implemented in the \jax\ framework that enables the automatic computation of the derivatives of cosmological likelihoods with respect to their parameters, and greatly speeds up likelihood evaluations thanks to automatic parallelisation and just-in-time compilation on GPUs.
+Currently, \jaxcosmo\ contains a small set of features corresponding to a DES-Y1 3x2pt analysis.
+Being an open source project, contributions of new features for additional scientific areas such as CMB or spectroscopic galaxy clustering are warmly welcome.
 
 To demonstrate the value of an automatically differentiable library, we have illustrated with concrete examples how Fisher matrices, which are notoriously unstable and require extensive and careful fine tuning, can now be computed robustly and at much lower cost.
-In addition, Fisher matrices becomes themselves differentiable, which allows for  Figure of Merit optimization by gradient descent, making survey optimization extremely simple.
+In addition, Fisher matrices becomes themselves differentiable, which allows for  Figure of Merit optimization by gradient descent, making survey optimization extremely straightforward.
 
 Going beyond Fisher forecasts, we have also compared simple Metropolis-Hastings to several gradient-based inference techniques (Hamiltonian Monte-Carlo, No-U-Turn-Sampler, and Stochastic Variational Inference).
-We show that the posterior samples with gradient-based methods can reproduce classic methods very efficiently, and can provide approximate posteriors very rapidly.
-These inference techniques can scale to hundreds of dimensions, and may become necessary in Stage IV analysis as the number of nuisance parameters is likely to become very large.
-
-The next extensions of this framework will be the inclusion of additional cosmological probes, as well as the integration of emulators for the matter power spectrum trained on CAMB or CLASS, as a means to go beyond the current analytic Eisenstein \& Hu model.
-
-In the spirit of reproducible research, all results presented in this paper can be reproduced from the following GitHub repository:
+We have shown that the posterior samples with gradient-based methods can reproduce classical methods very efficiently, and can provide approximate posteriors very rapidly.
+These inference techniques can scale to hundreds of dimensions, and may become necessary in Stage IV analysis, as the number of nuisance parameters is likely to become large.
+
+The next extensions to this framework will be the inclusion of additional cosmological probes, as well as the integration of emulators for the matter power spectrum trained on CAMB or CLASS, as a means to go beyond the current analytic Eisenstein \& Hu model.
+
+In the spirit of reproducible research, all results presented in this paper can be reproduced with code contained in the following GitHub repository:
 
@@
 \textbf{J.E Campagne:} Conceptualization, Methodology, Software, Validation, Writing, Visualization.
+\textbf{D.~Kirkby:} Software and validation for sparse linear algebra.
 \textbf{F. Lanusse:} Conceptualization, Methodology, Software, Validation, Writing, Project administration.
 \textbf{D. Lanzieri:} Software and validation for redshift distribution.
+\textbf{S. Casas:} Software contribution for growth rate and power spectra.
+\textbf{Y. Li:} Software contributions.
+\textbf{A. Peel:} Software and validation for spline interpolations in \jax.
 \textbf{J. Zuntz:} Investigation, Writing.
-\textbf{D.~Kirkby:} Software and validation for sparse linear algebra.
-\textbf{A. Peel:} Software and validation for spline interpolations in JAX.
+\textbf{M.~Karamanis:} Software for Gaussian likelihood computation.
 
@@
 
-%In case of the MVN  $\nabla_\mu \mathbb{H}(q)=0$, $\nabla_L \mathbb{H}(q)=(L^{-1})^T$ while the other gradients are computed by the \texttt{JAX} \texttt{autodiff} mechanism.
+%In case of the MVN  $\nabla_\mu \mathbb{H}(q)=0$, $\nabla_L \mathbb{H}(q)=(L^{-1})^T$ while the other gradients are computed by the \jax\ \texttt{autodiff} mechanism.
 %\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.}

### 2023-02-10 paper_OJA.tex
before submission

@@
 
-\title{\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library}
+\title{\texttt{JAX-COSMO}: An End-to-End Differentiable and GPU Accelerated Cosmology Library}
 
@@
 %
-As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), \jax\ exposes an \textit{automatic differentiation} (AD) mechanism.
-Thanks to AD, \jaxcosmo\ gives access to the derivatives of cosmological likelihoods with respect to any of their parameters, and thus enables a range of powerful %  AD can take a Python function that uses \jax\ and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn.
+As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), \jax\ exposes an \textit{automatic differentiation} (autodiff) mechanism.
+Thanks to autodiff, \jaxcosmo\ gives access to the derivatives of cosmological likelihoods with respect to any of their parameters, and thus enables a range of powerful %  AD can take a Python function that uses \jax\ and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn.
 This gives us the opportunity to apply a range of powerful
@@
 Hamiltonian Monte Carlo).
-In particular, we show how Fisher matrices are now fast, exact, no longer require any fine tuning, and are themselves differentiable with respect to parameters of the likelihood enabling complex survey optimization by simple gradient descent.
-Finally, using a Dark Energy Survey Year 1 3x2pt analysis as a benchmark, we demonstrate how \jaxcosmo\ can be combined with Probabilistic Programming Languages such as \numpyro\ to perform posterior inference with state-of-the-art algorithms including a No U-Turn Sampler (NUTS), Automatic Differentiation Variational Inference (ADVI), and Neural Transport HMC (NeuTra).% We discuss algorithms made possible by this library, present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical inference.
-%
-We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that Normalizing Flows using the Neural Transport are a promising methodology.\FrL{we need a better statement than this, what are our conclusions?
-Are we faster/better than Cobaya?}
+In particular, we show how Fisher matrices are now fast, exact, no longer require any fine tuning, and are themselves differentiable with respect to parameters of the likelihood, enabling complex survey optimization by simple gradient descent.
+Finally, using a Dark Energy Survey Year 1 3x2pt analysis as a benchmark, we demonstrate how \jaxcosmo\ can be combined with Probabilistic Programming Languages such as \numpyro\ to perform posterior inference with state-of-the-art algorithms including a No U-Turn Sampler (NUTS), Automatic Differentiation Variational Inference (ADVI), and Neural Transport HMC (NeuTra).
+% We discuss algorithms made possible by this library, present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical inference.
+We show that thee effective sample size per node (1 GPU or 32 CPUs) per hour of wall time is about 5 times better for a JAX NUTS sampler compared to the well optimized Cobaya Metropolis-Hasting sampler.
+We further demonstrate that Normalizing Flows using Neural Transport are a promising methodology for model validation in the early stages of analysis.
+%
+\github
 
@@
 A recent entrant to this field is the \jax\ library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} which has undergone rapid development and can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions, offering a speed up to the development process and indeed code runtimes.
-\jax\ offers an easy parallelization mechanism (\texttt{vmap}), just-in-time compilation (\texttt{jit}), and optimization targeting CPU, GPU, and TPU hardware thanks to the \texttt{XLA} library that powers TensorFlow.
+\jax\ offers an easy parallelization mechanism (\texttt{vmap}), just-in-time compilation (\texttt{jit}), and optimization targeting CPU, GPU, and TPU hardware thanks to the \texttt{XLA} library (which also powers TensorFlow).
 These attractive features have driven wide adoption of \jax\ in computational research, and motivate us to consider its usage in cosmology.
@@
 
-In this context we have developed the open source \jaxcosmo\ library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
+To explore alternative inference methods to the well-known Metropolis-Hasting likelihood sampler, and in order to use GPU devices in the context of JAX framework, we have developed the open source \jaxcosmo\ library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper.
 The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable \jax\ functions.
+
+%\FrL{Missing a paragraph on what we propose to do with this framework in this paper, and why that's interesting in the context of the very long time that it takes to sample common likelihood.}
 
@@
 
-These numerical differences are mostly due to different choices of integration methods and accuracy parameters (eg.
+These numerical differences are mostly due to different choices of integration methods and accuracy parameters (e.g.
 number of quadrature points).
@@
     \includegraphics[width=\columnwidth]{figures/halofit_pk.png}
-    \caption{Comparison of the non-linear matter power spectrum (\textit{halofit} function) between CCL and \jaxcosmo. Is also shown the linear power spectrum.}     \label{fig:halofit_comparison}
+    \caption{Comparison of the non-linear matter power spectrum (\textit{halofit} function) between CCL and \jaxcosmo. Also shown is the linear power spectrum.}     \label{fig:halofit_comparison}
 \end{figure}
@@
 % \section{Unleashing the Power of Fisher Forecasts with Automatic Differentiation}
-\section{Fisher Forecasts \& Data Compression}
+\section{Fisher Information Matrices Made Easy}
 \label{sec-fisher-forecast}
@@
     \includegraphics[width=0.7\columnwidth]{figures/simple_fisher_1.png}
-    \caption{Comparison of the two methods to compute the Fisher matrix: method 1 (method 2) is using  Eq.~\ref{eq:fisher_way1} (Eq.~\ref{eq:fisher_way2}).} 
+    \caption{Comparison of the two methods to compute the Fisher matrix: Eq.~\ref{eq:fisher_way1} (method 1) and Eq.~\ref{eq:fisher_way2} (method 2).} 
     \label{fig:simple_fisher_1}
@@
 is the idea rather that if we dont have a perfect matrix at that stage the worse that can happen is sub optimal compression?}
-%JEC{Well there are several ways that $C$ can be questioned: parameter dependence (ie.
+%JEC{Well there are several ways that $C$ can be questioned: parameter dependence (i.e.
 fiducial model/true model), non-gaussianities, errors...
@@
 We demonstrate a range of gradient-based methods, from Hamiltonian Monte Carlo (HMC), and its \textit{No-U-Turn Sampler} (NUTS) variant, to Stochastic Variational Inference (SVI).
-We further report on their respective computational costs in a DES-Y1 like analysis.
+We further report their respective computational costs in a DES-Y1 like analysis.
 All  the methods have been implemented using the \numpyro\ probabilistic programming language (PPL).
@@
 
-with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} (ie.
+with \texttt{cosmo} an instance of the \texttt{jc.Cosmology} setting the cosmological parameters generated with the priors, and \texttt{ell} (i.e.
 $\ell$) a series of 50 angular modes.
 After encapsulating the code above with input sampled parameters (using \numpyro\ distribution classes) in a function \texttt{model}, we can generate our mock data (\texttt{cl\_obs}).
-This comes comes from this model function evaluated at a fiducial cosmology, with \numpyro\ dealing with the random number generation that adds noise:
+This comes from this model function evaluated at a fiducial cosmology, with random noise generated by \numpyro:
 
@@
 
-We ran the NUTS sampler using \texttt{numpyro.infer.NUTS} on the DES Y1 likelihood,  with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, with the \texttt{max\_tree\_depth} set to seven (ie.
+We ran the NUTS sampler using \texttt{numpyro.infer.NUTS} on the DES Y1 likelihood,  with 16 chains of 1,000 samples each after a warm-up phase consisting of 200 samples, with the \texttt{max\_tree\_depth} set to seven (i.e.
 128 steps for each iteration).
@@
 There is a dramatic improvement of the ESS by about a factor of 10 using the NUTS sampler compared to Cobaya, with very good agreement between the CL contours.
-It is worth noting that the mass matrix structure described above increases the sampling efficiency by about a factor of two.
-
-The speed of the sampling could be improved: we have tested using the parameter \texttt{max\_tree\_depth=5} and found convergence in five hours, showing a linear scaling in this parameter while keeping the sampling efficiencies at a high level; the user is highly encouraged to tune this critical parameter.
+It is worth mentioning that the mass matrix structure described above increases the sampling efficiency by about a factor of two.
+
+The speed of the sampling could be further improved: we have tested using the parameter \texttt{max\_tree\_depth=5} and found convergence in five hours, showing a linear scaling in this parameter while keeping the sampling efficiencies at a high level; the user is highly encouraged to tune this critical parameter.
 
@@
 If we  write $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to Bayes theorem we have $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ as the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$.
-Variational Inference (VI) aims to find an approximation to this distribution, ie.
+Variational Inference (VI) aims to find an approximation to this distribution, i.e.
 $p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution.
@@
 samples from the $q(z,\lambda^\ast)$ distribution as shown on the previous section, the \textit{Neural Transport MCMC} method \citep{Parno2018,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution with unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed joint distributions for a subset of variables.
-From SVI, one obtains a first approximation of the target distribution, and this approximation is used to choose a better transform $T$ to map the parameter space to a more convenient one $z=F_\lambda(\zeta)$ (eg.
+From SVI, one obtains a first approximation of the target distribution, and this approximation is used to choose a better transform $T$ to map the parameter space to a more convenient one $z=F_\lambda(\zeta)$ (e.g.
 $F_\lambda=T^{-1}\circ S^{-1}_\lambda$) is such that
@@
 where $F_\lambda$ with the optimal $\lambda^\ast$ maps the best-fitting $q(z;\lambda^\ast)$ to a geometrically simple function like a unit multivariate normal distribution.
-So, one can use a HMC sampler (eg.
+So, one can use a HMC sampler (e.g.
 NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$.
@@
 With better tuning of the sampling parameters we would expect the $\eta$ values for both methods to become more comparable, but at this intermediate dimensionality the gain from HMC/NUTS compared to a standard MCMC sampler is small.
-The power of these approaches will become most evident at higher dimensionality still, such has when marginalizing over increasingly complex systematic models.
+The power of these approaches will become most evident at higher dimensionality still, such as when marginalizing over increasingly complex systematic models.
 NUTS also makes post-processing simpler, since samples are nearly uncorrelated, removing the need for a \textit{thinning} step, which is a rather delicate procedure, needing know-how to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, Owen2017}.
@@
 \JZ{I don't understand the next point.
-Doesn't the HMC/NUTS take a long time too on dedicated resources?} One should probably have in mind that the low sampling efficiency of a standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie.
+Doesn't the HMC/NUTS take a long time too on dedicated resources?} One should probably have in mind that the low sampling efficiency of a standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, i.e.
 several days on dedicated infrastructure.
@@
 As the dimensionality of cosmological models increases, methods like HMC/NUTS that by construction are more efficient will become increasingly important.
-Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of a model with rather modest sample set (eg.
+Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of a model with rather modest sample set (e.g.
 starting with 200 samples) producing good enough marginal contours in few hours.
@@
 The authors have not investigated higher dimensional ($O(10^2)$ parameters) or multi-modal problems, but the key argument in favour of HMC/NUTS sampling is that it exploits the geometry of the typical set of the posterior distribution automatically, unlike the standard random walk of Metropolis-Hasting sampling.
-Furthermore, using reparametrisation  one can adapt to poor geometry cases (eg.
+Furthermore, using reparametrisation  one can adapt to poor geometry cases (e.g.
 \citealp{2019arXiv190303704H}).
@@
 \textbf{J.E Campagne:} Conceptualization, Methodology, Software, Validation, Writing, Visualization.
+\textbf{S. Casas:} Software contribution for growth rate and power spectra.
+\textbf{M.~Karamanis:} Software for Gaussian likelihood computation.
 \textbf{D.~Kirkby:} Software and validation for sparse linear algebra.
@@
 \textbf{D. Lanzieri:} Software and validation for redshift distribution.
-\textbf{S. Casas:} Software contribution for growth rate and power spectra.
 \textbf{Y. Li:} Software contributions.
@@
 \textbf{J. Zuntz:} Investigation, Writing.
-\textbf{M.~Karamanis:} Software for Gaussian likelihood computation.
 

### 2023-04-17 paper_OJA.tex
JZ Changes implementing reviewer comments. Comment 1: - Added text to start of "Growth of Perturbations" section - Added two BaccoEmu citations to list in "General Discussion" section Comment 2: - Added text to start of "Angular power spectra" section - Added text to start of "General Discussion" section Comment 3 - Added text on chi interpolation - Updated Limber equation - Added text on Limber equation limit - Added text on Simpson for WL kernel - Added text on non-limber requirement - Fixed 70,000 -> 140,000 in several places - Added Kostic citation

@@
 \newcommand{\JEC}[1]{{\color{magenta}JEC: #1}}
+\newcommand{\resubnote}[1]{{\color{red}#1}}
 
@@
 % We discuss algorithms made possible by this library, present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical inference.
-We show that thee effective sample size per node (1 GPU or 32 CPUs) per hour of wall time is about 5 times better for a JAX NUTS sampler compared to the well optimized Cobaya Metropolis-Hasting sampler.
+We show that the effective sample size per node (1 GPU or 32 CPUs) per hour of wall time is about 5 times better for a JAX NUTS sampler compared to the well optimized Cobaya Metropolis-Hasting sampler.
 We further demonstrate that Normalizing Flows using Neural Transport are a promising methodology for model validation in the early stages of analysis.
@@
 \paragraph{\textbf{Automatic Differentiation}} Traditionally, two different approaches have been used in cosmology to obtain derivatives of given numerical expressions.
-The first approach is to derive analytically the formula for the derivatives of interest \citep[e.g.][]{2013MNRAS.432..894J}, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}.
+The first approach is to derive analytically the formula for the derivatives of interest \citep[e.g.][]{2013MNRAS.432..894J, kostic}, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}.
 This is typically only practical, however, for simple analytical models.
@@
 with $R_H$ the Hubble radius.
+\resubnote{$\chi(a)$ evaluation is performed with a linear interpolation of pre-computed values obtained over a logarithmic grid of 256 knots regularly spaced in the range [-3,0] using a Runge-Kutta 4th order solver.}
+
 %Using the scale factor ($a$) redshift ($z$) relationship, $\chi$ can be viewed as a function of $z$. 
@@
 \subsubsection{Growth of perturbations}
-%
-Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function $T$ which transforms the primordial matter power spectrum to its late-time non-linear value:
+\resubnote{Accurate calculations of the growth of cosmic structure perturbations require solving the Boltzmann equation describing interactions between different density species.
+This is a complicated process to achieve at high accuracy, and is implemented in dedicated codes such as CAMB \citep{camb} or CLASS \citep{2011JCAP...07..034B}.
+At this stage we have not attempted to implement a Boltzmann solver in JAX, but instead use the analytic approximation presented in \citet{Eisenstein_1998}.
+This defines a transfer function $T$} which relates the primordial matter power spectrum to its late-time non-linear value:
 \begin{equation}
@@
 \subsubsection{Angular power spectra}
+
+\resubnote{The bias and number density functions that enter the kernel terms have been modelled in many different ways to try to capture possible behavior of these key systematic errors.
+Exposing them in JAX-Cosmo allows us easily to deal with highly complicated models for either of them, opening up many useful avenues of exploration, as shown in \citet{ruiz-zapatero}}.
+
 %
@@
 \begin{align}
-    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{\mathrm{d}\chi}{c^2\chi^2}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
+    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{\mathrm{d}a}{c^2\chi^2}\frac{\mathrm{d}\chi}{\mathrm{d}a}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
 \end{align}
-The $m_i$ factors are $(0,-2)$ for the galaxy clustering and weak lensing, respectively, and each $K(z)$ function represents a single tomographic redshift bin's number density.
-These tracers are implemented as two kernel functions:
+The $m_i$ factors are $(0,-2)$ for the galaxy clustering and weak lensing, respectively, and each $K(z)$ function represents a single tomographic redshift bin's number density \resubnote{in $z=(1/a)-1$.
+The integration lower bound $a_{ij}$ is defined from the maximum redshift involved  and  Simpson's rule is used  with a regular grid of 512 knots.
+The $K(z)$} tracers are implemented as two kernel functions:
 
@@
 with $C\approx 0.0134$ being a dimensionless constant and $D(z)$ the growth factor of linear perturbations.
+\resubnote{The integration is performed using Simpson's rule with an upper bound given by the maximum redshift involved.}
 \end{description}
@@
 Since these differences are likely to be within the tolerance of the current generation of cosmological surveys, this trade-off is an acceptable one.
+
+\resubnote{Despite this high precision with respect to the CCL library, the current \jaxcosmo\ implementation will have to be reviewed to tackle Stage IV accuracy requirements.
+For instance the non-limber $C_\ell$  computation \cite{angpow, n5k} would be a limitation needing revision in a future release.}
 
@@
 \includegraphics[width=1.5\columnwidth]{figures/fig_Cobaya-NUTS-SVI200_v1.pdf}
-\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue, 70,000 samples), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red; 16,000 samples), and a SVI MVN optimisation followed by a Neural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green, 200 samples).}
+\caption{Constraints (30\%, 68\%, 90\%) on 4 of the 21 parameters of a simulated DES-Y1 3x2pt likelihood, using the \texttt{Cobaya} Metropolis-Hastings sampler (full curve in blue, \resubnote{140,000} samples), the  NUTS sampling described in section \ref{sec-NUTS} (dashed curve in red; 16,000 samples), and a SVI MVN optimisation followed by a Neural Transport parametrisation to run a NUTS sampling (dashed-dotted curve in green, 200 samples).}
 \label{fig_cobaya_NUTS_SVI}
@@
 Once the optimisation is done one can obtain  i.i.d.
-$z$ samples from the $q(z,\lambda^\ast)$ distribution  applying the inverse of the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
+$z$ samples from the $q(z,\lambda^\ast)$ distribution applying the inverse of the $S_{\lambda^\ast}$ and $T$ transformations.
 Using the same DES-Y1 simulation as in previous section, we use both a Multivariate Normal distribution (MVN) and a Block Neural Autoregressive Flow (B-NAF) \citep{pmlr-v115-de-cao20a} as \textit{guides} to approximate the true posterior (\textit{model}).
@@
 %}
-(a): Cobaya 70,000 samples;
+(a): Cobaya \resubnote{140,000} samples;
 (b): NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Section~\ref{sec-NUTS}); 
@@
 
-%
 \section{General discussion}
@@
 %
-Having demonstrated the utility of \jaxcosmo\ as a differentiable cosmology library, we now discuss several limitations, and raise a few questions.
+\resubnote{We have demonstrated that the derivatives made available in JAX-Cosmo can be applied in a range of different methodologies to accelerate and improve the stability of cosmological parameter estimation.
+We expect these improvements to be increasingly important over the coming years, as higher statistical precision pushes us towards more complicated models.
+Even though the standard cosmological model has now remained static for several decades, models for systematics such as redshift distribution uncertainty and galaxy bias are becoming more and more complex.
+Using derivative-based methods to constrain them can ensure that we propagate their uncertainties in full.}
+
+\resubnote{With these successes in mind, we now turn to important limitations of \jaxcosmo, and to questions about its place in the cosmology software ecosystem.}
 
@@
 Emulators based on neural networks or Gaussian processes are themselves automatically differentiable with respect to cosmological parameters.
-In fact, the literature is now rich in examples of such emulators \citep[e.g.][and references therein]{Gunther_2022, nygaard,cosmopower,cosmicnet, emucmb}.
+In fact, the literature is now rich in examples of such emulators \citep[e.g.][and references therein]{Gunther_2022, nygaard,cosmopower,cosmicnet, emucmb, bacco1, bacco2}.
 After validating their accuracy against a reference CAMB or CLASS implementation, they could be directly integrated as a plug-and-play replacement for the computation of the matter power spectrum.

### 2023-04-19 paper_OJA.tex
minor update

@@
 \begin{align}
-    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int\frac{\mathrm{d}a}{c^2\chi^2}\frac{\mathrm{d}\chi}{\mathrm{d}a}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
+    C_\ell^{i,j} \approx \left(\ell+\frac{1}{2}\right)^{m_i+m_j}\int_{a_{i,j}}^1\frac{\mathrm{d}a}{c^2\chi^2}\frac{\mathrm{d}\chi}{\mathrm{d}a}K_i(\chi)K_j(\chi)\,P\left(k=\frac{\ell+1/2}{\chi},z\right),\label{eq:Cell_limber}
 \end{align}
@@
 \resubnote{Despite this high precision with respect to the CCL library, the current \jaxcosmo\ implementation will have to be reviewed to tackle Stage IV accuracy requirements.
-For instance the non-limber $C_\ell$  computation \cite{angpow, n5k} would be a limitation needing revision in a future release.}
+For instance the non-limber $C_\ell$  computation \citep{n5k, angpow} would be a limitation needing revision in a future release.}
 
@@
 Emulators based on neural networks or Gaussian processes are themselves automatically differentiable with respect to cosmological parameters.
-In fact, the literature is now rich in examples of such emulators \citep[e.g.][and references therein]{Gunther_2022, nygaard,cosmopower,cosmicnet, emucmb, bacco1, bacco2}.
+In fact, the literature is now rich in examples of such emulators \citep[e.g.][and references therein]{bacco1, bacco2, Gunther_2022, nygaard,cosmopower,cosmicnet, emucmb}.
 After validating their accuracy against a reference CAMB or CLASS implementation, they could be directly integrated as a plug-and-play replacement for the computation of the matter power spectrum.
